### 2022-10-19 Manuscripts/deNovoRates.tex
Structural changes + code reorganization

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino}
\usepackage[hidelinks]{hyperref}

\begin{document}

\onehalfspacing

\section{Introduction}

\section{Results}

In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss. A \textit{de novo} gene emerges from a non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation. Both transcription and translation are complex processes involving many biomolecular complexes that work in concert. Here, we focus on the minimal requirements for these processes to occur. Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence. For translation, we only require the gene to have an open reading frame (ORF). Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost via mutagenesis. Using a similar approach, we estimate some fundamental biochemical properties of proteins that arise from random DNA sequences. Our models thus define a null hypothesis under which DNA sequences evolve neutrally through mutation pressure alone. A significant deviation from this null hypothesis may be interpreted as evidence in support of natural selection.

\subsection{How many times can a gene be lost?}

If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species. We asked what is the chance that the gene was present in a common ancestor, but was lost in all lineages except one. To this end we calculated the probability of gene emergence and gene loss (Materials and Methods). Specifically, the probability of gene gain can be defined as the sum of three probabilities. First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$). Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$) while transcription stays intact ($P_\textit{RNA-stay}$). Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible). The following equation describes the probability of gene gain\footnote{\autoref{genegaineq} describes total gene gain probability. Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$. Using this value does not change the results significantly.}:

\begin{equation}
P_\textit{gene-gain} = P_\textit{ORF-stay}\times P_\textit{RNA-gain} + P_\textit{RNA-stay}\times P_\textit{ORF-gain} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
\label{genegaineq}
\end{equation}

Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$) due to mutations. Specifically:

\begin{equation}
P_\textit{gene-loss} = P_\textit{RNA-loss} + P_\textit{ORF-loss}
\label{genelosseq}
\end{equation}

The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$. To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms of gene gain and gene loss: $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. For example, the ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event. We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF. We found that for all ORF lengths, the gene can be lost at least two times independently, in the time required to gain it (\autoref{gainlossprob}).

\begin{figure}[!t]
\centering
\includegraphics[scale=0.6]{../Figures/geneGainLoss.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis shows the number of independent gene loss events than can happen relative to one gene gain event i.e. $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF.}
\label{gainlossprob}

\vspace{1em}
\hrule
\end{figure}

We performed the same analysis to just the ORF (and not the whole genes). That is, we calculated the ratio of logarithms of ORF gain and ORF loss. We found that for ORFs longer than 143 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.

Overall, our analysis suggests that \textit{de novo} gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.

\subsection{Does \textit{de novo} gene emergence have a preferred trajectory?}

\begin{figure}[!b]
\hrule
\vspace{1ex}
\centering
\includegraphics[scale=0.6]{../Figures/first_ORF_RNA.pdf}
\caption{Shorter \textit{de novo} genes ($\leq$42 codons) preferentially emerge ORF-first whereas longer \textit{de novo} genes emerge RNA-first. The vertical axis shows the log transformed ratio of the probabilites of ORF-first ($P_\textit{ORF-stay}\times P_\textit{RNA-gain}$) and RNA-first ($P_\textit{RNA-stay}\times P_\textit{ORF-gain}$), trajectories of \textit{de novo} emergence. A positive value denotes ORF-first hypothesis and a negative value supports RNA-first hypothesis. The horizontal axis shows the number of codons in the ORF.}
\label{whoisfirst}
\end{figure}

For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge. That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain. Thus it may appear that the order of the occurrence of these two events does not matter. Gene gain is much more likely when one of the two features already exist (\autoref{genegaineq}) and hence there are two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first). To test the hypothesis that the order of feature gain events influences \textit{de novo} emergence, we first asked how much more likely it is to lose transcription before gaining an ORF. Specifically, we calculated the fold difference between the probability values of transcription loss and ORF gain. We found that transcription loss is $10^3$ to $10^{10}$ times higher than ORF gain, depending on the ORF length. This suggests that a mutation has a higher chance of disrupting an existing untranscribed ORF than cause a gain of transcription features. Next, we calculated the fold difference between transcription gain and ORF loss probability values, and found that ORF loss is $10^4$ to $10^5$ times more probable than transcription gain, suggesting that in a transcribed DNA region, a new mutation will more likely disrupt the transcription features than cause a gain of ORF. More generally features are more easily lost due to mutations than they are gained. However, this effect of mutations is not uniform for the two evolutionary trajectories and for different length of ORFs. For example, gain of an 40 codon long ORF in a transcribed region of DNA is $9.2\times10^3$ less likely than the transcription being lost. However, if the ORF already exists, then it is $2.7\times10^4$ times more likely to be lost before transcription emerges. This suggests that the two trajectories of \textit{de novo} emergence have different dynamics.

To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated the fold difference between two probabilities. First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF ($P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}$). This probability denotes the trajectory where ORF emerges first. The second probability denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription ($P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}$). Specifically, we calculated this ratio:

$$\log\left(\frac{P_\textit{ORF-stay}\times P_\textit{RNA-gain}}{P_\textit{RNA-stay}\times P_\textit{ORF-gain}}\right)$$

A positive value means that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in a transcribed DNA (RNA-first). With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF first. \textit{De novo} genes with longer coding region, emerge transcription first (\autoref{whoisfirst}).

\subsection{Would extensive transcription loss suggest negative selection of toxic proteins?}

\textit{De novo} genes that do not provide any fitness benefit to an organism are not fixed in populations via natural selection. These genes can be lost due to mutation pressure. Some newborn \textit{de novo} genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way. These genes would thus be eliminated from the population genomes via negative selection. As we discussed in previous sections, gene loss can occur due to loss of ORF, transcription, or both. To understand what is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$). We found that ORF loss is more probable than transcription loss, for all ORF lengths we investigated ($\geq$30 codons).

We remind the reader that an ORF is lost if the start codon is mutated, the stop codon is mutated an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation). It is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively). Furthermore, non-stop mutations can also lead to cellular toxicity. Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss is the best way to inactivate the associated genes. Since

\begin{itemize}
\item protein can be toxic
\item orf loss can extend or truncate orf (this can still cause expression of some protein)
\item transcription loss is the best way to shut down toxic protein synthesis
\item distribution of expected vs observed promoter sequences
\item extensive transcription loss can suggest protein toxicity
\end{itemize}

\subsection{Does mutation bias shape protein composition?}

\begin{itemize}
\item expected hydropathy index
\item expected number of specific mutations
\item higher than expected number of mutations would suggest positive selection
\end{itemize}

\section{Discussion}

\end{document}

### 2022-10-19 Modeling/DeNovoProb.jl

@@ -7,15 +7,15 @@ using LaTeXStrings
 include("nucleotidefuncts.jl")
 
 cm2pt = (cm) -> 28.3465*cm
-figdir = "~/Documents/EBB/StochasticModel/Figures/"
+figdir = "~/Documents/DeNovo/StochasticModel/Figures/"
 
-default(linecolor = :black, linewidth = 2, tickfont = font(10), guidefontsize = 12,
-        framestyle = :box, legend = false)
+default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
+guidefont = font(13,"Helvetica"),framestyle = :box, legend = false)
 
 
 gccontent = 0.42
 
-function transvals(gccontent)
+function rnavals(gccontent)
     ## TATA box
     # Consensus sequence = TATAWAWR
     # is TSS - 25 - 8nt (transcript must be +33nt longer)
@@ -69,20 +69,20 @@ function transvals(gccontent)
     ## Transcription
     # (Initiator or TATA) and polyA
 
-    transprob = (tataprob + inrprob)*polyaprob
-    transgain = (tatagain + inrgain)*polyastay + (tatastay + inrstay)*polyagain
+    rnaprob = (tataprob + inrprob)*polyaprob
+    rnagain = (tatagain + inrgain)*polyastay + (tatastay + inrstay)*polyagain
     tfpinr = inrprob*(1-tataprob)/(tataprob + inrprob - inrprob*tataprob)
     tfptata = tataprob*(1-inrprob)/(tataprob + inrprob - inrprob*tataprob)
     tfpboth = inrprob*tataprob/(tataprob + inrprob - inrprob*tataprob)
     tfpboth = 1 - tfptata - tfpinr
 
-    transloss = ( tfpinr*inrloss2*notatastay +
+    rnaloss = ( tfpinr*inrloss2*notatastay +
       tfptata*tataloss2*noinrstay +
       tfpboth*tataloss2*inrloss2 +
       polyaloss2
     )
-    transstay = (inrstay + tatastay)*polyastay
-    return [transprob; transgain; transloss; transstay]
+    rnastay = (inrstay + tatastay)*polyastay
+    return [rnaprob; rnagain; rnaloss; rnastay]
 end
 
 function ATGvals(gccontent)
[…]

### 2022-10-19 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -160,6 +160,10 @@ geneloss = orfloss .+ rnaloss
 rnafirst = rnastay .* orfgain
 orffirst = orfstay .* rnagain
 
+rnafirst2 = rnafirst.*rnagain
+orffirst2 = orffirst.*orfgain
+
+
 plot_genegain_geneloss = plot(ncodons,log.(genegain2)./log.(geneloss),
     xlabel = "ORF length (codons)",
     ylabel = "# Gene Losses \n per Gene Gain",
@@ -175,10 +179,18 @@ plot_rnafist_orffirst = plot(ncodons[1:50],log.(orffirst./rnafirst)[1:50],
 );
 savefig(plot_rnafist_orffirst, figdir*"first_ORF_RNA.pdf")
 
+plot_rnafist_orffirst = plot(ncodons[1:50],log.(orffirst2./rnafirst2)[1:50],
+    xlabel = "ORF length (codons)",
+    ylabel = "P_{ORF-first}\nP_RNA-first",
+    size = (width = cm2pt(12), height = cm2pt(11)),
+    xticks = [35:7:80;]
+);
+savefig(plot_rnafist_orffirst, figdir*"first_ORF_RNA.pdf")
+
 tl = log10.(rnaloss./orfgain)
 ol = log10.(orfloss/rnagain)
 
-plot_Loss = plot(ncodons,rnaloss./orfloss,
+plot_Loss = plot(ncodons,orfloss.rnaloss,
     xlabel = "ORF length (codons)",
     ylabel = "P_RNA-loss\nP_ORF-loss",
     size = (width = cm2pt(12), height = cm2pt(11))

### 2022-10-19 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -7,7 +7,7 @@ using LaTeXStrings
 include("nucleotidefuncts.jl")
 
 cm2pt = (cm) -> 28.3465*cm
-figdir = "~/Documents/DeNovo/StochasticModel/Figures/"
+figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/")
 
 default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
 guidefont = font(13,"Helvetica"),framestyle = :box, legend = false)
@@ -179,21 +179,22 @@ plot_rnafist_orffirst = plot(ncodons[1:50],log.(orffirst./rnafirst)[1:50],
 );
 savefig(plot_rnafist_orffirst, figdir*"first_ORF_RNA.pdf")
 
-plot_rnafist_orffirst = plot(ncodons[1:50],log.(orffirst2./rnafirst2)[1:50],
+plot_rnafist_orffirst2 = plot(ncodons[1:50],log.(orffirst2./rnafirst2)[1:50],
     xlabel = "ORF length (codons)",
     ylabel = "P_{ORF-first}\nP_RNA-first",
     size = (width = cm2pt(12), height = cm2pt(11)),
     xticks = [35:7:80;]
 );
-savefig(plot_rnafist_orffirst, figdir*"first_ORF_RNA.pdf")
+savefig(plot_rnafist_orffirst2, figdir*"first_ORF_RNA2.pdf")
 
 tl = log10.(rnaloss./orfgain)
 ol = log10.(orfloss/rnagain)
 
-plot_Loss = plot(ncodons,orfloss.rnaloss,
+plot_Loss = plot(ncodons[1:85],log.(orfloss./rnaloss)[1:85],
     xlabel = "ORF length (codons)",
     ylabel = "P_RNA-loss\nP_ORF-loss",
-    size = (width = cm2pt(12), height = cm2pt(11))
+    size = (width = cm2pt(12), height = cm2pt(11)),
+    xticks = [33:11:110;]
 );
 
 savefig(plot_Loss, figdir*"pLoss.pdf")

### 2022-10-19 Modeling/DFE.jl
modified:   Modeling/DFE.jl

@@ -11,6 +11,9 @@ allcodons = kmers(3)
 
 aasubmat[1:20,1:20] = (aasubmat[1:20,1:20] + aasubmat[1:20,1:20]')./2
 
+aas = unique(gencode[:,3])
+aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas]
+
 function nostop(codonset)
         codonset[codonset .∉ Ref(["TAA","TAG","TGA"])]
 end

### 2022-10-20 Modeling/DFE.jl
modified:   Modeling/DFE.jl,DeNovoProb.jl,nucleotidefuncts.jl

@@ -11,9 +11,6 @@ allcodons = kmers(3)
 
 aasubmat[1:20,1:20] = (aasubmat[1:20,1:20] + aasubmat[1:20,1:20]')./2
 
-aas = unique(gencode[:,3])
-aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas]
-
 function nostop(codonset)
         codonset[codonset .∉ Ref(["TAA","TAG","TGA"])]
 end

### 2022-10-20 Modeling/DeNovoProb.jl

@@ -198,3 +198,28 @@ plot_Loss = plot(ncodons[1:85],log.(orfloss./rnaloss)[1:85],
 );
 
 savefig(plot_Loss, figdir*"pLoss.pdf")
+
+aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas]
+
+aasubprob = zeros(20,20)
+
+for i = 1:20
+        for j = 1:20
+                if(i==j)
+                        continue
+                end
+                set1 = gencode[gencode[:,3] .== aas[i],1]
+                set2 = gencode[gencode[:,3] .== aas[j],1]
+                aasubprob[i,j] = featuregain(set1,set2,gccontent)
+        end
+end
+
+plot_aasubprob = heatmap(aas,aas,log10.(aasubprob./aaprob), 
+    xticks = (0.5:20, aas), 
+    yticks = (0.5:20,aas),
+    size = (width = cm2pt(15), height = cm2pt(14)),
+    colorbar_title = "Substitution Probability",
+    legend = :bottom
+);
+
+savefig(plot_aasubprob, figdir*"pAASub.pdf")

### 2022-10-20 Modeling/nucleotidefuncts.jl

@@ -8,6 +8,8 @@ nucv = ['A','T','G','C']
 nucnames = Dict('A'=>1,'T'=>2,'G'=>3,'C'=>4)
 ncomp = Dict('A'=>'T', 'G'=>'C', 'T'=>'A', 'C'=> 'G')
 
+aas = ["A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","V","W","Y"]
+
 """
 # Generate all nucleotide sequences with length `k`
 """

### 2022-10-20 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -7,10 +7,10 @@ using LaTeXStrings
 include("nucleotidefuncts.jl")
 
 cm2pt = (cm) -> 28.3465*cm
-figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/")
+figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
 
 default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
-guidefont = font(13,"Helvetica"),framestyle = :box, legend = false)
+guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
 
 
 gccontent = 0.42
@@ -140,10 +140,10 @@ end
 # minRNAlen = (k) -> 3*k + 4
 
 
-ncodons = [30:300;]
+ncodons = [30:300;];
 
-ATGvalues = ATGvals(gccontent)
-stopvalues = stopvals(gccontent)
+ATGvalues = ATGvals(gccontent);
+stopvalues = stopvals(gccontent);
 
 
 (rnaprob, rnagain, rnaloss, rnastay) = rnavals(gccontent)
@@ -153,15 +153,18 @@ for k in 1:length(ncodons)
     (orfprob[k], orfgain[k], orfloss[k], orfstay[k]) = orfvals(gccontent,ATGvalues,stopvalues,ncodons[k])
 end
 
-genegain = ((rnagain + rnastay) .* orfgain + orfstay .* rnagain)
-genegain2 = genegain./(1-rnaprob .- orfprob)
-geneloss = orfloss .+ rnaloss
+genegain = ((rnagain + rnastay) .* orfgain + orfstay .* rnagain);
+genegain2 = genegain./(1-rnaprob .- orfprob);
+geneloss = orfloss .+ rnaloss;
 
-rnafirst = rnastay .* orfgain
-orffirst = orfstay .* rnagain
+rnafirst = rnastay .* orfgain;
+orffirst = orfstay .* rnagain;
 
-rnafirst2 = rnafirst.*rnagain
-orffirst2 = orffirst.*orfgain
+onlyrnagain = (1 .- orfprob .- orfgain).*rnagain;
+onlyorfgain = (1 -rnaprob - rnagain).*orfgain;
+
+rnafirst2 = onlyrnagain*(1 - rnaloss).*(orfgain./(1 .-orfprob));
+orffirst2 = onlyorfgain.*(1 .-orfloss).*(rnagain/(1-rnaprob));
 
 
[…]

### 2022-10-20 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino, subcaption}
 \usepackage[hidelinks]{hyperref}
+
+\captionsetup[figure]{labelfont=bf,textfont={small,normalfont}}
+\captionsetup[subfigure]{labelsep=quad, labelfont={bf,small}, textfont=small, singlelinecheck=off, justification=raggedright, position=top}
+\renewcommand{\thesubfigure}{\Alph{subfigure}}
 
@@
 \centering
-\includegraphics[scale=0.6]{../Figures/geneGainLoss.pdf}
+\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
@@
 \centering
-\includegraphics[scale=0.6]{../Figures/first_ORF_RNA.pdf}
-\caption{Shorter \textit{de novo} genes ($\leq$42 codons) preferentially emerge ORF-first whereas longer \textit{de novo} genes emerge RNA-first.
-The vertical axis shows the log transformed ratio of the probabilites of ORF-first ($P_\textit{ORF-stay}\times P_\textit{RNA-gain}$) and RNA-first ($P_\textit{RNA-stay}\times P_\textit{ORF-gain}$), trajectories of \textit{de novo} emergence.
+\begin{subfigure}{0.48\textwidth}
+\centering
+\caption{One-step}
+\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
+\end{subfigure}\hfill%
+\begin{subfigure}{0.48\textwidth}
+\centering
+\caption{Two-step}
+\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
+\end{subfigure}
+\caption{Shorter \textit{de novo} genes preferentially emerge ORF-first whereas longer \textit{de novo} genes emerge RNA-first.
+In both panels, the horizontal axis denote the number of codons in the ORF.
+The vertical axes show the log transformed ratio of the probabilites of ORF-first and RNA-first, trajectories of \textit{de novo} emergence.
 A positive value denotes ORF-first hypothesis and a negative value supports RNA-first hypothesis.
-The horizontal axis shows the number of codons in the ORF.}
+Panel \textbf{(A)} shows the log-ratio of one-step probabilities defined in Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}, whereas panel \textbf{(B)} shows the log-ratio of two-step probabilities defined in Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}.}
 \label{whoisfirst}
@@
 
-To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated the fold difference between two probabilities.
-First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF ($P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}$).
+To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
+First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF.
 This probability denotes the trajectory where ORF emerges first.
-The second probability denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription ($P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}$).
-Specifically, we calculated this ratio:
 
-$$\log\left(\frac{P_\textit{ORF-stay}\times P_\textit{RNA-gain}}{P_\textit{RNA-stay}\times P_\textit{ORF-gain}}\right)$$
+\begin{equation}
+P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}
+\label{orffirsteq1}
+\end{equation}
 
-A positive value means that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in a transcribed DNA (RNA-first).
-With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF first.
-\textit{De novo} genes with longer coding region, emerge transcription first (\autoref{whoisfirst}).
+The second probability that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription.
+
+\begin{equation}
+P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}
+\label{rnafirsteq1}
+\end{equation}
+
+We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in a transcribed DNA (RNA-first).
+In other words, ORF-first hypothesis is more feasible.
+Likewise, a negative value of the ratio would suggest that RNA-first hypothesis is more feasible.
+With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first.
+\textit{De novo} genes with longer coding region, emerge transcription first (\autoref{whoisfirst}A).
+
+A more stringent definition of ORF-first hypothesis would describe a two-step process.
+In the first step, an ORF emerges in a non-genic DNA, and in the second step transcription emerges, while ORF stays intact.
+RNA-first hypothesis can be defined in a similar manner; in the first step transcription emerges in a non-genic DNA, and in the second step an ORF emerges, while transcription remains intact.
+The probabilities of these two trajectories can be defined as follows:
+
+\begin{align}
+P_\textit{ORF-first}' & = P_\textit{ORF-gain} \times P_\textit{ORF-stay}\times P_\textit{RNA-gain} \label{orffirsteq2}\\[1em]
+P_\textit{RNA-first}' & = P_\textit{RNA-gain} \times P_\textit{RNA-stay}\times P_\textit{ORF-gain}
+\label{rnafirsteq2}
+\end{align}
+
+We calculated the log-transformed ratio of the two step ORF-first and RNA-first probabilities, and found that \textit{de novo} genes with ORFs smaller than 57 codons preferentially emerge ORF-first (log ratio greater than 1, \autoref{whoisfirst}B).
+
+Overall, our analysis suggests that \textit{de novo} genes with long ORFs would arise from DNA regions that are already transcribed (for example, non-coding RNAs).
 
@@
 To understand what is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, for all ORF lengths we investigated ($\geq$30 codons).
+We found that ORF loss is more probable than transcription loss, for all ORF longer than 33 codons.
+However, the difference between these probabilities is not very large.
+
+To apply this theory to genomics data we asked a more specific question.
+To explain this question we use a hypothetical observation where for a DNA locus, there is evidence for transcription in only one species, whereas ORF is retained in all species.
+Two evolutionary events could describe this observation.
+First, a \textit{de novo} gene existed in an ancestor and transcription was lost.
+Alternatively, an ORF existed first and transcription was subsequently gained.
+Event 1 can be described by the probability $P_\textit{ORF-stay}\times P_\textit{RNA-loss}$.
+Event 2 can be described by the probability $(1-P_\textit{RNA} - P_\textit{RNA-gain})\times P_\textit{ORF-gain}$
 
@@
 \item extensive transcription loss can suggest protein toxicity
+\item if I compare P(nornagain)*P(orfloss) and P(rnaloss).P(orfstay) then differences are small. So such inferences cannot be made from species data.
 \end{itemize}

### 2022-10-21 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

--- 2022-10-19 Manuscripts/deNovoRates.tex
@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino, subcaption}
 \usepackage[hidelinks]{hyperref}
+
+\captionsetup[figure]{labelfont=bf,textfont={small,normalfont}}
+%\captionsetup[subfigure]{labelsep=quad, labelfont={bf,small}, textfont=small, singlelinecheck=off, justification=raggedright, position=top}
+%\renewcommand{\thesubfigure}{\Alph{subfigure}}
+
+\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
+\setcounter{footnote}{1}
 
@@
 
-\vspace{1em}
+\vspace{1ex}
 \hrule
@@
 \centering
-\includegraphics[scale=0.6]{../Figures/first_ORF_RNA.pdf}
-\caption{Shorter \textit{de novo} genes ($\leq$42 codons) preferentially emerge ORF-first whereas longer \textit{de novo} genes emerge RNA-first.
-The vertical axis shows the log transformed ratio of the probabilites of ORF-first ($P_\textit{ORF-stay}\times P_\textit{RNA-gain}$) and RNA-first ($P_\textit{RNA-stay}\times P_\textit{ORF-gain}$), trajectories of \textit{de novo} emergence.
-A positive value denotes ORF-first hypothesis and a negative value supports RNA-first hypothesis.
-The horizontal axis shows the number of codons in the ORF.}
+%\begin{subfigure}{0.48\textwidth}
+%\centering
+%\caption{One-step}
+\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
+%\end{subfigure}\hfill%
+%\begin{subfigure}{0.48\textwidth}
+%\centering
+%\caption{Two-step}
+%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
+%\end{subfigure}
+\caption{\textit{De novo} genes with shorter ORFs ($<$ 42 codons) preferentially emerge from existing ORFs (ORF-first) whereas \textit{de novo} genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
+The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
+A positive value suggests a ORF-first trajectory and a negative value supports RNA-first trajectory.
+Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
 \label{whoisfirst}
@@
 
-To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated the fold difference between two probabilities.
-First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF ($P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}$).
+To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
+First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF.
 This probability denotes the trajectory where ORF emerges first.
-The second probability denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription ($P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}$).
-Specifically, we calculated this ratio:
 
-$$\log\left(\frac{P_\textit{ORF-stay}\times P_\textit{RNA-gain}}{P_\textit{RNA-stay}\times P_\textit{ORF-gain}}\right)$$
+\begin{equation}
+P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}
+\label{orffirsteq1}
+\end{equation}
 
-A positive value means that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in a transcribed DNA (RNA-first).
-With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF first.
+The second probability that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription.
+
+\begin{equation}
+P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}
+\label{rnafirsteq1}
+\end{equation}
+
+We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
+In other words, ORF-first hypothesis is more feasible.
+Likewise, a negative value of the ratio would suggest that RNA-first hypothesis is more feasible.
+With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first.
 \textit{De novo} genes with longer coding region, emerge transcription first (\autoref{whoisfirst}).
+
+A more stringent definition of ORF-first hypothesis would describe a two-step process.
+In the first step, an ORF emerges in an untranscribed region of DNA, but transcription does not emerge.
+In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2})\footnote{$P_\textit{ORF-gain}$ describes the total probability of ORF emerging from a DNA region that does not have already have an ORF.
+Conditional probability of ORF gain would be $P_\textit{ORF-gain}/(1-P_\textit{ORF})$.
+Conditional probability of transcription gain can be described similarly.} .
+Likewise, RNA-first hypothesis can be defined by a two-step probability.
+In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).
+
+With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}$).
+However, we still find that \textit{de novo} genes with shorter ORFs ($<$ 34 codons) have a slightly higher chance of emerging ORF first.
+
+Overall, our analysis suggests that \textit{de novo} genes with long ORFs are more likely arise from DNA regions that are already transcribed (for example, non-coding RNAs).
+
+\begin{align}
+P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
+P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
+\end{align}
 
@@
 These genes would thus be eliminated from the population genomes via negative selection.
-As we discussed in previous sections, gene loss can occur due to loss of ORF, transcription, or both.
-To understand what is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, for all ORF lengths we investigated ($\geq$30 codons).
-
-We remind the reader that an ORF is lost if the start codon is mutated, the stop codon is mutated an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation).
+We remind the reader that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation).
 It is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively).
 Furthermore, non-stop mutations can also lead to cellular toxicity.
-Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss is the best way to inactivate the associated genes.
-Since
+Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may be the better way to inactivate the associated genes.
 
-\begin{itemize}
-\item protein can be toxic
-\item orf loss can extend or truncate orf (this can still cause expression of some protein)
-\item transcription loss is the best way to shut down toxic protein synthesis
-\item distribution of expected vs observed promoter sequences
-\item extensive transcription loss can suggest protein toxicity
-\end{itemize}
+To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
+We found that ORF loss is more probable than transcription loss, for all ORF longer than 33 codons (\autoref{lossprob}).
+
+We reframed our question to make it more relatable to commonly observed findings from genomics experiments.
+For example, a hypothetical experiment involves DNA sequencing of a genomic locus in different related species.
+This experiment shows that transcription features are present in only one species, whereas the ORF is present in all species.
+Two evolutionary events could describe this observation.
+First possibility is that a \textit{de novo} gene existed in an ancestor and transcription was lost in multiple species.
+For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$ \footnote{Note that $P_\textit{RNA-loss}$ is a conditional probability dependent on the presence of RNA.}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.6]{Figures/pLoss.pdf}
+\caption{ORF loss is more probable than transcription loss, for \textit{de novo} genes containing more than 33 codons.
+The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
+\label{lossprob}
+
+\vspace{1ex}
+\hrule
+\end{figure}
+
+Alternatively, an ORF could have existed in all species and transcription was subsequently gained in one species.
+The probability that this event happens in any one species is: $P_\textit{ORF-stay}\times P_\textit{RNA-gain}$ (see \autoref{orffirsteq1}).
+
+The ratio of the probabilities of the two evolutionary trajectories would simply be: $P_\textit{RNA} \times P_\textit{RNA-loss}/P_\textit{RNA-gain}$.
+We found the value of this ratio to be 0.914, suggesting that transcription loss in a \textit{de novo} gene would be more probable than transcription gain for an untranscribed ORF.
+However, this difference is small and hence the real evolutionary mechanism behind the observation from our hypothetical sequencing experiment, would be difficult to infer.
+
+In sum, transcription loss cannot be reliably inferred as an evidence for negative selection of protein expression.
 
@@
 \begin{itemize}
-\item expected hydropathy index
-\item expected number of specific mutations
-\item higher than expected number of mutations would suggest positive selection
+\item expected hydropathy index = $-0.08628$
+\item acid to basic rate is $1.3074\times$ higher than \textit{vice versa}. Conditional prob = $1.80543$
+\item mutations do not preferentially create hydrophobic amino acid. Total prob ratio = 1.025, Conditional prob ratio = 0.971
 \end{itemize}

### 2022-10-21 Modeling/DeNovoProb.jl

@@ -149,7 +149,7 @@ stopvalues = stopvals(gccontent);
 (rnaprob, rnagain, rnaloss, rnastay) = rnavals(gccontent)
 (orfprob, orfgain, orfloss, orfstay) = [zeros(size(ncodons)) for i = 1:4 ]
 
-for k in 1:length(ncodons)
+for k in eachindex(ncodons)
     (orfprob[k], orfgain[k], orfloss[k], orfstay[k]) = orfvals(gccontent,ATGvalues,stopvalues,ncodons[k])
 end
 
@@ -165,7 +165,6 @@ onlyorfgain = (1 -rnaprob - rnagain).*orfgain;
 
 rnafirst2 = onlyrnagain*(1 - rnaloss).*(orfgain./(1 .-orfprob));
 orffirst2 = onlyorfgain.*(1 .-orfloss).*(rnagain/(1-rnaprob));
-
 
 plot_genegain_geneloss = plot(ncodons,log.(genegain2)./log.(geneloss),
     xlabel = "ORF length (codons)",
@@ -202,7 +201,27 @@ plot_Loss = plot(ncodons[1:85],log.(orfloss./rnaloss)[1:85],
 
 savefig(plot_Loss, figdir*"pLoss.pdf")
 
+onlyrnaloss = orfstay .*(rnaprob*rnaloss)
+
 aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];
+aaprob2 = aaprob./sum(aaprob);
+exp_hydropathy = sum(hydropathy.*aaprob2)
+
+acidic = indexin(["D","E"],aas);
+basic = indexin(["K","R"],aas);
+hydrophobic = hydropathy .>0;
+
+pHydrophobic = sum(aaprob[hydrphobic])
+
+plot_aafreq = bar(aas,aaprob2,
+    xticks = (0.5:20, aas),
+    fill = :black,
+    xlabel = "Amino acid",
+    ylabel = "Expected frequency",
+    size = (width = cm2pt(15), height = cm2pt(11))
+);
+savefig(plot_aafreq, figdir*"pAAfreq.pdf")
+
 
 aasubprob = zeros(20,20);
 
@@ -217,6 +236,18 @@ for i = 1:20
         end
 end
 
+aasubprob_conditional = aasubprob./aaprob2
+
+asymmetric_transitions = log2.(aasubprob./aasubprob');
+
+asymmetric_conditional = log2.(aasubprob_conditional./aasubprob_conditional');
+
[…]

### 2022-10-21 Modeling/nucleotidefuncts.jl

@@ -8,7 +8,9 @@ nucv = ['A','T','G','C']
 nucnames = Dict('A'=>1,'T'=>2,'G'=>3,'C'=>4)
 ncomp = Dict('A'=>'T', 'G'=>'C', 'T'=>'A', 'C'=> 'G')
 
-aas = ["A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","V","W","Y"]
+aas = ["A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","V","W","Y"];
+
+hydropathy = [1.8, 2.5, -3.5, -3.5, 2.8, -0.4, -3.2, 4.5, -3.9, 3.8, 1.9, -3.5, -1.6, -3.5, -4.5, -0.8, -0.7, 4.2, -0.9, -1.3]; # Kyte and Doolittle 
 
 """
 # Generate all nucleotide sequences with length `k`

### 2022-10-21 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 
+In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss.
+We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of \textit{de novo} proteins.
+To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform.
+Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids.
+On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}).
+This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
+However, GC content also plays a role.
+For example, leucine, arginine and serine, all have six codons each.
+However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}).
+
 \begin{itemize}
@@
 
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.6]{Figures/pAAfreq.pdf}
+\caption{ORF loss is more probable than transcription loss, for \textit{de novo} genes containing more than 33 codons.
+The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
+\label{aafreq}
+
+\vspace{1ex}
+\hrule
+\end{figure}
+
 \section{Discussion}

### 2022-10-21 Modeling/DeNovoProb.jl

@@ -242,11 +242,13 @@ asymmetric_transitions = log2.(aasubprob./aasubprob');
 
 asymmetric_conditional = log2.(aasubprob_conditional./aasubprob_conditional');
 
-toHydrophobic = sum(aasubprob[:,hydrophobic]);
-frHydrophobic = sum(aasubprob[hydrophobic,:]);
+codonshydrophobic = gencode[(in).(gencode[:,3],Ref(aas[hydrophobic])),1]
 
-toHydrophobic_cond = sum(aasubprob_conditional[:,hydrophobic]);
-frHydrophobic_cond = sum(aasubprob_conditional[hydrophobic,:]);
+codonshydrophilic = kmers(3)[(!in).(kmers(3),Ref(codonshydrophobic))]
+
+toHydrophobic = featuregain(codonshydrophilic,codonshydrophobic, gccontent);
+
+frHydrophobic = featuregain(codonshydrophobic,codonshydrophilic, gccontent);
 
 plot_aasubprob = heatmap(aas,aas,log10.(aasubprob./aaprob), 
     xticks = (0.5:20, aas), 

### 2022-10-21 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 
+Previous studies have argued that hydrophobic amino acids are comprise 40\% of amino acids and can also easily arise due to mutation pressure.
+They based these conclusions on mutation rate bias estimated in bacteria.
+We asked if our model also makes similar predictions.
+To this end we calculated expected hydropathy in a protein sequence, based on expected frequency of different amino acids (\autoref{aafreq}) and their hydropathic index values (Kyte and Doolittle).
+We found that the expected hydropathy is $-0.08628$, which suggests that random protein sequences are not mostly hydrophobic.
+
+Next, we calculated the probability that any
+
 \begin{itemize}

### 2022-10-28 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 \captionsetup[figure]{labelfont=bf,textfont={small,normalfont}}
-%\captionsetup[subfigure]{labelsep=quad, labelfont={bf,small}, textfont=small, singlelinecheck=off, justification=raggedright, position=top}
-%\renewcommand{\thesubfigure}{\Alph{subfigure}}
+\captionsetup[subfigure]{labelsep=quad, labelfont={bf,small}, textfont=small, singlelinecheck=off, justification=raggedright, position=top}
+\renewcommand{\thesubfigure}{\Alph{subfigure}}
 
@@
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible).
-The following equation describes the probability of gene gain\footnote{\autoref{genegaineq} describes total gene gain probability.
-Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
-Using this value does not change the results significantly.}:
+The following equation describes the probability of gene gain:
 
@@
 \caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis shows the number of independent gene loss events than can happen relative to one gene gain event i.e.
+The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e.
 $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF.}
+The horizontal axis shows the number of codons in the ORF.
+\autoref{genegaineq} describes total gene gain probability.
+Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
+Using the conditional probability of gene gain does not change these results.}
 \label{gainlossprob}
@@
 In the first step, an ORF emerges in an untranscribed region of DNA, but transcription does not emerge.
-In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2})\footnote{$P_\textit{ORF-gain}$ describes the total probability of ORF emerging from a DNA region that does not have already have an ORF.
-Conditional probability of ORF gain would be $P_\textit{ORF-gain}/(1-P_\textit{ORF})$.
-Conditional probability of transcription gain can be described similarly.} .
+In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
 Likewise, RNA-first hypothesis can be defined by a two-step probability.
@@
 
-With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}$).
+With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}\ \%$).
 However, we still find that \textit{de novo} genes with shorter ORFs ($<$ 34 codons) have a slightly higher chance of emerging ORF first.
@@
 First possibility is that a \textit{de novo} gene existed in an ancestor and transcription was lost in multiple species.
-For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$ \footnote{Note that $P_\textit{RNA-loss}$ is a conditional probability dependent on the presence of RNA.}
+For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$.
+(Note that $P_\textit{RNA-loss}$ is a conditional probability dependent on the presence of RNA).
 
@@
 
+Next, we tried to find out if some amino acid substitutions are more probable, than the others.
+To this end, we calculated the gain probability ($G$) for every amino acid.
+Specifically for every amino acid $i$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
+More precisely, the total probability that an amino acid $j$ will mutate to an amino acid $i$ is the product of the expected frequency of amino acid $j$ ($P_j$), and the substitution rate from $i$ to $j$ ($\mu_{j\to i}$).
+The gain probability for amino acid $i$ is thus defined as:
+
+$$G_i = \sum_{j \neq i} P_j \times \mu_{j\to i}$$
+
+We found that amino gain probabilities were also non-uniform across different amino acids but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
+
+Next, we calculated how easily is an amino acid mutated to any other amino acid. Specifically, the conditional loss probability ($L$), given the amino acid already exists, is defined as:
+
+$$L_i = \sum_{j \neq i} \mu_{i\to j}$$
+
+The loss probability did not significantly correlate with the expected frequency (Pearson $\rho = -0.255, P = 0.2818$).
+
 Previous studies have argued that hydrophobic amino acids are comprise 40\% of amino acids and can also easily arise due to mutation pressure.
@@
 We asked if our model also makes similar predictions.
-To this end we calculated expected hydropathy in a protein sequence, based on expected frequency of different amino acids (\autoref{aafreq}) and their hydropathic index values (Kyte and Doolittle).
+To this end we calculated expected average hydropathy ($\bar{\alpha}$) of a protein sequence, based on expected frequency of different amino acids ($P_i$, \autoref{aafreq}) and their hydropathic index values ($\alpha$, Kyte and Doolittle).
+Specifically:
+
+$$\bar{\alpha} = \sum_{i} P_i \alpha_i$$
+
 We found that the expected hydropathy is $-0.08628$, which suggests that random protein sequences are not mostly hydrophobic.
 
-Next, we calculated the probability that any
-
-\begin{itemize}
-\item expected hydropathy index = $-0.08628$
-\item acid to basic rate is $1.3074\times$ higher than \textit{vice versa}. Conditional prob = $1.80543$
-\item mutations do not preferentially create hydrophobic amino acid. Total prob ratio = 1.025, Conditional prob ratio = 0.971
-\end{itemize}
+Next, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$).
+We defined hydrophobic amino acids as those with hydropathy index greater than 0 (i.e A, C, F, I, L, M \& V).
+Likewise, we calculated the probability that any mutation causes a hydrophobic to non-hydrophobic amino acid substitution ($P_{\alpha\textit{-loss}}$).
+Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it was 1.053.
+This shows that random mutations do not have a strong bias towards causing hydrophobic amino acids to accumulate in protein sequence.
 
@@
 \centering
-\includegraphics[scale=0.6]{Figures/pAAfreq.pdf}
+\begin{subfigure}{0.49\textwidth}
+\caption{}
+\centering
+\includegraphics[scale=0.5]{Figures/pAAfreq.pdf}
+\end{subfigure}\hfill%
+\begin{subfigure}{0.49\textwidth}
+\caption{}
+\centering
+\includegraphics[scale=0.5]{Figures/pToAA.pdf}
+\end{subfigure}
+
+\begin{subfigure}{0.49\textwidth}
+\caption{}
+\centering
+\includegraphics[scale=0.5]{Figures/pFrAA.pdf}
+\end{subfigure}
+
 \caption{ORF loss is more probable than transcription loss, for \textit{de novo} genes containing more than 33 codons.

### 2022-10-28 Modeling/DeNovoProb.jl

@@ -1,8 +1,7 @@
-using Base.Iterators
 using Combinatorics
 using Plots
 using Measures
-using LaTeXStrings
+using HypothesisTests
 
 include("nucleotidefuncts.jl")
 
@@ -211,7 +210,7 @@ acidic = indexin(["D","E"],aas);
 basic = indexin(["K","R"],aas);
 hydrophobic = hydropathy .>0;
 
-pHydrophobic = sum(aaprob[hydrphobic])
+pHydrophobic = sum(aaprob[hydrophobic]);
 
 plot_aafreq = bar(aas,aaprob2,
     xticks = (0.5:20, aas),
@@ -236,23 +235,36 @@ for i = 1:20
         end
 end
 
-aasubprob_conditional = aasubprob./aaprob2
+aasubhighprob = aasubprob.*(aasubprob .>mutrate^2)
 
-asymmetric_transitions = log2.(aasubprob./aasubprob');
+aasubprob_conditional = aasubhighprob./aaprob2;
 
+asymmetric_transitions = log2.(aasubhighprob./aasubhighprob');
 asymmetric_conditional = log2.(aasubprob_conditional./aasubprob_conditional');
 
-codonshydrophobic = gencode[(in).(gencode[:,3],Ref(aas[hydrophobic])),1]
+# c = kmeans(asymmetric_transitions,3);
+# idx = sortperm(assignments(c));
+hcl = hclust(aasubhighprob .+ aasubhighprob');
+idx = hcl.order;
 
-codonshydrophilic = kmers(3)[(!in).(kmers(3),Ref(codonshydrophobic))]
+gainAA = Vector{Float64}(vec(sum(aasubprob, dims=1)'));
+lossAA = Vector{Float64}(vec(sum(aasubprob, dims=2)./aaprob2));
+stayAA = [sum(featurestay(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];
+
+
+codonshydrophobic = gencode[(in).(gencode[:,3],Ref(aas[hydrophobic])),1];
+
+codonshydrophilic = kmers(3)[(!in).(kmers(3),Ref(codonshydrophobic))];
 
 toHydrophobic = featuregain(codonshydrophilic,codonshydrophobic, gccontent);
 
 frHydrophobic = featuregain(codonshydrophobic,codonshydrophilic, gccontent);
 
-plot_aasubprob = heatmap(aas,aas,log10.(aasubprob./aaprob), 
-    xticks = (0.5:20, aas), 
[…]

### 2022-10-28 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 To this end, we calculated the gain probability ($G$) for every amino acid.
-Specifically for every amino acid $i$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
-More precisely, the total probability that an amino acid $j$ will mutate to an amino acid $i$ is the product of the expected frequency of amino acid $j$ ($P_j$), and the substitution rate from $i$ to $j$ ($\mu_{j\to i}$).
-The gain probability for amino acid $i$ is thus defined as:
+Specifically for every amino acid $a$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
+More precisely, the total probability that an amino acid $b$ will mutate to an amino acid $a$ is the product of the expected frequency of amino acid $b$ ($P_b$), and the substitution rate from $a$ to $b$ ($\mu_{b\to a}$).
+The gain probability for amino acid $a$ is thus defined as:
 
-$$G_i = \sum_{j \neq i} P_j \times \mu_{j\to i}$$
+$$G_a = \sum_{b \neq a} P_b \times \mu_{b\to a}$$
 
@@
 
-$$L_i = \sum_{j \neq i} \mu_{i\to j}$$
+$$L_a = \sum_{b \neq a} \mu_{a\to b}$$
 
 The loss probability did not significantly correlate with the expected frequency (Pearson $\rho = -0.255, P = 0.2818$).
+
+Next, we investigated if some amino acid substitutions are more common than the others.
+As expected, amino acid substitutions that require more than one nucleotide change were much less probable than those needing just a single nucleotide change.
+Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutation, but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities).
+Although double nucleotide mutations are less probable than single mutations, the mutation bias makes causes different single mutations to occur with different probabilities.
+Thus it is very likely that an amino acid substitution (for example K$\to$E) may not be as probable as the reverse substitution (E$\to$K).
+To identify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{ab}$, for every pair of amino acids, $a$ and $b$.
+A positive value of $D_{ab}$ means that $a$ to $b$ substitution is more likely than $B$ to $a$ substitution, and \textit{vice versa}.
+
+$$ \textit{D}_{ab} = \log_2\left(\frac{\mu_{a\to b}}{\mu_{b\to a}}\right)$$
+
+We restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
+We found that most substitutions indeed had a preferred direction $\textit{D}_{ab} \neq 0$.
+The median absolute value of $\textit{D}_{ab}$ was 0.905 and the maximum value was 2.836.
+This means that for 50\% of the pairs we analysed, one substitution direction was 1.87 times more probable than the other.
+For example, E$\to$K is 1.8 times more probable than K$\to$E.
+This may appear as a small number but many such directional substitutions occurring in an evolving protein could indicate a directional selection.
 

### 2022-10-28 Modeling/DeNovoProb.jl

@@ -272,9 +272,9 @@ plot_aasubprob = heatmap(aas[idx],aas[idx],log2.(aasubhighprob[idx,idx]),
 
 savefig(plot_aasubprob, figdir*"pAASub.pdf")
 
-plot_aasymprob = heatmap(aas[idx],aas[idx],asymmetric_transitions[idx,idx], 
-    xticks = (0.5:20, aas[idx]), 
-    yticks = (0.5:20,aas[idx]),
+plot_aasymprob = heatmap(aas,aas,asymmetric_transitions, 
+    xticks = (0.5:20, aas), 
+    yticks = (0.5:20,aas),
     size = (width = cm2pt(15), height = cm2pt(14)),
     colorbar_title = "Substitution likelihood",
     legend = :bottom,

### 2022-11-04 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

--- 2022-10-28 Manuscripts/deNovoRates.tex
@@
 
-Next, we tried to find out if some amino acid substitutions are more probable, than the others.
-To this end, we calculated the gain probability ($G$) for every amino acid.
-Specifically for every amino acid $i$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
-More precisely, the total probability that an amino acid $j$ will mutate to an amino acid $i$ is the product of the expected frequency of amino acid $j$ ($P_j$), and the substitution rate from $i$ to $j$ ($\mu_{j\to i}$).
-The gain probability for amino acid $i$ is thus defined as:
+Next, we aimed to find out if some amino acid substitutions are more probable, than the others.
+To this end, we calculated the gain probability (\autoref{aagaineq}) for every amino acid.
+Specifically for every amino acid $x$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
+More precisely, the total probability that an amino acid $y$ will mutate to an amino acid $x$ is the product of the expected frequency of amino acid $y$ ($P_y$), and the substitution rate from $x$ to $y$ ($\mu_{y\to x}$).
+The gain probability for an amino acid $x$ is then defined as:
 
-$$G_i = \sum_{j \neq i} P_j \times \mu_{j\to i}$$
+\begin{equation}
+G_x = \sum_{b \neq a} P_y \times \mu_{y\to x}
+\label{aagaineq}
+\end{equation}
 
@@
 
-Next, we calculated how easily is an amino acid mutated to any other amino acid. Specifically, the conditional loss probability ($L$), given the amino acid already exists, is defined as:
+Next, we calculated how easily is an amino acid mutated to any other amino acid.
+Specifically, the conditional loss probability for an amino acid ($L_x$), given the amino acid already exists, is defined as:
 
-$$L_i = \sum_{j \neq i} \mu_{i\to j}$$
+\begin{equation}
+L_x = \sum_{b \neq a} \mu_{x\to y}
+\end{equation}
 
-The loss probability did not significantly correlate with the expected frequency (Pearson $\rho = -0.255, P = 0.2818$).
+The loss probability was also not uniformly distributed but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
 
-Previous studies have argued that hydrophobic amino acids are comprise 40\% of amino acids and can also easily arise due to mutation pressure.
+Next, we investigated if some amino acid substitutions are more common than the others.
+As expected, amino acid substitutions that require more than one nucleotide change were much less probable than those needing just a single nucleotide change.
+Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutation, but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities).
+Although single nucleotide mutations are more probable than double mutations, the different single mutations do not occur at the same rate due to mutation rate bias.
+Thus an amino acid substitution (for example K$\to$E) may not be as likely to occur as the reverse substitution (E$\to$K).
+To verify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{xy}$, for every pair of amino acids, $x$ and $y$.
+A positive value of $D_{xy}$ means that $x$ to $y$ substitution is more likely than $y$ to $x$ substitution, and \textit{vice versa}.
+
+\begin{equation}
+\textit{D}_{xy} = \log_2\left(\frac{\mu_{x\to y}}{\mu_{y\to x}}\right)
+\end{equation}
+
+As we mentioned in previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
+We found that most substitutions indeed had a preferred direction $\textit{D}_{xy} \neq 0$.
+The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836.
+That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other.
+For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution.
+This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection.
+We also performed a supporting analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}).
+
+Previous studies have argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure.
 They based these conclusions on mutation rate bias estimated in bacteria.
 We asked if our model also makes similar predictions.
-To this end we calculated expected average hydropathy ($\bar{\alpha}$) of a protein sequence, based on expected frequency of different amino acids ($P_i$, \autoref{aafreq}) and their hydropathic index values ($\alpha$, Kyte and Doolittle).
-Specifically:
+To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic.
+We classified an amino acid hydrophobic if its hydropathy index (Kyte and Doolittle) is greater than 0 (i.e A, C, F, I, L, M \& V).
+Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic.
+Next, we asked if a protein sequence in general tends to be hydrophobic.
+Specifically, we calculated expected hydropathy ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydropathic index values ($\alpha_x$).
 
-$$\bar{\alpha} = \sum_{i} P_i \alpha_i$$
+\begin{equation}
+\bar{\alpha} = \sum_{x} P_x \alpha_x
+\label{expHydropathy}
+\end{equation}
 
-We found that the expected hydropathy is $-0.08628$, which suggests that random protein sequences are not mostly hydrophobic.
+We found that the expected hydropathy is $-0.08628$, which suggests that random protein sequences are not generally hydrophobic.
 
-Next, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$).
-We defined hydrophobic amino acids as those with hydropathy index greater than 0 (i.e A, C, F, I, L, M \& V).
-Likewise, we calculated the probability that any mutation causes a hydrophobic to non-hydrophobic amino acid substitution ($P_{\alpha\textit{-loss}}$).
+Next, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$, \autoref{alphagain}).
+Likewise, we calculated the probability that any mutation causes a hydrophobic to non-hydrophobic amino acid substitution ($P_{\alpha\textit{-loss}}$, \autoref{alphaloss}).
 Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it was 1.053.
 This shows that random mutations do not have a strong bias towards causing hydrophobic amino acids to accumulate in protein sequence.
+
+\begin{align}
+P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\ |\ \alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\ |\ \alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
+%
+P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\ |\ \alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\ |\ \alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
+\end{align}
 

### 2022-11-04 Modeling/DeNovoProb.jl

@@ -244,8 +244,8 @@ asymmetric_conditional = log2.(aasubprob_conditional./aasubprob_conditional');
 
 # c = kmeans(asymmetric_transitions,3);
 # idx = sortperm(assignments(c));
-hcl = hclust(aasubhighprob .+ aasubhighprob');
-idx = hcl.order;
+# hcl = hclust(aasubhighprob .+ aasubhighprob');
+# idx = hcl.order;
 
 gainAA = Vector{Float64}(vec(sum(aasubprob, dims=1)'));
 lossAA = Vector{Float64}(vec(sum(aasubprob, dims=2)./aaprob2));

### 2022-11-04 Manuscripts/deNovoRates.tex
new file:   Manuscripts/refs.bib

@@
 \usepackage{amsmath,amssymb,graphicx,setspace,palatino, subcaption}
-\usepackage[hidelinks]{hyperref}
+\usepackage[colorlinks, allcolors = blue]{hyperref}
+\usepackage{natbib}
+\citestyle{egu}
 
@@
 Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost via mutagenesis.
-Using a similar approach, we estimate some fundamental biochemical properties of proteins that arise from random DNA sequences.
-Our models thus define a null hypothesis under which DNA sequences evolve neutrally through mutation pressure alone.
+These probability models are essentially described by two kinds of probability.
+First is the probability of finding a DNA sequence feature (such as TATA box or an ORF).
+This probability depends on the nucleotide distribution given by the GC content.
+The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another.
+Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC content.
+
+For all calculations we used a GC content of 0.42 which is reasonably close to the global GC content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes.
+We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
+We used mutation bias data from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
+shows the values of different mutation rates.
+
+Using a similar approach, we predict how the nucleotide distribution and random mutations shape protein composition and evolution.
+
+Overall our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.
 A significant deviation from this null hypothesis may be interpreted as evidence in support of natural selection.
@@
 \subsection{How many times can a gene be lost?}
+
+\begin{figure}[!b]
+\hrule
+\vspace{1ex}
+\centering
+\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
+\caption{Genes are more likely to be independently lost twice, than being born once.
+The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e.
+$\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the number of codons in the ORF.
+\autoref{genegaineq} describes total gene gain probability.
+Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
+Using the conditional probability of gene gain does not change these results.}
+\label{gainlossprob}
+\end{figure}
 
@@
 We found that for all ORF lengths, the gene can be lost at least two times independently, in the time required to gain it (\autoref{gainlossprob}).
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
-\caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e.
-$\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF.
-\autoref{genegaineq} describes total gene gain probability.
-Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
-Using the conditional probability of gene gain does not change these results.}
-\label{gainlossprob}
-
-\vspace{1ex}
-\hrule
-\end{figure}
 
@@
 
+\begin{align}
+P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
+P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
+\end{align}
+
 With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}\ \%$).
@@
 Overall, our analysis suggests that \textit{de novo} genes with long ORFs are more likely arise from DNA regions that are already transcribed (for example, non-coding RNAs).
-
-\begin{align}
-P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
-P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
-\end{align}
 
@@
 
+\begin{figure}[!t]
+\centering
+\begin{subfigure}{0.47\textwidth}
+\caption{}
+\flushleft
+\includegraphics[scale=0.5]{Figures/pAAfreq.pdf}
+\end{subfigure}\hfill%
+\begin{subfigure}{0.47\textwidth}
+\caption{}
+\flushleft
+\includegraphics[scale=0.5]{Figures/pToAA.pdf}
+\end{subfigure}
+
+\vspace{1em}
+
+\begin{subfigure}[t]{0.47\textwidth}
+\caption{}
+\flushleft
+\includegraphics[scale=0.5]{Figures/pFrAA.pdf}
+\end{subfigure}\hfill%
+\begin{subfigure}[t]{0.47\textwidth}
+\caption{}
+\flushright
+\includegraphics[scale=0.5]{Figures/pAAsubSymCond.pdf}
+\end{subfigure}
+
+\caption{Mutation bias and GC content shape protein composition.
+Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes).
+The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one.
+Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis).
+Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.}
+\label{aafreq}
+
+\vspace{1ex}
+\hrule
+\end{figure}
+
 In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss.
@@
 Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids.
-On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}).
+On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}A).
 This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
@@
 For example, leucine, arginine and serine, all have six codons each.
-However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}).
+However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}A).
 
@@
 
-We found that amino gain probabilities were also non-uniform across different amino acids but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
+We found that amino gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}B) but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
 
@@
 L_x = \sum_{b \neq a} \mu_{x\to y}
+\label{aalosseq}
 \end{equation}
 
-The loss probability was also not uniformly distributed but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
+The loss probability was also not uniformly distributed (\autoref{aafreq}C) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
 
@@
 Thus an amino acid substitution (for example K$\to$E) may not be as likely to occur as the reverse substitution (E$\to$K).
-To verify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{xy}$, for every pair of amino acids, $x$ and $y$.
-A positive value of $D_{xy}$ means that $x$ to $y$ substitution is more likely than $y$ to $x$ substitution, and \textit{vice versa}.
+To verify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{x\to y}$, for every pair of amino acids, $x$ and $y$.
+A positive value of $D_{x\to y}$ means that $x$ to $y$ substitution is more likely than $y$ to $x$ substitution, and \textit{vice versa}.
 
 \begin{equation}
-\textit{D}_{xy} = \log_2\left(\frac{\mu_{x\to y}}{\mu_{y\to x}}\right)
+\textit{D}_{x\to y} = \log_2\left(\frac{\mu_{x\to y}}{\mu_{y\to x}}\right)
+\label{directionality}
 \end{equation}
@@
 As we mentioned in previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
-We found that most substitutions indeed had a preferred direction $\textit{D}_{xy} \neq 0$.
+We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}D).
 The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836.
@@
 This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection.
-We also performed a supporting analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}).
+We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}).
+Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that the median value is 0.38.
+Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.
 
@@
 Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic.
+This finding is in agreement with the previous studies.
 Next, we asked if a protein sequence in general tends to be hydrophobic.
@@
 
-Next, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$, \autoref{alphagain}).
+To find out if random mutations indeed cause hydrobphobic amino acids to occur in protein sequences, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$, \autoref{alphagain}).
 Likewise, we calculated the probability that any mutation causes a hydrophobic to non-hydrophobic amino acid substitution ($P_{\alpha\textit{-loss}}$, \autoref{alphaloss}).
-Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it was 1.053.
-This shows that random mutations do not have a strong bias towards causing hydrophobic amino acids to accumulate in protein sequence.
 
@@
 
-\begin{figure}[!t]
-\centering
-\begin{subfigure}{0.49\textwidth}
-\caption{}
-\centering
-\includegraphics[scale=0.5]{Figures/pAAfreq.pdf}
-\end{subfigure}\hfill%
-\begin{subfigure}{0.49\textwidth}
-\caption{}
-\centering
-\includegraphics[scale=0.5]{Figures/pToAA.pdf}
-\end{subfigure}
-
-\begin{subfigure}{0.49\textwidth}
-\caption{}
-\centering
-\includegraphics[scale=0.5]{Figures/pFrAA.pdf}
-\end{subfigure}
-
-\caption{ORF loss is more probable than transcription loss, for \textit{de novo} genes containing more than 33 codons.
-The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
-\label{aafreq}
-
-\vspace{1ex}
-\hrule
-\end{figure}
+Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it was 1.053.
+This shows that random mutations do not preferentially cause hydrophobic amino acids to accumulate in protein sequence.
 
@@
 
+\section{Materials and Methods}
+
+\bibliographystyle{plainnat}
+\bibliography{refs}
+
 \end{document}

### 2022-11-04 Modeling/DeNovoProb.jl

@@ -214,9 +214,10 @@ pHydrophobic = sum(aaprob[hydrophobic]);
 
 plot_aafreq = bar(aas,aaprob2,
     xticks = (0.5:20, aas),
+    yticks = (0:0.03:0.9),
     fill = :black,
     xlabel = "Amino acid",
-    ylabel = "Expected frequency",
+    ylabel = "Expected frequency\n(Normalized)",
     size = (width = cm2pt(15), height = cm2pt(11))
 );
 savefig(plot_aafreq, figdir*"pAAfreq.pdf")
@@ -299,9 +300,10 @@ savefig(plot_aasymprob2, figdir*"pAAsubSymCond.pdf")
 
 plot_toaa = bar(aas,gainAA./sum(gainAA),
     xticks = (0.5:20, aas),
+    yticks = (0:0.03:0.9),
     fill = :black,
     xlabel = "Amino acid",
-    ylabel = "Gain Probability",
+    ylabel = "Gain Probability\n(Normalized)",
     size = (width = cm2pt(15), height = cm2pt(11))
 );
 
@@ -311,7 +313,7 @@ plot_fraa = bar(aas,lossAA./sum(lossAA),
     xticks = (0.5:20, aas),
     fill = :black,
     xlabel = "Amino acid",
-    ylabel = "Loss Probability",
+    ylabel = "Loss Probability\n(Normalized)",
     size = (width = cm2pt(15), height = cm2pt(11))
 );
 

### 2022-11-07 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino, subcaption}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs}
 \usepackage[colorlinks, allcolors = blue]{hyperref}
@@
 \renewcommand{\thesubfigure}{\Alph{subfigure}}
+
+\renewcommand{\subsectionautorefname}{Section}
+\renewcommand{\subsubsectionautorefname}{Section}
 
@@
 Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence.
+[here a rationale for this choice].
 For translation, we only require the gene to have an open reading frame (ORF).
+
 Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost via mutagenesis.
@@
 The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another.
-Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC content.
+Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.
 
-For all calculations we used a GC content of 0.42 which is reasonably close to the global GC content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes.
+For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes.
 We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
@@
 \begin{equation}
-L_x = \sum_{b \neq a} \mu_{x\to y}
+L_x = \sum_{x \neq y} \mu_{x\to y}
 \label{aalosseq}
@@
 To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic.
-We classified an amino acid hydrophobic if its hydropathy index (Kyte and Doolittle) is greater than 0 (i.e A, C, F, I, L, M \& V).
+We classified an amino acid hydrophobic if its hydropathy index \citep{hydropathy} is greater than 0 (i.e A, C, F, I, L, M \& V).
 Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic.
@@
 
+\subsection{Mutation probabilities}
+
+We calculated nucleotide substitution probabilities based on mutation rate and mutation rate bias data.
+Specifically, we  used a mutation rate ($\lambda$) of $7.8\times10^{-9}$ mutations per nucleotide position per generation \citep{drosophilamutrate}.
+We derived our mutation bias parameters from two published studies, the first on \textit{Drosophila melanogaster} \citep{drosophilamutrate}, and the second on humans \citep{humanmutrate}.
+\autoref{mutbias} shows the exact values of mutation bias probabilities that we used in this study.
+
+\begin{table}[H]
+\centering
+\begin{tabular}{c c}
+\toprule
+\textbf{Substitution} & Probability($\mu$) \\\midrule
+\texttt{A:T}$\to$\texttt{T:A} & 0.056 \\\midrule
+\texttt{A:T}$\to$\texttt{G:C} & 0.243 \\\midrule
+\texttt{A:T}$\to$\texttt{C:G} & 0.074 \\\midrule
+\texttt{G:C}$\to$\texttt{A:T} & 0.483 \\\midrule
+\texttt{G:C}$\to$\texttt{T:A} & 0.075 \\\midrule
+\texttt{G:C}$\to$\texttt{C:G} & 0.069 \\\bottomrule
+\end{tabular}
+\caption{Mutation bias probabilities.}
+\label{mutbias}
+\end{table}
+
+\subsection{Probabilities of finding, gaining, and losing a DNA sequence}
+
+\subsubsection{Probability of finding a DNA sequence}
+
+We calculated the probability of finding a DNA sequence based on global nucleotide frequency distributions, given by the GC-content.
+We used a GC content of 0.42 for our calculations.
+Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is: $S = 0.5\times$GC-content.
+The probability of finding an \texttt{A} or a \texttt{T} is: $W = 1 - 0.5\times$GC-content.
+Using these values, we calculated the probability of finding a DNA sequence motif by chance.
+For example, the probability of finding the sequence \texttt{ATG} would be: $W\times W \times S$.
+
+\subsubsection{Probability of gaining a DNA sequence}
+\label{methgain}
+
+We calculated the probability of gaining a DNA sequence due to mutations using GC-content, mutation rate and mutation bias.
+Specifically, we calculated the probability that a DNA sequence does not exist, and it emerges due to specific nucleotide mutations.
+More precisely, this probability is the product of two other probabilities.
+The first is the probability of finding a DNA sequence ($x$) that is not the sequence of interest.
+The second probability is that this sequence $x$ mutates to the sequence of interest.
+To explain this calculation better, we use the example of \texttt{CTA} mutating to \texttt{ATG}.
+The first probability i.e.
+the probability of finding \texttt{CTA} by chance is $SW^2$.
+\texttt{CTA} mutates to \texttt{ATG} via two nucleotide mutations (\texttt{C}$\to$\texttt{A} and \texttt{A}$\to$\texttt{G}).
+Thus the probability of this DNA change would be the probability of two nucleotide mutations ($\lambda^2$) multiplied by two mutation bias probabilities (\texttt{G:C}$\to$\texttt{T:A} and \texttt{A:T}$\to$\texttt{G:C}).
+Overall, the chance of \texttt{CTA} mutating to \texttt{ATG} would be:
+
+$$SW^2 \times \lambda^2 \times \mu_{\texttt{G:C}\to\texttt{T:A}} \times \mu_{\texttt{A:T}\to\texttt{G:C}}$$
+
+Next, we calculated the probability that every nucleotide triplet that is not \texttt{ATG}, mutates to \texttt{ATG}.
+This can happen via one, two, or three nucleotide mutations.
+The sum of all these mutation probabilities is the probability of \texttt{ATG} gain.
+
+Using the same principle we calculated the gain probability of any DNA sequence motif (of any length or composition).
+We excluded insertions and deletions as a mechanism of gain of small DNA sequences that we analysed in this study.
+
+\subsubsection{Probability of losing a DNA sequence}
+
+We calculated the loss of a DNA sequence motif using the same principle we used for calculating gain probabilities.
+However, we defined loss probability as a conditional probability i.e.
+we assume that the DNA sequence of interest already exists in the genome.
+For example, the loss probability of a specific \texttt{ATG} sequence would be the sum of probabilities of \texttt{ATG} mutating to any of the other 63 nucleotide triplets (via one, two, or three nucleotide mutations).
+We use conditional loss probabilities by default, because usually one is interested in finding out how quickly can an existing DNA sequence erode.
+
+We used this method to calculate the loss probability of any DNA sequence motif, and we excluded insertions and deletions from this calculation.
+
+\subsection{Probabilities of finding, gaining, and losing DNA features}
+
+Usually, a specific function is encoded in DNA by several DNA sequences.
+For example, translation stop is encoded by three codons (\texttt{TGA}, \texttt{TAG}, \texttt{TAA}).
+We use the term DNA features to mean a set of DNA sequences that are associated with the same function.
+For every such DNA feature set, there is a complementary set of non-features i.e DNA sequences that are not associated with the feature's function.
+For example, the non-feature set of stop codons would be all the other 61 codons.
+
+The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set.
+
+The probability that a DNA feature is gained via mutations, is the sum of probabilities of every non-feature sequence mutating to any feature sequence.
+If $F$ denotes the feature set, and $\mu_{y\to x}$ denotes the probability of a DNA sequence $y$ mutating to a DNA sequence $x$ (see \autoref{methgain}), then:
+
+\begin{equation}
+P_\textit{feature-gain} = \sum_{x \in F} \sum_{y \notin F} P_y \times \mu_{y\to x}
+\end{equation}
+
+The probability that a DNA feature is lost via mutations is a conditional probability that given a feature exists, it mutates to any of the non-feature sequences.
+
+\begin{equation}
+P_\textit{feature-loss} = \frac{\displaystyle\sum_{x \in F} \sum_{y \notin F} P_x \times \mu_{x\to y}}{\displaystyle\sum_{x \in F} P_x}
+\end{equation}
+
+Because a feature set usually has many DNA sequences, a mutation can change a feature sequence such that the resulting sequence is also a part of the feature set.
+Thus we defined the probability ($P_\textit{feature-stay}$) that a feature does not erode because of mutations.
+Specifically, it is the the sum of two probabilities.
+First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where mutated sequence is still a part of the feature set.
+
+\begin{equation}
+P_\textit{feature-stay} = P_0 + \sum_{x \in F} \sum_{\genfrac{}{}{0pt}{1}{y \in F}{y\neq x}} \mu_{x\to y}
+\end{equation}
+
+The probability that no mutation occurs ($P_0$) in a DNA sequence of length $k$ is described by Poisson distribution.
+
+\vspace{-1ex}
+
+$$P_0 = 1-e^{-k\lambda}$$
+
+Because the mutation rate is biased (\autoref{mutbias}), the probability that a mutation does not occur in a DNA sequence depends on the composition of the sequence.
+
+The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is described by the mutation bias (\autoref{mutbias}) as:
+
+\vspace{-1ex}
+
+$$\lambda_\texttt{AT} = 2\times\lambda\times(\mu_{\texttt{A:T}\to\texttt{T:A}} + \mu_{\texttt{A:T}\to\texttt{G:C}} + \mu_{\texttt{A:T}\to\texttt{C:G}})$$
+
+Likewise, the probability that a \texttt{G} or a \texttt{C} mutates ($\lambda_\texttt{GC}$) is:
+
+\vspace{-1ex}
+
+$$\lambda_\texttt{GC} = 2\times\lambda\times(\mu_{\texttt{G:C}\to\texttt{A:T}} + \mu_{\texttt{G:C}\to\texttt{T:A}} + \mu_{\texttt{G:C}\to\texttt{C:G}})$$
+
+(Note that the general mutation rate, $\lambda$, is an average of $\lambda_\texttt{AT}$ and $\lambda_\texttt{GC}$.)
+
+\vspace{1\baselineskip}
+
+Thus the probability that a sequence of length $k$, containing $m$ number of \texttt{A} and \texttt{T}, does not mutate is:
+
+$$P_0 = (1-e^{-m\lambda_\texttt{AT}})\times(1-e^{-(k-m)\lambda_\texttt{GC}})$$
+
+\subsection{Probabilities of finding, gaining, and losing transcription}
+
+To calculate the probabilities of finding, gaining, and losing transcription, we first defined the minimum requirements for transcription.
+These requirements are a promoter, either a TATA-box or an Initiator (Inr) element, and a polyadenylation signal.
+Each of these three DNA sequence motifs consists of more than one DNA sequence.
+
+\subsubsection{Definition of TATA-box, Inr and poly-A signal}
+
+\paragraph{TATA box:}
+
+The consensus sequence of TATA-box is \texttt{TATA [T/A] A [T/A] [G/A]} \citep{tata2}.
+Square brackets mean that, any of the nucleotides enclosed by them, can exist in that specific position.
+Another study used X-ray crystallography to identify mutations that are tolerated in TATA-box consensus sequence \cite{tata1}.
+Thus there are eight consensus sequences, and at seven positions, a non-consensus nucleotide is tolerated (\autoref{tatatab}).
+This makes a TATA-box feature set that contains 136 sequences.
+We defined the corresponding non-feature set by generating all 8-mer DNA sequences that do not belong to the TATA-box feature set.
+
+\begin{table}[!t]
+\centering
+\begin{tabular}{@{}ccc@{}}
+\toprule
+Position  & Accepted nucleotides & \begin{tabular}[c]{@{}c@{}}Number of Variants\\ (consensus + suboptimal)\end{tabular} \\ \midrule
+1         & T$\gg$C$>$A,G        & $1+3$                                                                                 \\ \midrule
+2         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
+3         & T$\gg$A,C            & $1+2$                                                                                 \\ \midrule
+4         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
+5         & T/A                  & $2+0$                                                                                 \\ \midrule
+6         & A$\gg$G$>$C/T        & $1+3$                                                                                 \\ \midrule
+7         & A/T$>$G$>$C          & $2+2$                                                                                 \\ \midrule
+8         & G/A$>$C/T            & $2+2$                                                                                                                                                                 \\ \bottomrule
+\end{tabular}
+\caption{Nucleotide residues that are tolerated at different positions in a TATA-box \citep[based on][]{tata1}}
+\label{tatatab}
+\end{table}
+
+\paragraph{Inr:} The consensus sequence of Inr is \texttt{[T/G/C] [T/G/C] CA [T/G/C] [T/A]} \citep{Inr}, which gives 54 sequences in the Inr feature set.
+We generated the corresponding non-feature set by taking all DNA 6-mers that are distinct from the 54 Inr sequences.
+
+\paragraph{poly-A signal:} There are four known poly-A signal variants -- \texttt{AATAAA}, \texttt{ATTAAA}, \texttt{AGTAAA}, and \texttt{TATAAA} that compose the poly-A signal feature set \citep{polyA}.
+The non-feature set consists of all other DNA 6-mers.
+
+We calculated probabilities of finding, gaining and losing these features as we described in the previous section.
+
+\subsubsection{Definition of the probabilities}
+
+Probability that a DNA region is transcribed by chance ($P_\textit{RNA}$), is the probability of finding a poly-A signal multiplied by the sum of probabilities of finding a TATA-box, and an Inr element.
+
+\begin{equation}
+P_\textit{RNA} = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr})
+\end{equation}
+
+Gain of transcription can happen via two mechanisms.
+In the first mechanism a promoter (TATA-box or Inr) already exists, and mutations cause a poly-A signal to emerge ($P_\textit{polyA-gain}$) but do not destroy the promoter ($P_\textit{TATA-stay}$,  $P_\textit{Inr-stay}$).
+In the second mechanism, a poly-A signal exists and the promoter emerges due to mutations ($P_\textit{TATA-gain}$,  $P_\textit{Inr-gain}$), while the poly-A signal remains undisrupted ($P_\textit{polyA-stay}$).
+Thus the total probability of transcription gain ($P_\textit{RNA-gain}$) would be the sum of the probabilities of the two mechanisms.
+
+\begin{equation}
+P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
+\end{equation}
+
+It is also possible that both the promoter and the poly-A signal were initially absent, and emerge via mutations. However, this probability is very small and is negligible.
+
+Transcription is lost when either the promoter is lost or when the poly-A signal is lost.
+Since two kinds of promoters can initiate transcription, there can be three kinds of transcription units.
+First that has only TATA-box, second that has only Inr, and the third that has both these promoters.
+All these three transcription units must also have a poly-A signal.
+To calculate the conditional probability of promoter loss, we calculated the fraction of transcription units ($Q$) that belong to each of these three classes.
+For example, the fraction of transcription units that are TATA-only is ($Q_\textit{TATA-only}$):
+
+$$Q_\textit{TATA-only} = \frac{P_\textit{TATA}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})}$$
+
+Here, the numerator denotes probability of finding a TATA-box but not an Inr, and the denominator denotes the probability of finding a promoter that can be a TATA-box, Inr, or both.
+Likewise, the fraction of Inr-only ($Q_\textit{Inr-only}$) and TATA-Inr promoters ($Q_\textit{both}$) are given by:
+
+\vspace{-1em}
+
+\begin{align*}
+Q_\textit{Inr-only} & = \frac{P_\textit{Inr}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})} \\[1em]
+Q_\textit{both} & = 1 - Q_\textit{TATA-only} -Q_\textit{Inr-only}
+\end{align*}
+
+For a TATA-only transcription unit to lose a promoter, it should lose the TATA-box but simultaneously not gain an Inr.
+Likewise, an Inr-only transcription unit will lose a promoter when it loses the Inr but not gain a TATA-box.
+A transcription unit that has both the promoters must lose both of them.
+For all the three kinds of transcription units, transcription would be lost if poly-A signal is lost.
+
+The conditional probability of transcription loss ($P_\textit{RNA-loss}$), given that transcription exists, is thus defined as:
+
+\begin{align}
+P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1ex]
+ & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1ex]
+ & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
+\end{align}
+
 \bibliographystyle{plainnat}

### 2022-11-07 Modeling/DeNovoProb.jl

@@ -31,55 +31,55 @@ function rnavals(gccontent)
 [degen(s,7,nucs) for s in tataboxes],
 [degen(s,8,nucs) for s in tataboxes],
         ])
-    ))
+    ));
 
-    notata = setdiff(kmers(8),tatavars)
+    notata = setdiff(kmers(8),tatavars);
 
-    tataprob = sum([nucprob(x,gccontent) for x in tatavars])
-    tatagain = featuregain(notata,tatavars,gccontent)
-    tataloss1 = featuregain(tatavars,notata,gccontent)
-    tataloss2 = tataloss1/tataprob
-    tatastay = featurestay(tatavars,gccontent)
-    notatastay = 1 - tatagain - tataprob
+    tataprob = sum([nucprob(x,gccontent) for x in tatavars]);
+    tatagain = featuregain(notata,tatavars,gccontent);
+    tataloss1 = featuregain(tatavars,notata,gccontent);
+    tataloss2 = tataloss1/tataprob;
+    tatastay = featurestay(tatavars,gccontent);
+    notatastay = 1 - tatagain;
 
     ## Initiator element
     # Consensus sequence = "BBCABW";  DOI:10.1101/gad.295980.117
 
-    mers6 = kmers(6)
-    inrvars = vec(join.(product("TGC","TGC","C", "A","TGC","TA")))
-    noinrs = setdiff(mers6, inrvars)
-    inrprob = sum([nucprob(x,gccontent) for x in inrvars])
-    inrgain = featuregain(noinrs,inrvars,gccontent)
-    inrloss1 = featuregain(inrvars,noinrs,gccontent)
-    inrloss2 = inrloss1/inrprob
-    inrstay = featurestay(inrvars,gccontent)
-    noinrstay = 1- inrgain -inrprob
+    mers6 = kmers(6);
+    inrvars = vec(join.(product("TGC","TGC","C", "A","TGC","TA")));
+    noinrs = setdiff(mers6, inrvars);
+    inrprob = sum([nucprob(x,gccontent) for x in inrvars]);
+    inrgain = featuregain(noinrs,inrvars,gccontent);
+    inrloss1 = featuregain(inrvars,noinrs,gccontent);
+    inrloss2 = inrloss1/inrprob;
+    inrstay = featurestay(inrvars,gccontent);
+    noinrstay = 1- inrgain;
 
     ## PolyA signal
 
[…]

### 2022-11-08 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
 \usepackage[colorlinks, allcolors = blue]{hyperref}
@@
 \renewcommand{\subsubsectionautorefname}{Section}
-
-\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
-\setcounter{footnote}{1}
+\titleformat{\paragraph}[hang]{\normalfont\normalsize\itshape}{\theparagraph}{0em}{}
+\titlespacing*{\paragraph}{0pt}{1\baselineskip}{3pt}
+\setlength{\parindent}{0pt}
 
@@
 \onehalfspacing
+
+\setlength{\abovedisplayskip}{0pt}
+\setlength{\belowdisplayskip}{1em}
 
@@
 
-\vspace{-1ex}
-
 $$P_0 = 1-e^{-k\lambda}$$
 
-Because the mutation rate is biased (\autoref{mutbias}), the probability that a mutation does not occur in a DNA sequence depends on the composition of the sequence.
+Because the mutation rate is biased (\autoref{mutbias}), the probability that no mutation occurs in a DNA sequence depends on its composition.
 
-The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is described by the mutation bias (\autoref{mutbias}) as:
-
-\vspace{-1ex}
+The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is thus described as:
 
@@
 
+\vspace{-1ex}
+
 $$P_0 = (1-e^{-m\lambda_\texttt{AT}})\times(1-e^{-(k-m)\lambda_\texttt{GC}})$$
@@
 
-The consensus sequence of TATA-box is \texttt{TATA [T/A] A [T/A] [G/A]} \citep{tata2}.
-Square brackets mean that, any of the nucleotides enclosed by them, can exist in that specific position.
-Another study used X-ray crystallography to identify mutations that are tolerated in TATA-box consensus sequence \cite{tata1}.
+The consensus sequence of TATA-box is \texttt{TATA [T/A] A [T/A] [G/A]} \citep{tata2} such that, any of the nucleotides enclosed within the square brackets, can exist in that specific position in the sequence.
+Another study used X-ray crystallography to identify mutations that are tolerated in TATA-box consensus sequence \citep{tata1}.
 Thus there are eight consensus sequences, and at seven positions, a non-consensus nucleotide is tolerated (\autoref{tatatab}).
@@
 
-\paragraph{Inr:} The consensus sequence of Inr is \texttt{[T/G/C] [T/G/C] CA [T/G/C] [T/A]} \citep{Inr}, which gives 54 sequences in the Inr feature set.
-We generated the corresponding non-feature set by taking all DNA 6-mers that are distinct from the 54 Inr sequences.
+\paragraph{Inr:} The consensus sequence of Inr is \texttt{[T/G/C] [T/G/C] CA [T/G/C] [T/A]}, which gives 54 sequences in the Inr feature set \citep{Inr}.
+We generated the corresponding non-feature set with all DNA 6-mers that are distinct from the 54 Inr sequences.
 
@@
 The non-feature set consists of all other DNA 6-mers.
+
+\vspace{1\baselineskip}
 
@@
 
-$$Q_\textit{TATA-only} = \frac{P_\textit{TATA}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})}$$
+\begin{equation*}
+Q_\textit{TATA-only} = \frac{P_\textit{TATA}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})}
+\end{equation*}
+
+\vspace{1ex}
 
@@
 
-\vspace{-1em}
-
 \begin{align*}
-Q_\textit{Inr-only} & = \frac{P_\textit{Inr}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})} \\[1em]
+Q_\textit{Inr-only} & = \frac{P_\textit{Inr}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})} \\[1ex]
 Q_\textit{both} & = 1 - Q_\textit{TATA-only} -Q_\textit{Inr-only}
@@
 \begin{align}
-P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1ex]
- & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1ex]
+P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
+ & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
  & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
 \end{align}
+
+\subsection{Probabilities of finding, gaining, and losing an ORF}
+
+\subsubsection{Probability of finding an ORF}
+
+An open reading frame (ORF) is a nucleotide sequence with a length that is a multiple of three, that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}).
+This necessarily means that there are no stop codons within the sequence.
+Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:
+
+\begin{equation}
+P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
+\end{equation}
+
+Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.
+
+\subsubsection{Probability of gaining an ORF}
+
+As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence).
+Thus an ORF can emerge due to mutations via three mechanisms.
+In each of these mechanisms one requirement is absent whereas the other two are present.
 

### 2022-11-08 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 We used mutation bias data from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
-shows the values of different mutation rates.
+shows the values of different mutation rates (\autoref{mutbias}).
 
@@
 Specifically, the probability of gene gain can be defined as the sum of three probabilities.
-First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$).
-Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$) while transcription stays intact ($P_\textit{RNA-stay}$).
+First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{eqrnagain}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{eqorfstay}).
+Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{eqorfgain}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{eqrnastay}).
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible).
@@
 
-Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$) due to mutations.
+Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{eqrnaloss}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{eqorfloss}) due to mutations.
 Specifically:
@@
 Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids.
-On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}A).
+On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}{\color{blue}A}).
 This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
@@
 For example, leucine, arginine and serine, all have six codons each.
-However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}A).
+However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}{\color{blue}A}).
 
@@
 
-We found that amino gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}B) but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
+We found that amino gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
 
@@
 
-The loss probability was also not uniformly distributed (\autoref{aafreq}C) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
+The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
 
@@
 As we mentioned in previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
-We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}D).
+We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}).
 The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836.
@@
 \begin{align}
-P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\ |\ \alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\ |\ \alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
+P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
 %
-P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\ |\ \alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\ |\ \alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
+P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
 \end{align}
@@
 \section{Discussion}
+
+In this work, we addressed some fundamental questions about \textit{de novo} gene evolution.
+Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory, and how the composition of their protein sequence is determined.
+To this end, we developed mathematical models and used them to address specific questions.
+These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection.
+Thus our models provide an opportunity to test the alternative hypothesis that some \textit{de novo} genes are evolving under selection.
+
+We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the expected time to gain it.
+Therefore, presence of a \textit{de novo} gene in one species (or population) cannot be interpreted as an evidence of gene emergence in this species.
+
+We answer the long standing question of what is the sequence of \textit{de novo} emergence.
+That is, whether ORF emerges first or transcription emerges first.
+We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
+
+We show that transcription loss is less likely than the loss of long ORFs, but this difference is not very large.
+Thus one cannot reliably infer that multiple independent transcription losses for a given \textit{de novo} gene, indicates a negative selection on the protein expression.
+
+Finally, we show that some amino acids substitutions are more likely to occur than the others, and frequent  substitutions that are expected to occur less frequently under neutral model, can suggest directional evolution.
+
+Like all theoretical models, our model is not without limitations.
+First, we assume that nucleotides are uniformly distributed according to GC content.
+This is not true as nucleotide composition can vary significantly across the genome.
+Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}.
+However, our model can be adapted to investigate specific genomic loci, where information on the local GC content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
+Second, our model is based on parameters such as mutation rate and mutation bias, which we obtained from published data from humans and \textit{D.melanogaster} \citep{humanmutrate,drosophilamutrate}.
+The accuracy of the model will depend on the accuracy of parameters.
+More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other.
+Thus our inferences based on our chosen parameters will not universally apply to all species.
+Third, we assume that all mutations are independent.
+This is not the case for tandem mutations, where multinucleotide mutations occur can more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
+Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation.
+Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels transpositions cause larger sequence alterations than point mutations.
+However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.
+
+As the famous saying goes, ``all models are wrong but some are useful'' \citep{GEBox}.
+This is true for our models too.
+They are not accurate but they can be useful for making predictions that can lead a more focused experimental validation.
+These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution.
+Therefore our work opens up an opportunity for theoretical biologists and computational biologists to develop models that are accurately describe their system of interest.
 
@@
 \subsection{Probabilities of finding, gaining, and losing a DNA sequence}
+\label{methbasic}
 
 \subsubsection{Probability of finding a DNA sequence}
+\label{methprob}
 
@@
 \subsection{Probabilities of finding, gaining, and losing DNA features}
+\label{methfeatures}
 
@@
 
-The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set.
+The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set (\autoref{methprob}).
 
@@
 
-\begin{table}[!t]
+\begin{table}[H]
 \centering
@@
 P_\textit{RNA} = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr})
+\label{eqrnaprob}
 \end{equation}
@@
 P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
+\label{eqrnagain}
 \end{equation}
@@
  & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
+\label{eqrnaloss}
 \end{align}
+
+\subsubsection{Probability that transcription remains intact}
+
+Transcription remains intact if neither the promoter, nor the poly-A signal is lost. More specifically, we define this probability ($P_\textit{RNA-stay}$) as:
+
+\begin{equation}
+P_\textit{RNA-stay} = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
+\label{eqrnastay}
+\end{equation}
 
@@
 P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
+\label{eqorfprob}
 \end{equation}
@@
 Thus an ORF can emerge due to mutations via three mechanisms.
-In each of these mechanisms one requirement is absent whereas the other two are present.
+In each of these mechanisms, one requirement is initially absent whereas the other two are present.
+Then mutations cause the missing requirement to emerge while not disrupting the other two requirements.
+More specifically, the ORF can be gained via the following three mechanisms:
+\begin{enumerate}
+\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of the ORF ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the ORF ($1- P_\textit{stop} - P_\textit{stop-gain}$).
+\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of the ORF ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the ORF.
+\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the ORF ($P_\textit{stop-gain}$).
+At the same time the start codon and the stop codon remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
+\end{enumerate}
+
+Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:
+
+\begin{align}
+P_\textit{ORF-gain} & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
+\label{eqorfgain}
+\end{align}
+
+\subsubsection{Probability of ORF loss}
+
+ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:
+
+\begin{equation}
+P_\textit{ORF-loss} = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
+\label{eqorfloss}
+\end{equation}
+
+The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.
+
+\subsubsection{Probability that ORF remains intact}
+
+We assumed that an ORF of a certain length remains intact if none of the necessary features are lost.
+However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence.
+This condition applies to all the three ORF probabilities described above.
+We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:
+
+\begin{equation}
+P_\textit{ORF-stay} = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
+\label{eqorfstay}
+\end{equation}
+
+\subsection{Probability of amino acid substitutions}
+
+We calculated the probability of amino acid substitutions based on nucleotide substitution in the corresponding codons.
+In this case, multiple codons that code for the same amino acid constitute a feature set (\autoref{methfeatures}).
+Similarly when analysing hydrobhobic to non-hydrophobic substitutions, all codons that represent hydrophobic amino acids form a feature set (and \textit{vice versa}).
+
+\section{Data availability}
+We performed all calculations using Julia programming language, and the all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
 

### 2022-11-08 Modeling/DeNovoProb.jl

@@ -70,14 +70,13 @@ function rnavals(gccontent)
 
     rnaprob = (tataprob + inrprob)*polyaprob;
     rnagain = (tatagain + inrgain)*polyastay + (tatastay + inrstay)*polyagain;
-    tfpinr = inrprob*(1-tataprob)/(tataprob + inrprob - inrprob*tataprob);
-    tfptata = tataprob*(1-inrprob)/(tataprob + inrprob - inrprob*tataprob);
-    tfpboth = inrprob*tataprob/(tataprob + inrprob - inrprob*tataprob);
-    tfpboth = 1 - tfptata - tfpinr;
+    Qinr = inrprob*(1-tataprob)/(tataprob + inrprob - inrprob*tataprob);
+    Qtata = tataprob*(1-inrprob)/(tataprob + inrprob - inrprob*tataprob);
+    Qboth = 1 - Qtata - Qinr;
 
-    rnaloss = (tfpinr*inrloss2*notatastay +
-      tfptata*tataloss2*noinrstay +
-      tfpboth*tataloss2*inrloss2 +
+    rnaloss = (Qinr*inrloss2*notatastay +
+      Qtata*tataloss2*noinrstay +
+      Qboth*tataloss2*inrloss2 +
       polyaloss2
     );
     rnastay = (inrstay + tatastay)*polyastay

### 2022-11-09 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 
+\renewenvironment{abstract}{%
+  \normalsize
+  \textbf{\large\abstractname\\[1em]} % with a normal space
+}
+
+\title{Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory}
+\author{Bharat Ravi Iyengar, Erich Bornberg-Bauer}
+\date{\small Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany}
+
 \begin{document}
@@
 
+\maketitle
+
+\begin{abstract}
+New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process known as \textit{de novo} gene emergence.
+To synthesize a protein, DNA must be transcribed as well as translated, both of which need some DNA sequence features.
+Stable transcription requires promoters and polydenylation signal, and translation requires at least an open reading frame (ORF).
+We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly do genes emerge and are lost.
+We also investigate whether different DNA features evolve in a sequential manner, and if sequence composition is biased by mutation rate.
+We find out that genes are lost much more rapidly than they emerge, and genes with long ORFs preferentially arise in regions that are already transcribed.
+Our study not only answers some fundamental question on the topic, but also provides a modeling framework for future studies.
+\end{abstract}
+
 \section{Introduction}
+
+Organisms evolve new traits by expanding their functional genome.
+Evolution of new genes is one of the ways by which new traits can emerge.
+The definition of a gene is complicated, and has been changing constantly \citep{genepostencode}.
+We use the following working definition of a gene: a gene is a region of the genome that gives rise to a functional product, that is an RNA or a protein.
+The next big challenge is to define what a function is, and how to classify gene functions.
+One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions.
+Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins (mRNA) or participate in protein synthesis (rRNA, tRNA), are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.
+
+In this study we focus on the evolution of genes that encode new proteins.
+New protein coding genes frequently arise via duplication of existing protein coding genes.
+These duplicated genes then genetically and functionally diverge from their parent genes by accumulating mutations \citep{Long2003,IAD}.
+Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein.
+This phenomenon is called \textit{de novo} gene emergence \citep{EBB-F1000,vanOss}.
+\textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{vanOss}.
+To express a protein, a genomic sequence should be transcribed as well as translated.
+Therefore, the gene needs sequence features that enable both these processes.
+
+The primary requirement for transcription is recruitment of RNA polymerase at the gene's promoter \citep{Promoters}.
+A part of the promoter, called the core promoter is the region that determines the start of transcription \citep{corepromoters}.
+The second requirement for a productive transcription is its termination at appropriate sites.
+Even though transcription is widespread in eukaryotic genomes, a phenomenon known as pervasive transcription, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}.
+Most eukaryotic mRNAs and long non-coding RNAs, are polyadenylated at their 3' termini, which makes them resistant to degradation.
+Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals.
+In prokaryotes, end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
+Thus the synthesis of a stable RNA, requires promoters, as well as termination/polyadenylation signal sequences.
+
+The second major requirement for protein expression is translation of the transcribed mRNA.
+The most fundamental requirement for translation is the presence of an open reading frame (ORF).
+Usually, an mRNA needs additional features to initiate protein synthesis.
+These features include ribosome binding sites in prokaryotes \citep{RBS}, and Kozak consensus sequences \citep{kozak} in eukaryotes.
+
+Although gain of transcription and translation features guarantees \textit{de novo} birth of a protein coding gene, it does not ensure that the gene would persist in the genome for many generations.
+The newly born gene can lose the features as easily as it gained them unless it is fixed in the genome via evolutionary selection.
+If the protein synthesized by the \textit{de novo} gene provides some fitness advantage to the host organism then it will undergo positive selection.
+Simulataneously, the protein should have low toxicity and cost of synthesis, to survive purifying selection.
+A common mechanism by which protein mediated toxicity occurs is by misfolding and aggregation of proteins \citep{misfold1,misfold2}.
+
+In this study we develop probabilistic models of \textit{de novo} gene emergence.
+We focus on \textit{de novo} genes emerging from intergenic sequences, also known as ``protogenes'' \citep{protogenes}.
+We model \textit{de novo} gene emergence as a two step process.
+Specifically, the protogene can first gain transcription features and produce an untranslated RNA, which in turn acquires translation features subsequently.
+Alternatively, the protogene can first gain translational features, and later gain transcription.
+Our models are based on neutral evolution where natural selection does not affect the rate of gene emergence, or loss.
+Using these models, we predict that \textit{de novo} gene emergence has a preferred evolutionary trajectory, even under the assumption of neutrality.
 
@@
 Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence.
-[here a rationale for this choice].
+We chose these two promoter motifs because they are the most abundant type of core promoters across diverse eukaryotes \citep{Promoters,corepromoters}.
+We defined the poly-A signal as a transcription requirement because it is crucial for transcript stability and its export to cytoplasm\citep{polyAterm,polyAexp}.
 For translation, we only require the gene to have an open reading frame (ORF).
@@
 These probability models are essentially described by two kinds of probability.
-First is the probability of finding a DNA sequence feature (such as TATA box or an ORF).
+First is the probability of finding a DNA sequence feature (such as TATA box or an ORF; \autoref{methprob}).
 This probability depends on the nucleotide distribution given by the GC content.
-The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another.
+The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another (\autoref{methgain}).
 Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.
@@
 We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
-We used mutation bias data from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
-shows the values of different mutation rates (\autoref{mutbias}).
+We used mutation bias data (\autoref{mutbias}) from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
 
@@
 Overall our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.
-A significant deviation from this null hypothesis may be interpreted as evidence in support of natural selection.
 
 \subsection{How many times can a gene be lost?}
-
-\begin{figure}[!b]
-\hrule
-\vspace{1ex}
-\centering
-\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
-\caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e.
-$\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF.
-\autoref{genegaineq} describes total gene gain probability.
-Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
-Using the conditional probability of gene gain does not change these results.}
-\label{gainlossprob}
-\end{figure}
 
@@
 
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
+\caption{Genes are more likely to be independently lost twice, than being born once.
+The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e.
+$\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the number of codons in the ORF.
+\autoref{genegaineq} describes total gene gain probability.
+Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
+Using the conditional probability of gene gain does not change these results.}
+\label{gainlossprob}
+\vspace{1ex}
+\hrule
+\end{figure}
+
 The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$.
@@
 
-\begin{figure}[!b]
-\hrule
-\vspace{1ex}
-\centering
-%\begin{subfigure}{0.48\textwidth}
-%\centering
-%\caption{One-step}
-\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
-%\end{subfigure}\hfill%
-%\begin{subfigure}{0.48\textwidth}
-%\centering
-%\caption{Two-step}
-%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
-%\end{subfigure}
-\caption{\textit{De novo} genes with shorter ORFs ($<$ 42 codons) preferentially emerge from existing ORFs (ORF-first) whereas \textit{de novo} genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
-The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
-A positive value suggests a ORF-first trajectory and a negative value supports RNA-first trajectory.
-Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
-\label{whoisfirst}
-\end{figure}
-
 For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge.
@@
 Thus it may appear that the order of the occurrence of these two events does not matter.
-Gene gain is much more likely when one of the two features already exist (\autoref{genegaineq}) and hence there are two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
-To test the hypothesis that the order of feature gain events influences \textit{de novo} emergence, we first asked how much more likely it is to lose transcription before gaining an ORF.
-Specifically, we calculated the fold difference between the probability values of transcription loss and ORF gain.
-We found that transcription loss is $10^3$ to $10^{10}$ times higher than ORF gain, depending on the ORF length.
-This suggests that a mutation has a higher chance of disrupting an existing untranscribed ORF than cause a gain of transcription features.
-Next, we calculated the fold difference between transcription gain and ORF loss probability values, and found that ORF loss is $10^4$ to $10^5$ times more probable than transcription gain, suggesting that in a transcribed DNA region, a new mutation will more likely disrupt the transcription features than cause a gain of ORF.
-More generally features are more easily lost due to mutations than they are gained.
-However, this effect of mutations is not uniform for the two evolutionary trajectories and for different length of ORFs.
-For example, gain of an 40 codon long ORF in a transcribed region of DNA is $9.2\times10^3$ less likely than the transcription being lost.
-However, if the ORF already exists, then it is $2.7\times10^4$ times more likely to be lost before transcription emerges.
-This suggests that the two trajectories of \textit{de novo} emergence have different dynamics.
-
-To better understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
-First is the probability that ORF exists and a mutation causes gain of transcription but no disruption of the ORF.
-This probability denotes the trajectory where ORF emerges first.
+However, gene emergence is much more likely when one of the two features already exist (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
+To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
+First is the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF.
+This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$; \autoref{orffirsteq1}).
 
@@
 
-The second probability that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and a mutation causes gain of ORF but does not abolish transcription.
+The second probability ($P_\textit{RNA-first}$; \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.
 
@@
 
+\begin{figure}[!t]
+\centering
+%\begin{subfigure}{0.48\textwidth}
+%\centering
+%\caption{One-step}
+\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
+%\end{subfigure}\hfill%
+%\begin{subfigure}{0.48\textwidth}
+%\centering
+%\caption{Two-step}
+%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
+%\end{subfigure}
+\caption{\textit{De novo} genes with shorter ORFs ($<$ 42 codons) preferentially emerge from existing ORFs (ORF-first) whereas \textit{de novo} genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
+The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
+A positive value suggests a ORF-first trajectory and a negative value supports RNA-first trajectory.
+Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
+\label{whoisfirst}
+\vspace{1ex}
+\hrule
+\end{figure}
+
 With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}\ \%$).
@@
 
-Overall, our analysis suggests that \textit{de novo} genes with long ORFs are more likely arise from DNA regions that are already transcribed (for example, non-coding RNAs).
+This preference of \textit{de novo} gene emergence trajectory exists, because the rates of feature gain and the rates of feature loss are different.
+For example, ORFs are very likely to be lost before transcription emerges.
+Specifically, ORF loss is $\sim10^4$ to $\sim10^5$ times more probable than transcription gain, such that this difference increases with the ORF length.
+The converse is also true, that is transcription loss is more probable than ORF gain.
+However, this difference is not the same as that between ORF loss and transcription gain.
+Transcription loss is $\sim10^3$ to $\sim10^{10}$ times more likely than ORF gain, such that this difference is larger for longer ORFs.
+Thus, long ORFs are more likely to be lost before transcription emerges, than short ORFs.
+Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.
 
@@
 For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$.
-(Note that $P_\textit{RNA-loss}$ is a conditional probability dependent on the presence of RNA).
+(Note that $P_\textit{RNA-loss}$ is a conditional probability, that is dependent on the presence of transcription).
 
@@
 We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
+Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs, but this evolutionary process may be constrained by a different set of factors.
+Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the \textit{de novo} gene.
+For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory.
+On the other hand, the selection would have no effect on an ORF that is not transcribed.
+Thus positive selection will intensify the preference for RNA-first trajectory.
 
@@
 
-As the famous saying goes, ``all models are wrong but some are useful'' \citep{GEBox}.
+A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}.
 This is true for our models too.
@@
 \bibliographystyle{plainnat}
+
+\small
 \bibliography{refs}

### 2022-11-09 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 It is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively).
-Furthermore, non-stop mutations can also lead to cellular toxicity.
+Furthermore, non-stop mutations can also lead to cellular toxicity \citep{nonstop}.
 Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may be the better way to inactivate the associated genes.
@@
 
-Previous studies have argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure.
-They based these conclusions on mutation rate bias estimated in bacteria.
+A revious studies has argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure \citep{hydrophobicRatchet}.
+They based these conclusions on mutation rate bias estimated in bacteria \citep{bacteriaMutbias}.
 We asked if our model also makes similar predictions.

### 2022-11-09 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 
-We reframed our question to make it more relatable to commonly observed findings from genomics experiments.
-For example, a hypothetical experiment involves DNA sequencing of a genomic locus in different related species.
-This experiment shows that transcription features are present in only one species, whereas the ORF is present in all species.
-Two evolutionary events could describe this observation.
-First possibility is that a \textit{de novo} gene existed in an ancestor and transcription was lost in multiple species.
-For any one species, the probability of this evolutionary trajectory is: $P_\textit{ORF-stay}\times P_\textit{RNA} \times P_\textit{RNA-loss}$.
-(Note that $P_\textit{RNA-loss}$ is a conditional probability, that is dependent on the presence of transcription).
+We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact \textit{de novo} gene.
+In the first scenario, the ORF remains intact whereas the transcription is lost.
+The probability of this scenario is described by:
+$P_\textit{ORF-stay} \times P_\textit{RNA-loss}/P_\textit{ORF}$.
+The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$.
+We find that the first scenario is more plausible for ORFs that have less than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}).
+Also, the probability values of scenario one and two, obtained from this stringent analysis, were not significantly different from the transcription ($P_\textit{RNA-loss}$) and ORF loss ($P_\textit{ORF-loss}$) probabilities, respectively.
+
+Thus widespread transcription loss can suggest a negative selection of
 
@@
 \end{figure}
-
-Alternatively, an ORF could have existed in all species and transcription was subsequently gained in one species.
-The probability that this event happens in any one species is: $P_\textit{ORF-stay}\times P_\textit{RNA-gain}$ (see \autoref{orffirsteq1}).
-
-The ratio of the probabilities of the two evolutionary trajectories would simply be: $P_\textit{RNA} \times P_\textit{RNA-loss}/P_\textit{RNA-gain}$.
-We found the value of this ratio to be 0.914, suggesting that transcription loss in a \textit{de novo} gene would be more probable than transcription gain for an untranscribed ORF.
-However, this difference is small and hence the real evolutionary mechanism behind the observation from our hypothetical sequencing experiment, would be difficult to infer.
-
-In sum, transcription loss cannot be reliably inferred as an evidence for negative selection of protein expression.
 
@@
 The accuracy of the model will depend on the accuracy of parameters.
+Thus our inferences based on our chosen parameters will not universally apply to all species.
+However, mutation rate itself only determines the values of different probabilities.
+Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate.
 More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other.
-Thus our inferences based on our chosen parameters will not universally apply to all species.
 Third, we assume that all mutations are independent.

### 2022-11-10 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 \begin{abstract}
-New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process known as \textit{de novo} gene emergence.
-To synthesize a protein, DNA must be transcribed as well as translated, both of which need some DNA sequence features.
+New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence.
+To synthesize a protein, DNA must be transcribed as well as translated, both of which need certain DNA sequence features.
 Stable transcription requires promoters and polydenylation signal, and translation requires at least an open reading frame (ORF).
@@
 We find out that genes are lost much more rapidly than they emerge, and genes with long ORFs preferentially arise in regions that are already transcribed.
-Our study not only answers some fundamental question on the topic, but also provides a modeling framework for future studies.
+Our study not only answers some fundamental question on the topic of \textit{de novo} emergence, but also provides a modeling framework for future studies.
 \end{abstract}
@@
 One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions.
-Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins (mRNA) or participate in protein synthesis (rRNA, tRNA), are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.
+Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins like mRNAs, or participate in protein synthesis like rRNAs and tRNAs, are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.
 
@@
 In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss.
-A \textit{de novo} gene emerges from a non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation.
+A \textit{de novo} gene emerges from non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation.
 Both transcription and translation are complex processes involving many biomolecular complexes that work in concert.
@@
 We chose these two promoter motifs because they are the most abundant type of core promoters across diverse eukaryotes \citep{Promoters,corepromoters}.
-We defined the poly-A signal as a transcription requirement because it is crucial for transcript stability and its export to cytoplasm\citep{polyAterm,polyAexp}.
+We defined the poly-A signal as a transcription requirement because it is crucial for transcript stability and its export to cytoplasm \citep{polyAterm,polyAexp}.
 For translation, we only require the gene to have an open reading frame (ORF).
@@
 First is the probability of finding a DNA sequence feature (such as TATA box or an ORF; \autoref{methprob}).
-This probability depends on the nucleotide distribution given by the GC content.
+This probability depends on the nucleotide distribution given by the GC-content.
 The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another (\autoref{methgain}).
@@
 
-For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes.
+For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes \citep{allgccontents, bionumbers}.
 We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
@@
 
-We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
-In other words, ORF-first hypothesis is more feasible.
-Likewise, a negative value of the ratio would suggest that RNA-first hypothesis is more feasible.
-With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first.
-\textit{De novo} genes with longer coding region, emerge transcription first (\autoref{whoisfirst}).
-
-A more stringent definition of ORF-first hypothesis would describe a two-step process.
-In the first step, an ORF emerges in an untranscribed region of DNA, but transcription does not emerge.
-In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
-Likewise, RNA-first hypothesis can be defined by a two-step probability.
-In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).
-
-\begin{align}
-P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
-P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
-\end{align}
-
 \begin{figure}[!t]
@@
 \end{figure}
+We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
+In other words, ORF-first hypothesis is more feasible.
+Likewise, a negative value of the ratio would suggest that RNA-first hypothesis is more feasible.
+With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\autoref{whoisfirst}).
+With a higher GC-content, ORF length above which RNA-first hypothesis is more feasible than ORF-first hypothesis, increases.
+For example, this number is 48 codons for a GC-content of 0.45 (data not shown).
+
+A more stringent definition of ORF-first hypothesis would describe a two-step process.
+In the first step, an ORF emerges in an untranscribed region of DNA, but transcription does not emerge.
+In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
+Likewise, RNA-first hypothesis can be defined by a two-step probability.
+In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).
+
+\begin{align}
+P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
+P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
+\end{align}
 
@@
 
-This preference of \textit{de novo} gene emergence trajectory exists, because the rates of feature gain and the rates of feature loss are different.
-For example, ORFs are very likely to be lost before transcription emerges.
-Specifically, ORF loss is $\sim10^4$ to $\sim10^5$ times more probable than transcription gain, such that this difference increases with the ORF length.
-The converse is also true, that is transcription loss is more probable than ORF gain.
-However, this difference is not the same as that between ORF loss and transcription gain.
-Transcription loss is $\sim10^3$ to $\sim10^{10}$ times more likely than ORF gain, such that this difference is larger for longer ORFs.
-Thus, long ORFs are more likely to be lost before transcription emerges, than short ORFs.
+Overall, long ORFs are more likely to be lost before transcription emerges, than short ORFs, and hence they preferentially emerge in existing transcripts.
 Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.
@@
 
-Thus widespread transcription loss can suggest a negative selection of
+Thus widespread transcription loss can suggest a negative selection on protein expression.
 
@@
 
-\caption{Mutation bias and GC content shape protein composition.
+\caption{Mutation bias and GC-content shape protein composition.
 Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes).
@@
 This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
-However, GC content also plays a role.
+However, GC-content also plays a role.
 For example, leucine, arginine and serine, all have six codons each.
@@
 This shows that random mutations do not preferentially cause hydrophobic amino acids to accumulate in protein sequence.
+Inclusion of proline and tryptophan in the list of hydrophobic amino acids does not change this finding.
 
@@
 We answer the long standing question of what is the sequence of \textit{de novo} emergence.
-That is, whether ORF emerges first or transcription emerges first.
+That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
 We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
 Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs, but this evolutionary process may be constrained by a different set of factors.
+A previous study has shown that new transcripts frequently arise in regions overlapping existing genes.
+These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}.
+This is an example of RNA-first trajectory.
+However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences.
+Therefore, they may be less frequently mutated or lost.
+
 Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the \textit{de novo} gene.
@@
 On the other hand, the selection would have no effect on an ORF that is not transcribed.
+This is based on the assumption that the ORF sequence has no other function (for e.g.
+regulation), or is not under indirect selection (for e.g.
+if it overlaps with an existing gene).
 Thus positive selection will intensify the preference for RNA-first trajectory.
 
-We show that transcription loss is less likely than the loss of long ORFs, but this difference is not very large.
-Thus one cannot reliably infer that multiple independent transcription losses for a given \textit{de novo} gene, indicates a negative selection on the protein expression.
+We show that transcription loss is less likely than the loss of long ORFs, However, it may be difficult to find direct evidence for such a phenomenon.
+Specifically, it is difficult to infer from data if transcription was lost or it never emerged.
+If the protein encoded by an ORF is indeed toxic, it would be silenced rapidly in microevolutionary time scales.
+One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status.
+Mutations within ORF sequence are much more probable than mutations that disrupt transcription or ORF.
+This is so because the entire ORF sequence has higher number of mutable sites.
+If an ORF variant is untranscribed in many populations (species or taxa), then it is possibly toxic.
 
@@
 
-Like all theoretical models, our model is not without limitations.
-First, we assume that nucleotides are uniformly distributed according to GC content.
+Like all theoretical models, our models are not without limitations.
+First, we assume that nucleotides are uniformly distributed according to GC-content.
 This is not true as nucleotide composition can vary significantly across the genome.
 Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}.
-However, our model can be adapted to investigate specific genomic loci, where information on the local GC content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
-Second, our model is based on parameters such as mutation rate and mutation bias, which we obtained from published data from humans and \textit{D.melanogaster} \citep{humanmutrate,drosophilamutrate}.
-The accuracy of the model will depend on the accuracy of parameters.
-Thus our inferences based on our chosen parameters will not universally apply to all species.
+However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
+Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}.
+The accuracy of the models' prediction will depend on the accuracy of parameters.
 However, mutation rate itself only determines the values of different probabilities.
@@
 More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other.
+Thus our inferences based on our chosen parameters will not universally apply to all species.
 Third, we assume that all mutations are independent.
@@
 This is true for our models too.
-They are not accurate but they can be useful for making predictions that can lead a more focused experimental validation.
+They are not 100\% accurate but they are useful for making predictions that can lead a more focused experimental validation.
 These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution.
-Therefore our work opens up an opportunity for theoretical biologists and computational biologists to develop models that are accurately describe their system of interest.
+Therefore our work opens up an opportunity for theoretical and computational biologists to develop models using our modeling framework, that more accurately describe their system of interest.
 
@@
 We calculated the probability of finding a DNA sequence based on global nucleotide frequency distributions, given by the GC-content.
-We used a GC content of 0.42 for our calculations.
+We used a GC-content of 0.42 for our calculations.
 Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is: $S = 0.5\times$GC-content.
@@
 \begin{enumerate}
-\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of the ORF ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the ORF ($1- P_\textit{stop} - P_\textit{stop-gain}$).
-\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of the ORF ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the ORF.
-\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the ORF ($P_\textit{stop-gain}$).
-At the same time the start codon and the stop codon remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
+\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of a reading frame ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the reading frame ($1- P_\textit{stop} - P_\textit{stop-gain}$).
+\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of a reading frame ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the reading frame.
+\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the reading frame ($P_\textit{stop-gain}$).
+At the same time start and stop codons remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
 \end{enumerate}

### 2022-11-10 Modeling/DeNovoProb.jl

@@ -317,4 +317,3 @@ plot_fraa = bar(aas,lossAA./sum(lossAA),
 );
 
 savefig(plot_fraa, figdir*"pFrAA.pdf")
-

### 2022-11-10 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex 	modified:   Manuscripts/refs.bib

@@
 Usually, an mRNA needs additional features to initiate protein synthesis.
-These features include ribosome binding sites in prokaryotes \citep{RBS}, and Kozak consensus sequences \citep{kozak} in eukaryotes.
+These features include ribosome binding sites in prokaryotes \citep{RBS}, and Kozak consensus sequences in eukaryotes \citep{kozak,kozakDroso,kozakHuman}.
 
@@
 Thus there are eight consensus sequences, and at seven positions, a non-consensus nucleotide is tolerated (\autoref{tatatab}).
-This makes a TATA-box feature set that contains 136 sequences.
+This makes a TATA-box feature set that contains 104 sequences.
 We defined the corresponding non-feature set by generating all 8-mer DNA sequences that do not belong to the TATA-box feature set.
@@
 \centering
-\begin{tabular}{@{}ccc@{}}
+\begin{tabular}{c c}
 \toprule
-Position  & Accepted nucleotides & \begin{tabular}[c]{@{}c@{}}Number of Variants\\ (consensus + suboptimal)\end{tabular} \\ \midrule
-1         & T$\gg$C$>$A,G        & $1+3$                                                                                 \\ \midrule
-2         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
-3         & T$\gg$A,C            & $1+2$                                                                                 \\ \midrule
-4         & A$\gg$T              & $1+1$                                                                                 \\ \midrule
-5         & T/A                  & $2+0$                                                                                 \\ \midrule
-6         & A$\gg$G$>$C/T        & $1+3$                                                                                 \\ \midrule
-7         & A/T$>$G$>$C          & $2+2$                                                                                 \\ \midrule
-8         & G/A$>$C/T            & $2+2$                                                                                                                                                                 \\ \bottomrule
+Position  & Accepted nucleotides \\ \midrule
+1         & T$\gg$C$>$A,G        \\ \midrule
+2         & A$\gg$T               \\ \midrule
+3         & T$\gg$A,C             \\ \midrule
+4         & A$\gg$T               \\ \midrule
+5         & T/A                   \\ \midrule
+6         & A$\gg$G$>$C/T         \\ \midrule
+7         & A/T$>$G$>$C           \\ \midrule
+8         & G/A$>$C/T             \\ \bottomrule
 \end{tabular}

### 2022-11-10 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{eqorfgain}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{eqrnastay}).
-Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small, and negligible).
+Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
 The following equation describes the probability of gene gain:
@@
 First is the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF.
-This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$; \autoref{orffirsteq1}).
+This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$, \autoref{orffirsteq1}).
 
@@
 
-The second probability ($P_\textit{RNA-first}$; \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.
+The second probability ($P_\textit{RNA-first}$, \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.
 
@@
 
-\begin{table}[H]
+\begin{table}[!t]
 \centering

### 2022-11-11 Manuscripts/deNovoRates.tex
modified:   Modeling/nucleotidefuncts.jl

@@
 To synthesize a protein, DNA must be transcribed as well as translated, both of which need certain DNA sequence features.
-Stable transcription requires promoters and polydenylation signal, and translation requires at least an open reading frame (ORF).
-We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly do genes emerge and are lost.
+Stable transcription requires promoters and polydenylation signal, while translation requires at least an open reading frame (ORF).
+We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly genes emerge and are lost.
 We also investigate whether different DNA features evolve in a sequential manner, and if sequence composition is biased by mutation rate.
 We find out that genes are lost much more rapidly than they emerge, and genes with long ORFs preferentially arise in regions that are already transcribed.
-Our study not only answers some fundamental question on the topic of \textit{de novo} emergence, but also provides a modeling framework for future studies.
+Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence, but also provides a modeling framework for future studies.
 \end{abstract}
@@
 
-Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{eqrnaloss}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{eqorfloss}) due to mutations.
+Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{eqrnaloss}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{eqorfloss}).
 Specifically:
@@
 
-We performed the same analysis to just the ORF (and not the whole genes).
+We performed the same analysis on just the ORFs (and not the whole genes).
 That is, we calculated the ratio of logarithms of ORF gain and ORF loss.
@@
 
-Overall, our analysis suggests that \textit{de novo} gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.
+Overall, our analysis suggests that a \textit{de novo} gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.
 
@@
 To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
-First is the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF.
+First, the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF.
 This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$, \autoref{orffirsteq1}).
@@
 The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
-A positive value suggests a ORF-first trajectory and a negative value supports RNA-first trajectory.
+A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory.
 Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
@@
 We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
-In other words, ORF-first hypothesis is more feasible.
-Likewise, a negative value of the ratio would suggest that RNA-first hypothesis is more feasible.
+In other words, the ORF-first hypothesis is more feasible.
+Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible.
 With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\autoref{whoisfirst}).
@@
 
-Overall, long ORFs are more likely to be lost before transcription emerges, than short ORFs, and hence they preferentially emerge in existing transcripts.
+Overall, long ORFs are more likely to be lost before transcription emerges than short ORFs, and hence they preferentially emerge in existing transcripts.
 Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.
@@
 
-Thus widespread transcription loss can suggest a negative selection on protein expression.
+Thus, widespread transcription loss can suggest a negative selection on protein expression.
 
@@
 \includegraphics[scale=0.6]{Figures/pLoss.pdf}
-\caption{ORF loss is more probable than transcription loss, for \textit{de novo} genes containing more than 33 codons.
+\caption{ORF loss is more probable than transcription loss for \textit{de novo} genes containing more than 33 codons.
 The horizontal axis denotes the number of codons in the ORF.
@@
 Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is: $S = 0.5\times$GC-content.
-The probability of finding an \texttt{A} or a \texttt{T} is: $W = 1 - 0.5\times$GC-content.
+The probability of finding an \texttt{A} or a \texttt{T} is: $W = 0.5 - S$.
 Using these values, we calculated the probability of finding a DNA sequence motif by chance.

### 2022-11-11 Modeling/DFE.jl

@@ -11,46 +11,60 @@ allcodons = kmers(3)
 
 aasubmat[1:20,1:20] = (aasubmat[1:20,1:20] + aasubmat[1:20,1:20]')./2
 
+stop = ["TAA","TAG","TGA"]
+
 function nostop(codonset)
-        codonset[codonset .∉ Ref(["TAA","TAG","TGA"])]
+	codonset[codonset .∉ Ref(stop)]
 end
 
 function dfecdn(codon1, codon2)
-        mp = nucprob(codon1,gccontent)*mprob(codon1,codon2)
-        fe = aasubmat[aanames[transnucs(codon1)],aanames[transnucs(codon2)]]
-        return [-fe mp]
+	mp = nucprob(codon1,gccontent)*mprob(codon1,codon2)
+	fe = aasubmat[aanames[transnucs(codon1)],aanames[transnucs(codon2)]]
+	return [-fe mp]
 end
 
 function dfedistgen(codonset1,codonset2)
-        return vcat(reduce(vcat,[[dfecdn(y,x) for x in codonset2] for y in codonset1])...)
+	return vcat(reduce(vcat,[[dfecdn(y,x) for x in codonset2] for y in codonset1])...)
 end
 
 function combinedfe(dfeset)
-        return reduce(vcat,[[i sum(dfeset[dfeset[:,1] .== i,2])] for i in unique(dfeset[:,1])])
+	return reduce(vcat,[[i sum(dfeset[dfeset[:,1] .== i,2])] for i in unique(dfeset[:,1])])
 end
 
 function expdfe(dfeset)
-        return sum(dfeset[:,1].*dfeset[:,2])/sum(dfeset[:,2])
+	return sum(dfeset[:,1].*dfeset[:,2])/sum(dfeset[:,2])
 end
 
 function dfeframeX(cdn,frm)
-        fset0 = frameXcodons(cdn,frm)
-        # cset1 = singlemutant(cdn)
-        cset1 = nostopcodons[nostopcodons .!= cdn]
-        fset1 = hcat([frameXcodons(x,frm) for x in cset1]...)
-        dfe1 = combinedfe(vcat(vcat([[dfecdn(fset0[y],x) for x in fset1[y,:]] for y in 1:20]...)...))
-        return dfe1
+	fset0 = frameXcodons(cdn,frm)
+	# cset1 = singlemutant(cdn)
[…]

### 2022-11-11 Modeling/DeNovoProb.jl

@@ -317,3 +317,4 @@ plot_fraa = bar(aas,lossAA./sum(lossAA),
 );
 
 savefig(plot_fraa, figdir*"pFrAA.pdf")
+

### 2022-11-11 Modeling/nucleotidefuncts.jl

@@ -75,6 +75,20 @@ function frameXcodons(codon::String,frame::Int)
     end
 end
 
+function frameXcpairs(codon::String,frame::Int)
+    if(frame ∉ [-2,-1,1,2])
+            error("Frame should be ±1, or ±2")
+    end
+    frabs = abs(frame)
+    z = 3-frabs
+    if(frame<0)
+        codon = reverse(comp(codon))
+    end
+    down = codon[1+frabs:3].*vec(join.(product(repeated(nucs,frabs)...)))
+    up = vec(join.(product(repeated(nucs,z)...))).*codon[1:frabs]
+    return permutedims(hcat(collect.(product(up,down))...))
+end
+
 """
 `comp(s)`
 ...

### 2022-11-11 Manuscripts/deNovoRates.tex
modified:   Modeling/DFE.jl

@@
 We find out that genes are lost much more rapidly than they emerge, and genes with long ORFs preferentially arise in regions that are already transcribed.
-Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence, but also provides a modeling framework for future studies.
+Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.
 \end{abstract}
@@
 If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species.
-We asked what is the chance that the gene was present in a common ancestor, but was lost in all lineages except one.
-To this end we calculated the probability of gene emergence and gene loss (Materials and Methods).
+We asked what is the chance that the gene was present in a common ancestor but was lost in all lineages except one.
+To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}).
 Specifically, the probability of gene gain can be defined as the sum of three probabilities.
@@
 \caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) than can happen relative to one gene gain event (\autoref{genegaineq}) i.e.
-$\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
 The horizontal axis shows the number of codons in the ORF.
@@
 With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\autoref{whoisfirst}).
-With a higher GC-content, ORF length above which RNA-first hypothesis is more feasible than ORF-first hypothesis, increases.
+This minimum ORF length above which RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content.
 For example, this number is 48 codons for a GC-content of 0.45 (data not shown).
@@
 A more stringent definition of ORF-first hypothesis would describe a two-step process.
-In the first step, an ORF emerges in an untranscribed region of DNA, but transcription does not emerge.
+In the first step, an ORF emerges in an untranscribed region of DNA but transcription does not emerge.
 In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
@@
 To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, for all ORF longer than 33 codons (\autoref{lossprob}).
+We found that ORF loss is more probable than transcription loss, for all ORFs longer than 33 codons (\autoref{lossprob}).
 
@@
 
-Next, we aimed to find out if some amino acid substitutions are more probable, than the others.
+Next, we aimed to find out if some amino acid substitutions are more probable than the others.
 To this end, we calculated the gain probability (\autoref{aagaineq}) for every amino acid.
@@
 
-We found that amino gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but had a good positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
+We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
 
@@
 As expected, amino acid substitutions that require more than one nucleotide change were much less probable than those needing just a single nucleotide change.
-Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutation, but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities).
+Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutations but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities).
 Although single nucleotide mutations are more probable than double mutations, the different single mutations do not occur at the same rate due to mutation rate bias.
@@
 
-As we mentioned in previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
+As we mentioned in the previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
 We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}).
@@
 
-A revious studies has argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure \citep{hydrophobicRatchet}.
+A previous study has argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure \citep{hydrophobicRatchet}.
 They based these conclusions on mutation rate bias estimated in bacteria \citep{bacteriaMutbias}.
@@
 To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic.
-We classified an amino acid hydrophobic if its hydropathy index \citep{hydropathy} is greater than 0 (i.e A, C, F, I, L, M \& V).
+We classified an amino acid hydrophobic if its hydropathy index \citep{hydropathy} is greater than 0 (that is A, C, F, I, L, M \& V).
 Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic.
@@
 
-We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the expected time to gain it.
+We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge.
 Therefore, presence of a \textit{de novo} gene in one species (or population) cannot be interpreted as an evidence of gene emergence in this species.
@@
 We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
-Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs, but this evolutionary process may be constrained by a different set of factors.
+Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs but this evolutionary process may be constrained by a different set of factors.
 A previous study has shown that new transcripts frequently arise in regions overlapping existing genes.
@@
 On the other hand, the selection would have no effect on an ORF that is not transcribed.
-This is based on the assumption that the ORF sequence has no other function (for e.g.
-regulation), or is not under indirect selection (for e.g.
-if it overlaps with an existing gene).
+This is based on the assumption that the ORF sequence has no other function (for example, regulation), or is not under indirect selection (for example, if it overlaps with an existing gene).
 Thus positive selection will intensify the preference for RNA-first trajectory.
 
-We show that transcription loss is less likely than the loss of long ORFs, However, it may be difficult to find direct evidence for such a phenomenon.
-Specifically, it is difficult to infer from data if transcription was lost or it never emerged.
+We show that transcription loss is less likely than the loss of long ORFs.
+However, it may be difficult to find direct evidence for such a phenomenon.
+Specifically, it is difficult to infer from data if transcription was lost or never emerged.
 If the protein encoded by an ORF is indeed toxic, it would be silenced rapidly in microevolutionary time scales.
@@
 This is so because the entire ORF sequence has higher number of mutable sites.
-If an ORF variant is untranscribed in many populations (species or taxa), then it is possibly toxic.
+If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.
 
-Finally, we show that some amino acids substitutions are more likely to occur than the others, and frequent  substitutions that are expected to occur less frequently under neutral model, can suggest directional evolution.
+Finally, we show that some amino acid substitutions are more likely to occur than others, and frequent substitutions that are expected to occur less frequently under neutral model, can suggest directional evolution.
 
@@
 Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate.
-More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially so for species that are highly distant from each other.
+More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other.
 Thus our inferences based on our chosen parameters will not universally apply to all species.
 Third, we assume that all mutations are independent.
-This is not the case for tandem mutations, where multinucleotide mutations occur can more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
+This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
 Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation.
-Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels transpositions cause larger sequence alterations than point mutations.
+Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations.
 However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.
@@
 This is true for our models too.
-They are not 100\% accurate but they are useful for making predictions that can lead a more focused experimental validation.
+They are not 100\% accurate but they are useful for making predictions that can lead to a more focused experimental validation.
 These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution.
@@
 To explain this calculation better, we use the example of \texttt{CTA} mutating to \texttt{ATG}.
-The first probability i.e.
-the probability of finding \texttt{CTA} by chance is $SW^2$.
+The first probability, that is the probability of finding \texttt{CTA} by chance is $SW^2$.
 \texttt{CTA} mutates to \texttt{ATG} via two nucleotide mutations (\texttt{C}$\to$\texttt{A} and \texttt{A}$\to$\texttt{G}).
@@
 We calculated the loss of a DNA sequence motif using the same principle we used for calculating gain probabilities.
-However, we defined loss probability as a conditional probability i.e.
-we assume that the DNA sequence of interest already exists in the genome.
+However, we defined loss probability as a conditional probability, that is we assume that the DNA sequence of interest already exists in the genome.
 For example, the loss probability of a specific \texttt{ATG} sequence would be the sum of probabilities of \texttt{ATG} mutating to any of the other 63 nucleotide triplets (via one, two, or three nucleotide mutations).
-We use conditional loss probabilities by default, because usually one is interested in finding out how quickly can an existing DNA sequence erode.
+We use conditional loss probabilities by default, because usually one is interested in finding out how quickly an existing DNA sequence can erode.
 
@@
 We use the term DNA features to mean a set of DNA sequences that are associated with the same function.
-For every such DNA feature set, there is a complementary set of non-features i.e DNA sequences that are not associated with the feature's function.
+For every such DNA feature set, there is a complementary set of non-features, that is DNA sequences that are not associated with the feature's function.
 For example, the non-feature set of stop codons would be all the other 61 codons.
@@
 Since two kinds of promoters can initiate transcription, there can be three kinds of transcription units.
-First that has only TATA-box, second that has only Inr, and the third that has both these promoters.
+First that contains only a TATA-box, second that contains only an Inr, and the third that contains both these promoters.
 All these three transcription units must also have a poly-A signal.
@@
 
-An open reading frame (ORF) is a nucleotide sequence with a length that is a multiple of three, that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}).
+A reading frame is a nucleotide sequence with a length that is a multiple of three.
+A reading frame that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}) is an open reading frame (ORF).
 This necessarily means that there are no stop codons within the sequence.
@@
 \section{Data availability}
-We performed all calculations using Julia programming language, and the all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
+We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
 

### 2022-11-11 Modeling/DFE.jl

@@ -44,7 +44,7 @@ function dfeframeX(cdn,frm)
 	return dfe1
 end
 
-gccontent = 0.42
+gccontent = 0.53942 # all CDS #
 nostopcodons = nostop(allcodons);
 
 f1cdn = permutedims(hcat(frameXcodons.(nostopcodons,1)...));
@@ -55,16 +55,18 @@ r2cdn = reverse.(comp.(f2cdn));
 
 
 
-stopneighborsR1 = unique(permutedims(hcat(frameXcodons.(stop,-1)...)));
-stopneighborsR2 = unique(permutedims(hcat(frameXcodons.(stop,-2)...)));
+stopneighborsR1 = unique(vcat(frameXcpairs.(stop,-1)...), dims = 1);
+stopneighborsR1 = stopneighborsR1[(stopneighborsR1[:,1].∉ Ref(stop)) .& (stopneighborsR1[:,2].∉ Ref(stop)),:];
+
+stopneighborsR2 = unique(vcat(frameXcpairs.(stop,-2)...), dims = 1);
+stopneighborsR2 = stopneighborsR2[(stopneighborsR2[:,1].∉ Ref(stop)) .& (stopneighborsR2[:,2].∉ Ref(stop)),:];
+
 stopneighborsR0 = reverse.(comp.(stop));
 
-stopneighborsR = unique(vcat(vcat(stopneighborsR1,stopneighborsR2,stopneighborsR0)...));
+probSNR1 = sum(nucprob.(stopneighborsR1[:,1],gccontent).*nucprob.(stopneighborsR1[:,2],gccontent));
+probSNR2 = sum(nucprob.(stopneighborsR2[:,1],gccontent).*nucprob.(stopneighborsR2[:,2],gccontent));
 
-probSNR1 = sum(nucprob.(stopneighborsR1,gccontent));
-probSNR2 = sum(nucprob.(stopneighborsR2,gccontent));
 probSNR0 = sum(nucprob.(stopneighborsR0,gccontent));
-probSNR = sum(nucprob.(stopneighborsR,gccontent));
 
 dfedist0 = combinedfe(dfedistgen(nostopcodons,allcodons))
 

### 2022-11-11 Modeling/DFE.jl
Modeling/antisenseGenes.jl

@@ -53,20 +53,7 @@ f2cdn = permutedims(hcat(frameXcodons.(nostopcodons,2)...));
 r1cdn = reverse.(comp.(f1cdn));
 r2cdn = reverse.(comp.(f2cdn));
 
-
-
-stopneighborsR1 = unique(vcat(frameXcpairs.(stop,-1)...), dims = 1);
-stopneighborsR1 = stopneighborsR1[(stopneighborsR1[:,1].∉ Ref(stop)) .& (stopneighborsR1[:,2].∉ Ref(stop)),:];
-
-stopneighborsR2 = unique(vcat(frameXcpairs.(stop,-2)...), dims = 1);
-stopneighborsR2 = stopneighborsR2[(stopneighborsR2[:,1].∉ Ref(stop)) .& (stopneighborsR2[:,2].∉ Ref(stop)),:];
-
-stopneighborsR0 = reverse.(comp.(stop));
-
-probSNR1 = sum(nucprob.(stopneighborsR1[:,1],gccontent).*nucprob.(stopneighborsR1[:,2],gccontent));
-probSNR2 = sum(nucprob.(stopneighborsR2[:,1],gccontent).*nucprob.(stopneighborsR2[:,2],gccontent));
-
-probSNR0 = sum(nucprob.(stopneighborsR0,gccontent));
+probATGR0 = sum(nucprob("ATG",gccontent));
 
 dfedist0 = combinedfe(dfedistgen(nostopcodons,allcodons))
 

### 2022-11-11 Modeling/antisenseGenes.jl
new file:   Modeling/antisenseGenes.jl

@@ -1,0 +1,32 @@
+cd(Base.source_dir())
+include("nucleotidefuncts.jl")
+stop = ["TAA","TAG","TGA"]
+
+function nostop(codonset)
+	codonset[codonset .∉ Ref(stop)]
+end
+
+gccontent = 0.53942 # all CDS #
+nostopcodons = nostop(allcodons);
+
+stopneighborsR1 = unique(vcat(frameXcpairs.(stop,-1)...), dims = 1);
+stopneighborsR1 = stopneighborsR1[(stopneighborsR1[:,1].∉ Ref(stop)) .& (stopneighborsR1[:,2].∉ Ref(stop)),:];
+
+stopneighborsR2 = unique(vcat(frameXcpairs.(stop,-2)...), dims = 1);
+stopneighborsR2 = stopneighborsR2[(stopneighborsR2[:,1].∉ Ref(stop)) .& (stopneighborsR2[:,2].∉ Ref(stop)),:];
+
+stopneighborsR0 = reverse.(comp.(stop));
+
+probstopR1 = sum(nucprob.(stopneighborsR1[:,1],gccontent).*nucprob.(stopneighborsR1[:,2],gccontent));
+probstopR2 = sum(nucprob.(stopneighborsR2[:,1],gccontent).*nucprob.(stopneighborsR2[:,2],gccontent));
+
+probstopR0 = sum(nucprob.(stopneighborsR0,gccontent));
+
+ATGneighborsR1 = unique(frameXcpairs("ATG",-1), dims = 1);
+ATGneighborsR1 = ATGneighborsR1[(ATGneighborsR1[:,1].!= Ref(stop)) .& (ATGneighborsR1[:,2].∉ Ref(stop)),:];
+
+ATGneighborsR2 = unique(frameXcpairs("ATG",-2), dims = 1);
+ATGneighborsR2 = ATGneighborsR2[(ATGneighborsR2[:,1].∉ Ref(stop)) .& (ATGneighborsR2[:,2].∉ Ref(stop)),:];
+
+probATGR1 = sum(nucprob.(ATGneighborsR1[:,1],gccontent).*nucprob.(ATGneighborsR1[:,2],gccontent));
+probATGR2 = sum(nucprob.(ATGneighborsR2[:,1],gccontent).*nucprob.(ATGneighborsR2[:,2],gccontent));

### 2022-11-11 Modeling/antisenseGenes.jl
modified:   Modeling/antisenseGenes.jl

@@ -30,3 +30,4 @@ ATGneighborsR2 = ATGneighborsR2[(ATGneighborsR2[:,1].∉ Ref(stop)) .& (ATGneigh
 
 probATGR1 = sum(nucprob.(ATGneighborsR1[:,1],gccontent).*nucprob.(ATGneighborsR1[:,2],gccontent));
 probATGR2 = sum(nucprob.(ATGneighborsR2[:,1],gccontent).*nucprob.(ATGneighborsR2[:,2],gccontent));
+

### 2022-11-14 Modeling/antisenseGenes.jl
modified:   Modeling/antisenseGenes.jl

@@ -9,25 +9,42 @@ end
 gccontent = 0.53942 # all CDS #
 nostopcodons = nostop(allcodons);
 
+codonpairprob = (cpair) -> sum(nucprob.(cpair[:,1],gccontent) .* nucprob.(cpair[:,2],gccontent))
+
 stopneighborsR1 = unique(vcat(frameXcpairs.(stop,-1)...), dims = 1);
-stopneighborsR1 = stopneighborsR1[(stopneighborsR1[:,1].∉ Ref(stop)) .& (stopneighborsR1[:,2].∉ Ref(stop)),:];
+
+stopNbrR1stopUP = stopneighborsR1[(stopneighborsR1[:,1] .∈ Ref(stop)),:];
+
+stopNbrR1stopDN = stopneighborsR1[(stopneighborsR1[:,1].∉ Ref(stop)) .& (stopneighborsR1[:,2].∈ Ref(stop)),:];
+
+stopNbrR1within = stopneighborsR1[(stopneighborsR1[:,1].∉ Ref(stop)) .& (stopneighborsR1[:,2].∉ Ref(stop)),:];
 
 stopneighborsR2 = unique(vcat(frameXcpairs.(stop,-2)...), dims = 1);
-stopneighborsR2 = stopneighborsR2[(stopneighborsR2[:,1].∉ Ref(stop)) .& (stopneighborsR2[:,2].∉ Ref(stop)),:];
+
+stopNbrR2stopUP = stopneighborsR2[(stopneighborsR2[:,1] .∈ Ref(stop)),:];
+
+stopNbrR2stopDN = stopneighborsR2[(stopneighborsR2[:,1].∉ Ref(stop)) .& (stopneighborsR2[:,2].∈ Ref(stop)),:];
+
+stopNbrR2within = stopneighborsR2[(stopneighborsR2[:,1].∉ Ref(stop)) .& (stopneighborsR2[:,2].∉ Ref(stop)),:];
 
 stopneighborsR0 = reverse.(comp.(stop));
 
-probstopR1 = sum(nucprob.(stopneighborsR1[:,1],gccontent).*nucprob.(stopneighborsR1[:,2],gccontent));
-probstopR2 = sum(nucprob.(stopneighborsR2[:,1],gccontent).*nucprob.(stopneighborsR2[:,2],gccontent));
+p_stopR1within = codonpairprob(stopnbrR1within);
+p_stopR1stopUP = codonpairprob(stopnbrR1stopUP);
[…]

### 2022-11-14 Modeling/nucleotidefuncts.jl

@@ -165,10 +165,14 @@ function mprob(s1,s2)
 end
 
 function nucprob(x,gccontent)
+    if(isempty(x) || occursin(r"[^ATGC]",x))
+        @warn "non-nucleotide string input, returning 0"
+        return 0
+    end
     S = 0.5*gccontent
     W = 0.5 - S
     nw = count(r"[AT]", x)
-    return W^nw * S^(length(x)-nw)
+    return W^nw * S^(count(r"[GC]", x))
 end
 
 function numAT(x)

### 2022-12-02 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 
-\title{Neutral models of \textit{de novo} gene emergence show that gene evolution has a preferred trajectory}
+\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
 \author{Bharat Ravi Iyengar, Erich Bornberg-Bauer}
@@
 New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence.
-To synthesize a protein, DNA must be transcribed as well as translated, both of which need certain DNA sequence features.
-Stable transcription requires promoters and polydenylation signal, while translation requires at least an open reading frame (ORF).
+To synthesize a protein, DNA must be transcribed as well as translated.
+Both processes need certain DNA sequence features.
+Stable transcription requires promoters and a polydenylation signal, while translation requires at least an open reading frame (ORF).
 We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly genes emerge and are lost.
-We also investigate whether different DNA features evolve in a sequential manner, and if sequence composition is biased by mutation rate.
-We find out that genes are lost much more rapidly than they emerge, and genes with long ORFs preferentially arise in regions that are already transcribed.
+We also investigate the effect of the order by which DNA features evolve, and if sequence composition is biased by mutation rate.
+We rationalize how genes are lost much more rapidly than they emerge, and how genes with long ORFs preferentially arise in regions that are already transcribed.
 Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.
@@
 The next big challenge is to define what a function is, and how to classify gene functions.
-One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions.
+One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions \citep{deNovoFunction}.
 Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins like mRNAs, or participate in protein synthesis like rRNAs and tRNAs, are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.
@@
 The second requirement for a productive transcription is its termination at appropriate sites.
-Even though transcription is widespread in eukaryotic genomes, a phenomenon known as pervasive transcription, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}.
+Even though pervasive transcription is widespread in eukaryotic genomes, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}.
 Most eukaryotic mRNAs and long non-coding RNAs, are polyadenylated at their 3' termini, which makes them resistant to degradation.
-Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals.
-In prokaryotes, end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
-Thus the synthesis of a stable RNA, requires promoters, as well as termination/polyadenylation signal sequences.
+Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals \cite{polyA}.
+In prokaryotes, the end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
+Thus the synthesis of a stable RNA requires promoters, as well as termination or polyadenylation signal sequences.
 
@@
 Although gain of transcription and translation features guarantees \textit{de novo} birth of a protein coding gene, it does not ensure that the gene would persist in the genome for many generations.
-The newly born gene can lose the features as easily as it gained them unless it is fixed in the genome via evolutionary selection.
-If the protein synthesized by the \textit{de novo} gene provides some fitness advantage to the host organism then it will undergo positive selection.
-Simulataneously, the protein should have low toxicity and cost of synthesis, to survive purifying selection.
+The newly born gene can lose the features as easily as it gained them unless it has been fixed in the genome, for example via evolutionary selection.
+Specifically, if the protein synthesized by the \textit{de novo} gene provides a fitness advantage to the host organism, it will undergo positive selection \citep{deNovoFunction}.
+Simultaneously, the protein should have low toxicity and cost of synthesis, to survive purifying selection.
 A common mechanism by which protein mediated toxicity occurs is by misfolding and aggregation of proteins \citep{misfold1,misfold2}.
@@
 In this study we develop probabilistic models of \textit{de novo} gene emergence.
-We focus on \textit{de novo} genes emerging from intergenic sequences, also known as ``protogenes'' \citep{protogenes}.
+We focus on \textit{de novo} genes emerging from non-genic sequences, also known as ``proto-genes'' \citep{protogenes,vanOss}.
 We model \textit{de novo} gene emergence as a two step process.
-Specifically, the protogene can first gain transcription features and produce an untranslated RNA, which in turn acquires translation features subsequently.
-Alternatively, the protogene can first gain translational features, and later gain transcription.
-Our models are based on neutral evolution where natural selection does not affect the rate of gene emergence, or loss.
-Using these models, we predict that \textit{de novo} gene emergence has a preferred evolutionary trajectory, even under the assumption of neutrality.
+In the first step, a non-genic DNA sequence gains transcriptional features that allows it to express an untranslated (non-coding) RNA.
+In the second step this non-coding RNA gene acquires translational features that allows it to express a protein.
+Alternatively, translational features can emerge in the non-genic DNA before transcriptional features.
+Specifically, we calculate the probabilities by which promoters, polyadenylation signals and ORFs emerge, and are lost, based on the rates at which different kinds of DNA mutations naturally occur (mutation bias).
+With these probabilities we estimate whether transcriptional and translational features preferentially evolve in a specific order.
+Using a similar approach, we calculate how random mutations affect protein composition.
+In our models, we assume that the proteins expressed from these proto-genes provide no fitness advantage and are not toxic to the host organism.
+Thus our models are based on neutral evolution.
+Our models predict that there is indeed a preferred sequence of DNA feature evolution during \textit{de novo} emergence, even under the assumption of neutrality.
 
@@
 In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss.
-A \textit{de novo} gene emerges from non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation.
+A proto-gene emerges from non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation \citep{protogenes,vanOss}.
 Both transcription and translation are complex processes involving many biomolecular complexes that work in concert.
@@
 For translation, we only require the gene to have an open reading frame (ORF).
+Furthermore, we assume that the ORF is uninterrupted by introns.
 
-Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost via mutagenesis.
+Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost due to random mutations.
 These probability models are essentially described by two kinds of probability.
@@
 Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.
+Our model of DNA sequence evolution is in principle similar to a theoretical model described in a previous study \citep{promoterWaiting}.
+In this study, the authors have calculated the time required for different transcription factor binding sites to emerge in gene promoters.
+We extend this principle to calculate not only the probability of sequence emergence but also the likelihood that an emerged sequence will remain intact or be lost (\autoref{methfeatures}).
 
-For all calculations we used a GC-content of 0.42 which is reasonably close to the global GC-content of human, \textit{Drosophila melanogaster}, and \textit{Saccharomyces cerevisiae} genomes \citep{allgccontents, bionumbers}.
+For all calculations we used a GC-content of 0.42 which is reasonably close to the total GC-content of human \citep[0.41,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[0.416,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[0.38,][]{yeastGC} genomes.
 We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
@@
 
-Overall our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.
+Overall, our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.
 
-\subsection{How many times can a gene be lost?}
+\subsection{How likely is gene loss?}
 
@@
 Specifically, the probability of gene gain can be defined as the sum of three probabilities.
-First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{eqrnagain}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{eqorfstay}).
-Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{eqorfgain}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{eqrnastay}).
+First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{methORF}).
+Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{methORF}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{methRNA}).
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
@@
 \begin{equation}
-P_\textit{gene-gain} = P_\textit{ORF-stay}\times P_\textit{RNA-gain} + P_\textit{RNA-stay}\times P_\textit{ORF-gain} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
+P_\textit{gene-gain} = P_\textit{RNA-gain} \times P_\textit{ORF-stay} + P_\textit{ORF-gain} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
 \label{genegaineq}
@@
 
-Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{eqrnaloss}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{eqorfloss}).
+Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
 Specifically:
@@
 We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for all ORF lengths, the gene can be lost at least two times independently, in the time required to gain it (\autoref{gainlossprob}).
+We found that for all ORF lengths, at least two independent gene loss events can occur in the time frame of one gene gain event (\autoref{gainlossprob}).
 
@@
 
-\subsection{Does \textit{de novo} gene emergence have a preferred trajectory?}
+\subsection{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}
 
@@
 Thus it may appear that the order of the occurrence of these two events does not matter.
-However, gene emergence is much more likely when one of the two features already exist (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
+However, gene emergence is much more likely when one of the two features already exists (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
 To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
@@
 With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\autoref{whoisfirst}).
-This minimum ORF length above which RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content.
+This minimum ORF length above which the RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content.
 For example, this number is 48 codons for a GC-content of 0.45 (data not shown).
@@
 
-\textit{De novo} genes that do not provide any fitness benefit to an organism are not fixed in populations via natural selection.
-These genes can be lost due to mutation pressure.
+\begin{figure}[!b]
+\centering
+\hrule
+\vspace{1ex}
+\includegraphics[scale=0.6]{Figures/pLoss.pdf}
+\caption{ORF loss is more probable than transcription loss for \textit{de novo} genes containing more than 33 codons.
+The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
+\label{lossprob}
+\end{figure}
+
+\textit{De novo} genes that do not provide any fitness benefit to an organism may not be fixed in populations via natural selection.
+These genes may be lost due to mutation pressure.
 Some newborn \textit{de novo} genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way.
 These genes would thus be eliminated from the population genomes via negative selection.
-We remind the reader that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation).
-It is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively).
+We note that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation).
+However, it is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively).
 Furthermore, non-stop mutations can also lead to cellular toxicity \citep{nonstop}.
-Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may be the better way to inactivate the associated genes.
+Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may more effectively inactivate the associated genes.
 
@@
 The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$.
-We find that the first scenario is more plausible for ORFs that have less than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}).
+We find that the first scenario is more plausible for ORFs that have fewer than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}).
 Also, the probability values of scenario one and two, obtained from this stringent analysis, were not significantly different from the transcription ($P_\textit{RNA-loss}$) and ORF loss ($P_\textit{ORF-loss}$) probabilities, respectively.
@@
 Thus, widespread transcription loss can suggest a negative selection on protein expression.
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.6]{Figures/pLoss.pdf}
-\caption{ORF loss is more probable than transcription loss for \textit{de novo} genes containing more than 33 codons.
-The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
-\label{lossprob}
-
-\vspace{1ex}
-\hrule
-\end{figure}
 
@@
 For example, leucine, arginine and serine, all have six codons each.
-However, leucine is more probable than serine which in turn is more probable than arginine (\autoref{aafreq}{\color{blue}A}).
+However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 0.42 (\autoref{aafreq}{\color{blue}A}).
+Overall, our analysis is roughly in agreement with many previous studies \citep{aasubOhta,aafreq3,aafreq4}.
+However, notable differences exist between our analysis and that of the previous studies because the latter primarily focused on characterized proteins and not random proto-gene derived protein sequences.
 
@@
 Thus an amino acid substitution (for example K$\to$E) may not be as likely to occur as the reverse substitution (E$\to$K).
+That is substitutions between a pair of amino acids may have a directionality.
+Although many previous studies have explored the likelihood of different amino acid substitutions \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}, none to our knowledge have focused primarily on the directionality of these substitutions.
 To verify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{x\to y}$, for every pair of amino acids, $x$ and $y$.
@@
 We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}).
-Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that the median value is 0.38.
+Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that its median value is 0.38.
 Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.
 
-A previous study has argued that proteins that haven't been shaped via selection, tend to be hydrophobic because hydrophobic amino acids comprise 40\% of amino acids and can also easily arise in a protein sequence due to mutation pressure \citep{hydrophobicRatchet}.
-They based these conclusions on mutation rate bias estimated in bacteria \citep{bacteriaMutbias}.
+A previous study found that mutations that cause hydrophobic amino acids to appear on protein surfaces, can lead to evolution of protein dimers \citep{hydrophobicRatchet}.
+The authors \citep{hydrophobicRatchet} also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids.
+They based this argument on another study that estimated mutation rate bias in bacteria \citep{bacteriaMutbias}.
+The authors also argued that hydrophobic amino acids are more frequently found in random protein sequences.
+
 We asked if our model also makes similar predictions.
 To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic.
-We classified an amino acid hydrophobic if its hydropathy index \citep{hydropathy} is greater than 0 (that is A, C, F, I, L, M \& V).
-Based on this definition and the expected frequency of amino acids, we found that $\sim$39\% of amino acids in a protein sequence are expected to be hydrophobic.
-This finding is in agreement with the previous studies.
+A popular hydrophobicity scale of amino acids is based on solubility of an amino acid in water or ethanol \citep{hydropathy}.
+However, this scale does not classify tryptophan as a hydrophobic amino acid.
+Another hydrophobicity scale estimated by a different study \citep{WWhydropathy} is more biologically realistic and is based on rates of transfer of different amino acids between water and a hydrophobic medium.
+This study analysed two hydrophobic media -- a lipid bilayer and octanol, and also considered the effect of peptide bonds in the calculations.
+Therefore we used the hydrophobicity scale by \cite{WWhydropathy} for classifying amino acids such that hydrophobic amino acids have a negative free energy change of transfer from water to hydrophobic media.
+Conversely, a hydrophilic amino acid would have a positive value of the free energy change.
+Based the octanol hydrophobicity scale, we classified the following amino acids as hydrophobic -- cysteine (C), phenylalanine (F), isoleucine (I), leucine (L), methionine (M), valine (V), tryptophan (W) and tyrosine (Y).
+With the lipid bilayer hydrophobicity scale, valine is not classified as hydrophobic, possibly because it does not integrate well in the bilayer.
+Therefore we used the octanol scale for our calculations.
+
+We found that $\sim$37.5\% of amino acids in a protein sequence are expected to be hydrophobic.
+This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}.
 Next, we asked if a protein sequence in general tends to be hydrophobic.
-Specifically, we calculated expected hydropathy ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydropathic index values ($\alpha_x$).
+Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).
 
@@
 
-We found that the expected hydropathy is $-0.08628$, which suggests that random protein sequences are not generally hydrophobic.
+We found that the expected hydropathy is higher than zero ($0.39$), which suggests that random protein sequences are not  on an average, hydrophobic.
 
@@
 
-Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it was 1.053.
-This shows that random mutations do not preferentially cause hydrophobic amino acids to accumulate in protein sequence.
-Inclusion of proline and tryptophan in the list of hydrophobic amino acids does not change this finding.
+Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is 1.114, which suggests that gain of a hydrophobic amino acid is slightly more probable than its loss.
+We performed a complementary analysis where we calculated the average change in hydrophobicity due to any random mutation ($\bar{\Delta\alpha}$), defined by:
+
+\begin{equation}
+\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y}}
+\end{equation}
+
+We found that the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity.
+If we exclude amino acid pairs where both amino acids are either hydrophobic or are hydrophilic, then the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).
+
+Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect is very small.
 
@@
 
+\label{methRNA}
 To calculate the probabilities of finding, gaining, and losing transcription, we first defined the minimum requirements for transcription.
@@
 
+\label{methORF}
+
 \subsubsection{Probability of finding an ORF}
@@
 We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
+Specifically, our model is implemented in the script \texttt{DeNovoProb.jl} which in turn uses the script \texttt{nucleotidefuncts.jl} for some basic functions.
 
-\bibliographystyle{plainnat}
+\bibliographystyle{mybst}
 

### 2022-12-02 Modeling/DeNovoProb.jl

@@ -203,13 +203,16 @@ onlyrnaloss = orfstay .*(rnaprob*rnaloss)
 
 aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];
 aaprob2 = aaprob./sum(aaprob);
-exp_hydropathy = sum(hydropathy.*aaprob2)
+exp_hydropathy = sum(hydropathy3.*aaprob2)
 
 acidic = indexin(["D","E"],aas);
 basic = indexin(["K","R"],aas);
-hydrophobic = hydropathy .>0;
+hydrophobic = hydropathy2 .<0;
 
-pHydrophobic = sum(aaprob[hydrophobic]);
+allcodons = kmers(3);
+nostopcodons = allcodons[allcodons .∉ Ref(["TAA","TAG","TGA"])]
+
+pHydrophobic = sum(aaprob[hydrophobic3]);
 
 plot_aafreq = bar(aas,aaprob2,
     xticks = (0.5:20, aas),
@@ -251,10 +254,31 @@ gainAA = Vector{Float64}(vec(sum(aasubprob, dims=1)'));
 lossAA = Vector{Float64}(vec(sum(aasubprob, dims=2)./aaprob2));
 stayAA = [sum(featurestay(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];
 
+getHP = (a) -> hydropathy3[aas .== a]
 
 codonshydrophobic = gencode[(in).(gencode[:,3],Ref(aas[hydrophobic])),1];
 
 codonshydrophilic = kmers(3)[(!in).(kmers(3),Ref(codonshydrophobic))];
+
+hmp = 0
+mp=0
+for i in nostopcodons
+    for j in nostopcodons
+        x = mprob(i,j)
+        yj = getHP(transnucs(j))[1]
+        yi = getHP(transnucs(i))[1]
+        if(sign(yi)!=sign(yj))
+            y = yj - yi
+        else
+            y = 0
+        end
+        z = nucprob(i,gccontent)
+        hmp += x*y*z
+        mp += x
+    end
+end
+MutΔHydrophobic = hmp/mp
+
 
 toHydrophobic = featuregain(codonshydrophilic,codonshydrophobic, gccontent);
 

### 2022-12-02 Modeling/nucleotidefuncts.jl

@@ -12,6 +12,11 @@ aas = ["A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","V","
 
 hydropathy = [1.8, 2.5, -3.5, -3.5, 2.8, -0.4, -3.2, 4.5, -3.9, 3.8, 1.9, -3.5, -1.6, -3.5, -4.5, -0.8, -0.7, 4.2, -0.9, -1.3]; # Kyte and Doolittle 
 
+hydropathy2 = [0.17, -0.24, 1.23, 2.02, -1.13, 0.01, 0.96, -0.31, 0.99, -0.56, -0.23, 0.42, 0.45, 0.58, 0.81, 0.13, 0.14, 0.07, -1.85, -0.94
+]; # Wimley and White
+
+hydropathy3 = [0.5, -0.02, 3.64, 3.63, -1.71, 1.15, 2.33, -1.12, 2.8, -1.25, -0.67, 0.85, 0.14, 0.77, 1.81, 0.46, 0.25, -0.46, -2.09, -0.71
+]; # Wimley and White
 """
 # Generate all nucleotide sequences with length `k`
 """

### 2022-12-02 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -207,12 +207,12 @@ exp_hydropathy = sum(hydropathy3.*aaprob2)
 
 acidic = indexin(["D","E"],aas);
 basic = indexin(["K","R"],aas);
-hydrophobic = hydropathy2 .<0;
+hydrophobic = hydropathy3 .<0;
 
 allcodons = kmers(3);
 nostopcodons = allcodons[allcodons .∉ Ref(["TAA","TAG","TGA"])]
 
-pHydrophobic = sum(aaprob[hydrophobic3]);
+pHydrophobic = sum(aaprob[hydrophobic]);
 
 plot_aafreq = bar(aas,aaprob2,
     xticks = (0.5:20, aas),

### 2022-12-04 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -1,7 +1,7 @@
 using Combinatorics
 using Plots
 using Measures
-using HypothesisTests
+# using HypothesisTests
 
 include("nucleotidefuncts.jl")
 
@@ -11,7 +11,8 @@ figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
 default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
 guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
 
-
+aaorder = ["W","F","Y","I","V","L","M","C","D","E","G","A","P","H","K","R","S","T","N","Q"];
+ixaaorder = indexin(aaorder,aas);
 gccontent = 0.42
 
 function rnavals(gccontent)
@@ -214,8 +215,8 @@ nostopcodons = allcodons[allcodons .∉ Ref(["TAA","TAG","TGA"])]
 
 pHydrophobic = sum(aaprob[hydrophobic]);
 
-plot_aafreq = bar(aas,aaprob2,
-    xticks = (0.5:20, aas),
+plot_aafreq = bar(aaorder,aaprob2[ixaaorder],
+    xticks = (0.5:20, aaorder),
     yticks = (0:0.03:0.9),
     fill = :black,
     xlabel = "Amino acid",
@@ -286,9 +287,9 @@ frHydrophobic = featuregain(codonshydrophobic,codonshydrophilic, gccontent);
 
 stayHydrophobic = featurestay(codonshydrophobic, gccontent);
 
-plot_aasubprob = heatmap(aas[idx],aas[idx],log2.(aasubhighprob[idx,idx]),
-    xticks = (0.5:20, aas[idx]), 
-    yticks = (0.5:20,aas[idx]),
+plot_aasubprob = heatmap(aaorder,aaorder,log2.(aasubhighprob[ixaaorder,ixaaorder]),
+    xticks = (0.5:20, aaorder), 
+    yticks = (0.5:20, aaorder),
     size = (width = cm2pt(15), height = cm2pt(14)),
     colorbar_title = "Substitution Probability",
     legend = :bottom
@@ -296,9 +297,9 @@ plot_aasubprob = heatmap(aas[idx],aas[idx],log2.(aasubhighprob[idx,idx]),
 
 savefig(plot_aasubprob, figdir*"pAASub.pdf")
 
[…]

### 2022-12-04 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis).
-Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.}
+Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.
+In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
 \label{aafreq}
@@
 A previous study found that mutations that cause hydrophobic amino acids to appear on protein surfaces, can lead to evolution of protein dimers \citep{hydrophobicRatchet}.
-The authors \citep{hydrophobicRatchet} also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids.
+The authors also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids.
 They based this argument on another study that estimated mutation rate bias in bacteria \citep{bacteriaMutbias}.
-The authors also argued that hydrophobic amino acids are more frequently found in random protein sequences.
+The authors \citep{hydrophobicRatchet} further suggested that hydrophobic amino acids are more frequently found in random protein sequences.
 
@@
 
-We found that the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity.
-If we exclude amino acid pairs where both amino acids are either hydrophobic or are hydrophilic, then the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).
+We found the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity.
+We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein.
+To focus exclusively on non-hydrophobic to hydrophobic amino acid substitutions, we excluded amino acid pairs where both amino acids are either hydrophobic or are non-hydrophobic.
+That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity.
+With this focused analysis, we find that the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).
 
@@
 
-In this work, we addressed some fundamental questions about \textit{de novo} gene evolution.
-Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory, and how the composition of their protein sequence is determined.
+In this work, we addressed some fundamental questions about \textit{de novo} emergence of protogenes.
+Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory of mutational events, and how the composition of their protein sequence is determined.
 To this end, we developed mathematical models and used them to address specific questions.
@@
 
-We answer the long standing question of what is the sequence of \textit{de novo} emergence.
+Witin the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence.
 That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.

### 2022-12-04 Modeling/DeNovoProb.jl

@@ -320,7 +320,7 @@ plot_aasymprob2 = heatmap(aaorder,aaorder,asymmetric_conditional[ixaaorder,ixaao
     xlabel = "Substituted amino acid",
     c = :bluesreds
 );
-savefig(plot_aasymprob2, figdir*"pAAsubSymCond.pdf")
+savefig(plot_aasymprob2, figdir*"pAAsubSymCond.svg")
 
 plot_toaa = bar(aaorder,gainAA[ixaaorder]./sum(gainAA),
     xticks = (0.5:20, aaorder),

### 2022-12-05 Manuscripts/deNovoRates.tex
modified:   Modeling/DeNovoProb.jl

@@
 \title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
-\author{Bharat Ravi Iyengar, Erich Bornberg-Bauer}
-\date{\small Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany}
+\author{Bharat Ravi Iyengar$^1$, Erich Bornberg-Bauer$^{1,2}$}
+\date{\small $^1$Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany\\ $^2$Max Planck-Institute for Biology T\"{u}bingen, T\"{u}bingen, Germany}
 
@@
 Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.
+We only show substitutions that can be realized by a single nucleotide mutation but we include all possible nucleotide mutations in the directionality calculations for these substitutions.
 In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
@@
 This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}.
+We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}.
 Next, we asked if a protein sequence in general tends to be hydrophobic.
@@
 We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge.
-Therefore, presence of a \textit{de novo} gene in one species (or population) cannot be interpreted as an evidence of gene emergence in this species.
+In a hypothetical case, a proto-gene may be present in a species (A) but is absent in two sister taxa (B and C).
+Two evolutionary scenarios could explain this observation.
+In the first scenario, the proto-gene simply emerged for the first time in species A.
+The second scenario posits that the proto-gene was present in the common ancestor of the three species but was lost in both species B and species C.
+Our models suggest that the second scenario is more likely.
+Future phylogenetic studies on proto-genes should consider this finding for making inferences on the dynamics of gene gain and loss.
 
-Witin the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence.
+Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence.
 That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
@@
 Specifically, it is difficult to infer from data if transcription was lost or never emerged.
-If the protein encoded by an ORF is indeed toxic, it would be silenced rapidly in microevolutionary time scales.
+If the protein encoded by an ORF is indeed toxic, it would be silenced or purged rapidly in microevolutionary time scales.
 One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status.
-Mutations within ORF sequence are much more probable than mutations that disrupt transcription or ORF.
-This is so because the entire ORF sequence has higher number of mutable sites.
+Mutations within an ORF sequence are much more probable than mutations that disrupt transcription or an ORF.
+This is so because the entire ORF sequence can have a higher number of mutable sites than a promoter.
 If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.
 
-Finally, we show that some amino acid substitutions are more likely to occur than others, and frequent substitutions that are expected to occur less frequently under neutral model, can suggest directional evolution.
+A large volume of work exists on the likelihood of amino acid substitutions and many of these models are widely used for phylogenetics \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}.
+However commonly used amino acid substitution models such as PAM and BLOSUM \citep{PAM,blosum} do not focus on directionality of mutations (the substitution matrices are symmetric).
+Our aim was to highlight how proteins can evolve under mutation pressure and that mutation bias causes some substitutions to occur more frequently than the others.
+Furthermore, we do not base our substitution probability calculations on known protein sequences but rather on random protein sequences that are more likely to to be encoded in proto-genes.
+Using our models, we show that for a pair of amino acids, one direction of substitution can be significantly more likely than the other.
+If a substitution occurs more frequently than it is expected to occur in our neutral model, then directional evolution can be a possible explanation for the observation.
+We also show that random mutations can lead to a small increase in a protein's hydrophobicity.
+Hydrophobic amino acids can facilitate folding \citep{Dill1985}, but can also cause proteins to aggregate \cite{hydrophobicRatchet}.
+Thus the effect of hydrophobicity increasing mutations on a protein's stability and toxicity, is dependent on where they occur in the protein structure.
 
@@
 This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
+Fourth, we did not model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes.
+For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}.
+We instead focused on minimal requirements for transcription and translation.
+However, our modeling framework allows inclusion of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}).
 Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation.
 Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations.
-However, our modeling methodology allows incorporation of these mechanisms as a source of mutations.
+Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.
 
 A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}.
-This is true for our models too.
+Our models are not an exception.
 They are not 100\% accurate but they are useful for making predictions that can lead to a more focused experimental validation.
@@
 \subsection{Mutation probabilities}
-
-We calculated nucleotide substitution probabilities based on mutation rate and mutation rate bias data.
-Specifically, we  used a mutation rate ($\lambda$) of $7.8\times10^{-9}$ mutations per nucleotide position per generation \citep{drosophilamutrate}.
-We derived our mutation bias parameters from two published studies, the first on \textit{Drosophila melanogaster} \citep{drosophilamutrate}, and the second on humans \citep{humanmutrate}.
-\autoref{mutbias} shows the exact values of mutation bias probabilities that we used in this study.
 
@@
 \end{table}
+
+We calculated nucleotide substitution probabilities based on mutation rate and mutation rate bias data.
+Specifically, we  used a mutation rate ($\lambda$) of $7.8\times10^{-9}$ mutations per nucleotide position per generation \citep{drosophilamutrate}.
+We derived our mutation bias parameters from two published studies, the first on \textit{Drosophila melanogaster} \citep{drosophilamutrate}, and the second on humans \citep{humanmutrate}.
+\autoref{mutbias} shows the exact values of mutation bias probabilities that we used in this study.
 
@@
 Thus we defined the probability ($P_\textit{feature-stay}$) that a feature does not erode because of mutations.
-Specifically, it is the the sum of two probabilities.
-First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where mutated sequence is still a part of the feature set.
+Specifically, it is the sum of two probabilities.
+First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where the mutated sequence remains a part of the feature set.
 

### 2022-12-05 Modeling/DeNovoProb.jl

@@ -144,6 +144,8 @@ ncodons = [30:300;];
 ATGvalues = ATGvals(gccontent);
 stopvalues = stopvals(gccontent);
 
+gcrange = [0.3:0.05:0.6;];
+ovalsgc = hcat([orfvals(x,ATGvals(x),stopvals(x),100) for x in gcrange]...)
 
 (rnaprob, rnagain, rnaloss, rnastay) = rnavals(gccontent)
 (orfprob, orfgain, orfloss, orfstay) = [zeros(size(ncodons)) for i = 1:4 ]
@@ -170,7 +172,22 @@ plot_genegain_geneloss = plot(ncodons,log.(genegain2)./log.(geneloss),
     ylabel = "# Gene Losses \n per Gene Gain",
     size = (width = cm2pt(12), height = cm2pt(11)),
 );
-savefig(plot_genegain_geneloss, figdir*"geneGainLoss.pdf")
+savefig(plot_genegain_geneloss, figdir*"geneGainLoss100GC.pdf")
+
+gcrange = [0.3:0.05:0.6;];
+rvalsgc = hcat([rnavals(x) for x in gcrange]);
+ovalsgc = [hcat([orfvals(x,ATGvals(x),stopvals(x),y) for x in gcrange]...) for y in ncodons];
+ovalsgc50 = ovalsgc[findfirst(ncodons .==50)];
+genegain50 = ((rvalsgc[2,:] + rvalsgc[4,:]) .* ovalsgc50[2,:] .+ ovalsgc50[4,:] .* rvalsgc[2,:]);
+genegain2_50 = genegain50./(1 .-rvalsgc[1,:] - ovalsgc50[1,:]);
+geneloss50 = ovalsgc50[3,:] .+ rvalsgc[3,:]
+
+plot_genegain_geneloss_50_gcrange = plot(gcrange,log.(genegain2_50)./log.(geneloss50),
+    xlabel = "GC-content",
+    ylabel = "# Gene Losses \n per Gene Gain",
+    size = (width = cm2pt(12), height = cm2pt(11)),
+);
+savefig(plot_genegain_geneloss_50_gcrange, figdir*"geneGainLoss50gcr.pdf")
 
 plot_rnafist_orffirst = plot(ncodons[1:50],log.(orffirst./rnafirst)[1:50],
     xlabel = "ORF length (codons)",
[…]

### 2022-12-05 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -234,6 +234,13 @@ plot_Loss = plot(ncodons[1:85],log.(orfloss./rnaloss)[1:85],
 
 savefig(plot_Loss, figdir*"pLoss.pdf")
 
+plot_loss_50_gcrange = plot(gcrange,log.(ovalsgc50[3,:]./rvalsgc[3,:]),
+    xlabel = "GC-content",
+    ylabel = "P_RNA-loss\nP_ORF-loss",
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
+);
+savefig(plot_loss_50_gcrange, figdir*"pLoss50gcr.pdf")
+
 onlyrnaloss = orfstay .*(rnaprob*rnaloss)
 
 aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];

### 2022-12-05 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 
-For all calculations we used a GC-content of 0.42 which is reasonably close to the total GC-content of human \citep[0.41,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[0.416,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[0.38,][]{yeastGC} genomes.
+For all calculations we used a default GC-content of 0.42 which is reasonably close to the total genomic GC-content of human \citep[0.41,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[0.416,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[0.38,][]{yeastGC}.
 We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
@@
 \centering
-\includegraphics[scale=0.6]{Figures/geneGainLoss.pdf}
+\includegraphics[scale=0.6]{Figures/geneGainLossComplete.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF.
+The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.
 \autoref{genegaineq} describes total gene gain probability.
@@
 We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for all ORF lengths, at least two independent gene loss events can occur in the time frame of one gene gain event (\autoref{gainlossprob}).
+We found that for all ORF lengths, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
+This was also the case for a proto-gene containing 50 codons in its ORF and having different amounts of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
 
@@
 
-Overall, our analysis suggests that a \textit{de novo} gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.
+Overall, our analysis suggests that a proto-gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.
 
@@
 
-For a \textit{de novo} gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge.
+For a proto-gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge.
 That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain.
@@
 %\caption{One-step}
-\includegraphics[scale=0.6]{Figures/first_ORF_RNA.pdf}
+\includegraphics[scale=0.6]{Figures/whichfirstComplete.pdf}
 %\end{subfigure}\hfill%
@@
 %\end{subfigure}
-\caption{\textit{De novo} genes with shorter ORFs ($<$ 42 codons) preferentially emerge from existing ORFs (ORF-first) whereas \textit{de novo} genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
-The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
+\caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
+\textbf{(A)} The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence, with a GC-content of 0.42 (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
 A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory.
-Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).}
+Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
+\textbf{(B)} Minimum ORF length (codons) that makes RNA-first trajectory more feasible (vertical axis) for different GC-content (horizontal axis).
+For a GC-content of 0.3 and 0.35, proto-genes of every analysed ORF length ($\geq33$ codons) preferentially emerge RNA-first.}
 \label{whoisfirst}
@@
 Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible.
-With the specific parameter values we used, it turns out that \textit{de novo} genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\autoref{whoisfirst}).
+With the specific parameter values we used, it turns out that proto-genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
 This minimum ORF length above which the RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content.
-For example, this number is 48 codons for a GC-content of 0.45 (data not shown).
+For example, this number is 48 codons for a GC-content of 0.45 (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
 
@@
 With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}\ \%$).
-However, we still find that \textit{de novo} genes with shorter ORFs ($<$ 34 codons) have a slightly higher chance of emerging ORF first.
+However, we still find that proto-genes with shorter ORFs ($<$ 34 codons) have a slightly higher chance of emerging ORF first.
 
@@
 \vspace{1ex}
-\includegraphics[scale=0.6]{Figures/pLoss.pdf}
-\caption{ORF loss is more probable than transcription loss for \textit{de novo} genes containing more than 33 codons.
-The horizontal axis denotes the number of codons in the ORF.
+\includegraphics[scale=0.6]{Figures/pLossComplete.pdf}
+\caption{ORF loss is more probable than transcription loss for proto-genes containing more than 33 codons.
+The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
 The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
@@
 
-\textit{De novo} genes that do not provide any fitness benefit to an organism may not be fixed in populations via natural selection.
+Proto-genes that do not provide any fitness benefit to an organism may not be fixed in populations via natural selection.
 These genes may be lost due to mutation pressure.
-Some newborn \textit{de novo} genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way.
+Some newborn proto-genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way.
 These genes would thus be eliminated from the population genomes via negative selection.
@@
 To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, for all ORFs longer than 33 codons (\autoref{lossprob}).
+We found that ORF loss is more probable than transcription loss, for all ORFs longer than 33 codons (\hyperref[lossprob]{\autoref{lossprob}A}).
+Loss of a longer ORF (50 codons) was more probable than transcription loss for a wide range of GC-content (\hyperref[lossprob]{\autoref{lossprob}B}).
 
-We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact \textit{de novo} gene.
+We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact proto-gene.
 In the first scenario, the ORF remains intact whereas the transcription is lost.
@@
 In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss.
-We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of \textit{de novo} proteins.
+We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of proteins encoded by proto-genes.
 To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform.
@@
 These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection.
-Thus our models provide an opportunity to test the alternative hypothesis that some \textit{de novo} genes are evolving under selection.
+Thus our models provide an opportunity to test the alternative hypothesis that some proto-genes are evolving under selection.
 
@@
 We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
-Long non-coding RNAs may be a potential source of \textit{de novo} genes with long ORFs but this evolutionary process may be constrained by a different set of factors.
+Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors.
 A previous study has shown that new transcripts frequently arise in regions overlapping existing genes.
@@
 
-Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the \textit{de novo} gene.
+Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
 For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory.

### 2022-12-06 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 \captionsetup[figure]{labelfont=bf,textfont={small,normalfont}}
-\captionsetup[subfigure]{labelsep=quad, labelfont={bf,small}, textfont=small, singlelinecheck=off, justification=raggedright, position=top}
-\renewcommand{\thesubfigure}{\Alph{subfigure}}
+%\captionsetup[subfigure]{labelsep=quad, labelfont={bf,small}, textfont=small, singlelinecheck=off, justification=raggedright, position=top}
+%\renewcommand{\thesubfigure}{\Alph{subfigure}}
 
@@
 Most eukaryotic mRNAs and long non-coding RNAs, are polyadenylated at their 3' termini, which makes them resistant to degradation.
-Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals \cite{polyA}.
+Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals \citep{polyA}.
 In prokaryotes, the end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
@@
 \centering
-\includegraphics[scale=0.6]{Figures/geneGainLossComplete.pdf}
+\includegraphics[scale=0.5]{Figures/geneGainLossComplete.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
@@
 %\caption{One-step}
-\includegraphics[scale=0.6]{Figures/whichfirstComplete.pdf}
+\includegraphics[scale=0.5]{Figures/whichfirstComplete.pdf}
 %\end{subfigure}\hfill%
@@
 \vspace{1ex}
-\includegraphics[scale=0.6]{Figures/pLossComplete.pdf}
+\includegraphics[scale=0.5]{Figures/pLossComplete.pdf}
 \caption{ORF loss is more probable than transcription loss for proto-genes containing more than 33 codons.
@@
 \centering
-\begin{subfigure}{0.47\textwidth}
-\caption{}
-\flushleft
-\includegraphics[scale=0.5]{Figures/pAAfreq.pdf}
-\end{subfigure}\hfill%
-\begin{subfigure}{0.47\textwidth}
-\caption{}
-\flushleft
-\includegraphics[scale=0.5]{Figures/pToAA.pdf}
-\end{subfigure}
-
-\vspace{1em}
-
-\begin{subfigure}[t]{0.47\textwidth}
-\caption{}
-\flushleft
-\includegraphics[scale=0.5]{Figures/pFrAA.pdf}
-\end{subfigure}\hfill%
-\begin{subfigure}[t]{0.47\textwidth}
-\caption{}
-\flushright
-\includegraphics[scale=0.5]{Figures/pAAsubSymCond.pdf}
-\end{subfigure}
+\includegraphics[scale=0.5]{Figures/proteinComposition.pdf}
 
@@
 
-A large volume of work exists on the likelihood of amino acid substitutions and many of these models are widely used for phylogenetics \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}.
-However commonly used amino acid substitution models such as PAM and BLOSUM \citep{PAM,blosum} do not focus on directionality of mutations (the substitution matrices are symmetric).
+A large volume of work exists on the likelihood of amino acid substitutions and many of these models are widely used in phylogenetics studies \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}.
+However, commonly used amino acid substitution models such as PAM and BLOSUM \citep{PAM,blosum} do not focus on directionality of mutations (the substitution matrices are symmetric).
 Our aim was to highlight how proteins can evolve under mutation pressure and that mutation bias causes some substitutions to occur more frequently than the others.
@@
 We also show that random mutations can lead to a small increase in a protein's hydrophobicity.
-Hydrophobic amino acids can facilitate folding \citep{Dill1985}, but can also cause proteins to aggregate \cite{hydrophobicRatchet}.
+Hydrophobic amino acids can facilitate folding \citep{Dill1985}, but can also cause proteins to aggregate \citep{hydrophobicRatchet}.
 Thus the effect of hydrophobicity increasing mutations on a protein's stability and toxicity, is dependent on where they occur in the protein structure.
@@
 To calculate the conditional probability of promoter loss, we calculated the fraction of transcription units ($Q$) that belong to each of these three classes.
-For example, the fraction of transcription units that are TATA-only is ($Q_\textit{TATA-only}$):
+For example, the fraction of transcription units that are TATA-only ($Q_\textit{TATA-only}$) is:
 

### 2022-12-06 Modeling/DeNovoProb2.jl
new file:   Modeling/DeNovoProb2.jl

@@ -1,0 +1,392 @@
+using Combinatorics
+using Plots
+using Measures
+# using HypothesisTests
+
+include("nucleotidefuncts.jl")
+
+cm2pt = (cm) -> 28.3465*cm
+figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
+
+default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
+guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
+
+aaorder = ["W","F","Y","I","V","L","M","C","D","E","G","A","P","H","K","R","S","T","N","Q"];
+ixaaorder = indexin(aaorder,aas);
+gccontent = 0.42
+
+(xProb, xGain, xLoss, xStay) = [i for i in 1:4];
+
+
+function rnavals(gccontent,orflen)
+    n_6mersORF = 3*orflen - 6 + 1
+    n_nopolyaORF = n_6mersORF - trunc(Int,n_6mersORF/6)
+    ## TATA box
+    # Consensus sequence = TATAWAWR
+    # is TSS - 25 - 8nt (transcript must be +33nt longer)
+
+    tataboxes = vec(join.(product(("TATA" .* collect("TA") .* "A"), "TA", "GA")))
+
+    tatavars = unique(reduce(vcat,
+        reduce(vcat, [
+[degen(s,1,nucs) for s in tataboxes],
+[degen(s,2,"TA") for s in tataboxes],
+[degen(s,3,"TAC") for s in tataboxes],
+[degen(s,4,"AT") for s in tataboxes],
+[degen(s,6,nucs) for s in tataboxes],
+[degen(s,7,nucs) for s in tataboxes],
+[degen(s,8,nucs) for s in tataboxes],
+        ])
+    ));
+
+    notata = setdiff(kmers(8),tatavars);
+
+    tataprob = sum([nucprob(x,gccontent) for x in tatavars]);
+    tatagain = featuregain(notata,tatavars,gccontent);
+    tataloss1 = featuregain(tatavars,notata,gccontent);
+    tataloss2 = tataloss1/tataprob;
+    tatastay = featurestay(tatavars,gccontent);
+    notatastay = 1 - tatagain;
+
+    ## Initiator element
[…]

### 2022-12-06 Modeling/DeNovoProb2.jl
modified:   Modeling/DeNovoProb2.jl

@@ -76,7 +76,10 @@ function rnavals(gccontent,orflen)
     # (Initiator or TATA) and polyA
 
     rnaprob = (tataprob*nopolyaprob^27 + inrprob)*polyaprob*nopolyaprob^n_nopolyaORF;
-    rnagain = (tatagain + inrgain)*polyastay + (tatastay + inrstay)*polyagain;
+    rnagain = (tatagain + inrgain)*polyastay*nopolyastay^n_nopolyaORF + 
+    (tatastay + inrstay)*polyagain*nopolyastay^n_nopolyaORF +
+    (n_nopolyaORF-1)*(tatastay + inrstay)*polyastay*polyaloss1*nopolyastay^(n_nopolyaORF-1);
+
     Qinr = inrprob*(1-tataprob)/(tataprob + inrprob - inrprob*tataprob);
     Qtata = tataprob*(1-inrprob)/(tataprob + inrprob - inrprob*tataprob);
     Qboth = 1 - Qtata - Qinr;
@@ -84,9 +87,9 @@ function rnavals(gccontent,orflen)
     rnaloss = (Qinr*inrloss2*notatastay +
       Qtata*tataloss2*noinrstay +
       Qboth*tataloss2*inrloss2 +
-      polyaloss2
+      polyaloss2 + n_nopolyaORF*polyagain
     );
-    rnastay = (inrstay + tatastay)*polyastay
+    rnastay = (inrstay + tatastay)*polyastay*nopolyastay^(n_nopolyaORF-1);
     return [rnaprob; rnagain; rnaloss; rnastay]
 end
 

### 2022-12-07 Modeling/DeNovoProb2.jl
modified:   Modeling/DeNovoProb2.jl

@@ -17,25 +17,24 @@ gccontent = 0.42
 
 (xProb, xGain, xLoss, xStay) = [i for i in 1:4];
 
+mers6 = kmers(6);
+allcodons = kmers(3)
 
-function rnavals(gccontent,orflen)
-    n_6mersORF = 3*orflen - 6 + 1
-    n_nopolyaORF = n_6mersORF - trunc(Int,n_6mersORF/6)
+function tataprobs(gccontent)
     ## TATA box
     # Consensus sequence = TATAWAWR
     # is TSS - 25 - 8nt (transcript must be +33nt longer)
-
     tataboxes = vec(join.(product(("TATA" .* collect("TA") .* "A"), "TA", "GA")))
 
     tatavars = unique(reduce(vcat,
         reduce(vcat, [
-[degen(s,1,nucs) for s in tataboxes],
-[degen(s,2,"TA") for s in tataboxes],
-[degen(s,3,"TAC") for s in tataboxes],
-[degen(s,4,"AT") for s in tataboxes],
-[degen(s,6,nucs) for s in tataboxes],
-[degen(s,7,nucs) for s in tataboxes],
-[degen(s,8,nucs) for s in tataboxes],
+            [degen(s,1,nucs) for s in tataboxes],
+            [degen(s,2,"TA") for s in tataboxes],
+            [degen(s,3,"TAC") for s in tataboxes],
+            [degen(s,4,"AT") for s in tataboxes],
+            [degen(s,6,nucs) for s in tataboxes],
+            [degen(s,7,nucs) for s in tataboxes],
+            [degen(s,8,nucs) for s in tataboxes],
         ])
     ));
 
@@ -43,185 +42,223 @@ function rnavals(gccontent,orflen)
 
     tataprob = sum([nucprob(x,gccontent) for x in tatavars]);
     tatagain = featuregain(notata,tatavars,gccontent);
-    tataloss1 = featuregain(tatavars,notata,gccontent);
-    tataloss2 = tataloss1/tataprob;
+    tataloss = featuregain(tatavars,notata,gccontent)/tataprob;
     tatastay = featurestay(tatavars,gccontent);
-    notatastay = 1 - tatagain;
+    return [tataprob,tatagain,tataloss,tatastay]
+end
 
[…]

### 2022-12-18 Manuscripts/deNovoRates2.tex
new file:   Manuscripts/deNovoRates2.tex 	modified:   Modeling/DeNovoProb2.jl

--- 2022-12-06 Manuscripts/deNovoRates.tex
@@
 \centering
-\includegraphics[scale=0.5]{Figures/geneGainLossComplete.pdf}
+\includegraphics[scale=0.5]{Figures/geneGainLossComplete_new.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
@@
 %\caption{One-step}
-\includegraphics[scale=0.5]{Figures/whichfirstComplete.pdf}
+\includegraphics[scale=0.5]{Figures/whichfirstComplete_new.pdf}
 %\end{subfigure}\hfill%
@@
 \vspace{1ex}
-\includegraphics[scale=0.5]{Figures/pLossComplete.pdf}
+\includegraphics[scale=0.5]{Figures/pLossComplete_new.pdf}
 \caption{ORF loss is more probable than transcription loss for proto-genes containing more than 33 codons.
@@
 
-\subsubsection{Definition of the probabilities}
+\subsubsection{Probability of transcription}
 
-Probability that a DNA region is transcribed by chance ($P_\textit{RNA}$), is the probability of finding a poly-A signal multiplied by the sum of probabilities of finding a TATA-box, and an Inr element.
+For a a DNA region of a length, to be transcribed it should have a promoter the beginning of the DNA region, and a poly-A signal at the end of it.
+The transcribed DNA region should also not have any poly-A signal sequences within it except the one at the end of it.
+A transcribed DNA of length $l$ nucleotides, contains $l-5$ 6-mers.
+Each of these 6-mers can be a poly-A signal with a probability, $P_\textit{polyA}$.
+Probability that none of these 6-mers is a poly-A signal is $(1-P_\textit{polyA})^{l-5}$.
+Since we focus on protein coding genes in this work, we describe the length of transcript in terms of the length of protein it can encode.
+Assuming that untranslated regions have no effect on protein synthesis, the length of a transcript in number of nucleotides should be at least three times the length of the encoded protein in number of amino acids.
+Thus the probability that a DNA region produces a transcript that can harbor an ORF with $k$ codons ($P_\textit{RNA}$) is:
 
 \begin{equation}
-P_\textit{RNA} = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr})
+P_\textit{RNA}(k) = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr}) (1-P_\textit{polyA})^{3k-5}
 \label{eqrnaprob}
@@
 
-Gain of transcription can happen via two mechanisms.
-In the first mechanism a promoter (TATA-box or Inr) already exists, and mutations cause a poly-A signal to emerge ($P_\textit{polyA-gain}$) but do not destroy the promoter ($P_\textit{TATA-stay}$,  $P_\textit{Inr-stay}$).
-In the second mechanism, a poly-A signal exists and the promoter emerges due to mutations ($P_\textit{TATA-gain}$,  $P_\textit{Inr-gain}$), while the poly-A signal remains undisrupted ($P_\textit{polyA-stay}$).
-Thus the total probability of transcription gain ($P_\textit{RNA-gain}$) would be the sum of the probabilities of the two mechanisms.
+\subsubsection{Probability of gaining transcription}
 
-\begin{equation}
-P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
+As we mentioned in the previous section, three requirements need to be met to produce a transcript of any given length.
+First, a promoter needs to be present.
+Second, a poly-A signal needs to be present at the end of the DNA region to be transcribed, and third, no poly-A signal should exist within this DNA region.
+Transcription can be gained if one of two of these requirements are already met and the third required feature emerges due to mutations.
+At the same time mutations do not destroy the existing features.
+It is possible that two or all the required features are missing and they emerge due to mutations.
+However such an event is highly improbable.
+Thus transcription gain can occur via three mechanisms.
+
+\vspace{\baselineskip}
+In the first mechanism:
+\begin{itemize}
+\item A promoter (TATA-box or Inr) exists at the beginning of a DNA region and remains intact ($P_\textit{TATA-stay}$, $P_\textit{Inr-stay}$)
+\item No poly-A signal sequences are present within the DNA region and none emerges ($P_\textit{nopolyA-stay} = 1 - P_\textit{polyA} - P_\textit{polyA-gain}$)
+\item A poly-A signal emerges at the end of the DNA region ($P_\textit{polyA-gain}$)
+\end{itemize}
+
+\vspace{\baselineskip}
+In the second mechanism:
+\begin{itemize}
+\item A poly-A signal is present at the end of the DNA region remains intact 
+\item No poly-A signal sequences are present within the DNA region and none emerges
+\item A promoter emerges at the beginning of the DNA region ($P_\textit{TATA-gain}$, $P_\textit{Inr-gain}$)
+\end{itemize}
+
+\vspace{\baselineskip}
+Finally, in the third mechanism:
+\begin{itemize}
+\item A promoter exists at the beginning of a DNA region and remains intact
+\item A poly-A site is present at the end of the DNA region remains intact
+\item One poly-A site that was present in the DNA region, is lost. ($P_\textit{polyA-loss}$)
+\item At every other site in the DNA region, no poly-A signal exists and none emerges
+\end{itemize}
+
+The total gain probability ($P_\textit{RNA-gain}$) of a transcript that can harbor an ORF with $k$ codons  would be the sum of the probabilities of the three gain mechanisms:
+
+\begin{align}
+P_\textit{RNA-gain}(k) &  = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber\\[1ex]
+%
+& + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber \\[1ex]
+& + (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
 \label{eqrnagain}
-\end{equation}
+\end{align}
 
 It is also possible that both the promoter and the poly-A signal were initially absent, and emerge via mutations. However, this probability is very small and is negligible.
+
+\subsubsection{Probability of losing transcription}
 
@@
 \begin{align}
-P_\textit{RNA-loss} & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
+P_\textit{RNA-loss}(k) & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
  & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
@@
 \begin{equation}
-P_\textit{RNA-stay} = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
+P_\textit{RNA-stay}(k) = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
 \label{eqrnastay}
@@
 \begin{equation}
-P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
+P_\textit{ORF}(k) = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
 \label{eqorfprob}
@@
 \begin{align}
-P_\textit{ORF-gain} & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+P_\textit{ORF-gain}(k) & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
 & \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
@@
 \begin{equation}
-P_\textit{ORF-loss} = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
+P_\textit{ORF-loss}(k) = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
 \label{eqorfloss}
@@
 \begin{equation}
-P_\textit{ORF-stay} = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
+P_\textit{ORF-stay}(k) = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
 \label{eqorfstay}
 \end{equation}
+
+\subsection{Independence of transcription and ORF probabilities}
+
+\label{rnaorfindependence}
+
+All the four poly-A signal variant contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon.
+Since stop codons cannot exist inside an ORF, presence of an ORF reduces the possible number of poly-A signal sites in a DNA region.
+Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame.
+Thus, out of $3k-5$ positions in an ORF with $k$ codons, $k-2$ positions cannot harbor a poly-A signal, which means that number of possible poly-A sites is $2k-3$.
+
+Conversely, stop codons will be less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
+However, this difference is very small because poly-A signals are only 0.5\% of all 6-mer sequences that can harbor a stop codon.
+Thus presence of transcription has a negligible effect on ORF gain.
+
+Overall the evolutionary dynamics of transcription and ORF, are not independent of each other.
+The presence of an ORF has a significant effect on the gain of transcription especially if the ORF is long, whereas the presence of transcription does not affect ORF gain significantly.
 

### 2022-12-18 Modeling/DeNovoProb2.jl

@@ -70,16 +70,15 @@ function polyaprobs(gccontent)
     return [polyaprob,polyagain,polyaloss,polyastay]
 end
 
-function rnaprobs(tata,inr,polya,orflen)
+function rnaprobs(tata,inr,polya,orflen,ptype::Bool)
     
-    # Total number of 6-mers in an ORF
-    n_6mersORF = 3*orflen - 6 + 1
-
     # Number of possible polyA signals given ORF is present
     # TAA within polyA signal cannot exist with intact ORF
-
-    noPAsites = n_6mersORF - trunc(Int,n_6mersORF/6)
-
+    if(ptype)
+        noPAsites = 3*orflen - 5 # total 6-mers
+    else
+        noPAsites = 2*orflen - 3 # 6-mers not in with ORF
+    end
     nopolyaprob = 1 - polya[xProb]
     nopolyastay = nopolyaprob - polya[xGain]
 
@@ -87,7 +86,7 @@ function rnaprobs(tata,inr,polya,orflen)
     # (Initiator or TATA) and polyA
 
     # Probability of finding a transcript of length ≥ orflen
-    rnaprob = (tata[xProb]*nopolyaprob^27 + inr[xProb]) * polya[xProb] * nopolyaprob^noPAsites;
+    rnaprob = (tata[xProb] + inr[xProb]) * polya[xProb] * nopolyaprob^noPAsites;
     
     # RNA gain mechanism 1
     # Gain of any promoter
@@ -177,19 +176,21 @@ inrvals = inrprobs(gccontent);
 polyavals = polyaprobs(gccontent);
 
 (rnaprob, rnagain, rnaloss, rnastay) = [zeros(size(ncodons)) for i = 1:4 ];
+(crnaprob, crnagain, crnaloss, crnastay) = [zeros(size(ncodons)) for i = 1:4 ];
 (orfprob, orfgain, orfloss, orfstay) = [zeros(size(ncodons)) for i = 1:4 ];
 
 for k in eachindex(ncodons)
     (orfprob[k], orfgain[k], orfloss[k], orfstay[k]) = orfprobs(ATGvals,stopvals,ncodons[k])
[…]

### 2022-12-19 Modeling/DeNovoProb2.jl
modified:   Modeling/DeNovoProb2.jl

@@ -152,10 +152,13 @@ end
 
 
 
-function orfprobs(ATG,stop,k)
+function orfprobs(ATG,stop,k,ptype::Bool,polya)
 
+    if(!ptype)
+        stop[xProb] -= polya[xProb]
+        stop[xStay] -= polya[x]
+    end
     nostopstay = 1 - stop[xProb] - stop[xGain]
-
     orfprob = ATG[xProb]*stop[xProb]*(1 - stop[xProb])^(k-2)
 
     orfgain = nostopstay^(k-2)*(

### 2022-12-19 Modeling/DeNovoProb2.jl
modified:   Modeling/DeNovoProb2.jl

@@ -77,7 +77,7 @@ function rnaprobs(tata,inr,polya,orflen,ptype::Bool)
     if(ptype)
         noPAsites = 3*orflen - 5 # total 6-mers
     else
-        noPAsites = 2*orflen - 3 # 6-mers not in with ORF
+        noPAsites = 2*orflen - 5 # 6-mers not in with ORF
     end
     nopolyaprob = 1 - polya[xProb]
     nopolyastay = nopolyaprob - polya[xGain]
@@ -155,51 +155,69 @@ end
 function orfprobs(ATG,stop,k,ptype::Bool,polya)
 
     if(!ptype)
-        stop[xProb] -= polya[xProb]
-        stop[xStay] -= polya[x]
+        stopgainc = stop[xGain] - polya[xGain]
+        stopprobc = stop[xProb] - polya[xProb]
+        stoplossc = stop[xLoss] - polya[xLoss]
+    else
+        stopgainc = stop[xGain]
+        stopprobc = stop[xProb]
+        stoplossc = stop[xLoss]
     end
-    nostopstay = 1 - stop[xProb] - stop[xGain]
-    orfprob = ATG[xProb]*stop[xProb]*(1 - stop[xProb])^(k-2)
 
-    orfgain = nostopstay^(k-2)*(
-    stop[xStay]*ATG[xGain] + ATG[xStay]*stop[xGain] +
-    (k-2)*(stop[xStay]*ATG[xStay]*stop[xProb]*stop[xLoss]/nostopstay)
-    )
-    orfloss = stop[xLoss] + ATG[xLoss] + (k-2)*stop[xGain]/(1-stop[xProb])
+    nostopstay = 1 - stopprobc - stopgainc
+
+    orfprob = ATG[xProb]*stop[xProb]*(1 - stopprobc)^(k-2)
+
+    gainmech1 = stop[xStay]*ATG[xGain]*nostopstay^(k-2);
+    gainmech2 = ATG[xStay]*stop[xGain]*nostopstay^(k-2);
+    gainmech3 = (k-2)*stop[xStay]*ATG[xStay]*stop[xProb]*stoplossc*nostopstay^(k-1)
+
+    orfgain =  gainmech1 + gainmech2 + gainmech3
+
+    orfloss = stop[xLoss] + ATG[xLoss] + (k-2)*stopgainc/(1-stopprobc)
[…]

### 2022-12-19 Modeling/DeNovoProb2.jl
modified:   Modeling/DeNovoProb2.jl

@@ -186,6 +186,7 @@ gcrange = [0.3:0.01:0.6;];
 (rnaprob, rnagain, rnaloss, rnastay) = [zeros(length(gcrange),length(ncodons)) for i = 1:4 ];
 (crnaprob, crnagain, crnaloss, crnastay) = [zeros(length(gcrange),length(ncodons)) for i = 1:4 ];
 (orfprob, orfgain, orfloss, orfstay) = [zeros(length(gcrange),length(ncodons)) for i = 1:4 ];
+(corfprob, corfgain, corfloss, corfstay) = [zeros(length(gcrange),length(ncodons)) for i = 1:4 ];
 
 for g in eachindex(gcrange)
     ATGvalsG = ATGprobs(gcrange[g]);

### 2022-12-19 Modeling/DeNovoProb2.jl
modified:   Modeling/DeNovoProb2.jl

@@ -157,11 +157,9 @@ function orfprobs(ATG,stop,k,ptype::Bool,polya)
     if(!ptype)
         stopgainc = stop[xGain] - polya[xGain]
         stopprobc = stop[xProb] - polya[xProb]
-        stoplossc = stop[xLoss] - polya[xLoss]
     else
         stopgainc = stop[xGain]
         stopprobc = stop[xProb]
-        stoplossc = stop[xLoss]
     end
 
     nostopstay = 1 - stopprobc - stopgainc
@@ -170,11 +168,12 @@ function orfprobs(ATG,stop,k,ptype::Bool,polya)
 
     gainmech1 = stop[xStay]*ATG[xGain]*nostopstay^(k-2);
     gainmech2 = ATG[xStay]*stop[xGain]*nostopstay^(k-2);
-    gainmech3 = (k-2)*stop[xStay]*ATG[xStay]*stop[xProb]*stoplossc*nostopstay^(k-1)
+    gainmech3 = (k-2)*stop[xStay]*ATG[xStay]*stopprobc*stop[xLoss]*nostopstay^(k-1)
 
     orfgain =  gainmech1 + gainmech2 + gainmech3
 
     orfloss = stop[xLoss] + ATG[xLoss] + (k-2)*stopgainc/(1-stopprobc)
+
     orfstay = ATG[xStay]*stop[xStay]*(nostopstay)^(k-2)
     return [orfprob; orfgain; orfloss; orfstay]
 end
@@ -203,19 +202,19 @@ for g in eachindex(gcrange)
     println("Done: ",g)
 end
 
-dGC = which(gcrange.==gccontent)
+dGC = findfirst(gcrange.==gccontent)
 
 genegain = rnastay[dGC,:] .* corfgain[dGC,:] + orfstay[dGC,:] .* crnagain[dGC,:] + orfgain[dGC,:].*rnagain[dGC,:];
 
 genegain2 = genegain./(1 .-rnaprob[dGC,:] .- orfprob[dGC,:]);
 
-geneloss = orfloss .+ rnaloss;
+geneloss = orfloss[dGC,:].+ rnaloss[dGC,:];
 
 rnafirst = rnastay[dGC,:] .* orfgain[dGC,:];
 orffirst = orfstay[dGC,:] .* crnagain[dGC,:];
 
[…]

### 2022-12-20 Manuscripts/deNovoRates2.tex
modified:   Modeling/DeNovoProb2.jl

@@
 The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$.
-To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms of gene gain and gene loss: $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-For example, the ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event.
+To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms (log-log ratio) of gene gain and gene loss for proto-genes with an ORF ranging from 30 -- 300 codons in length.
+
+$$ \text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$$
+
+For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event.
 We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for all ORF lengths, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-This was also the case for a proto-gene containing 50 codons in its ORF and having different amounts of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
+We found that for ORF longer than 44 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
+To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for a proto-gene containing an ORF with 50 codons, at different values of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
 
-We performed the same analysis on just the ORFs (and not the whole genes).
+We performed an analogous analysis on just the ORFs (and not the whole genes).
 That is, we calculated the ratio of logarithms of ORF gain and ORF loss.
-We found that for ORFs longer than 143 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.
+We found that for ORFs longer than 142 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.
 

### 2022-12-20 Modeling/DeNovoProb2.jl

@@ -266,17 +266,16 @@ whichfirstgcr[isnothing.(whichfirstgcr)] .= 1;
 plot_whichfirstgcr = plot(gcrange,ncodons[whichfirstgcr],
     xlabel = "GC-content",
     ylabel = "Minimum codons for \n RNA-first trajectory",
-    size = (width = cm2pt(13), height = cm2pt(11)),
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
     yticks = [30:2:40;]
 );
 
 savefig(plot_whichfirstgcr, figdir*"whoisfirstgcr_new.pdf")
 
-plot_rnafist_orffirst2 = plot(ncodons[1:50],log.(orffirst2./rnafirst2)[1:50],
+plot_rnafist_orffirst2 = plot(ncodons,log.(orffirst2./rnafirst2),
     xlabel = "ORF length (codons)",
     ylabel = "P_{ORF-first}\nP_RNA-first",
-    size = (width = cm2pt(12), height = cm2pt(11)),
-    xticks = [35:7:80;]
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
 );
 savefig(plot_rnafist_orffirst2, figdir*"first_ORF_RNA2_new.pdf")
 

### 2022-12-20 Manuscripts/deNovoRates2.tex
modified:   Manuscripts/deNovoRates2.tex

@@
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
-The following equation describes the probability of gene gain:
+We note that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present.
+This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence.
+This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}).
+Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed.
+Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively.
+We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:
 
 \begin{equation}
-P_\textit{gene-gain} = P_\textit{RNA-gain} \times P_\textit{ORF-stay} + P_\textit{ORF-gain} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
+P_\textit{gene-gain}^* = P_\textit{RNA-gain \vline\ ORF} \times P_\textit{ORF-stay} + P_\textit{ORF-gain \vline\ RNA} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
 \label{genegaineq}
@@
 
-Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
+\autoref{genegaineq} describes the total probability of gene gain that includes the probability that the gene was not previously present.
+To more precisely address the question of whether a gene that was absent in an ancestor but emerges in a descendant, we calculated the corresponding conditional probability of gene gain:
+
+\begin{equation}
+P_\textit{gene-gain} = P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})
+\label{genegaineq2}
+\end{equation}
+
+We note that the difference between the total probability ($P_\textit{gene-gain}^*$) and the conditional probability ($P_\textit{gene-gain}$) of gene gain, is very small.
+
+Next, we calculated the probability of gene loss, given the gene (transcription and ORF) is already present($P_\textit{gene-loss}$).
+It is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
 Specifically:
@@
 \caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
 The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.
-\autoref{genegaineq} describes total gene gain probability.
-Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
-Using the conditional probability of gene gain does not change these results.}
+We note that \autoref{genegaineq} describes total gene gain probability.}
 \label{gainlossprob}
@@
 To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for a proto-gene containing an ORF with 50 codons, at different values of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
+For this proto-gene, we found that the log-log ratio initially decreases with GC-content.
+It is minimum at a GC-content of 0.41, above which it increases.
+More generally, the ratio remains above 2 for all GC-content values for this proto-gene.
 
@@
 \centering
-%\begin{subfigure}{0.48\textwidth}
-%\centering
-%\caption{One-step}
-\includegraphics[scale=0.5]{Figures/whichfirstComplete_new.pdf}
-%\end{subfigure}\hfill%
-%\begin{subfigure}{0.48\textwidth}
-%\centering
-%\caption{Two-step}
-%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
-%\end{subfigure}
+\includegraphics[scale=0.5]{Figures/whichfirstComplete_with2step.pdf}
 \caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).

### 2022-12-21 Manuscripts/deNovoRates2.tex
modified:   Modeling/DeNovoProb2.jl

@@
 \begin{equation}
-P_\textit{gene-gain} = P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})
+P_\textit{gene-gain} = P_\textit{gene-gain}^*/(1-P_\textit{ORF} - P_\textit{RNA})
 \label{genegaineq2}
@@
 \centering
-\includegraphics[scale=0.5]{Figures/geneGainLossComplete_new.pdf}
+\includegraphics[scale=0.5]{Figures/geneGainLossComplete2_new.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
 The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.
-We note that \autoref{genegaineq} describes total gene gain probability.}
+The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
 \label{gainlossprob}
@@
 We found that for ORF longer than 44 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for a proto-gene containing an ORF with 50 codons, at different values of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
-For this proto-gene, we found that the log-log ratio initially decreases with GC-content.
-It is minimum at a GC-content of 0.41, above which it increases.
-More generally, the ratio remains above 2 for all GC-content values for this proto-gene.
+To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons).
+We found that the effect of GC-content on the log-log ratio depends on the length of the ORF.
+Specifically, the log-log ratio steadily increases with GC-content ranging from 0.3 to 0.6, when the proto-gene contains an ORF with 30 codons.
+When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases.
+For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range).
+Overall, beyond a certain ORF length, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.
 
@@
 \begin{equation}
-P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}
+P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}
 \label{orffirsteq1}
@@
 \begin{equation}
-P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}
+P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
 \label{rnafirsteq1}
@@
 \centering
-\includegraphics[scale=0.5]{Figures/whichfirstComplete_with2step.pdf}
+\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
 \caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).

### 2022-12-21 Modeling/DeNovoProb2.jl

@@ -227,12 +227,14 @@ plot_genegain_geneloss = plot(ncodons,log.(genegain2)./log.(geneloss),
 );
 savefig(plot_genegain_geneloss, figdir*"geneGainLoss_new.pdf")
 
-k50 = findfirst(ncodons .==50)
+cxlen = [30:30:150;]
 
-genegain50 = rnastay[:,k50] .* corfgain[:,k50] + orfstay[:,k50] .* crnagain[:,k50] + orfgain[:,k50].*rnagain[:,k50];
+klen = ncodons .∈ Ref(cxlen)
 
-genegain2_50 = genegain50./(1 .-rnaprob[:,k50] .- orfprob[:,k50]);
-geneloss50 = orfloss[:,k50] .+ rnaloss[:,k50]
+genegainklen = rnastay[:,klen] .* corfgain[:,klen] + orfstay[:,klen] .* crnagain[:,klen] + orfgain[:,klen].*rnagain[:,klen];
+
+genegain2_klen = genegainklen./(1 .-rnaprob[:,klen] .- orfprob[:,klen]);
+genelossklen = orfloss[:,klen] .+ rnaloss[:,klen]
 
 # genegainX = ((rnagain .+ rnastay) .* orfgain .+ orfstay .* crnagain);
 # genegainX2 = genegainX./(1 .-rnaprob .- orfprob);
@@ -242,12 +244,13 @@ geneloss50 = orfloss[:,k50] .+ rnaloss[:,k50]
 
 # whichGCloss = [findfirst(x.>=2) for x in eachcol(genegainlossratio)];
 
-plot_genegain_geneloss_50_gcrange = plot(gcrange,log.(genegain2_50)./log.(geneloss50),
+plot_genegain_geneloss_klen_gcrange = plot(gcrange,log.(genegain2_klen)./log.(genelossklen),
     xlabel = "GC-content",
     ylabel = "# Gene Losses \n per Gene Gain",
-    size = (width = cm2pt(12.4), height = cm2pt(11)),
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
+    xlims = [0.29,0.65]
 );
-savefig(plot_genegain_geneloss_50_gcrange, figdir*"geneGainLoss50gcr_new.pdf")
+savefig(plot_genegain_geneloss_klen_gcrange, figdir*"geneGainLoss_klengcr_new.pdf")
 
[…]

### 2022-12-21 Manuscripts/deNovoRates2.tex
modified:   Modeling/DeNovoProb2.jl

@@
 Specifically, the probability of gene gain can be defined as the sum of three probabilities.
-First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{methORF}).
-Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{methORF}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{methRNA}).
+First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}).
+Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}).
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
@@
 We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for ORF longer than 44 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
+We found that for ORF longer than 30 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
 To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons).
@@
 For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range).
-Overall, beyond a certain ORF length, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.
+Overall, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.
 
@@
 \includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
-\caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
-\textbf{(A)} The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence, with a GC-content of 0.42 (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
-A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory.
-Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
-\textbf{(B)} Minimum ORF length (codons) that makes RNA-first trajectory more feasible (vertical axis) for different GC-content (horizontal axis).
-For a GC-content of 0.3 and 0.35, proto-genes of every analysed ORF length ($\geq33$ codons) preferentially emerge RNA-first.}
+\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first.
+The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
+A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
+In both panels, the horizontal axis denotes the number of codons in the ORF.
+\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and GC-content values at which ORF-first trajectory is more likely than RNA-first trajectory.}
 \label{whoisfirst}
@@
 \end{figure}
+
 We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
@@
 Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible.
-With the specific parameter values we used, it turns out that proto-genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
-This minimum ORF length above which the RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content.
-For example, this number is 48 codons for a GC-content of 0.45 (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
+We found that proto-genes with an ORF containing 60 codons or more, preferentially emerge RNA-first, whereas proto-genes with shorter ORFs emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
+More specifically, the likelihood of the ORF-first trajectory relative to that of the RNA-first trajectory  decreases with increasing ORF length u a minimum value, and then increases.
+To understand if GC-content determines the evolutionary trajectory of \textit{de novo} emergence, we calculated the probabilities of the ORF-first and the RNA-first trajectory for different values of GC-content.
+We found that proto-genes with short ORF can emerge ORF-first at high GC-content.
+For example, the ORF-first trajectory is more likely at a GC-content of 0.6, for ORFs with 30 -- 36 codons (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
 
-A more stringent definition of ORF-first hypothesis would describe a two-step process.
+A more stringent definition of the ORF-first hypothesis would describe a two-step process.
 In the first step, an ORF emerges in an untranscribed region of DNA but transcription does not emerge.
 In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
-Likewise, RNA-first hypothesis can be defined by a two-step probability.
+Likewise, a RNA-first hypothesis can be defined by a two-step probability.
 In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).
@@
 \begin{align}
-P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
-P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
+P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
+P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
 \end{align}
 
-With this stringent definition, the difference between the probabilities of the two trajectories becomes very small ($< 3\times10^{-5}\ \%$).
-However, we still find that proto-genes with shorter ORFs ($<$ 34 codons) have a slightly higher chance of emerging ORF first.
+With this stringent definition, proto-genes with every analysed ORF length (30 -- 300) preferentially emerge via the RNA-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}C}).
+Furthermore, the two-step RNA-first trajectory becomes increasingly more probable than the two-step ORF-first trajectory, with increasing ORF length.
 
-Overall, long ORFs are more likely to be lost before transcription emerges than short ORFs, and hence they preferentially emerge in existing transcripts.
+Overall, long ORFs are likely to be lost before transcription emerges, and hence they preferentially emerge in existing transcripts.
 Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.
@@
 \vspace{1ex}
-\includegraphics[scale=0.5]{Figures/pLossComplete_new.pdf}
-\caption{ORF loss is more probable than transcription loss for proto-genes containing more than 33 codons.
+\includegraphics[scale=0.5]{Figures/pLossComplete2_new.pdf}
+\caption{ORF loss in proto-genes is more probable than transcription loss.
 The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
-The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
+The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
 \label{lossprob}
@@
 To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, for all ORFs longer than 33 codons (\hyperref[lossprob]{\autoref{lossprob}A}).
-Loss of a longer ORF (50 codons) was more probable than transcription loss for a wide range of GC-content (\hyperref[lossprob]{\autoref{lossprob}B}).
+We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}).
+At low GC-content (0.3 -- 0.34) ORF loss is less likely than RNA-loss, for short ORFs (30 codons; \hyperref[lossprob]{\autoref{lossprob}B}).
 
@@
 The probability of this scenario is described by:
-$P_\textit{ORF-stay} \times P_\textit{RNA-loss}/P_\textit{ORF}$.
+$P_\textit{ORF-stay \vline\ RNA} \times P_\textit{RNA-loss \vline\ ORF}/P_\textit{ORF}$.
 The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$.
@@
 That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
-We show that, in the absence of selection, long ORFs are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
+We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
 Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors.
@@
 \end{tabular}
-\caption{Mutation bias probabilities.}
+\caption{Mutation bias probabilities for different nucleotide mutations.
+\texttt{A:T} denotes an \texttt{A}-\texttt{T} base pair in a double stranded DNA.
+Thus \texttt{A}$\to$\texttt{G} mutation on one DNA strand would cause a \texttt{T}$\to$\texttt{C} mutation on the complementary strand.
+We describe the other mutations in the same way.}
 \label{mutbias}
@@
 $$P_0 = (1-e^{-m\lambda_\texttt{AT}})\times(1-e^{-(k-m)\lambda_\texttt{GC}})$$
+
+\subsection{Probabilities of finding, gaining, and losing an ORF}
+
+\label{methORF}
+
+\subsubsection{Probability of finding an ORF}
+
+A reading frame is a nucleotide sequence with a length that is a multiple of three.
+A reading frame that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}) is an open reading frame (ORF).
+This necessarily means that there are no stop codons within the sequence.
+Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:
+
+\begin{equation}
+P_\textit{ORF}(k) = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
+\label{eqorfprob}
+\end{equation}
+
+Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.
+
+\subsubsection{Probability of gaining an ORF}
+
+As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence).
+Thus an ORF can emerge due to mutations via three mechanisms.
+In each of these mechanisms, one requirement is initially absent whereas the other two are present.
+Then mutations cause the missing requirement to emerge while not disrupting the other two requirements.
+More specifically, the ORF can be gained via the following three mechanisms:
+\begin{enumerate}
+\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of a reading frame ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the reading frame ($1- P_\textit{stop} - P_\textit{stop-gain}$).
+\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of a reading frame ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the reading frame.
+\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the reading frame ($P_\textit{stop-gain}$).
+At the same time start and stop codons remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
+\end{enumerate}
+
+Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:
+
+\begin{align}
+P_\textit{ORF-gain}(k) & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
+\label{eqorfgain}
+\end{align}
+
+\subsubsection{Probability of ORF loss}
+
+ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:
+
+\begin{equation}
+P_\textit{ORF-loss}(k) = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
+\label{eqorfloss}
+\end{equation}
+
+The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.
+
+\subsubsection{Probability that ORF remains intact}
+
+We assumed that an ORF of a certain length remains intact if none of the necessary features are lost.
+However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence.
+This condition applies to all the three ORF probabilities described above.
+We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:
+
+\begin{equation}
+P_\textit{ORF-stay}(k) = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-gain})^{k-2}
+\label{eqorfstay}
+\end{equation}
 
@@
 
-\subsection{Probabilities of finding, gaining, and losing an ORF}
-
-\label{methORF}
-
-\subsubsection{Probability of finding an ORF}
-
-A reading frame is a nucleotide sequence with a length that is a multiple of three.
-A reading frame that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}) is an open reading frame (ORF).
-This necessarily means that there are no stop codons within the sequence.
-Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:
-
-\begin{equation}
-P_\textit{ORF}(k) = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
-\label{eqorfprob}
-\end{equation}
-
-Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.
-
-\subsubsection{Probability of gaining an ORF}
-
-As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence).
-Thus an ORF can emerge due to mutations via three mechanisms.
-In each of these mechanisms, one requirement is initially absent whereas the other two are present.
-Then mutations cause the missing requirement to emerge while not disrupting the other two requirements.
-More specifically, the ORF can be gained via the following three mechanisms:
-\begin{enumerate}
-\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of a reading frame ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the reading frame ($1- P_\textit{stop} - P_\textit{stop-gain}$).
-\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of a reading frame ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the reading frame.
-\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the reading frame ($P_\textit{stop-gain}$).
-At the same time start and stop codons remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
-\end{enumerate}
-
-Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:
-
-\begin{align}
-P_\textit{ORF-gain}(k) & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
-& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
-& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
-\label{eqorfgain}
-\end{align}
-
-\subsubsection{Probability of ORF loss}
-
-ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:
-
-\begin{equation}
-P_\textit{ORF-loss}(k) = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
-\label{eqorfloss}
-\end{equation}
-
-The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.
-
-\subsubsection{Probability that ORF remains intact}
-
-We assumed that an ORF of a certain length remains intact if none of the necessary features are lost.
-However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence.
-This condition applies to all the three ORF probabilities described above.
-We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:
-
-\begin{equation}
-P_\textit{ORF-stay}(k) = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
-\label{eqorfstay}
-\end{equation}
-
 \subsection{Independence of transcription and ORF probabilities}
@@
 Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame.
-Thus, out of $3k-5$ positions in an ORF with $k$ codons, $k-2$ positions cannot harbor a poly-A signal, which means that number of possible poly-A sites is $2k-3$.
+Thus, out of $3k-5$ positions in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}).
+Furthermore a poly-A signal cannot overlap with the start codon, thus reducing two more positions from all sites where a poly-A signal can exist (one position is already counted -- if a poly-A signal overlaps with the first codon in the second frame, then the second codon becomes a stop codon).
+Overall, the possible number poly-A sites on an ORF with $k$ codons is $2k-5$.
+Thus for a gene to emerge via gain of transcription, then the DNA should not contain poly-A sites in these $2k-5$ positions.
+This probability is higher than the transcription gain probability for an RNA of equal length emerging from a DNA that does not contain an ORF (\autoref{eqrnagain}).
 
 Conversely, stop codons will be less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
-However, this difference is very small because poly-A signals are only 0.5\% of all 6-mer sequences that can harbor a stop codon.
-Thus presence of transcription has a negligible effect on ORF gain.
+Specifically, the probability that a 6-mer contains a stop codon in the third position is same as the probability of finding a stop codon ($P_\textit{stop}$).
+If poly-A signals are excluded from these 6-mers, then the probability of finding a stop codon is $P_\textit{stop} - P_\textit{poly-A}$.
+Conversely, lack of poly-A signals will increase the probability of amino acid coding codons and thereby their probability of mutating into a stop codon.
+Furthermore, the likelihood that an ORF is undisrupted my premature nonsense mutations, is lower when the DNA is already transcribed.
+Specifically:
 
-Overall the evolutionary dynamics of transcription and ORF, are not independent of each other.
-The presence of an ORF has a significant effect on the gain of transcription especially if the ORF is long, whereas the presence of transcription does not affect ORF gain significantly.
+$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} - P_\textit{poly-A})}{1- P_\textit{stop}}$$
+
+Overall the evolutionary dynamics of transcription and ORF gain, are not independent of each other.
+That is, presence of one feature makes the gain of the other more likely.
+The dependence becomes more prominent with increasing number of codons in the ORF of the proto-gene.
 

### 2022-12-21 Modeling/DeNovoProb2.jl

@@ -55,7 +55,7 @@ function inrprobs(gccontent)
     inrprob = sum([nucprob(x,gccontent) for x in inrvars]);
     inrgain = featuregain(noinrs,inrvars,gccontent);
     inrloss = featuregain(inrvars,noinrs,gccontent);
-    inrstay = featurestay(inrvars,gccontent)/inrprob;
+    inrstay = featurestay(inrvars,gccontent);
     return [inrprob,inrgain,inrloss,inrstay]
 end
 
@@ -155,8 +155,8 @@ end
 function orfprobs(ATG,stop,k,ptype::Bool,polya)
 
     if(!ptype)
-        stopgainc = stop[xGain] - polya[xGain]
         stopprobc = stop[xProb] - polya[xProb]
+        stopgainc = stop[xGain]*(1-stopprobc)/(1-stop[xProb])
     else
         stopgainc = stop[xGain]
         stopprobc = stop[xProb]
@@ -222,7 +222,6 @@ orffirst2 = onlyorfgain.*(1 .- orfloss[dGC,:]).*(crnagain[dGC,:]./(1 .-rnaprob[d
 plot_genegain_geneloss = plot(ncodons,log.(genegain2)./log.(geneloss),
     xlabel = "ORF length (codons)",
     ylabel = "# Gene Losses \n per Gene Gain",
-    yticks = [2:0.4:3.2;],
     size = (width = cm2pt(12), height = cm2pt(11)),
 );
 savefig(plot_genegain_geneloss, figdir*"geneGainLoss_new.pdf")
@@ -256,8 +255,17 @@ plot_rnafist_orffirst = plot(ncodons,log2.(orffirst./rnafirst),
     xlabel = "ORF length (codons)",
     ylabel = "P_{ORF-first}\nP_RNA-first",
     size = (width = cm2pt(12.5), height = cm2pt(11)),
+    yticks = [-0.2:0.2:0.4;]
 );
 savefig(plot_rnafist_orffirst, figdir*"whoisfirst_new.pdf")
+
+plot_rnafist_orffirst2 = plot(ncodons,log.(orffirst2./rnafirst2),
+    xlabel = "ORF length (codons)",
+    ylabel = "P_{ORF-first}\nP_RNA-first",
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
[…]

### 2022-12-22 Manuscripts/deNovoRates2.tex
modified:   Manuscripts/deNovoRates2.tex

@@
 
-\subsection{Independence of transcription and ORF probabilities}
+\subsection{Interdependence of transcription and ORF probabilities}
 
@@
 
-All the four poly-A signal variant contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon.
+All the four poly-A signal variants contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon.
 Since stop codons cannot exist inside an ORF, presence of an ORF reduces the possible number of poly-A signal sites in a DNA region.
 Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame.
-Thus, out of $3k-5$ positions in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}).
+Thus, out of $3k-5$ positions for a 6-mer (length of a poly-A signal) in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}).
 Furthermore a poly-A signal cannot overlap with the start codon, thus reducing two more positions from all sites where a poly-A signal can exist (one position is already counted -- if a poly-A signal overlaps with the first codon in the second frame, then the second codon becomes a stop codon).
 Overall, the possible number poly-A sites on an ORF with $k$ codons is $2k-5$.
-Thus for a gene to emerge via gain of transcription, then the DNA should not contain poly-A sites in these $2k-5$ positions.
-This probability is higher than the transcription gain probability for an RNA of equal length emerging from a DNA that does not contain an ORF (\autoref{eqrnagain}).
+Thus for a gene to emerge via gain of transcription, the DNA should not contain poly-A sites at these $2k-5$ positions.
+As a consequence, the probability of transcription gain for a specific transcript length, is higher when the DNA region contains an ORF than when it does not (\autoref{eqrnagain}).
+We have ignored the probability of poly-A signal overlapping with a stop codon because it makes a very small difference in the probability of RNA gain.
+This is so because at least one stop codon (\texttt{TAA}) allows a poly-A signal to overlap with it in three frames.
 
-Conversely, stop codons will be less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
+Conversely to the effect of an existing ORF on transcription gain, stop codons are less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
 Specifically, the probability that a 6-mer contains a stop codon in the third position is same as the probability of finding a stop codon ($P_\textit{stop}$).

### 2022-12-22 Manuscripts/deNovoRates2.tex
modified:   Manuscripts/deNovoRates2.tex

@@
 \begin{align}
-P_\textit{RNA-gain}(k) &  = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber\\[1ex]
+P_\textit{RNA-gain}(k)  = & \ (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber\\[1ex]
 %

### 2022-12-22 Modeling/DeNovoProb.jl
modified:   Manuscripts/deNovoRates.tex 	deleted:    Manuscripts/deNovoRates2.tex     	modified:   Modeling/DeNovoProb.jl 	deleted:    Modeling/DeNovoProb2.jl

@@ -15,22 +15,26 @@ aaorder = ["W","F","Y","I","V","L","M","C","D","E","G","A","P","H","K","R","S","
 ixaaorder = indexin(aaorder,aas);
 gccontent = 0.42
 
-function rnavals(gccontent)
+(xProb, xGain, xLoss, xStay) = [i for i in 1:4];
+
+mers6 = kmers(6);
+allcodons = kmers(3)
+
+function tataprobs(gccontent)
     ## TATA box
     # Consensus sequence = TATAWAWR
     # is TSS - 25 - 8nt (transcript must be +33nt longer)
-
     tataboxes = vec(join.(product(("TATA" .* collect("TA") .* "A"), "TA", "GA")))
 
     tatavars = unique(reduce(vcat,
         reduce(vcat, [
-[degen(s,1,nucs) for s in tataboxes],
-[degen(s,2,"TA") for s in tataboxes],
-[degen(s,3,"TAC") for s in tataboxes],
-[degen(s,4,"AT") for s in tataboxes],
-[degen(s,6,nucs) for s in tataboxes],
-[degen(s,7,nucs) for s in tataboxes],
-[degen(s,8,nucs) for s in tataboxes],
+            [degen(s,1,nucs) for s in tataboxes],
+            [degen(s,2,"TA") for s in tataboxes],
+            [degen(s,3,"TAC") for s in tataboxes],
+            [degen(s,4,"AT") for s in tataboxes],
+            [degen(s,6,nucs) for s in tataboxes],
+            [degen(s,7,nucs) for s in tataboxes],
+            [degen(s,8,nucs) for s in tataboxes],
         ])
     ));
 
@@ -38,210 +42,287 @@ function rnavals(gccontent)
 
     tataprob = sum([nucprob(x,gccontent) for x in tatavars]);
     tatagain = featuregain(notata,tatavars,gccontent);
-    tataloss1 = featuregain(tatavars,notata,gccontent);
-    tataloss2 = tataloss1/tataprob;
+    tataloss = featuregain(tatavars,notata,gccontent)/tataprob;
     tatastay = featurestay(tatavars,gccontent);
-    notatastay = 1 - tatagain;
+    return [tataprob,tatagain,tataloss,tatastay]
+end
 
[…]

### 2022-12-22 Modeling/DeNovoProb2.jl

@@ -1,467 +1,0 @@
-using Combinatorics
-using Plots
-using Measures
-# using HypothesisTests
-
-include("nucleotidefuncts.jl")
-
-cm2pt = (cm) -> 28.3465*cm
-figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
-
-default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
-guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
-
-aaorder = ["W","F","Y","I","V","L","M","C","D","E","G","A","P","H","K","R","S","T","N","Q"];
-ixaaorder = indexin(aaorder,aas);
-gccontent = 0.42
-
-(xProb, xGain, xLoss, xStay) = [i for i in 1:4];
-
-mers6 = kmers(6);
-allcodons = kmers(3)
-
-function tataprobs(gccontent)
-    ## TATA box
-    # Consensus sequence = TATAWAWR
-    # is TSS - 25 - 8nt (transcript must be +33nt longer)
-    tataboxes = vec(join.(product(("TATA" .* collect("TA") .* "A"), "TA", "GA")))
-
-    tatavars = unique(reduce(vcat,
-        reduce(vcat, [
-            [degen(s,1,nucs) for s in tataboxes],
-            [degen(s,2,"TA") for s in tataboxes],
-            [degen(s,3,"TAC") for s in tataboxes],
-            [degen(s,4,"AT") for s in tataboxes],
-            [degen(s,6,nucs) for s in tataboxes],
-            [degen(s,7,nucs) for s in tataboxes],
-            [degen(s,8,nucs) for s in tataboxes],
-        ])
-    ));
-
-    notata = setdiff(kmers(8),tatavars);
-
-    tataprob = sum([nucprob(x,gccontent) for x in tatavars]);
-    tatagain = featuregain(notata,tatavars,gccontent);
-    tataloss = featuregain(tatavars,notata,gccontent)/tataprob;
-    tatastay = featurestay(tatavars,gccontent);
-    return [tataprob,tatagain,tataloss,tatastay]
-end
-
-function inrprobs(gccontent)
-    ## Initiator element
[…]

### 2022-12-22 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 Specifically, the probability of gene gain can be defined as the sum of three probabilities.
-First is the probability that an ORF already exists, and a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}) but doesn't disrupt the ORF ($P_\textit{ORF-stay}$, \autoref{methORF}).
-Second is the probability that the DNA is transcribed, and an ORF emerges due to mutations ($P_\textit{ORF-gain}$, \autoref{methORF}) while transcription stays intact ($P_\textit{RNA-stay}$, \autoref{methRNA}).
+First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}).
+Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}).
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
-The following equation describes the probability of gene gain:
+We note that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present.
+This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence.
+This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}).
+Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed.
+Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively.
+We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:
 
 \begin{equation}
-P_\textit{gene-gain} = P_\textit{RNA-gain} \times P_\textit{ORF-stay} + P_\textit{ORF-gain} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
+P_\textit{gene-gain}^* = P_\textit{RNA-gain \vline\ ORF} \times P_\textit{ORF-stay} + P_\textit{ORF-gain \vline\ RNA} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
 \label{genegaineq}
@@
 
-Likewise, the probability of gene loss, given the gene (transcription and ORF) is already present, is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
+\autoref{genegaineq} describes the total probability of gene gain that includes the probability that the gene was not previously present.
+To more precisely address the question of whether a gene that was absent in an ancestor but emerges in a descendant, we calculated the corresponding conditional probability of gene gain:
+
+\begin{equation}
+P_\textit{gene-gain} = P_\textit{gene-gain}^*/(1-P_\textit{ORF} - P_\textit{RNA})
+\label{genegaineq2}
+\end{equation}
+
+We note that the difference between the total probability ($P_\textit{gene-gain}^*$) and the conditional probability ($P_\textit{gene-gain}$) of gene gain, is very small.
+
+Next, we calculated the probability of gene loss, given the gene (transcription and ORF) is already present($P_\textit{gene-loss}$).
+It is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
 Specifically:
@@
 \centering
-\includegraphics[scale=0.5]{Figures/geneGainLossComplete.pdf}
+\includegraphics[scale=0.5]{Figures/geneGainLossComplete2_new.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.
-\autoref{genegaineq} describes total gene gain probability.
-Gene gain probability conditioned on the gene being absent would be $P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})$.
-Using the conditional probability of gene gain does not change these results.}
+The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
 \label{gainlossprob}
@@
 The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$.
-To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms of gene gain and gene loss: $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-For example, the ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event.
+To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms (log-log ratio) of gene gain and gene loss for proto-genes with an ORF ranging from 30 -- 300 codons in length.
+
+$$ \text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$$
+
+For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event.
 We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for all ORF lengths, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-This was also the case for a proto-gene containing 50 codons in its ORF and having different amounts of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
+We found that for ORF longer than 30 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
+To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons).
+We found that the effect of GC-content on the log-log ratio depends on the length of the ORF.
+Specifically, the log-log ratio steadily increases with GC-content ranging from 0.3 to 0.6, when the proto-gene contains an ORF with 30 codons.
+When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases.
+For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range).
+Overall, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.
 
-We performed the same analysis on just the ORFs (and not the whole genes).
+We performed an analogous analysis on just the ORFs (and not the whole genes).
 That is, we calculated the ratio of logarithms of ORF gain and ORF loss.
-We found that for ORFs longer than 143 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.
+We found that for ORFs longer than 142 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.
 
@@
 \centering
-%\begin{subfigure}{0.48\textwidth}
-%\centering
-%\caption{One-step}
-\includegraphics[scale=0.5]{Figures/whichfirstComplete.pdf}
-%\end{subfigure}\hfill%
-%\begin{subfigure}{0.48\textwidth}
-%\centering
-%\caption{Two-step}
-%\includegraphics[scale=0.6]{Figures/first_ORF_RNA2.pdf}
-%\end{subfigure}
-\caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first).
-\textbf{(A)} The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence, with a GC-content of 0.42 (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}).
-A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory.
-Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
-\textbf{(B)} Minimum ORF length (codons) that makes RNA-first trajectory more feasible (vertical axis) for different GC-content (horizontal axis).
-For a GC-content of 0.3 and 0.35, proto-genes of every analysed ORF length ($\geq33$ codons) preferentially emerge RNA-first.}
+\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
+\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first.
+The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
+A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
+In both panels, the horizontal axis denotes the number of codons in the ORF.
+\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and GC-content values at which ORF-first trajectory is more likely than RNA-first trajectory.}
 \label{whoisfirst}
@@
 \vspace{1ex}
-\includegraphics[scale=0.5]{Figures/pLossComplete.pdf}
-\caption{ORF loss is more probable than transcription loss for proto-genes containing more than 33 codons.
+\includegraphics[scale=0.5]{Figures/pLossComplete2_new.pdf}
+\caption{ORF loss in proto-genes is more probable than transcription loss.
 The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
-The vertical axis shows the log transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF-loss is more probable than RNA loss, and \textit{vice versa}.}
+The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
 \label{lossprob}
@@
 
+\subsection{Probabilities of finding, gaining, and losing an ORF}
+
+\label{methORF}
+
+\subsubsection{Probability of finding an ORF}
+
+A reading frame is a nucleotide sequence with a length that is a multiple of three.
+A reading frame that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}) is an open reading frame (ORF).
+This necessarily means that there are no stop codons within the sequence.
+Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:
+
+\begin{equation}
+P_\textit{ORF}(k) = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
+\label{eqorfprob}
+\end{equation}
+
+Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.
+
+\subsubsection{Probability of gaining an ORF}
+
+As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence).
+Thus an ORF can emerge due to mutations via three mechanisms.
+In each of these mechanisms, one requirement is initially absent whereas the other two are present.
+Then mutations cause the missing requirement to emerge while not disrupting the other two requirements.
+More specifically, the ORF can be gained via the following three mechanisms:
+\begin{enumerate}
+\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of a reading frame ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the reading frame ($1- P_\textit{stop} - P_\textit{stop-gain}$).
+\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of a reading frame ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the reading frame.
+\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the reading frame ($P_\textit{stop-gain}$).
+At the same time start and stop codons remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
+\end{enumerate}
+
+Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:
+
+\begin{align}
+P_\textit{ORF-gain}(k) & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
+& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
+\label{eqorfgain}
+\end{align}
+
+\subsubsection{Probability of ORF loss}
+
+ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:
+
+\begin{equation}
+P_\textit{ORF-loss}(k) = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
+\label{eqorfloss}
+\end{equation}
+
+The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.
+
+\subsubsection{Probability that ORF remains intact}
+
+We assumed that an ORF of a certain length remains intact if none of the necessary features are lost.
+However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence.
+This condition applies to all the three ORF probabilities described above.
+We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:
+
+\begin{equation}
+P_\textit{ORF-stay}(k) = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-gain})^{k-2}
+\label{eqorfstay}
+\end{equation}
+
 \subsection{Probabilities of finding, gaining, and losing transcription}
@@
 
-Gain of transcription can happen via two mechanisms.
-In the first mechanism a promoter (TATA-box or Inr) already exists, and mutations cause a poly-A signal to emerge ($P_\textit{polyA-gain}$) but do not destroy the promoter ($P_\textit{TATA-stay}$,  $P_\textit{Inr-stay}$).
-In the second mechanism, a poly-A signal exists and the promoter emerges due to mutations ($P_\textit{TATA-gain}$,  $P_\textit{Inr-gain}$), while the poly-A signal remains undisrupted ($P_\textit{polyA-stay}$).
-Thus the total probability of transcription gain ($P_\textit{RNA-gain}$) would be the sum of the probabilities of the two mechanisms.
+\subsubsection{Probability of gaining transcription}
 
-\begin{equation}
-P_\textit{RNA-gain} = (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain} + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}
+As we mentioned in the previous section, three requirements need to be met to produce a transcript of any given length.
+First, a promoter needs to be present.
+Second, a poly-A signal needs to be present at the end of the DNA region to be transcribed, and third, no poly-A signal should exist within this DNA region.
+Transcription can be gained if one of two of these requirements are already met and the third required feature emerges due to mutations.
+At the same time mutations do not destroy the existing features.
+It is possible that two or all the required features are missing and they emerge due to mutations.
+However such an event is highly improbable.
+Thus transcription gain can occur via three mechanisms.
+
+\vspace{\baselineskip}
+In the first mechanism:
+\begin{itemize}
+\item A promoter (TATA-box or Inr) exists at the beginning of a DNA region and remains intact ($P_\textit{TATA-stay}$, $P_\textit{Inr-stay}$)
+\item No poly-A signal sequences are present within the DNA region and none emerges ($P_\textit{nopolyA-stay} = 1 - P_\textit{polyA} - P_\textit{polyA-gain}$)
+\item A poly-A signal emerges at the end of the DNA region ($P_\textit{polyA-gain}$)
+\end{itemize}
+
+\vspace{\baselineskip}
+In the second mechanism:
+\begin{itemize}
+\item A poly-A signal is present at the end of the DNA region remains intact 
+\item No poly-A signal sequences are present within the DNA region and none emerges
+\item A promoter emerges at the beginning of the DNA region ($P_\textit{TATA-gain}$, $P_\textit{Inr-gain}$)
+\end{itemize}
+
+\vspace{\baselineskip}
+Finally, in the third mechanism:
+\begin{itemize}
+\item A promoter exists at the beginning of a DNA region and remains intact
+\item A poly-A site is present at the end of the DNA region remains intact
+\item One poly-A site that was present in the DNA region, is lost. ($P_\textit{polyA-loss}$)
+\item At every other site in the DNA region, no poly-A signal exists and none emerges
+\end{itemize}
+
+The total gain probability ($P_\textit{RNA-gain}$) of a transcript that can harbor an ORF with $k$ codons  would be the sum of the probabilities of the three gain mechanisms:
+
+\begin{align}
+P_\textit{RNA-gain}(k)  = & \ (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber\\[1ex]
+%
+& + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber \\[1ex]
+& + (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
 \label{eqrnagain}
-\end{equation}
+\end{align}
 
 It is also possible that both the promoter and the poly-A signal were initially absent, and emerge via mutations. However, this probability is very small and is negligible.
+
+\subsubsection{Probability of losing transcription}
 
@@
 
-\subsection{Probabilities of finding, gaining, and losing an ORF}
+\subsection{Interdependence of transcription and ORF probabilities}
 
-\label{methORF}
+\label{rnaorfindependence}
 
-\subsubsection{Probability of finding an ORF}
+All the four poly-A signal variants contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon.
+Since stop codons cannot exist inside an ORF, presence of an ORF reduces the possible number of poly-A signal sites in a DNA region.
+Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame.
+Thus, out of $3k-5$ positions for a 6-mer (length of a poly-A signal) in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}).
+Furthermore a poly-A signal cannot overlap with the start codon, thus reducing two more positions from all sites where a poly-A signal can exist (one position is already counted -- if a poly-A signal overlaps with the first codon in the second frame, then the second codon becomes a stop codon).
+Overall, the possible number poly-A sites on an ORF with $k$ codons is $2k-5$.
+Thus for a gene to emerge via gain of transcription, the DNA should not contain poly-A sites at these $2k-5$ positions.
+As a consequence, the probability of transcription gain for a specific transcript length, is higher when the DNA region contains an ORF than when it does not (\autoref{eqrnagain}).
+We have ignored the probability of poly-A signal overlapping with a stop codon because it makes a very small difference in the probability of RNA gain.
+This is so because at least one stop codon (\texttt{TAA}) allows a poly-A signal to overlap with it in three frames.
 
-A reading frame is a nucleotide sequence with a length that is a multiple of three.
-A reading frame that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}) is an open reading frame (ORF).
-This necessarily means that there are no stop codons within the sequence.
-Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:
+Conversely to the effect of an existing ORF on transcription gain, stop codons are less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
+Specifically, the probability that a 6-mer contains a stop codon in the third position is same as the probability of finding a stop codon ($P_\textit{stop}$).
+If poly-A signals are excluded from these 6-mers, then the probability of finding a stop codon is $P_\textit{stop} - P_\textit{poly-A}$.
+Conversely, lack of poly-A signals will increase the probability of amino acid coding codons and thereby their probability of mutating into a stop codon.
+Furthermore, the likelihood that an ORF is undisrupted my premature nonsense mutations, is lower when the DNA is already transcribed.
+Specifically:
 
-\begin{equation}
-P_\textit{ORF} = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
-\label{eqorfprob}
-\end{equation}
+$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} + P_\textit{poly-A})}{1- P_\textit{stop}}$$
 
-Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.
-
-\subsubsection{Probability of gaining an ORF}
-
-As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence).
-Thus an ORF can emerge due to mutations via three mechanisms.
-In each of these mechanisms, one requirement is initially absent whereas the other two are present.
-Then mutations cause the missing requirement to emerge while not disrupting the other two requirements.
-More specifically, the ORF can be gained via the following three mechanisms:
-\begin{enumerate}
-\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of a reading frame ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the reading frame ($1- P_\textit{stop} - P_\textit{stop-gain}$).
-\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of a reading frame ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the reading frame.
-\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the reading frame ($P_\textit{stop-gain}$).
-At the same time start and stop codons remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
-\end{enumerate}
-
-Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:
-
-\begin{align}
-P_\textit{ORF-gain} & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
-& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
-& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
-\label{eqorfgain}
-\end{align}
-
-\subsubsection{Probability of ORF loss}
-
-ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:
-
-\begin{equation}
-P_\textit{ORF-loss} = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
-\label{eqorfloss}
-\end{equation}
-
-The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.
-
-\subsubsection{Probability that ORF remains intact}
-
-We assumed that an ORF of a certain length remains intact if none of the necessary features are lost.
-However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence.
-This condition applies to all the three ORF probabilities described above.
-We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:
-
-\begin{equation}
-P_\textit{ORF-stay} = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-stay})^{k-2}
-\label{eqorfstay}
-\end{equation}
+Overall the evolutionary dynamics of transcription and ORF gain, are not independent of each other.
+That is, presence of one feature makes the gain of the other more likely.
+The dependence becomes more prominent with increasing number of codons in the ORF of the proto-gene.
 

### 2022-12-22 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 In the first scenario, the ORF remains intact whereas the transcription is lost.
-The probability of this scenario is described by:
-$P_\textit{ORF-stay \vline\ RNA} \times P_\textit{RNA-loss \vline\ ORF}/P_\textit{ORF}$.
-The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$.
-We find that the first scenario is more plausible for ORFs that have fewer than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}).
-Also, the probability values of scenario one and two, obtained from this stringent analysis, were not significantly different from the transcription ($P_\textit{RNA-loss}$) and ORF loss ($P_\textit{ORF-loss}$) probabilities, respectively.
+The probability of this scenario ($P_\textit{onlyORF-loss}$) is described by:
+
+\vspace{-1ex}
+
+\begin{equation}
+P_\textit{onlyRNA-loss} = \frac{P_\textit{ORF-stay \vline\ RNA}}{P_\textit{ORF \vline \ RNA}} \times P_\textit{RNA-loss \vline\ ORF}
+\end{equation}
+
+We note that $P_\textit{ORF-stay \vline \ RNA}$ includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
+To describe the conditional probability of ORF remaining undisrupted given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
+
+In the second scenario, transcription stays intact while ORF is lost. The probability of this scenario is:
+
+\vspace{-1em}
+\begin{equation}
+P_\textit{onlyORF-loss} = \frac{P_\textit{RNA-stay \vline\ ORF}}{P_\textit{RNA \vline \ ORF}} \times P_\textit{ORF-loss \vline\ RNA}
+\end{equation}
+
+We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths.
+Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.
 

### 2022-12-26 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 In both panels, the horizontal axis denotes the number of codons in the ORF.
-\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and GC-content values at which ORF-first trajectory is more likely than RNA-first trajectory.}
+\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and the GC-content values (horizontal axis) at which ORF-first trajectory is more likely than RNA-first trajectory.}
 \label{whoisfirst}
@@
 
-We note that $P_\textit{ORF-stay \vline \ RNA}$ includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
-To describe the conditional probability of ORF remaining undisrupted given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
+We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
+To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
 
@@
 We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
+However, if both transcription and ORF are absent in a genomic locus, then \textit{de novo} emergence is more likely to occur transcription-first.
 Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors.

### 2023-02-14 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -100,13 +100,13 @@ function rnaprobs(tata,inr,polya,orflen,ptype::Bool)
     # no polyA present in the sequence 
     gainmech2 = (tata[xStay] + inr[xStay]) * polya[xGain] * nopolyastay^noPAsites
 
+
     # RNA gain mechanism 3
     # Any one promoter stays
     # polyA stays at the end of sequence
     # A polyA present in the sequence is lost (for every 6mer in the sequence)
     gainmech3 = (noPAsites-1)*(tata[xStay] + inr[xStay]) * polya[xStay] * polya[xLoss] * polya[xProb] * nopolyastay^(noPAsites-1);
-
-
+   
     rnagain =  gainmech1 + gainmech2 + gainmech3
     
     # Fraction of Inr-only promoters
@@ -116,7 +116,7 @@ function rnaprobs(tata,inr,polya,orflen,ptype::Bool)
     Qtata = tata[xProb]*(1-inr[xProb])/(tata[xProb] + inr[xProb] - inr[xProb]*tata[xProb]);
 
     # Fraction of promoters containing both TATA and Inr
-    Qboth = 1 - Qtata - Qinr;
+    # Qboth = 1 - Qtata - Qinr;
     notatastay  = 1 - tata[xGain] - tata[xProb];
     noinrstay = 1 - inr[xGain] - inr[xProb]
     rnaloss = (Qinr*inr[xLoss]*notatastay +
@@ -125,8 +125,55 @@ function rnaprobs(tata,inr,polya,orflen,ptype::Bool)
       polya[xLoss] + noPAsites*polya[xGain]
     );
 
+    rnaloss = polya[xLoss] + noPAsites*polya[xGain];
+
     rnastay = (inr[xStay] + tata[xStay])*polya[xStay]*nopolyastay^noPAsites;
     return [rnaprob; rnagain; rnaloss; rnastay]
+end
+
+
+function rnaprobs_nopromoter(polya,orflen,ptype::Bool)
+    
+    # Number of possible polyA signals given ORF is present
+    # TAA within polyA signal cannot exist with intact ORF
+    if(ptype)
+        noPAsites = 3*orflen - 5 # total 6-mers
+    else
+        noPAsites = 2*orflen - 5 # 6-mers not in with ORF
+    end
+    nopolyaprob = 1 - polya[xProb]
[…]

### 2023-02-15 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
new file:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Modeling/DeNovoProb.jl 	new file:   Modeling/hexamers2.txt 	modified:   Modeling/nucleotidefuncts.jl 	new file:   Modeling/trimers2.txt

\documentclass[12pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
\usepackage[colorlinks, allcolors = blue]{hyperref}
\usepackage{natbib}
\citestyle{egu}

\titlespacing*{\paragraph}{0pt}{1\baselineskip}{3pt}
\setlength{\parindent}{0pt}

\begin{document}

\onehalfspacing

\section*{Reviewer: 1}

{\itshape Comments to the Author

One of the most valuable points of this paper is that it formalizes many principles that are intuitive, but which have not been explicitly discussed in the field of de novo gene emergence with theoretical rigor. The authors build a simple probabilistic model for the neutral evolution of de novo protein coding genes, which I think is important because it could inform about what to expect under a neutral scenario given the molecular knowledge we have gathered over the past 60 years.

Because of its theoretical approach this paper is a good addition to the de novo gene literature, however, I think the paper will be more relevant if it would either:

\begin{enumerate}
\item offer a more exhaustive contextualization of the model and the field, emphasizing why the model is showing results that cannot be addressed with empirical data, or

\item it would evaluate some of their results with empirical genetic diversity or experimental evolution datasets (such as https://academic.oup.com/mbe/article/39/2/msab368/6491261) to test whether the rates of ORF gain or loss within a population match their predictions. As it stands, although formalizing these probabilities theoretically has the potential to be an important contribution, given its simplifications, I believe the model is not too informative of the realistic potential of its predictions.
\end{enumerate}
}

{\itshape When it comes to the model itself, I had a problem with some limitations concerning basic molecular biology processes and biases that are relevant for the origination of de novo genes. While some of these points were touched upon in the discussion, I still think that their model could easily incorporate ways of addressing them, which would make this model much more realistic and relevant. I specifically find four points to be discussed in more detail or incorporated into the model:

\vspace{\baselineskip}

\textbf{1) Basal transcription:} The authors repeatedly mention that in order for transcription to take place there is a need for a promoter to evolve. Much empirical work in the field of de novo gene emergence has shown how basal levels of transcription or expression due to chromatin rearrangements might prime the functionalization of de novo genes

\begin{itemize}\setlength{\itemsep}{-5pt}
\item https://genome.cshlp.org/content/20/10/1313
\item https://elifesciences.org/articles/47138
\item https://academic.oup.com/mbe/article/37/4/1165/5679780
\item https://genome.cshlp.org/content/28/11/1675.short
\item https://elifesciences.org/articles/09977
\end{itemize}

Because the presence/absence of transcription is a fundamental aspect of the model in this paper, I think this paradigm needs to be integrated into it for this study to be relevant given the knowledge we now have.
}

\vspace{\baselineskip}

\textbf{2) Non-neutral background:} de novo ORFs might emerge in non-coding regions that still encode for important cellular functions, such as introns or enhancers. This would mean that some mutations that lead to the functionalization of a de novo peptide might be unviable because they temper with the functions of the genomic regions where they arise. This might make some sequences harder or easier to evolve than others because some specific genomic positions will be locked in to a given nucleotide or arrangement of nucleotide. The way this could be integrated into the model is by considering what is the percentage of sites that tends to be strongly constrained or conserved between different species.

\vspace{\baselineskip}

{\itshape \textbf{3) Structured probabilities:} The authors rightly used empirical estimations of CG content to study the probability of functionalizing or defunctionalizing mutations. But It's not just CG content, it might be important to consider the frequency or presence/absence of genomic patterns like CpG islands or microsatellites. The existence of these features will strongly influence the likelihood of getting specific kinds of mutations, which will be important for the biases in evolution as well as the propensities in protein evolution that the authors tried to study with their model. Not to mention that these features will have different probabilities of evolving transcription on top of the coding sequence.}

\vspace{\baselineskip}

We thank the reviewer for this valuable suggestion. We agree with them that nucleotide distribution is not uniform and therefore the probabilities of different DNA sequences cannot be estimated accurately from GC content alone. We also note that sequence composition can be highly variable between different species, as well as between different genomic regions in the same species. Thus the most accurate way to estimate the rate of gene evolutionary dynamics would be to analyse each genomic locus individually. Although such an analysis would be accurate, it would be difficult to generalize. To keep a balance between accuracy and generalizability of our results, we repeated our original calculations of various probability using nucleotide distribution estimates from intergenic regions of \textit{Drosophila melanogaster}  (flybase release 6.49). Specifically, we estimated the frequencies of all hexamers in the intergenic sequences using a sliding window. We ignored the hexamers that contained an ambiguous nucleotide character (non ATGC). From the frequency distribution of all the 4096 hexamers, we estimated the probability of finding (also gaining and losing) the polyadenylation signal sequences (that are 6 nucleotides long). Similarly, we estimated the probabilities of start and stop codons using a distribution of trimers estimated from the \textit{D.melanogaster} intergenic regions. With these probabilities, we repeated the analyses shown in the original figures 1A, 2A, 2C and 3A.

We found that the the results from actual nucleotide distributions indeed differed from the ones obtained using total GC content of \textit{D.melanogaster} intergenic genome (0.403). We now include these results in the main manuscript along with the old figures. We also explain the details of the re-analysis and its rationale (as described in this response), in the main text.

\vspace{\baselineskip}

\textbf{4) Population genetics:} Ultimately a de novo gene needs to increase in frequency in a population if it is going to be maintained in the genome. The probability of fixation of a neutral allele is 1/2N, but how is the birth/death mutational dynamics of de novo genes that the authors are exploring affected by the demography of species? I believe that drift might be the primary source of loss of de novo neutral genes, much more than mutational defunctionalization. Or do the authors want to disregard this and just focus on the macroevolutionary retention of de novo genes in phylogenies? If that is the case, this could be discussed, maybe not necessarily incorporated into the model.

\vspace{\baselineskip}

I think that if the authors would follow at least some of these recommendations then they would be walking on much more solid ground and they would not need to use the ``all models are wrong'' quote, in favor of a more grounded implications of their findings. This would also mean that we would count with a much more solid point of argument for the discussion of the likelihood of de novo gene emergence, a field which tends to be too empirical and with little theory, which is why this paper would be a good addition to the literature if improvements are made.

\section*{Reviewer: 2}

Comments to the Author
This paper describes a model to calculate the mutational probabilities of de novo gene origin and loss. Its conclusions are that certain modes of gene origin are more probable than others. I am no expert in these kinds of models or in de novo gene emergence, but I found these arguments interesting and valuable.

I have two major comments about the paper. The first is that I was surprised to see no arguments or formalism rooted in population genetics in this paper. The authors only calculate mutational probabilities, but no fixation probabilities. They state several times that they work under the assumption of total neutrality, so I guess an informed reader can put two and two together, knowing that in Kimura's model neutral fixation probabilities only depend on the mutation rate of neutral alleles. I think the paper would benefit from making this explicit -- the assumption of neutrality justifies only caring about mutational probabilities. Of course, the neutral model comes with its own assumptions (diploid organisms  etc). Perhaps this could be acknowledged somewhere. And while fixation probabilities only depend on the mutation rate, I think the time to fixation under neutrality does depend on the effective population size. So the authors' model as it stands does not make predictions about how long it takes for de novo genes to fix or be lost (though this extension may be easy to carry out).

Another difficulty for me was the argument about whether the existence of a lineage specific gene really implies it evolved only in that lineage. The alternative being an earlier gain with losses in all lineages but one. The paper justifies this by deriving that the probability of losing a de novo gene is much larger than gaining it. I struggle to see how that comparison alone is sufficient to make this argument. It seems to me one would have to tally up all relevant probabilities of gaining, losing, and keeping the gene along the relevant internal and terminal branches of the species tree. The length of these branches also seems like it might matter a lot. A gain-with-2-losses trajectory, like the authors suggest may be likely, requires the new gene to be gained and then kept along one internal and one terminal branch. It would seem to me that this is vastly less likely under strict neutrality than a lineage specific gain. I may of course be wrong, but I think a very simplistic argument based on only two probabilities with no treatment of the actual phylogeny is not sufficient to convince me of the authors' argument.

Another minor comment from me is that in the discussion it might be valuable to address that at least some de novo genes become essential quickly (I would for example point to work by the Long lab, which is cited already). This would allow purifying selection to quickly preserve such genes against losses.

Lastly, I couldn't find the actual parameter values used for the various probabilities in the paper. I know the authors will make the code available, but for the casual reader a table would be valuable.

\end{document}

### 2023-02-15 Modeling/DeNovoProb.jl

@@ -18,28 +18,27 @@ gccontent = 0.42
 (xProb, xGain, xLoss, xStay) = [i for i in 1:4];
 
 mers6 = kmers(6);
-allcodons = kmers(3)
+
+## TATA box
+# Consensus sequence = TATAWAWR
+# is TSS - 25 - 8nt (transcript must be +33nt longer)
+tataboxes = vec(join.(product(("TATA" .* collect("TA") .* "A"), "TA", "GA")))
+
+tatavars = unique(reduce(vcat,
+    reduce(vcat, [
+        [degen(s,1,nucs) for s in tataboxes],
+        [degen(s,2,"TA") for s in tataboxes],
+        [degen(s,3,"TAC") for s in tataboxes],
+        [degen(s,4,"AT") for s in tataboxes],
+        [degen(s,6,nucs) for s in tataboxes],
+        [degen(s,7,nucs) for s in tataboxes],
+        [degen(s,8,nucs) for s in tataboxes],
+    ])
+));
+
+notata = setdiff(kmers(8),tatavars);
 
 function tataprobs(gccontent)
-    ## TATA box
-    # Consensus sequence = TATAWAWR
-    # is TSS - 25 - 8nt (transcript must be +33nt longer)
-    tataboxes = vec(join.(product(("TATA" .* collect("TA") .* "A"), "TA", "GA")))
-
-    tatavars = unique(reduce(vcat,
-        reduce(vcat, [
-            [degen(s,1,nucs) for s in tataboxes],
-            [degen(s,2,"TA") for s in tataboxes],
-            [degen(s,3,"TAC") for s in tataboxes],
-            [degen(s,4,"AT") for s in tataboxes],
-            [degen(s,6,nucs) for s in tataboxes],
-            [degen(s,7,nucs) for s in tataboxes],
-            [degen(s,8,nucs) for s in tataboxes],
-        ])
-    ));
-
-    notata = setdiff(kmers(8),tatavars);
-
     tataprob = sum([nucprob(x,gccontent) for x in tatavars]);
     tatagain = featuregain(notata,tatavars,gccontent);
     tataloss = featuregain(tatavars,notata,gccontent)/tataprob;
@@ -47,11 +46,12 @@ function tataprobs(gccontent)
[…]

### 2023-02-15 Modeling/nucleotidefuncts.jl

@@ -209,3 +209,59 @@ function featurestay(set1,gccontent)
         )
     return v
 end
+
+using DataFrames
+# Data based calculations #
+
+data, header = readdlm(joinpath(Base.source_dir(),"hexamers2.txt"), '\t', header=true);
+header = replace.(header, r"[^A-Z,a-z]" => "");
+mers6freq = DataFrame(data[:,[1,3]], header[[1,3]]);
+
+data, header = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t', header=true);
+header = replace.(header, r"[^A-Z,a-z]" => "");
+mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
+
+nprob6 = x -> mers6freq.Frequency[mers6freq.kmer .== x][1]
+nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
+
+function featuregain3(set1,set2)
+    if(isempty(set1) || isempty(set2))
+        @warn "empty feature sets, returning 0"
+        return 0
+    end
+    v1 = sum([nprob3(s1)*sum([mprob(s1,s2) for s2 in set2])
+                for s1 in set1])
+    return v1
+end
+
+function featurestay3(set1)
+    v =  sum(
+        [nprob3(s1)*(
+            sum([mprob(s1,s2) for s2 in set1]) +
+            (1-µA(mutrate,nsub))^numAT(s1)*
+            (1-µG(mutrate,nsub))^(3-numAT(s1))
+            ) for s1 in set1]
+        )
+    return v
+end
+
+function featuregain6(set1,set2)
+    if(isempty(set1) || isempty(set2))
+        @warn "empty feature sets, returning 0"
+        return 0
+    end
+    v1 = sum([nprob6(s1)*sum([mprob(s1,s2) for s2 in set2])
+                for s1 in set1])
+    return v1
+end
+
+function featurestay6(set1)
+    v =  sum(
+        [nprob6(s1)*(
+            sum([mprob(s1,s2) for s2 in set1]) +
+            (1-µA(mutrate,nsub))^numAT(s1)*
+            (1-µG(mutrate,nsub))^(6-numAT(s1))
+            ) for s1 in set1]
+        )
+    return v
+end

### 2023-02-15 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -469,11 +469,16 @@ gainfix = hcat([genegain2X/(2*x) for x in popsizes]...);
 lossfix = hcat([genelossX/(2*x) for x in popsizes]...);
 fixprobs = log.(gainfix)./log.(lossfix);
 
-heatmap(ncodons,popsizes,fixprobs,
-    yaxis = :log
-)
 
-plot(ncodons,fixprobs)
+plot_fix = plot(ncodons,fixprobs[:,2:2:10],
+    xlabel = "ORF length (codons)",
+    ylabel = "#Gene loss with extinction \n per Gene gain with fixation",
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
+    xlim = [25,350]
+);
+
+savefig(plot_fix, figdir*"pFix_dmel.pdf")
+
 ## Protein properties ##
 
 aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];

### 2023-02-16 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

@@
 \setlength{\parindent}{0pt}
+\setlength{\parskip}{\baselineskip}
 
@@
 
-\vspace{\baselineskip}
-
 \textbf{1) Basal transcription:} The authors repeatedly mention that in order for transcription to take place there is a need for a promoter to evolve.
@@
 
-\vspace{\baselineskip}
+We thank the reviewer for this suggestion and for pointing us to relevant studies.
+It is indeed true that low level of transcription exists even in the absence of a core promoter motif.
+Furthermore enhancers can also initiate transcription in their proximal regions.
+However, we also note that vast majority of RNAs arising due to pervasive transcription (including the enhancer associated eRNAs), are degraded in the nucleus and thus do not reach the cytoplasm where they can be translated.
+Therefore, we think that a polyadenylation signal would still be a necessary feature for determining transcript length and stability.
+We further rationalize this argument by two observations.
+First, polyA signals are 10-fold less enriched in coding regions (0.0004) than in intergenic regions (0.004).
+This suggests that polyA signals within the gene body (including ORF) can be disruptive.
+Second, many studies on \textit{de novo} gene discovery to our knowledge, employ a RNAseq library preparation kit that amplifies polyadenylated RNAs (\href{https://doi.org/10.7554/eLife.09977}{Neme and Tautz, 2016}; \href{https://doi.org/10.7554/eLife.47138}{Witt \textit{et al.}, 2019}; \href{https://doi.org/10.1093/molbev/msz300}{Majic and Payne, 2019}; \href{https://doi.org/10.1038/s41467-021-20911-3}{Blevins \textit{et al.}, 2021}).
+Thus most of the discovered \textit{de novo} genes indeed have a polyA tail (and must also have a polyadenylation signal).
+We now mention both these observations in the main text (lines XXX).
 
-\textbf{2) Non-neutral background:} de novo ORFs might emerge in non-coding regions that still encode for important cellular functions, such as introns or enhancers.
+We next focus on sequences that initiate transcription.
+Although several DNA sequence motifs corresponding to enhancers and core promoters have been identified, there is no set of sequence motifs that is deemed to be absolutely necessary.
+The best understood, and the most common core promoter motifs are TATA-box and Inr, which we included in our original analysis.
+We further note that although some sequences are indeed associated with enhancers, the actual functional enhancers are largely determined by epigenetic marks that vary significantly in different cell types, developmental stages, and during different environmental conditions.
+One common feature of both promoters and enhancers is that they both are characterized by a nucleosome depleted region (\href{https://doi.org/10.1038/s41576-019-0173-8}{Andersson and Sandelin, 2020}).
+Overall, we surmise that transcription can be initiated at any nucleosome depleted region but the stability of such transcripts is determined by polyA signals.
+
+We assume that such nucleosome free regions are uniformly distributed in open chromatin.
+With the condition that a \textit{de novo} gene emerges in open chromatin regions, we now calculated the probability of transcript emergence based on poly-A signals alone, while ignoring the requirement of a promoter.
+We first asked if such an assumption would significantly change the results presented in our first draft (Figures 1 -- 3).
+We found that these results do not change significantly.
+The fundamental inferences remain the same.
+If we had data on every promoter/enhancer sequence, and if we incorporate them in the model, then the exact RNA probability values should lie in between those estimated by the promoterless model and TATA/Inr-only model.
+In the revised manuscript we now present the results from the promoterless model, and provide a suitable justification in the text (lines XXX).
+We include the old results based on TATA/Inr-only model in the supplementary data (Fig SXX).
+
+{\itshape \textbf{2) Non-neutral background:} de novo ORFs might emerge in non-coding regions that still encode for important cellular functions, such as introns or enhancers.
 This would mean that some mutations that lead to the functionalization of a de novo peptide might be unviable because they temper with the functions of the genomic regions where they arise.
 This might make some sequences harder or easier to evolve than others because some specific genomic positions will be locked in to a given nucleotide or arrangement of nucleotide.
-The way this could be integrated into the model is by considering what is the percentage of sites that tends to be strongly constrained or conserved between different species.
+The way this could be integrated into the model is by considering what is the percentage of sites that tends to be strongly constrained or conserved between different species.}
 
-\vspace{\baselineskip}
+The reviewer raised an important concern.
+To address this concern, we performed an analysis of total number of available sites and the probability of gene emergence, focusing on the model organism \textit{D.melanogaster}.
+To this end, we first downloaded the genomic coordinates of all \textit{cis}-regulatory motifs from \href{http://redfly.ccr.buffalo.edu/}{REDfly database}.
+Next, we calculated the lengths of intergenic sequences that are not interrupted by any \textit{cis}-regulatory motif, and estimated the total number of sites available for genes of different lengths to emerge.
+Finally, we calculated the total gene gain probability by multiplying the total number of feasible sites and per-site gene gain probability.
+We now briefly describe these findings in the main text, and in greater detail in the supplementary data (Figure SXX).
+
+We would like to note that in the original analysis, we calculate the probabilities of gene loss and gain, at one genomic locus.
+We also assume that this locus exists in an intergenic region devoid of any functional sequences.
+Therefore the original results will not be dependent on the number of feasible loci (or specific genomes).
+We have now clarified this assumption in the revised manuscript (lines XXX).
 
@@
 Not to mention that these features will have different probabilities of evolving transcription on top of the coding sequence.}
-
-\vspace{\baselineskip}
 
@@
 From the frequency distribution of all the 4096 hexamers, we estimated the probability of finding (also gaining and losing) the polyadenylation signal sequences (that are 6 nucleotides long).
-Similarly, we estimated the probabilities of start and stop codons using a distribution of trimers estimated from the \textit{D.melanogaster} intergenic regions.
+Similarly, we calculated the probabilities of start and stop codons, using a distribution of trimers estimated from the \textit{D.melanogaster} intergenic regions.
 With these probabilities, we repeated the analyses shown in the original figures 1A, 2A, 2C and 3A.
@@
 We also explain the details of the re-analysis and its rationale (as described in this response), in the main text.
+(lines XXX)
 
-\vspace{\baselineskip}
-
-\textbf{4) Population genetics:} Ultimately a de novo gene needs to increase in frequency in a population if it is going to be maintained in the genome.
+{\itshape \textbf{4) Population genetics:} Ultimately a de novo gene needs to increase in frequency in a population if it is going to be maintained in the genome.
 The probability of fixation of a neutral allele is 1/2N, but how is the birth/death mutational dynamics of de novo genes that the authors are exploring affected by the demography of species?
@@
 Or do the authors want to disregard this and just focus on the macroevolutionary retention of de novo genes in phylogenies?
-If that is the case, this could be discussed, maybe not necessarily incorporated into the model.
-
-\vspace{\baselineskip}
+If that is the case, this could be discussed, maybe not necessarily incorporated into the model.}
 

### 2023-02-17 Modeling/DeNovoProb.jl
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

@@ -13,7 +13,7 @@ guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
 
 aaorder = ["W","F","Y","I","V","L","M","C","D","E","G","A","P","H","K","R","S","T","N","Q"];
 ixaaorder = indexin(aaorder,aas);
-gccontent = 0.42
+gccontent = 0.41
 
 (xProb, xGain, xLoss, xStay) = [i for i in 1:4];
 
@@ -117,7 +117,7 @@ function rnaprobs(tata,inr,polya,orflen,ptype::Bool)
     Qtata = tata[xProb]*(1-inr[xProb])/(tata[xProb] + inr[xProb] - inr[xProb]*tata[xProb]);
 
     # Fraction of promoters containing both TATA and Inr
-    # Qboth = 1 - Qtata - Qinr;
+    Qboth = 1 - Qtata - Qinr;
     notatastay  = 1 - tata[xGain] - tata[xProb];
     noinrstay = 1 - inr[xGain] - inr[xProb]
     rnaloss = (Qinr*inr[xLoss]*notatastay +
@@ -257,6 +257,11 @@ for g in eachindex(gcrange)
     println("Done: ",g)
 end
 
+rnaprob = rnaprob*promprob;
+rnagain = rnagain*promprob;
+crnaprob = crnaprob*promprob;
+crnagain = crnagain*promprob;
+
 dGC = findfirst(gcrange.==gccontent)
 
 genegain = rnastay[dGC,:] .* corfgain[dGC,:] + orfstay[dGC,:] .* crnagain[dGC,:] + orfgain[dGC,:].*rnagain[dGC,:];
@@ -310,7 +315,7 @@ plot_rnafist_orffirst = plot(ncodons,log2.(orffirst./rnafirst),
     xlabel = "ORF length (codons)",
     ylabel = "P_{ORF-first}\nP_RNA-first",
     size = (width = cm2pt(12.5), height = cm2pt(11)),
-    yticks = [-0.2:0.2:0.4;]
+   # yticks = [-0.2:0.2:0.4;]
 );
 savefig(plot_rnafist_orffirst, figdir*"whoisfirst_new.pdf")
 
@@ -318,7 +323,7 @@ plot_rnafist_orffirst2 = plot(ncodons,log.(orffirst2./rnafirst2),
     xlabel = "ORF length (codons)",
     ylabel = "P_{ORF-first}\nP_RNA-first",
[…]

### 2023-02-17 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

@@
 
+anna's data (table 4). Can be estimated by total rnagain * total sites * generations per year * \% open chromatin (1-6)* \% open chromatin in intergenic region (35)
+
 {\itshape When it comes to the model itself, I had a problem with some limitations concerning basic molecular biology processes and biases that are relevant for the origination of de novo genes.
@@
 First, polyA signals are 10-fold less enriched in coding regions (0.0004) than in intergenic regions (0.004).
-This suggests that polyA signals within the gene body (including ORF) can be disruptive.
+PolyA signals are also 3 times less enriched in \textit{D.melanogaster} RNAs than in the intergenic regions.
+Furthermore, the polyA signals in RNAs are usually located at the end of the RNA.
+These observations suggests that polyA signals within the gene body (including ORF) can be disruptive.
 Second, many studies on \textit{de novo} gene discovery to our knowledge, employ a RNAseq library preparation kit that amplifies polyadenylated RNAs (\href{https://doi.org/10.7554/eLife.09977}{Neme and Tautz, 2016}; \href{https://doi.org/10.7554/eLife.47138}{Witt \textit{et al.}, 2019}; \href{https://doi.org/10.1093/molbev/msz300}{Majic and Payne, 2019}; \href{https://doi.org/10.1038/s41467-021-20911-3}{Blevins \textit{et al.}, 2021}).

### 2023-02-20 Modeling/DeNovoProb.jl
renamed:    Manuscripts/Figures/geneGainLossComplete2_new.svg -> Manuscripts/Figures/Old_figures/geneGainLossComplete2_new.svg

@@ -257,102 +257,141 @@ for g in eachindex(gcrange)
     println("Done: ",g)
 end
 
+promprob = 5331240/43164203; # CRMs in open chromatin / Total intergenic region
+
 rnaprob = rnaprob*promprob;
 rnagain = rnagain*promprob;
 crnaprob = crnaprob*promprob;
 crnagain = crnagain*promprob;
 
-dGC = findfirst(gcrange.==gccontent)
-
-genegain = rnastay[dGC,:] .* corfgain[dGC,:] + orfstay[dGC,:] .* crnagain[dGC,:] + orfgain[dGC,:].*rnagain[dGC,:];
-
-genegain2 = genegain./(1 .-rnaprob[dGC,:] .- orfprob[dGC,:]);
-
-geneloss = orfloss[dGC,:].+ rnaloss[dGC,:];
-
-rnafirst = rnastay[dGC,:] .* orfgain[dGC,:];
-orffirst = orfstay[dGC,:] .* crnagain[dGC,:];
-
-onlyrnagain = (1 .- orfprob[dGC,:] .- orfgain[dGC,:]).*rnagain[dGC,:];
-onlyorfgain = (1 .- rnaprob[dGC,:] .- rnagain[dGC,:]).*orfgain[dGC,:];
-
-rnafirst2 = onlyrnagain.*(1 .- rnaloss[dGC,:]).*(corfgain[dGC,:]./(1 .-orfprob[dGC,:]));
-orffirst2 = onlyorfgain.*(1 .- orfloss[dGC,:]).*(crnagain[dGC,:]./(1 .-rnaprob[dGC,:]));
-
-plot_genegain_geneloss = plot(ncodons,log.(genegain2)./log.(geneloss),
-    xlabel = "ORF length (codons)",
-    ylabel = "# Gene Losses \n per Gene Gain",
-    size = (width = cm2pt(12), height = cm2pt(11)),
-);
-savefig(plot_genegain_geneloss, figdir*"geneGainLoss_new.pdf")
-
+dGC = findall(gcrange .∈ Ref([0.34:0.08:0.5;]));
 cxlen = [30:30:150;]
-
 klen = ncodons .∈ Ref(cxlen)
 
-genegainklen = rnastay[:,klen] .* corfgain[:,klen] + orfstay[:,klen] .* crnagain[:,klen] + orfgain[:,klen].*rnagain[:,klen];
+# Gene gain/loss probabilities
 
[…]

### 2023-02-20 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -392,23 +392,38 @@ plot_whichfirstgcr2 = plot(100*gcrange,log2.(rnafirst2./orffirst2)[:,klen],
 
 savefig(plot_whichfirstgcr2, figdir*"whoisfirstgcr2_promoterless.pdf")
 
+# Plot ORF loss versus RNA loss
 
-plot_Loss = plot(ncodons,log2.(orfloss[dGC,:]./rnaloss[dGC,:]),
+plot_Loss = plot(ncodons,log2.(orfloss./rnaloss)[dGC[1],:],
     xlabel = "ORF length (codons)",
-    ylabel = "P_ORF-loss\nP_RNA-loss",
-    size = (width = cm2pt(12), height = cm2pt(11)),
+    ylabel = "P_{ORF-first}\nP_RNA-first",
+    size = (width = cm2pt(12.5), height = cm2pt(11)),
+    linestyle = :dash,
+    legend = :bottom,
+    label = Int(100*gcrange[dGC[1]])
+);
+    
+
+plot!(plot_Loss, ncodons,log2.(orfloss./rnaloss)[dGC[2],:],
+    linestyle = :solid,
+    label = Int(100*gcrange[dGC[2]])
 );
 
-savefig(plot_Loss, figdir*"pLoss_new.pdf")
 
-plot_loss_klen_gcrange = plot(gcrange,log2.(orfloss[:,klen]./rnaloss[:,klen]),
-    xlabel = "GC-content",
+plot!(plot_Loss, ncodons,log2.(orfloss./rnaloss)[dGC[3],:],
+    linestyle = :dot,
+    label = Int(100*gcrange[dGC[3]])
+);
+
+savefig(plot_Loss, figdir*"pLoss_promoterless.pdf")
+
+plot_loss_klen_gcrange = plot(100*gcrange,log2.(orfloss[:,klen]./rnaloss[:,klen]),
+    xlabel = "GC %",
     ylabel = "P_ORF-loss\nP_RNA-loss",
     size = (width = cm2pt(12.5), height = cm2pt(11)),
-    ylims = [-0.3,2.1],
-    xlims = [0.29,0.65]
+    xlims = [29,65]
 );
-savefig(plot_loss_klen_gcrange, figdir*"pLossklengcr_new.pdf")
+savefig(plot_loss_klen_gcrange, figdir*"pLossgcr_promoterless.pdf")
 
[…]

### 2023-02-21 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -506,6 +506,13 @@ plotX_genegain_geneloss = plot(ncodons,log.(genegain2X)./log.(genelossX),
 );
 savefig(plotX_genegain_geneloss, figdir*"geneGainLoss_dmel.pdf")
 
+plot!(plot_genegain_geneloss, ncodons,log.(genegain2X)./log.(genelossX),
+    linestyle = :solid,
+    linecolor = :blue,
+    label = "D.mel"
+);
+savefig(plot_genegain_geneloss, figdir*"geneGainLoss_promoterlessX.pdf")
+
 plotX_rnafist_orffirst = plot(ncodons,log2.(rnafirstX./orffirstX),
     xlabel = "ORF length (codons)",
     ylabel = "P_{ORF-first}\nP_RNA-first",
@@ -543,6 +550,13 @@ plotX_Loss = plot(ncodons,log2.(orflossX./rnalossX),
 );
 
 savefig(plotX_Loss, figdir*"pLoss_dmel.pdf")
+
+plot!(plot_Loss, ncodons,log2.(orflossX./rnalossX),
+    linestyle = :solid,
+    linecolor = :blue,
+    label = "D.mel"
+);
+savefig(plot_Loss, figdir*"pLoss_promoterlessX.pdf")
 
 freegenome = readdlm("freegenome_lendist_knownchr.txt", Int32);
 
@@ -619,6 +633,10 @@ aaprob = [sum(nucprob.(gencode[gencode[:,3] .== x,1],gccontent)) for x in aas];
 aaprob2 = aaprob./sum(aaprob);
 exp_hydropathy = sum(hydropathy3.*aaprob2)
 
+aaprob_dme = [sum(nprob3.(gencode[gencode[:,3] .== x,1])) for x in aas];
+aaprob2_dme = aaprob_dme./sum(aaprob_dme);
+exp_hydropathy_dme = sum(hydropathy3.*aaprob2)
+
 acidic = indexin(["D","E"],aas);
 basic = indexin(["K","R"],aas);
 hydrophobic = hydropathy3 .<0;
@@ -628,15 +646,28 @@ nostopvars = allcodons[allcodons .∉ Ref(["TAA","TAG","TGA"])]
 
 pHydrophobic = sum(aaprob[hydrophobic]);
 
-plot_aafreq = bar(aaorder,aaprob2[ixaaorder],
-    xticks = (0.5:20, aaorder),
[…]

### 2023-02-22 Manuscripts/deNovoRates.tex
renamed:    Manuscripts/Figures/pFix_dmel.svg -> Manuscripts/Figures/pGainLossFix.svg

@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec, xcolor}
 \usepackage[colorlinks, allcolors = blue]{hyperref}
@@
 \setlength{\parindent}{0pt}
+\setlength{\parskip}{\baselineskip}
+
+\newcommand{\cmnt}[1]{{\color{purple} #1}}
 
@@
 Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein.
-This phenomenon is called \textit{de novo} gene emergence \citep{EBB-F1000,vanOss}.
-\textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{vanOss}.
+This phenomenon is called \textit{de novo} gene emergence \citep{Tautz2011,Zhao2014,EBB-F1000,Vakirlis2017,vanOss}.
+\textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{Tautz2011,Zhao2014,Vakirlis2017,vanOss,Prabh2019}.
 To express a protein, a genomic sequence should be transcribed as well as translated.
@@
 
-The primary requirement for transcription is recruitment of RNA polymerase at the gene's promoter \citep{Promoters}.
+\cmnt{The primary requirement for transcription is recruitment of RNA polymerase.
+This process is facilitated by DNA sequences called \citep{Promoters}.
 A part of the promoter, called the core promoter is the region that determines the start of transcription \citep{corepromoters}.
-The second requirement for a productive transcription is its termination at appropriate sites.
+Transcription can also be initiated without the requirement of a defined promoter sequence, and this process can occur throughout the genome \cite{PervasiveTranscription}.
 Even though pervasive transcription is widespread in eukaryotic genomes, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}.
@@
 In prokaryotes, the end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
-Thus the synthesis of a stable RNA requires promoters, as well as termination or polyadenylation signal sequences.
+Termination at an appropriate site is another fundamental requirement for a productive transcription.}
 
@@
 
-In this study we develop probabilistic models of \textit{de novo} gene emergence.
+In this study we develop probabilistic models of \textit{de novo} gene emergence \cmnt{in eukaryotes}.
 We focus on \textit{de novo} genes emerging from non-genic sequences, also known as ``proto-genes'' \citep{protogenes,vanOss}.
@@
 Alternatively, translational features can emerge in the non-genic DNA before transcriptional features.
-Specifically, we calculate the probabilities by which promoters, polyadenylation signals and ORFs emerge, and are lost, based on the rates at which different kinds of DNA mutations naturally occur (mutation bias).
+Specifically, we calculate the probabilities by which polyadenylation signals and ORFs emerge, and are lost, based on the rates at which different kinds of DNA mutations naturally occur (mutation bias).
 With these probabilities we estimate whether transcriptional and translational features preferentially evolve in a specific order.
@@
 Here, we focus on the minimal requirements for these processes to occur.
-Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence.
-We chose these two promoter motifs because they are the most abundant type of core promoters across diverse eukaryotes \citep{Promoters,corepromoters}.
-We defined the poly-A signal as a transcription requirement because it is crucial for transcript stability and its export to cytoplasm \citep{polyAterm,polyAexp}.
-For translation, we only require the gene to have an open reading frame (ORF).
-Furthermore, we assume that the ORF is uninterrupted by introns.
+\cmnt{Experimental data show that transcription in eukaryotes can be initiated genome wide, especially in regions proximal to enhancers, and more specifically, in regions depleted of nucleosomes \citep{enhancers}.
+Thus transcription of many proto-genes can be initiated at a basal transcription rate, and may not require specific core promoters.
+Although transcription may be initiated without a core promoter, a stable RNA product often requires a poly-A tail that also facilitates RNA export to cytoplasm \citep{polyAterm,polyAexp}.
+This assumption is supported by experimental data, where the majority of detectably expressed proto-genes have polyadenylated RNAs \citep{Neme2016,Witt2019,PacoEnhancers,albaYeastdenovo}.
+Thus, we define a poly-A signal as the primary requirement for transcription such that it should exist but only at the end of a transcribed region.} For translation, we only require the gene to have an open reading frame (ORF).
+We only focus on intronless genes in this study.
 
-Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost due to random mutations.
+Using simple probability models, we calculate the likelihood of finding the sequence features that facilitate \textit{de novo} gene emergence by random chance, and the rate at which these features are gained and lost due to random mutations.
 These probability models are essentially described by two kinds of probability.
-First is the probability of finding a DNA sequence feature (such as TATA box or an ORF; \autoref{methprob}).
-This probability depends on the nucleotide distribution given by the GC-content.
+First is the probability of finding a DNA sequence feature (such as the poly-A signal or an ORF; \autoref{methprob}).
+This probability depends on the nucleotide distribution \cmnt{that can be roughly approximated by the GC-content}.
 The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another (\autoref{methgain}).
-Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.
+Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and \cmnt{nucleotide composition}.
 Our model of DNA sequence evolution is in principle similar to a theoretical model described in a previous study \citep{promoterWaiting}.
@@
 
-For all calculations we used a default GC-content of 0.42 which is reasonably close to the total genomic GC-content of human \citep[0.41,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[0.416,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[0.38,][]{yeastGC}.
-We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
+\cmnt{In our model we primarily focus on \textit{de novo} gene emergence at one genomic locus.
+We assume that transcription is initiated at this locus with a probability of 0.12.
+We based this estimate on the fraction of intergenic open chromatin in \textit{Drosophila melanogaster} that is occupied by \textit{cis}-regulatory elements \cite{RedFly2022} (XX cite method)}.
+We used a default GC-content of 42\% in our calculations which is reasonably close to the total genomic GC-content of human \citep[41\%,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[41.6\%,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[38\%,][]{yeastGC}.
+Because GC content can vary between different genomic loci, we also performed of our calculations with other values of GC-content.
+\cmnt{We note that nucleotide composition in any DNA region may not be uniformly distributed.
+Therefore, we also included calculations using a more realistic approximation of nucleotide distribution, using frequencies of short DNA sequences in the intergenic genome of \textit{D.melanogaster}.
+Specifically, we calculated the frequency of all DNA trimers (for codon probabilites), and of all DNA hexamers (for poly-A signal probabilities) (XX cite method).}
+
+We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{D.
+melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
 We used mutation bias data (\autoref{mutbias}) from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
@@
 Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
-We note that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present.
+We found that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present.
 This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence.
@@
 \centering
-\includegraphics[scale=0.5]{Figures/geneGainLossComplete2_new.pdf}
+\includegraphics[scale=0.5]{Figures/pGainLossFix.pdf}
 \caption{Genes are more likely to be independently lost twice, than being born once.
@@
 We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for ORF longer than 30 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons).
+We found that proto-genes with a GC-content of 42\% and ORFs longer than 40 codons, can be lost two times independently in the time frame of one gene gain event (log-log ratio = 2, \hyperref[gainlossprob]{\autoref{gainlossprob}A}).
+For proto-genes with GC-contents of 34\% and 50\%, the minimum number of codons required for a log-log ratio of 2, are 42 and 33 respectively.
+To better understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content ({\color{blue} Fig SXX}), and that contain ORFs of different lengths (codons).
 We found that the effect of GC-content on the log-log ratio depends on the length of the ORF.
-Specifically, the log-log ratio steadily increases with GC-content ranging from 0.3 to 0.6, when the proto-gene contains an ORF with 30 codons.
+Specifically, the log-log ratio steadily increases with GC-content ranging from 30\% to 60\%, when the proto-gene contains an ORF with 30 codons.
 When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases.
 For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range).
-Overall, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.
+\cmnt{Because GC-content may not accurately describe the nucleotide composition of a locus, we calculated gene gain and loss probabilities using frequencies of DNA trimers and hexamers from the intergenic genome of \textit{D.
+melanogaster}.
+Specifically, we used the frequencies of hexamers to estimate the stationary, gain and loss probabilities of poly-A signals (which is also a hexamer) and the RNA.
+Likewise, we used the frequencies of trimers to estimate the stationary, gain and loss probabilities of different codons and the ORF.
+Using these values, we found that protogenes with more than 49 codons can be independently lost twice, in the time frame of one gene gain event (orange line, \hyperref[gainlossprob]{\autoref{gainlossprob}A}).}
 
-We performed an analogous analysis on just the ORFs (and not the whole genes).
-That is, we calculated the ratio of logarithms of ORF gain and ORF loss.
-We found that for ORFs longer than 142 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.
+\cmnt{Reproduction and selection determine the rate at which new genes spread in a population.
+To understand the combined effect of mutations, reproduction and selection we calculated two fixation probabilities.
+First is the probability that one individual gains a gene ($P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via population dynamics ($P_\textit{fix}$).
+The second probability describes the event that all individuals in a population have a gene, that is lost in one individual ($P_\textit{gene-loss}$), and eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$).
+We calculated the log-log ratio of the two probabilities under three scenarios, wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$).
+We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously.
+We found that under the neutral scenario, gene loss with extinction is still more likely than gene gain with fixation.
+However, the minimum length of ORF that causes two independent gene extinction as likely as a single gene fixation event, is more than 150 codons.
+This minimum number of codons increases with decreasing effective population size ($N_e$).
+When the gene is beneficial then gene extinctions are less frequent.
+This is true even for a marginally beneficial gene, such that in populations larger than 6400 individuals extinction of genes of any length is less likely than their fixation.
+Conversely, when the gene is mildly deleterious, its independent loss in more than two populations is more likely than its fixation in one population.}
 
-Overall, our analysis suggests that a proto-gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.
+Overall, our analysis suggests that a proto-gene expressed in only one population may not necessarily mean that it emerged for the first time in this population.
 
@@
 However, gene emergence is much more likely when one of the two features already exists (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
+Furthermore, emergence of an ORF is more likely when the proto-gene is already transcribed, and \textit{vice versa} (\autoref{rnaorfindependence}).
 To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
-First, the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF.
-This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$, \autoref{orffirsteq1}).
+First, the probability that an RNA exists and mutations cause a gain of an ORF but no disruption of transcription.
+This probability denotes the trajectory where RNA emerges first ($P_\textit{RNA-first}$, \autoref{rnafirsteq1}).
+
+\begin{equation}
+P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
+\label{rnafirsteq1}
+\end{equation}
+
+The second probability ($P_\textit{ORF-first}$, \autoref{orffirsteq1}) that denotes the trajectory where ORF emerges first, is the probability that an ORF already exists and mutations cause a gain of transcription but do not disrupt the ORF.
 
@@
 
-The second probability ($P_\textit{RNA-first}$, \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.
-
-\begin{equation}
-P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
-\label{rnafirsteq1}
-\end{equation}
-
 \begin{figure}[!t]
 \centering
-\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
-\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first.
-The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
-A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
-In both panels, the horizontal axis denotes the number of codons in the ORF.
-\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and the GC-content values (horizontal axis) at which ORF-first trajectory is more likely than RNA-first trajectory.}
+\includegraphics[scale=0.5]{Figures/whoisfirstX.pdf}
+\caption{Proto-genes preferentially emerge RNA-first.
+The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the RNA-first and the ORF-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{rnafirsteq1} \& \ref{orffirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{rnafirsteq2} \& \ref{orffirsteq2}).
+A positive value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
+In both panels, the horizontal axis denotes the number of codons in the ORF.}
 \label{whoisfirst}
@@
 
-We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
-In other words, the ORF-first hypothesis is more feasible.
-Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible.
-We found that proto-genes with an ORF containing 60 codons or more, preferentially emerge RNA-first, whereas proto-genes with shorter ORFs emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
-More specifically, the likelihood of the ORF-first trajectory relative to that of the RNA-first trajectory  decreases with increasing ORF length u a minimum value, and then increases.
-To understand if GC-content determines the evolutionary trajectory of \textit{de novo} emergence, we calculated the probabilities of the ORF-first and the RNA-first trajectory for different values of GC-content.
-We found that proto-genes with short ORF can emerge ORF-first at high GC-content.
-For example, the ORF-first trajectory is more likely at a GC-content of 0.6, for ORFs with 30 -- 36 codons (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
+\cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first).
+In other words, the RNA-first hypothesis is more feasible.
+Likewise, a negative value of the ratio would suggest that the ORF-first hypothesis is more feasible.
+We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
+In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first (Fig SXXX).
+More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length.
+This is true for loci with three different values of GC content (34\%, 42\%, 50\%).
+However, our complementary analysis using DNA oligomer frequencies from \textit{D.
+melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length.
+Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.}
 
-A more stringent definition of the ORF-first hypothesis would describe a two-step process.
-In the first step, an ORF emerges in an untranscribed region of DNA but transcription does not emerge.
-In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
-Likewise, a RNA-first hypothesis can be defined by a two-step probability.
-In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).
+A more stringent definition of the RNA-first trajectory would describe a two-step process.
+In the first step, an RNA emerges in an untranscribed region of DNA but an ORF does not emerge.
+In the second step an ORF emerges, while transcription stays intact (\autoref{rnafirsteq2}).
+Likewise, a ORF-first trajectory can be defined by a two-step probability.
+In the first step ORF emerges in an untranscribed DNA region, and in the second step transcription emerges, while the ORF remains intact (\autoref{orffirsteq2}).
 
 \begin{align}
-P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
-P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
+P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}/(1-P_\textit{ORF})}_\textit{second step} \label{rnafirsteq2} \\[1em]
+P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}
 \end{align}
 
-With this stringent definition, proto-genes with every analysed ORF length (30 -- 300) preferentially emerge via the RNA-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}C}).
-Furthermore, the two-step RNA-first trajectory becomes increasingly more probable than the two-step ORF-first trajectory, with increasing ORF length.
-
-Overall, long ORFs are likely to be lost before transcription emerges, and hence they preferentially emerge in existing transcripts.
-Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.
+\cmnt{Even with this stringent definition, \textit{de novo} emergence via the RNA-first trajectory is more feasible than via the ORF-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
+For all the three GC-content values, as well the specific analysis with \textit{D.
+melanogaster} genome, the two-step RNA-first trajectory becomes increasingly more probable than the two-step ORF-first trajectory, with increasing ORF length.}
 
@@
 \vspace{1ex}
-\includegraphics[scale=0.5]{Figures/pLossComplete2_new.pdf}
+\includegraphics[scale=0.5]{Figures/pLoss_promoterlessX.pdf}
 \caption{ORF loss in proto-genes is more probable than transcription loss.

### 2023-02-22 Modeling/DeNovoProb.jl

@@ -498,13 +498,13 @@ onlyorfgainX = (1 .- rnaprobX .- rnagainX).*orfgainX;
 rnafirst2X = onlyrnagainX.*(1 .- rnalossX).*(corfgainX./(1 .-orfprobX));
 orffirst2X = onlyorfgainX.*(1 .- orflossX).*(crnagainX./(1 .-rnaprobX));
 
-plotX_genegain_geneloss = plot(ncodons,log.(genegain2X)./log.(genelossX),
-    xlabel = "ORF length (codons)",
-    ylabel = "# Gene Losses \n per Gene Gain",
-    size = (width = cm2pt(12), height = cm2pt(11)),
-    yticks = [2,2.5,3]
-);
-savefig(plotX_genegain_geneloss, figdir*"geneGainLoss_dmel.pdf")
+# plotX_genegain_geneloss = plot(ncodons,log.(genegain2X)./log.(genelossX),
+#     xlabel = "ORF length (codons)",
+#     ylabel = "# Gene Losses \n per Gene Gain",
+#     size = (width = cm2pt(12), height = cm2pt(11)),
+#     yticks = [2,2.5,3]
+# );
+# savefig(plotX_genegain_geneloss, figdir*"geneGainLoss_dmel.pdf")
 
 plot!(plot_genegain_geneloss, ncodons,log.(genegain2X)./log.(genelossX),
     linestyle = :solid,
@@ -513,13 +513,13 @@ plot!(plot_genegain_geneloss, ncodons,log.(genegain2X)./log.(genelossX),
 );
 savefig(plot_genegain_geneloss, figdir*"geneGainLoss_promoterlessX.pdf")
 
-plotX_rnafist_orffirst = plot(ncodons,log2.(rnafirstX./orffirstX),
-    xlabel = "ORF length (codons)",
-    ylabel = "P_{ORF-first}\nP_RNA-first",
-    size = (width = cm2pt(12.5), height = cm2pt(11)),
-    yticks = [3.5,3,2.5]
-);
-savefig(plotX_rnafist_orffirst, figdir*"whoisfirst_dmel.pdf")
+# plotX_rnafist_orffirst = plot(ncodons,log2.(rnafirstX./orffirstX),
+#     xlabel = "ORF length (codons)",
+#     ylabel = "P_{ORF-first}\nP_RNA-first",
[…]

### 2023-02-23 Manuscripts/deNovoRates.tex
modified:   Manuscripts/Figures/pGainLossFix.svg 	new file:   Manuscripts/Figures/pGainLossFix2.svg 	modified:   Manuscripts/deNovoRates.tex 	modified:   Modeling/DeNovoProb.jl

@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec, xcolor}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
 \usepackage[colorlinks, allcolors = blue]{hyperref}
@@
 \citestyle{egu}
+
+\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 
-If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species.
-We asked what is the chance that the gene was present in a common ancestor but was lost in all lineages except one.
+If a gene is found in one population but not populations, then the most parsimonious assumption is that this gene emerged only in one specific population.
+We asked what is the chance that the gene was present in an ancestral population but was lost in all lineages except one.
 To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}).
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/pGainLossFix.pdf}
-\caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
-\label{gainlossprob}
-\vspace{1ex}
-\hrule
-\end{figure}
-
 The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$.
@@
 
-\cmnt{Reproduction and selection determine the rate at which new genes spread in a population.
-To understand the combined effect of mutations, reproduction and selection we calculated two fixation probabilities.
-First is the probability that one individual gains a gene ($P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via population dynamics ($P_\textit{fix}$).
-The second probability describes the event that all individuals in a population have a gene, that is lost in one individual ($P_\textit{gene-loss}$), and eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$).
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.5]{Figures/pGainLossFix2.pdf}
+\caption{Genes are more likely to be independently lost twice, than being born once.
+The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
+\label{gainlossprob}
+\vspace{1ex}
+\hrule
+\end{figure}
+
+\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population.
+To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals.
+First is the probability that any any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$).
+The second probability describes the event that all individuals in a population have two copies of a gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$).
 We calculated the log-log ratio of the two probabilities under three scenarios, wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$).
 We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously.
-We found that under the neutral scenario, gene loss with extinction is still more likely than gene gain with fixation.
-However, the minimum length of ORF that causes two independent gene extinction as likely as a single gene fixation event, is more than 150 codons.
-This minimum number of codons increases with decreasing effective population size ($N_e$).
-When the gene is beneficial then gene extinctions are less frequent.
-This is true even for a marginally beneficial gene, such that in populations larger than 6400 individuals extinction of genes of any length is less likely than their fixation.
-Conversely, when the gene is mildly deleterious, its independent loss in more than two populations is more likely than its fixation in one population.}
+Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele.
+Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF.
+For example, a gene of any length is less likely to go extinct than it is fixed, in populations with at least 6400 individuals.
+Furthermore, a beneficial gene can be independently lost in more than two small populations with 400 individuals, only when it has more than 133 codons.
+When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population.}
 
@@
 \cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first).
-In other words, the RNA-first hypothesis is more feasible.
-Likewise, a negative value of the ratio would suggest that the ORF-first hypothesis is more feasible.
+In other words, the RNA-first trajectory is more feasible.
+Likewise, a negative value of the ratio would suggest that the ORF-first trajectory is more feasible.
 We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
@@
 
-To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
+\cmnt{To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
 We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}).
-At low GC-content (0.3 -- 0.34) ORF loss is less likely than RNA-loss, for short ORFs (30 codons; \hyperref[lossprob]{\autoref{lossprob}B}).
+If transcription is strictly dependent on a core promoter (TATA-box or Inr), small genes (30 codons) with low GC content ($>$34\%) can be preferentially lost by loss of transcription (Figure SXXX).
+When the organismal fitness is affected only by the protein product of a proto-gene, then both RNA loss and ORF loss would affect the fitness equally.
+Thus the likelihood of gene extinction through ORF loss or RNA loss depends solely on the probability of these events.
+
+Our analysis of gene loss mechanism considers the loss of one feature (for example transcription) without requiring the other feature to remain intact (for example the ORF).
+With more stringent analysis where we consider the loss of only one feature but not the other, we find a the that exclusive ORF loss is more likely than exclusive RNA loss (Supplementary section XXX).
+The difference between the likelihood of the two gene loss mechanisms is more remarkable when we consider exclusive loss of one feature.}
+
+Overall, our analysis suggests widespread transcription loss may be indicative of a negative selection on protein expression.
+
+{\color{PineGreen} \textbf{MOVE TO SUPPLEMENTARY DATA}
 
@@
 Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.
-
-Thus, widespread transcription loss can suggest a negative selection on protein expression.
+}
 
@@
 We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of proteins encoded by proto-genes.
-To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform.
+To this end, we first asked if the expected frequency of different amino acids is uniform.
+For a GC-content of 42\% and with DNA trimer frequencies from \textit{D.
+melanogaster} intergenic genome, we found that it is not uniform.
 Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids.
@@
 This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
-However, GC-content also plays a role.
+However, nucleotide composition also determines the frequency of an amino acid.
 For example, leucine, arginine and serine, all have six codons each.
-However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 0.42 (\autoref{aafreq}{\color{blue}A}).
+However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 42\% (\autoref{aafreq}{\color{blue}A}).
+The same is true for the frequencies of these amino acids estimated from the DNA trimer distribution.
 Overall, our analysis is roughly in agreement with many previous studies \citep{aasubOhta,aafreq3,aafreq4}.
@@
 
-We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
+We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (GC = 42\%, Pearson $\rho = 0.955, P<10^{-10}$; \textit{D.
+melanogaster} trimer frequencies, Pearson $\rho = 0.841, P<10^{-10}$).
 
@@
 
-The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
+The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (GC = 42\%, Pearson $\rho = -0.255, P = 0.2818$; \textit{D.
+melanogaster} trimer frequencies, Pearson $\rho = -0.356, P=0.1232$).
 
@@
 This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection.
-We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}).
-Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that its median value is 0.38.
-Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.
+\cmnt{We note that that actual frequencies of specific substitutions depend on the initial distribution of different codons in the ORF sequence.}
 
@@
 
-We found that $\sim$37.5\% of amino acids in a protein sequence are expected to be hydrophobic.
-This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}.
+We found that $\sim$40\% of amino acids in a protein sequence are expected to be hydrophobic (with a GC-content of 42\% as well as with \textit{D.
+melanogaster} trimer distribution).
+This finding is broadly in agreement with the previous study \citep{hydrophobicRatchet}.
 We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}.
@@
 
-We found that the expected hydropathy is higher than zero ($0.39$), which suggests that random protein sequences are not on an average, hydrophobic.
+We found that the expected hydropathy is higher than zero (0.38 with GC\% = 42; 0.36 with \textit{D.
+melanogaster} trimer frequencies), which suggests that random protein sequences are not on an average, hydrophobic.
 
@@
 
-Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is 1.114, which suggests that gain of a hydrophobic amino acid is slightly more probable than its loss.
+Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is slightly more than 1 (1.114 with GC\% = 42; 1.015 with \textit{D.
+melanogaster} trimer frequencies).
+This suggests that gain of a hydrophobic amino acid is slightly more probable than its loss.
 We performed a complementary analysis where we calculated the average change in hydrophobicity due to any random mutation ($\bar{\Delta\alpha}$), defined by:
@@
 \begin{equation}
-\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y}}
+\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} \mu_{x\to y}}
 \end{equation}
 
-We found the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity.
+We found the average change in hydrophobicity to be slightly less than zero ($-0.0798$), which suggests that on an average mutations cause a very small increase in hydrophobicity.
 We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein.
@@
 That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity.
-With this focused analysis, we find that the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).
+With this focused analysis, we find that the average change in hydrophobicity is comparable ($-0.0793$).
+For both the analyses, we note that the actual change in hydrophobicity will depend on the composition of the ancestral protein sequence.
 
-Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect is very small.
+Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect may be very small.
 
@@
 We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge.
-In a hypothetical case, a proto-gene may be present in a species (A) but is absent in two sister taxa (B and C).
+In a hypothetical case, a proto-gene may be present in a population (A) but is absent in two other geographically isolated populations (B and C).
 Two evolutionary scenarios could explain this observation.
-In the first scenario, the proto-gene simply emerged for the first time in species A.
-The second scenario posits that the proto-gene was present in the common ancestor of the three species but was lost in both species B and species C.
-Our models suggest that the second scenario is more likely.
-Future phylogenetic studies on proto-genes should consider this finding for making inferences on the dynamics of gene gain and loss.
+In the first scenario, the proto-gene simply emerged for the first time in population A.
+The second scenario posits that the proto-gene was present in the ancestral population but was lost in both population B and population C.
+Our models suggest that the second scenario is more likely than is assumed.
+\cmnt{This inference can be further extended to the level of closely related species where the divergence times are relatively shorter than the timescale of multiple gain and loss events within a lineage}.
+Future phylogenetic studies on proto-genes should consider this finding for inferring the dynamics of gene gain and loss.
 
@@
 That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
-We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
-However, if both transcription and ORF are absent in a genomic locus, then \textit{de novo} emergence is more likely to occur transcription-first.
-Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors.
+We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first).
+Long non-coding RNAs may be a potential source of proto-genes but this evolutionary process may be constrained by a different set of factors.
 A previous study has shown that new transcripts frequently arise in regions overlapping existing genes.
@@
 
-Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
+CHECK THIS XXX: Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
 For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory.
@@
 
-Like all theoretical models, our models are not without limitations.
+XXX: Like all theoretical models, our models are not without limitations.
 First, we assume that nucleotides are uniformly distributed according to GC-content.

### 2023-02-23 Modeling/DeNovoProb.jl

@@ -322,6 +322,12 @@ onlyorfgain = (1 .- rnaprob .- rnagain).*orfgain;
 rnafirst2 = onlyrnagain.*(1 .- rnaloss).*(corfgain./(1 .-orfprob));
 orffirst2 = onlyorfgain.*(1 .- orfloss).*(crnagain./(1 .-rnaprob));
 
+# Exclusive RNA/ORF loss
+
+onlyrnaloss = crnaloss.*corfstay./corfprob;
+onlyorfloss = corfloss.*crnastay./crnaprob;
+
+
 # Plot single step trajectory probability versus ORF length
 
 plot_rnafist_orffirst = plot(ncodons,log2.(rnafirst./orffirst)[dGC[1],:],
@@ -589,13 +595,12 @@ lossfix = hcat([genelossX/(2*x) for x in popsizes]...);
 
 selcoef = 0.001
 
-gainfixPS = hcat([genegain2X.*kimura(selcoef,2*x) for x in popsizes]...);
-gainfixNS = hcat([genegain2X.*kimura(-selcoef,2*x) for x in popsizes]...);
+gainfixPS = hcat([2*x*genegain2X.*kimura(selcoef,2*x) for x in popsizes]...);
+gainfixNS = hcat([2*x*genegain2X.*kimura(-selcoef,2*x) for x in popsizes]...);
 
-lossfixPS = hcat([genelossX.*kimura(-selcoef,2*x) for x in popsizes]...);
-lossfixNS = hcat([genelossX.*kimura(selcoef,2*x) for x in popsizes]...);
+lossfixPS = hcat([2*x*genelossX.*kimura(-selcoef,2*x) for x in popsizes]...);
+lossfixNS = hcat([2*x*genelossX.*kimura(selcoef,2*x) for x in popsizes]...);
 
-fixprobs = log.(gainfix)./log.(lossfix);
 fixprobsPS = log.(gainfixPS)./log.(lossfixPS);
 fixprobsNS = log.(gainfixNS)./log.(lossfixNS);
 
@@ -635,7 +640,7 @@ exp_hydropathy = sum(hydropathy3.*aaprob2)
 
 aaprob_dme = [sum(nprob3.(gencode[gencode[:,3] .== x,1])) for x in aas];
 aaprob2_dme = aaprob_dme./sum(aaprob_dme);
-exp_hydropathy_dme = sum(hydropathy3.*aaprob2)
[…]

### 2023-02-24 Manuscripts/SM_deNovoRates.tex
new file:   Manuscripts/SM_deNovoRates.tex

--- 2022-12-26 Manuscripts/deNovoRates.tex
@@
 
+\usepackage[usenames,dvipsnames]{xcolor}
+
 \captionsetup[figure]{labelfont=bf,textfont={small,normalfont}}
@@
 
-\renewcommand{\subsectionautorefname}{Section}
-\renewcommand{\subsubsectionautorefname}{Section}
 \titleformat{\paragraph}[hang]{\normalfont\normalsize\itshape}{\theparagraph}{0em}{}
@@
 \setlength{\parindent}{0pt}
+\setlength{\parskip}{\baselineskip}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\newcommand{\cmnt}[1]{{\color{purple} #1}}
 
@@
 
-\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
+\title{{\Large Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}\\ Supplementary Material}
 \author{Bharat Ravi Iyengar$^1$, Erich Bornberg-Bauer$^{1,2}$}
@@
 
-\begin{abstract}
-New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence.
-To synthesize a protein, DNA must be transcribed as well as translated.
-Both processes need certain DNA sequence features.
-Stable transcription requires promoters and a polydenylation signal, while translation requires at least an open reading frame (ORF).
-We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly genes emerge and are lost.
-We also investigate the effect of the order by which DNA features evolve, and if sequence composition is biased by mutation rate.
-We rationalize how genes are lost much more rapidly than they emerge, and how genes with long ORFs preferentially arise in regions that are already transcribed.
-Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.
-\end{abstract}
+\section{Model with core promoters}
 
-\section{Introduction}
+In this model we required a core promoter to exist for transcription to occur, along with the requirement of a single poly-A at the end of the transcribed region as defined in the main text.
+Specifically, we require a transcribed gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr).
+We chose these two promoter motifs because they are the most abundant type of core promoters across diverse eukaryotes \citep{Promoters,corepromoters}.
 
-Organisms evolve new traits by expanding their functional genome.
-Evolution of new genes is one of the ways by which new traits can emerge.
-The definition of a gene is complicated, and has been changing constantly \citep{genepostencode}.
-We use the following working definition of a gene: a gene is a region of the genome that gives rise to a functional product, that is an RNA or a protein.
-The next big challenge is to define what a function is, and how to classify gene functions.
-One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions \citep{deNovoFunction}.
-Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins like mRNAs, or participate in protein synthesis like rRNAs and tRNAs, are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.
-
-In this study we focus on the evolution of genes that encode new proteins.
-New protein coding genes frequently arise via duplication of existing protein coding genes.
-These duplicated genes then genetically and functionally diverge from their parent genes by accumulating mutations \citep{Long2003,IAD}.
-Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein.
-This phenomenon is called \textit{de novo} gene emergence \citep{EBB-F1000,vanOss}.
-\textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{vanOss}.
-To express a protein, a genomic sequence should be transcribed as well as translated.
-Therefore, the gene needs sequence features that enable both these processes.
-
-The primary requirement for transcription is recruitment of RNA polymerase at the gene's promoter \citep{Promoters}.
-A part of the promoter, called the core promoter is the region that determines the start of transcription \citep{corepromoters}.
-The second requirement for a productive transcription is its termination at appropriate sites.
-Even though pervasive transcription is widespread in eukaryotic genomes, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}.
-Most eukaryotic mRNAs and long non-coding RNAs, are polyadenylated at their 3' termini, which makes them resistant to degradation.
-Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals \citep{polyA}.
-In prokaryotes, the end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
-Thus the synthesis of a stable RNA requires promoters, as well as termination or polyadenylation signal sequences.
-
-The second major requirement for protein expression is translation of the transcribed mRNA.
-The most fundamental requirement for translation is the presence of an open reading frame (ORF).
-Usually, an mRNA needs additional features to initiate protein synthesis.
-These features include ribosome binding sites in prokaryotes \citep{RBS}, and Kozak consensus sequences in eukaryotes \citep{kozak,kozakDroso,kozakHuman}.
-
-Although gain of transcription and translation features guarantees \textit{de novo} birth of a protein coding gene, it does not ensure that the gene would persist in the genome for many generations.
-The newly born gene can lose the features as easily as it gained them unless it has been fixed in the genome, for example via evolutionary selection.
-Specifically, if the protein synthesized by the \textit{de novo} gene provides a fitness advantage to the host organism, it will undergo positive selection \citep{deNovoFunction}.
-Simultaneously, the protein should have low toxicity and cost of synthesis, to survive purifying selection.
-A common mechanism by which protein mediated toxicity occurs is by misfolding and aggregation of proteins \citep{misfold1,misfold2}.
-
-In this study we develop probabilistic models of \textit{de novo} gene emergence.
-We focus on \textit{de novo} genes emerging from non-genic sequences, also known as ``proto-genes'' \citep{protogenes,vanOss}.
-We model \textit{de novo} gene emergence as a two step process.
-In the first step, a non-genic DNA sequence gains transcriptional features that allows it to express an untranslated (non-coding) RNA.
-In the second step this non-coding RNA gene acquires translational features that allows it to express a protein.
-Alternatively, translational features can emerge in the non-genic DNA before transcriptional features.
-Specifically, we calculate the probabilities by which promoters, polyadenylation signals and ORFs emerge, and are lost, based on the rates at which different kinds of DNA mutations naturally occur (mutation bias).
-With these probabilities we estimate whether transcriptional and translational features preferentially evolve in a specific order.
-Using a similar approach, we calculate how random mutations affect protein composition.
-In our models, we assume that the proteins expressed from these proto-genes provide no fitness advantage and are not toxic to the host organism.
-Thus our models are based on neutral evolution.
-Our models predict that there is indeed a preferred sequence of DNA feature evolution during \textit{de novo} emergence, even under the assumption of neutrality.
-
-\section{Results}
-
-In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss.
-A proto-gene emerges from non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation \citep{protogenes,vanOss}.
-Both transcription and translation are complex processes involving many biomolecular complexes that work in concert.
-Here, we focus on the minimal requirements for these processes to occur.
-Specifically for transcription, we require a gene to have one of the two core promoter motifs, the TATA-box or the Initiator element (Inr), and a polyadenylation (poly-A) signal sequence.
-We chose these two promoter motifs because they are the most abundant type of core promoters across diverse eukaryotes \citep{Promoters,corepromoters}.
-We defined the poly-A signal as a transcription requirement because it is crucial for transcript stability and its export to cytoplasm \citep{polyAterm,polyAexp}.
-For translation, we only require the gene to have an open reading frame (ORF).
-Furthermore, we assume that the ORF is uninterrupted by introns.
-
-Using simple probability models, we calculate the likelihood of finding these features in a DNA sequence by random chance, and the rate at which these features are gained and lost due to random mutations.
-These probability models are essentially described by two kinds of probability.
-First is the probability of finding a DNA sequence feature (such as TATA box or an ORF; \autoref{methprob}).
-This probability depends on the nucleotide distribution given by the GC-content.
-The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another (\autoref{methgain}).
-Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and GC-content.
-Our model of DNA sequence evolution is in principle similar to a theoretical model described in a previous study \citep{promoterWaiting}.
-In this study, the authors have calculated the time required for different transcription factor binding sites to emerge in gene promoters.
-We extend this principle to calculate not only the probability of sequence emergence but also the likelihood that an emerged sequence will remain intact or be lost (\autoref{methfeatures}).
-
-For all calculations we used a default GC-content of 0.42 which is reasonably close to the total genomic GC-content of human \citep[0.41,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[0.416,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[0.38,][]{yeastGC}.
-We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{Drosophila melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
-We used mutation bias data (\autoref{mutbias}) from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
-
-Using a similar approach, we predict how the nucleotide distribution and random mutations shape protein composition and evolution.
-
-Overall, our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.
-
-\subsection{How likely is gene loss?}
-
-If a gene is found in one species but not in sister taxa, then the most common assumption is that this gene emerged only in one specific species.
-We asked what is the chance that the gene was present in a common ancestor but was lost in all lineages except one.
-To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}).
-Specifically, the probability of gene gain can be defined as the sum of three probabilities.
-First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}).
-Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}).
-Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
-We note that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present.
-This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence.
-This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}).
-Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed.
-Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively.
-We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:
-
-\begin{equation}
-P_\textit{gene-gain}^* = P_\textit{RNA-gain \vline\ ORF} \times P_\textit{ORF-stay} + P_\textit{ORF-gain \vline\ RNA} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
-\label{genegaineq}
-\end{equation}
-
-\autoref{genegaineq} describes the total probability of gene gain that includes the probability that the gene was not previously present.
-To more precisely address the question of whether a gene that was absent in an ancestor but emerges in a descendant, we calculated the corresponding conditional probability of gene gain:
-
-\begin{equation}
-P_\textit{gene-gain} = P_\textit{gene-gain}^*/(1-P_\textit{ORF} - P_\textit{RNA})
-\label{genegaineq2}
-\end{equation}
-
-We note that the difference between the total probability ($P_\textit{gene-gain}^*$) and the conditional probability ($P_\textit{gene-gain}$) of gene gain, is very small.
-
-Next, we calculated the probability of gene loss, given the gene (transcription and ORF) is already present($P_\textit{gene-loss}$).
-It is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
-Specifically:
-
-\begin{equation}
-P_\textit{gene-loss} = P_\textit{RNA-loss} + P_\textit{ORF-loss}
-\label{genelosseq}
-\end{equation}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/geneGainLossComplete2_new.pdf}
-\caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
-\label{gainlossprob}
-\vspace{1ex}
-\hrule
-\end{figure}
-
-The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$.
-To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms (log-log ratio) of gene gain and gene loss for proto-genes with an ORF ranging from 30 -- 300 codons in length.
-
-$$ \text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$$
-
-For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event.
-We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that for ORF longer than 30 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons).
-We found that the effect of GC-content on the log-log ratio depends on the length of the ORF.
-Specifically, the log-log ratio steadily increases with GC-content ranging from 0.3 to 0.6, when the proto-gene contains an ORF with 30 codons.
-When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases.
-For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range).
-Overall, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.
-
-We performed an analogous analysis on just the ORFs (and not the whole genes).
-That is, we calculated the ratio of logarithms of ORF gain and ORF loss.
-We found that for ORFs longer than 142 codons, the likelihood of two independent ORF loss events is more than or equal to a single ORF gain event.
-
-Overall, our analysis suggests that a proto-gene expressed in only one species may not necessarily mean that it emerged for the first time in this species.
-
-\subsection{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}
-
-For a proto-gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge.
-That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain.
-Thus it may appear that the order of the occurrence of these two events does not matter.
-However, gene emergence is much more likely when one of the two features already exists (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
-To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
-First, the probability that ORF exists and  mutations cause gain of transcription but no disruption of the ORF.
-This probability denotes the trajectory where ORF emerges first ($P_\textit{ORF-first}$, \autoref{orffirsteq1}).
-
-\begin{equation}
-P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}
-\label{orffirsteq1}
-\end{equation}
-
-The second probability ($P_\textit{RNA-first}$, \autoref{rnafirsteq1}) that denotes the trajectory where transcription emerges first, is the probability that DNA is already transcribed and mutations cause a gain of ORF but do not eliminate transcription.
-
-\begin{equation}
-P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
-\label{rnafirsteq1}
-\end{equation}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
-\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first.
-The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}).
-A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
-In both panels, the horizontal axis denotes the number of codons in the ORF.
-\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and the GC-content values (horizontal axis) at which ORF-first trajectory is more likely than RNA-first trajectory.}
-\label{whoisfirst}
-\vspace{1ex}
-\hrule
-\end{figure}
-
-We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first).
-In other words, the ORF-first hypothesis is more feasible.
-Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible.
-We found that proto-genes with an ORF containing 60 codons or more, preferentially emerge RNA-first, whereas proto-genes with shorter ORFs emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
-More specifically, the likelihood of the ORF-first trajectory relative to that of the RNA-first trajectory  decreases with increasing ORF length u a minimum value, and then increases.
-To understand if GC-content determines the evolutionary trajectory of \textit{de novo} emergence, we calculated the probabilities of the ORF-first and the RNA-first trajectory for different values of GC-content.
-We found that proto-genes with short ORF can emerge ORF-first at high GC-content.
-For example, the ORF-first trajectory is more likely at a GC-content of 0.6, for ORFs with 30 -- 36 codons (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
-
-A more stringent definition of the ORF-first hypothesis would describe a two-step process.
-In the first step, an ORF emerges in an untranscribed region of DNA but transcription does not emerge.
-In the second step transcription emerges, while ORF stays intact (\autoref{orffirsteq2}).
-Likewise, a RNA-first hypothesis can be defined by a two-step probability.
-In the first step transcription emerges in a non-genic DNA (that does not have an ORF), and in the second step an ORF emerges, while transcription remains intact (\autoref{rnafirsteq2}).
-
-\begin{align}
-P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}\\[1em]
-P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}/(1-P_\textit{ORF})}_\textit{second step}\label{rnafirsteq2}
-\end{align}
-
-With this stringent definition, proto-genes with every analysed ORF length (30 -- 300) preferentially emerge via the RNA-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}C}).
-Furthermore, the two-step RNA-first trajectory becomes increasingly more probable than the two-step ORF-first trajectory, with increasing ORF length.
-
-Overall, long ORFs are likely to be lost before transcription emerges, and hence they preferentially emerge in existing transcripts.
-Conversely, short ORFs are more likely to stay undisrupted until the emergence of transcription.
-
-\subsection{Would extensive transcription loss suggest negative selection of toxic proteins?}
-
-\begin{figure}[!b]
-\centering
-\hrule
-\vspace{1ex}
-\includegraphics[scale=0.5]{Figures/pLossComplete2_new.pdf}
-\caption{ORF loss in proto-genes is more probable than transcription loss.
-The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
-The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
-\label{lossprob}
-\end{figure}
-
-Proto-genes that do not provide any fitness benefit to an organism may not be fixed in populations via natural selection.
-These genes may be lost due to mutation pressure.
-Some newborn proto-genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way.
-These genes would thus be eliminated from the population genomes via negative selection.
-We note that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation).
-However, it is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively).
-Furthermore, non-stop mutations can also lead to cellular toxicity \citep{nonstop}.
-Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may more effectively inactivate the associated genes.
-
-To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}).
-At low GC-content (0.3 -- 0.34) ORF loss is less likely than RNA-loss, for short ORFs (30 codons; \hyperref[lossprob]{\autoref{lossprob}B}).
-
-We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact proto-gene.
-In the first scenario, the ORF remains intact whereas the transcription is lost.
-The probability of this scenario ($P_\textit{onlyORF-loss}$) is described by:
-
-\vspace{-1ex}
-
-\begin{equation}
-P_\textit{onlyRNA-loss} = \frac{P_\textit{ORF-stay \vline\ RNA}}{P_\textit{ORF \vline \ RNA}} \times P_\textit{RNA-loss \vline\ ORF}
-\end{equation}
-
-We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
-To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
-
-In the second scenario, transcription stays intact while ORF is lost. The probability of this scenario is:
-
-\vspace{-1em}
-\begin{equation}
-P_\textit{onlyORF-loss} = \frac{P_\textit{RNA-stay \vline\ ORF}}{P_\textit{RNA \vline \ ORF}} \times P_\textit{ORF-loss \vline\ RNA}
-\end{equation}
-
-We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths.
-Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.
-
-Thus, widespread transcription loss can suggest a negative selection on protein expression.
-
-\subsection{Does mutation bias shape protein composition?}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/proteinComposition.pdf}
-
-\caption{Mutation bias and GC-content shape protein composition.
-Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes).
-The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one.
-Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis).
-Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.
-We only show substitutions that can be realized by a single nucleotide mutation but we include all possible nucleotide mutations in the directionality calculations for these substitutions.
-In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
-\label{aafreq}
-
-\vspace{1ex}
-\hrule
-\end{figure}
-
-In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss.
-We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of proteins encoded by proto-genes.
-To this end, we first asked if the expected frequency of different amino acids is uniform, and found that it is not uniform.
-Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids.
-On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}{\color{blue}A}).
-This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
-However, GC-content also plays a role.
-For example, leucine, arginine and serine, all have six codons each.
-However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 0.42 (\autoref{aafreq}{\color{blue}A}).
-Overall, our analysis is roughly in agreement with many previous studies \citep{aasubOhta,aafreq3,aafreq4}.
-However, notable differences exist between our analysis and that of the previous studies because the latter primarily focused on characterized proteins and not random proto-gene derived protein sequences.
-
-Next, we aimed to find out if some amino acid substitutions are more probable than the others.
-To this end, we calculated the gain probability (\autoref{aagaineq}) for every amino acid.
-Specifically for every amino acid $x$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
-More precisely, the total probability that an amino acid $y$ will mutate to an amino acid $x$ is the product of the expected frequency of amino acid $y$ ($P_y$), and the substitution rate from $x$ to $y$ ($\mu_{y\to x}$).
-The gain probability for an amino acid $x$ is then defined as:
-
-\begin{equation}
-G_x = \sum_{b \neq a} P_y \times \mu_{y\to x}
-\label{aagaineq}
-\end{equation}
-
-We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (Pearson $\rho = 0.955, P<10^{-10}$).
-
-Next, we calculated how easily is an amino acid mutated to any other amino acid.
-Specifically, the conditional loss probability for an amino acid ($L_x$), given the amino acid already exists, is defined as:
-
-\begin{equation}
-L_x = \sum_{x \neq y} \mu_{x\to y}
-\label{aalosseq}
-\end{equation}
-
-The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (Pearson $\rho = -0.255, P = 0.2818$).
-
-Next, we investigated if some amino acid substitutions are more common than the others.
-As expected, amino acid substitutions that require more than one nucleotide change were much less probable than those needing just a single nucleotide change.
-Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutations but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities).
-Although single nucleotide mutations are more probable than double mutations, the different single mutations do not occur at the same rate due to mutation rate bias.
-Thus an amino acid substitution (for example K$\to$E) may not be as likely to occur as the reverse substitution (E$\to$K).
-That is substitutions between a pair of amino acids may have a directionality.
-Although many previous studies have explored the likelihood of different amino acid substitutions \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}, none to our knowledge have focused primarily on the directionality of these substitutions.
-To verify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{x\to y}$, for every pair of amino acids, $x$ and $y$.
-A positive value of $D_{x\to y}$ means that $x$ to $y$ substitution is more likely than $y$ to $x$ substitution, and \textit{vice versa}.
-
-\begin{equation}
-\textit{D}_{x\to y} = \log_2\left(\frac{\mu_{x\to y}}{\mu_{y\to x}}\right)
-\label{directionality}
-\end{equation}
-
-As we mentioned in the previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
-We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}).
-The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836.
-That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other.
-For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution.
-This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection.
-We also performed an analogous analysis where we defined the substitution directionality using the ratio of corresponding amino acid gain probabilities (\autoref{aagaineq}).
-Using this definition reduces the range of directionality values ($\textit{D}_{xy}$) such that its median value is 0.38.
-Although this more stringent definition reduces the intensity of the directionality, it does not change the qualitative results.
-
-A previous study found that mutations that cause hydrophobic amino acids to appear on protein surfaces, can lead to evolution of protein dimers \citep{hydrophobicRatchet}.
-The authors also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids.
-They based this argument on another study that estimated mutation rate bias in bacteria \citep{bacteriaMutbias}.
-The authors \citep{hydrophobicRatchet} further suggested that hydrophobic amino acids are more frequently found in random protein sequences.
-
-We asked if our model also makes similar predictions.
-To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic.
-A popular hydrophobicity scale of amino acids is based on solubility of an amino acid in water or ethanol \citep{hydropathy}.
-However, this scale does not classify tryptophan as a hydrophobic amino acid.
-Another hydrophobicity scale estimated by a different study \citep{WWhydropathy} is more biologically realistic and is based on rates of transfer of different amino acids between water and a hydrophobic medium.
-This study analysed two hydrophobic media -- a lipid bilayer and octanol, and also considered the effect of peptide bonds in the calculations.
-Therefore we used the hydrophobicity scale by \cite{WWhydropathy} for classifying amino acids such that hydrophobic amino acids have a negative free energy change of transfer from water to hydrophobic media.
-Conversely, a hydrophilic amino acid would have a positive value of the free energy change.
-Based the octanol hydrophobicity scale, we classified the following amino acids as hydrophobic -- cysteine (C), phenylalanine (F), isoleucine (I), leucine (L), methionine (M), valine (V), tryptophan (W) and tyrosine (Y).
-With the lipid bilayer hydrophobicity scale, valine is not classified as hydrophobic, possibly because it does not integrate well in the bilayer.
-Therefore we used the octanol scale for our calculations.
-
-We found that $\sim$37.5\% of amino acids in a protein sequence are expected to be hydrophobic.
-This finding is broadly in agreement with the previous study that places this percentage at 40\% \citep{hydrophobicRatchet}.
-We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}.
-Next, we asked if a protein sequence in general tends to be hydrophobic.
-Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).
-
-\begin{equation}
-\bar{\alpha} = \sum_{x} P_x \alpha_x
-\label{expHydropathy}
-\end{equation}
-
-We found that the expected hydropathy is higher than zero ($0.39$), which suggests that random protein sequences are not on an average, hydrophobic.
-
-To find out if random mutations indeed cause hydrobphobic amino acids to occur in protein sequences, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$, \autoref{alphagain}).
-Likewise, we calculated the probability that any mutation causes a hydrophobic to non-hydrophobic amino acid substitution ($P_{\alpha\textit{-loss}}$, \autoref{alphaloss}).
-
-\begin{align}
-P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
-%
-P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
-\end{align}
-
-Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is 1.114, which suggests that gain of a hydrophobic amino acid is slightly more probable than its loss.
-We performed a complementary analysis where we calculated the average change in hydrophobicity due to any random mutation ($\bar{\Delta\alpha}$), defined by:
-
-\begin{equation}
-\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} P_x \times \mu_{x\to y}}
-\end{equation}
-
-We found the average change in hydrophobicity to be $-0.033$, which suggests that on an average mutations cause a very small increase in hydrophobicity.
-We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein.
-To focus exclusively on non-hydrophobic to hydrophobic amino acid substitutions, we excluded amino acid pairs where both amino acids are either hydrophobic or are non-hydrophobic.
-That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity.
-With this focused analysis, we find that the average change in hydrophobicity is even smaller ($-3.7\times 10^{-4}$).
-
-Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect is very small.
-
-\section{Discussion}
-
-In this work, we addressed some fundamental questions about \textit{de novo} emergence of protogenes.
-Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory of mutational events, and how the composition of their protein sequence is determined.
-To this end, we developed mathematical models and used them to address specific questions.
-These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection.
-Thus our models provide an opportunity to test the alternative hypothesis that some proto-genes are evolving under selection.
-
-We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge.
-In a hypothetical case, a proto-gene may be present in a species (A) but is absent in two sister taxa (B and C).
-Two evolutionary scenarios could explain this observation.
-In the first scenario, the proto-gene simply emerged for the first time in species A.
-The second scenario posits that the proto-gene was present in the common ancestor of the three species but was lost in both species B and species C.
-Our models suggest that the second scenario is more likely.
-Future phylogenetic studies on proto-genes should consider this finding for making inferences on the dynamics of gene gain and loss.
-
-Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence.
-That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
-We show that, in the absence of selection, long ORFs ($\geq$60 codons) are likely to emerge in existing transcripts (transcription emerges first), whereas short ORFs are likely to emerge before transcription emerges in the genomic locus.
-However, if both transcription and ORF are absent in a genomic locus, then \textit{de novo} emergence is more likely to occur transcription-first.
-Long non-coding RNAs may be a potential source of proto-genes with long ORFs but this evolutionary process may be constrained by a different set of factors.
-A previous study has shown that new transcripts frequently arise in regions overlapping existing genes.
-These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}.
-This is an example of RNA-first trajectory.
-However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences.
-Therefore, they may be less frequently mutated or lost.
-
-Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
-For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory.
-On the other hand, the selection would have no effect on an ORF that is not transcribed.
-This is based on the assumption that the ORF sequence has no other function (for example, regulation), or is not under indirect selection (for example, if it overlaps with an existing gene).
-Thus positive selection will intensify the preference for RNA-first trajectory.
-
-We show that transcription loss is less likely than the loss of long ORFs.
-However, it may be difficult to find direct evidence for such a phenomenon.
-Specifically, it is difficult to infer from data if transcription was lost or never emerged.
-If the protein encoded by an ORF is indeed toxic, it would be silenced or purged rapidly in microevolutionary time scales.
-One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status.
-Mutations within an ORF sequence are much more probable than mutations that disrupt transcription or an ORF.
-This is so because the entire ORF sequence can have a higher number of mutable sites than a promoter.
-If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.
-
-A large volume of work exists on the likelihood of amino acid substitutions and many of these models are widely used in phylogenetics studies \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}.
-However, commonly used amino acid substitution models such as PAM and BLOSUM \citep{PAM,blosum} do not focus on directionality of mutations (the substitution matrices are symmetric).
-Our aim was to highlight how proteins can evolve under mutation pressure and that mutation bias causes some substitutions to occur more frequently than the others.
-Furthermore, we do not base our substitution probability calculations on known protein sequences but rather on random protein sequences that are more likely to to be encoded in proto-genes.
-Using our models, we show that for a pair of amino acids, one direction of substitution can be significantly more likely than the other.
-If a substitution occurs more frequently than it is expected to occur in our neutral model, then directional evolution can be a possible explanation for the observation.
-We also show that random mutations can lead to a small increase in a protein's hydrophobicity.
-Hydrophobic amino acids can facilitate folding \citep{Dill1985}, but can also cause proteins to aggregate \citep{hydrophobicRatchet}.
-Thus the effect of hydrophobicity increasing mutations on a protein's stability and toxicity, is dependent on where they occur in the protein structure.
-
-Like all theoretical models, our models are not without limitations.
-First, we assume that nucleotides are uniformly distributed according to GC-content.
-This is not true as nucleotide composition can vary significantly across the genome.
-Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}.
-However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
-Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}.
-The accuracy of the models' prediction will depend on the accuracy of parameters.
-However, mutation rate itself only determines the values of different probabilities.
-Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate.
-More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other.
-Thus our inferences based on our chosen parameters will not universally apply to all species.
-Third, we assume that all mutations are independent.
-This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
-Fourth, we did not model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes.
-For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}.
-We instead focused on minimal requirements for transcription and translation.
-However, our modeling framework allows inclusion of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}).
-Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation.
-Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations.
-Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.
-
-A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}.
-Our models are not an exception.
-They are not 100\% accurate but they are useful for making predictions that can lead to a more focused experimental validation.
-These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution.
-Therefore our work opens up an opportunity for theoretical and computational biologists to develop models using our modeling framework, that more accurately describe their system of interest.
-
-\section{Materials and Methods}
-
-\subsection{Mutation probabilities}
-
-\begin{table}[H]
-\centering
-\begin{tabular}{c c}
-\toprule
-\textbf{Substitution} & Probability($\mu$) \\\midrule
-\texttt{A:T}$\to$\texttt{T:A} & 0.056 \\\midrule
-\texttt{A:T}$\to$\texttt{G:C} & 0.243 \\\midrule
-\texttt{A:T}$\to$\texttt{C:G} & 0.074 \\\midrule
-\texttt{G:C}$\to$\texttt{A:T} & 0.483 \\\midrule
-\texttt{G:C}$\to$\texttt{T:A} & 0.075 \\\midrule
-\texttt{G:C}$\to$\texttt{C:G} & 0.069 \\\bottomrule
-\end{tabular}
-\caption{Mutation bias probabilities for different nucleotide mutations.
-\texttt{A:T} denotes an \texttt{A}-\texttt{T} base pair in a double stranded DNA.
-Thus \texttt{A}$\to$\texttt{G} mutation on one DNA strand would cause a \texttt{T}$\to$\texttt{C} mutation on the complementary strand.
-We describe the other mutations in the same way.}
-\label{mutbias}
-\end{table}
-
-We calculated nucleotide substitution probabilities based on mutation rate and mutation rate bias data.
-Specifically, we  used a mutation rate ($\lambda$) of $7.8\times10^{-9}$ mutations per nucleotide position per generation \citep{drosophilamutrate}.
-We derived our mutation bias parameters from two published studies, the first on \textit{Drosophila melanogaster} \citep{drosophilamutrate}, and the second on humans \citep{humanmutrate}.
-\autoref{mutbias} shows the exact values of mutation bias probabilities that we used in this study.
-
-\subsection{Probabilities of finding, gaining, and losing a DNA sequence}
-\label{methbasic}
-
-\subsubsection{Probability of finding a DNA sequence}
-\label{methprob}
-
-We calculated the probability of finding a DNA sequence based on global nucleotide frequency distributions, given by the GC-content.
-We used a GC-content of 0.42 for our calculations.
-Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is: $S = 0.5\times$GC-content.
-The probability of finding an \texttt{A} or a \texttt{T} is: $W = 0.5 - S$.
-Using these values, we calculated the probability of finding a DNA sequence motif by chance.
-For example, the probability of finding the sequence \texttt{ATG} would be: $W\times W \times S$.
-
-\subsubsection{Probability of gaining a DNA sequence}
-\label{methgain}
-
-We calculated the probability of gaining a DNA sequence due to mutations using GC-content, mutation rate and mutation bias.
-Specifically, we calculated the probability that a DNA sequence does not exist, and it emerges due to specific nucleotide mutations.
-More precisely, this probability is the product of two other probabilities.
-The first is the probability of finding a DNA sequence ($x$) that is not the sequence of interest.
-The second probability is that this sequence $x$ mutates to the sequence of interest.
-To explain this calculation better, we use the example of \texttt{CTA} mutating to \texttt{ATG}.
-The first probability, that is the probability of finding \texttt{CTA} by chance is $SW^2$.
-\texttt{CTA} mutates to \texttt{ATG} via two nucleotide mutations (\texttt{C}$\to$\texttt{A} and \texttt{A}$\to$\texttt{G}).
-Thus the probability of this DNA change would be the probability of two nucleotide mutations ($\lambda^2$) multiplied by two mutation bias probabilities (\texttt{G:C}$\to$\texttt{T:A} and \texttt{A:T}$\to$\texttt{G:C}).
-Overall, the chance of \texttt{CTA} mutating to \texttt{ATG} would be:
-
-$$SW^2 \times \lambda^2 \times \mu_{\texttt{G:C}\to\texttt{T:A}} \times \mu_{\texttt{A:T}\to\texttt{G:C}}$$
-
-Next, we calculated the probability that every nucleotide triplet that is not \texttt{ATG}, mutates to \texttt{ATG}.
-This can happen via one, two, or three nucleotide mutations.
-The sum of all these mutation probabilities is the probability of \texttt{ATG} gain.
-
-Using the same principle we calculated the gain probability of any DNA sequence motif (of any length or composition).
-We excluded insertions and deletions as a mechanism of gain of small DNA sequences that we analysed in this study.
-
-\subsubsection{Probability of losing a DNA sequence}
-
-We calculated the loss of a DNA sequence motif using the same principle we used for calculating gain probabilities.
-However, we defined loss probability as a conditional probability, that is we assume that the DNA sequence of interest already exists in the genome.
-For example, the loss probability of a specific \texttt{ATG} sequence would be the sum of probabilities of \texttt{ATG} mutating to any of the other 63 nucleotide triplets (via one, two, or three nucleotide mutations).
-We use conditional loss probabilities by default, because usually one is interested in finding out how quickly an existing DNA sequence can erode.
-
-We used this method to calculate the loss probability of any DNA sequence motif, and we excluded insertions and deletions from this calculation.
-
-\subsection{Probabilities of finding, gaining, and losing DNA features}
-\label{methfeatures}
-
-Usually, a specific function is encoded in DNA by several DNA sequences.
-For example, translation stop is encoded by three codons (\texttt{TGA}, \texttt{TAG}, \texttt{TAA}).
-We use the term DNA features to mean a set of DNA sequences that are associated with the same function.
-For every such DNA feature set, there is a complementary set of non-features, that is DNA sequences that are not associated with the feature's function.
-For example, the non-feature set of stop codons would be all the other 61 codons.
-
-The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set (\autoref{methprob}).
-
-The probability that a DNA feature is gained via mutations, is the sum of probabilities of every non-feature sequence mutating to any feature sequence.
-If $F$ denotes the feature set, and $\mu_{y\to x}$ denotes the probability of a DNA sequence $y$ mutating to a DNA sequence $x$ (see \autoref{methgain}), then:
-
-\begin{equation}
-P_\textit{feature-gain} = \sum_{x \in F} \sum_{y \notin F} P_y \times \mu_{y\to x}
-\end{equation}
-
-The probability that a DNA feature is lost via mutations is a conditional probability that given a feature exists, it mutates to any of the non-feature sequences.
-
-\begin{equation}
-P_\textit{feature-loss} = \frac{\displaystyle\sum_{x \in F} \sum_{y \notin F} P_x \times \mu_{x\to y}}{\displaystyle\sum_{x \in F} P_x}
-\end{equation}
-
-Because a feature set usually has many DNA sequences, a mutation can change a feature sequence such that the resulting sequence is also a part of the feature set.
-Thus we defined the probability ($P_\textit{feature-stay}$) that a feature does not erode because of mutations.
-Specifically, it is the sum of two probabilities.
-First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where the mutated sequence remains a part of the feature set.
-
-\begin{equation}
-P_\textit{feature-stay} = P_0 + \sum_{x \in F} \sum_{\genfrac{}{}{0pt}{1}{y \in F}{y\neq x}} \mu_{x\to y}
-\end{equation}
-
-The probability that no mutation occurs ($P_0$) in a DNA sequence of length $k$ is described by Poisson distribution.
-
-$$P_0 = 1-e^{-k\lambda}$$
-
-Because the mutation rate is biased (\autoref{mutbias}), the probability that no mutation occurs in a DNA sequence depends on its composition.
-
-The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is thus described as:
-
-$$\lambda_\texttt{AT} = 2\times\lambda\times(\mu_{\texttt{A:T}\to\texttt{T:A}} + \mu_{\texttt{A:T}\to\texttt{G:C}} + \mu_{\texttt{A:T}\to\texttt{C:G}})$$
-
-Likewise, the probability that a \texttt{G} or a \texttt{C} mutates ($\lambda_\texttt{GC}$) is:
-
-\vspace{-1ex}
-
-$$\lambda_\texttt{GC} = 2\times\lambda\times(\mu_{\texttt{G:C}\to\texttt{A:T}} + \mu_{\texttt{G:C}\to\texttt{T:A}} + \mu_{\texttt{G:C}\to\texttt{C:G}})$$
-
-(Note that the general mutation rate, $\lambda$, is an average of $\lambda_\texttt{AT}$ and $\lambda_\texttt{GC}$.)
-
-\vspace{1\baselineskip}
-
-Thus the probability that a sequence of length $k$, containing $m$ number of \texttt{A} and \texttt{T}, does not mutate is:
-
-\vspace{-1ex}
-
-$$P_0 = (1-e^{-m\lambda_\texttt{AT}})\times(1-e^{-(k-m)\lambda_\texttt{GC}})$$
-
-\subsection{Probabilities of finding, gaining, and losing an ORF}
-
-\label{methORF}
-
-\subsubsection{Probability of finding an ORF}
-
-A reading frame is a nucleotide sequence with a length that is a multiple of three.
-A reading frame that begins with a start codon (\texttt{ATG}), and ends with one of the three stop codons (\texttt{TAG}, \texttt{TGA}, \texttt{TAA}) is an open reading frame (ORF).
-This necessarily means that there are no stop codons within the sequence.
-Thus the probability of finding an ORF containing $k$ codons including start and stop codons ($P_\textit{ORF}$) is:
-
-\begin{equation}
-P_\textit{ORF}(k) = P_\textit{ATG} \times P_\textit{stop} \times (1 - P_\textit{stop})^{k-2}
-\label{eqorfprob}
-\end{equation}
-
-Here, $P_\textit{ATG}$ and $P_\textit{stop}$ are the probabilities of finding a start codon, and a stop codon by chance, respectively.
-
-\subsubsection{Probability of gaining an ORF}
-
-As we defined in the previous section, an ORF has three requirements (start codon, stop codon, and no premature stop codon in the sequence).
-Thus an ORF can emerge due to mutations via three mechanisms.
-In each of these mechanisms, one requirement is initially absent whereas the other two are present.
-Then mutations cause the missing requirement to emerge while not disrupting the other two requirements.
-More specifically, the ORF can be gained via the following three mechanisms:
-\begin{enumerate}
-\item Gain of a start codon ($P_\textit{ATG-gain}$) while a stop codon continues to exist at the end of a reading frame ($P_\textit{stop-stay}$), and there is no emergence of stop codon within the reading frame ($1- P_\textit{stop} - P_\textit{stop-gain}$).
-\item Gain of a stop codon ($P_\textit{stop-gain}$), while a start codon continues to exist at the beginning of a reading frame ($P_\textit{ATG-stay}$), and there is no emergence of stop codon within the reading frame.
-\item Loss of a premature stop codon, at any of the $k-2$ codon positions within the reading frame ($P_\textit{stop-gain}$).
-At the same time start and stop codons remain undisrupted by mutations, and no stop codon emerges at any of the other $k-3$ positions.
-\end{enumerate}
-
-Thus we define the probability of ORF gain ($P_\textit{ORF-gain}$) as:
-
-\begin{align}
-P_\textit{ORF-gain}(k) & = \quad P_\textit{ATG-gain}\times P_\textit{stop-stay} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
-& \quad + P_\textit{ATG-stay}\times P_\textit{stop-gain} \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-2} \nonumber \\[1pt]
-& \quad + P_\textit{ATG-stay}\times P_\textit{stop-stay} \times P_\textit{stop-loss}\times(k-2) \times (1- P_\textit{stop} - P_\textit{stop-gain})^{k-3} 
-\label{eqorfgain}
-\end{align}
-
-\subsubsection{Probability of ORF loss}
-
-ORF can be lost when any of its three requirements are lost. We thus define the conditional probability of ORF loss as:
-
-\begin{equation}
-P_\textit{ORF-loss}(k) = P_\textit{ATG-loss} + P_\textit{stop-loss} + (k-2)\times \frac{P_\textit{stop-gain}}{1-P_\textit{stop}}
-\label{eqorfloss}
-\end{equation}
-
-The last term in this equation describes the conditional probability of stop-gain, given the assumption that no stop codon exists within the ORF.
-
-\subsubsection{Probability that ORF remains intact}
-
-We assumed that an ORF of a certain length remains intact if none of the necessary features are lost.
-However, the ORF sequence can mutate to cause non-synonymous changes in the translated protein sequence.
-This condition applies to all the three ORF probabilities described above.
-We define the probability that an ORF remains intact ($P_\textit{ORF-stay}$) as:
-
-\begin{equation}
-P_\textit{ORF-stay}(k) = P_\textit{ATG-stay} \times P_\textit{stop-stay} \times (1 - P_\textit{stop} - P_\textit{stop-gain})^{k-2}
-\label{eqorfstay}
-\end{equation}
-
-\subsection{Probabilities of finding, gaining, and losing transcription}
-
-\label{methRNA}
-To calculate the probabilities of finding, gaining, and losing transcription, we first defined the minimum requirements for transcription.
-These requirements are a promoter, either a TATA-box or an Initiator (Inr) element, and a polyadenylation signal.
-Each of these three DNA sequence motifs consists of more than one DNA sequence.
-
-\subsubsection{Definition of TATA-box, Inr and poly-A signal}
+\subsection{Definition of TATA-box, Inr and poly-A signal}
 
@@
 
-We calculated probabilities of finding, gaining and losing these features as we described in the previous section.
+We calculated probabilities of finding, gaining and losing these features as we described in the main text.
 
-\subsubsection{Probability of transcription}
-
-For a a DNA region of a length, to be transcribed it should have a promoter the beginning of the DNA region, and a poly-A signal at the end of it.
-The transcribed DNA region should also not have any poly-A signal sequences within it except the one at the end of it.
-A transcribed DNA of length $l$ nucleotides, contains $l-5$ 6-mers.
-Each of these 6-mers can be a poly-A signal with a probability, $P_\textit{polyA}$.
-Probability that none of these 6-mers is a poly-A signal is $(1-P_\textit{polyA})^{l-5}$.
-Since we focus on protein coding genes in this work, we describe the length of transcript in terms of the length of protein it can encode.
-Assuming that untranslated regions have no effect on protein synthesis, the length of a transcript in number of nucleotides should be at least three times the length of the encoded protein in number of amino acids.
-Thus the probability that a DNA region produces a transcript that can harbor an ORF with $k$ codons ($P_\textit{RNA}$) is:
-
-\begin{equation}
-P_\textit{RNA}(k) = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr}) (1-P_\textit{polyA})^{3k-5}
-\label{eqrnaprob}
-\end{equation}
-
-\subsubsection{Probability of gaining transcription}
+\subsection{Probability of gaining transcription}
 
@@
 
-\subsubsection{Probability of losing transcription}
+\subsection{Probability of losing transcription}
 
@@
 A transcription unit that has both the promoters must lose both of them.
-For all the three kinds of transcription units, transcription would be lost if poly-A signal is lost.
+For all the three kinds of transcription units, transcription would be lost if poly-A signal is lost, or if one is gained within the transcription unit.
 
@@
  & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
- & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
+ & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss} + (3k-5)\times P_\textit{polyA-gain}
 \label{eqrnaloss}
@@
 
-\subsubsection{Probability that transcription remains intact}
+\subsection{Probability that transcription remains intact}
 
@@
 \begin{equation}
-P_\textit{RNA-stay}(k) = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
+P_\textit{RNA-stay}(k) = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
 \label{eqrnastay}
@@
 
-\subsection{Interdependence of transcription and ORF probabilities}
+\subsection{Gene gain versus gene loss}
 
-\label{rnaorfindependence}
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{Figures/Supp_figures/geneGainLossComplete2_new.pdf}
+\caption{Genes are more likely to be independently lost twice, than being born once.
+The vertical axis in both panels shows the number of independent gene loss events that can occur relative to one gene gain event -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
+\end{figure}
 
-All the four poly-A signal variants contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon.
-Since stop codons cannot exist inside an ORF, presence of an ORF reduces the possible number of poly-A signal sites in a DNA region.
-Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame.
-Thus, out of $3k-5$ positions for a 6-mer (length of a poly-A signal) in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}).
-Furthermore a poly-A signal cannot overlap with the start codon, thus reducing two more positions from all sites where a poly-A signal can exist (one position is already counted -- if a poly-A signal overlaps with the first codon in the second frame, then the second codon becomes a stop codon).
-Overall, the possible number poly-A sites on an ORF with $k$ codons is $2k-5$.
-Thus for a gene to emerge via gain of transcription, the DNA should not contain poly-A sites at these $2k-5$ positions.
-As a consequence, the probability of transcription gain for a specific transcript length, is higher when the DNA region contains an ORF than when it does not (\autoref{eqrnagain}).
-We have ignored the probability of poly-A signal overlapping with a stop codon because it makes a very small difference in the probability of RNA gain.
-This is so because at least one stop codon (\texttt{TAA}) allows a poly-A signal to overlap with it in three frames.
+\subsection{Transcription first or ORF first}
 
-Conversely to the effect of an existing ORF on transcription gain, stop codons are less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
-Specifically, the probability that a 6-mer contains a stop codon in the third position is same as the probability of finding a stop codon ($P_\textit{stop}$).
-If poly-A signals are excluded from these 6-mers, then the probability of finding a stop codon is $P_\textit{stop} - P_\textit{poly-A}$.
-Conversely, lack of poly-A signals will increase the probability of amino acid coding codons and thereby their probability of mutating into a stop codon.
-Furthermore, the likelihood that an ORF is undisrupted my premature nonsense mutations, is lower when the DNA is already transcribed.
-Specifically:
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{Figures/Supp_figures/whichfirstComplete2_new.pdf}
+\caption{Proto-genes with ORFs longer than 59 codons, emerge RNA-first, whereas proto-gens with shorter ORFs emerge ORF-first.
+The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the ORF-first and the RNA-first trajectories, described as \textbf{(A)} a single-step process and \textbf{(C)} a two-step process.
+A negative value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
+In both panels, the horizontal axis denotes the number of codons in the ORF.
+\textbf{(B)} Filled area denotes the ORF length in codons (vertical axis) and the GC-content values (horizontal axis) at which ORF-first trajectory is more likely than RNA-first trajectory.}
+\end{figure}
 
-$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} + P_\textit{poly-A})}{1- P_\textit{stop}}$$
+\subsection{ORF loss versus transcription loss}
 
-Overall the evolutionary dynamics of transcription and ORF gain, are not independent of each other.
-That is, presence of one feature makes the gain of the other more likely.
-The dependence becomes more prominent with increasing number of codons in the ORF of the proto-gene.
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{Figures/Supp_figures/pLossComplete2_new.pdf}
+\caption{ORF loss in proto-genes is more probable than transcription loss.
+The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
+The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
+\end{figure}
 
-\subsection{Probability of amino acid substitutions}
+\section{Effect on GC content on the probability of gene gain and loss}
 
-We calculated the probability of amino acid substitutions based on nucleotide substitution in the corresponding codons.
-In this case, multiple codons that code for the same amino acid constitute a feature set (\autoref{methfeatures}).
-Similarly when analysing hydrobhobic to non-hydrophobic substitutions, all codons that represent hydrophobic amino acids form a feature set (and \textit{vice versa}).
+We analysed the effect of effect on GC content on the number independent gene losses per gene gain, in the promoterless model.
 
-\section{Data availability}
-We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
-Specifically, our model is implemented in the script \texttt{DeNovoProb.jl} which in turn uses the script \texttt{nucleotidefuncts.jl} for some basic functions.
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{Figures/Supp_figures/geneGainLoss_klengcr_promoterless.pdf}
+\caption{Genes are more likely to be independently lost twice, than being born once.
+The vertical axis shows the number of independent gene loss events that can occur relative to one gene gain event -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
+The horizontal axis shows the GC-content.}
+\end{figure}
+
+\section{Exclusive ORF or transcription loss}
+
+In the promoterless model, we made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact proto-gene.
+In the first scenario, the ORF remains intact whereas the transcription is lost.
+The probability of this scenario ($P_\textit{onlyORF-loss}$) is described by:
+
+\vspace{-1ex}
+
+\begin{equation}
+P_\textit{onlyRNA-loss} = \frac{P_\textit{ORF-stay \vline\ RNA}}{P_\textit{ORF \vline \ RNA}} \times P_\textit{RNA-loss \vline\ ORF}
+\end{equation}
+
+We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
+To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
+
+In the second scenario, transcription stays intact while ORF is lost. The probability of this scenario is:
+
+\vspace{-1em}
+\begin{equation}
+P_\textit{onlyORF-loss} = \frac{P_\textit{RNA-stay \vline\ ORF}}{P_\textit{RNA \vline \ ORF}} \times P_\textit{ORF-loss \vline\ RNA}
+\end{equation}
+
+We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths.
+
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{Figures/Supp_figures/pLossOnly_promoterlessX.pdf}
+\caption{Exclusive ORF loss in proto-genes is more probable than exclusive transcription loss.
+The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
+The vertical axis shows the $\log_2$ transformed ratio of only ORF loss and only RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
+\end{figure}
 

### 2023-02-24 Manuscripts/deNovoRates.tex
new file:   Manuscripts/SM_deNovoRates.tex

@@
 We assume that transcription is initiated at this locus with a probability of 0.12.
-We based this estimate on the fraction of intergenic open chromatin in \textit{Drosophila melanogaster} that is occupied by \textit{cis}-regulatory elements \cite{RedFly2022} (XX cite method)}.
+We based this estimate on the fraction of intergenic open chromatin in \textit{Drosophila melanogaster} that is occupied by \textit{cis}-regulatory elements (\autoref{methRNA})}.
 We used a default GC-content of 42\% in our calculations which is reasonably close to the total genomic GC-content of human \citep[41\%,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[41.6\%,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[38\%,][]{yeastGC}.
@@
 Therefore, we also included calculations using a more realistic approximation of nucleotide distribution, using frequencies of short DNA sequences in the intergenic genome of \textit{D.melanogaster}.
-Specifically, we calculated the frequency of all DNA trimers (for codon probabilites), and of all DNA hexamers (for poly-A signal probabilities) (XX cite method).}
+Specifically, we calculated the frequency of all DNA trimers for codon probabilites, and of all DNA hexamers for poly-A signal probabilities (\autoref{methRNA}).}
 
@@
 For proto-genes with GC-contents of 34\% and 50\%, the minimum number of codons required for a log-log ratio of 2, are 42 and 33 respectively.
-To better understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content ({\color{blue} Fig SXX}), and that contain ORFs of different lengths (codons).
+To better understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content ({\color{blue} Figure S4}), and that contain ORFs of different lengths (codons).
 We found that the effect of GC-content on the log-log ratio depends on the length of the ORF.
@@
 The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
+The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.
+lack lines denote the probability values estimated from overall GC content and the orange lined denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
+melanogaster} intergenic genome.}
 \label{gainlossprob}
@@
 
-\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population.
+\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population \citep{KimuraFix}.
 To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals.
-First is the probability that any any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$).
-The second probability describes the event that all individuals in a population have two copies of a gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$).
-We calculated the log-log ratio of the two probabilities under three scenarios, wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$).
+First is the probability that any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$).
+The second probability describes the event that all individuals in a population have two copies of the proto-gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$).
+We calculated the log-log ratio of the two probabilities under three scenarios (\autoref{pFix}), wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$).
 We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously.
@@
 Overall, our analysis suggests that a proto-gene expressed in only one population may not necessarily mean that it emerged for the first time in this population.
+
+Despite low probability of gene gain at one particular locus ($\sim 10^{-15}$ per generation), the overall rate of gene gain throughout the genome could be much higher.
+To this end we calculated the total gain rate of genes of any length, at any locus in \textit{D.
+melanogaster} intergenic genome.
+We found this rate to be $2.8\times 10^{-5}$ new genes per generation, that corresponds to approximately $7.5\times 10^{-5}$ new genes per year \citep[assuming \textit{D.
+melanogaster} generation time of two weeks;][]{drosophilaModelOrg}.
+We performed an analogous analysis on the rate of gain of new transcripts and found this rate to be 1.53 transcripts per year.
+This estimate is slightly higher than the rate of transcript gain predicted from \textit{D.
+melanogaster} population transcriptomics data \citep[0.13 -- 0.34;][]{AnnaPeteTranscripts}.
 
@@
 \caption{Proto-genes preferentially emerge RNA-first.
-The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the RNA-first and the ORF-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{rnafirsteq1} \& \ref{orffirsteq1}) and \textbf{(C)} a two-step process (Equations \ref{rnafirsteq2} \& \ref{orffirsteq2}).
+The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the RNA-first and the ORF-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{rnafirsteq1} \& \ref{orffirsteq1}) and \textbf{(B)} a two-step process (Equations \ref{rnafirsteq2} \& \ref{orffirsteq2}).
 A positive value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
@@
 We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
-In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first (Fig SXXX).
+In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first ({\color{blue}Figure S2}).
 More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length.
@@
 
-\cmnt{Even with this stringent definition, \textit{de novo} emergence via the RNA-first trajectory is more feasible than via the ORF-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
+\cmnt{Even with this stringent definition, \textit{de novo} emergence via the RNA-first trajectory is more probable than via the ORF-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
 For all the three GC-content values, as well the specific analysis with \textit{D.
@@
 \caption{ORF loss in proto-genes is more probable than transcription loss.
-The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
-The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
+The horizontal axis denotes the number of codons in the ORF.
+The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.
+Black lines denote the probability values estimated from overall GC content and the orange lined denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
+melanogaster} intergenic genome.}
 \label{lossprob}
@@
 We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}).
-If transcription is strictly dependent on a core promoter (TATA-box or Inr), small genes (30 codons) with low GC content ($>$34\%) can be preferentially lost by loss of transcription (Figure SXXX).
+If transcription is strictly dependent on a core promoter (TATA-box or Inr), small genes (30 codons) with low GC content ($>$34\%) can be preferentially lost by loss of transcription ({\color{blue} Figure S3}).
 When the organismal fitness is affected only by the protein product of a proto-gene, then both RNA loss and ORF loss would affect the fitness equally.
@@
 Our analysis of gene loss mechanism considers the loss of one feature (for example transcription) without requiring the other feature to remain intact (for example the ORF).
-With more stringent analysis where we consider the loss of only one feature but not the other, we find a the that exclusive ORF loss is more likely than exclusive RNA loss (Supplementary section XXX).
+With more stringent analysis where we consider the loss of only one feature but not the other, we find a the that exclusive ORF loss is more likely than exclusive RNA loss ({\color{blue}Figure S5}).
 The difference between the likelihood of the two gene loss mechanisms is more remarkable when we consider exclusive loss of one feature.}
@@
 Overall, our analysis suggests widespread transcription loss may be indicative of a negative selection on protein expression.
-
-{\color{PineGreen} \textbf{MOVE TO SUPPLEMENTARY DATA}
-
-We made our analysis of gene loss more stringent by comparing the probabilities of two evolutionary scenarios, both of which begin with an intact proto-gene.
-In the first scenario, the ORF remains intact whereas the transcription is lost.
-The probability of this scenario ($P_\textit{onlyORF-loss}$) is described by:
-
-\vspace{-1ex}
-
-\begin{equation}
-P_\textit{onlyRNA-loss} = \frac{P_\textit{ORF-stay \vline\ RNA}}{P_\textit{ORF \vline \ RNA}} \times P_\textit{RNA-loss \vline\ ORF}
-\end{equation}
-
-We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
-To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
-
-In the second scenario, transcription stays intact while ORF is lost. The probability of this scenario is:
-
-\vspace{-1em}
-\begin{equation}
-P_\textit{onlyORF-loss} = \frac{P_\textit{RNA-stay \vline\ ORF}}{P_\textit{RNA \vline \ ORF}} \times P_\textit{ORF-loss \vline\ RNA}
-\end{equation}
-
-We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths.
-Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.
-}
 
@@
 
-CHECK THIS XXX: Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
-For example, if the gene produces a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the ORF emerges via RNA-first trajectory.
-On the other hand, the selection would have no effect on an ORF that is not transcribed.
-This is based on the assumption that the ORF sequence has no other function (for example, regulation), or is not under indirect selection (for example, if it overlaps with an existing gene).
-Thus positive selection will intensify the preference for RNA-first trajectory.
+Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
+For example, if an ORF encodes a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the it emerges via RNA-first trajectory.
+On the other hand, the selection would have no effect on an ORF that is not transcribed, as long as it is does not harbor a functional DNA sequence (such as regulatory elements), and is not under indirect selection (for example, if it overlaps with an existing gene).
+The ORF may mutate before transcription emerges and the corresponding mutated protein may not be beneficial anymore.
+Thus positive selection may further increase the likelihood of the RNA-first trajectory.
 
@@
 
-XXX: Like all theoretical models, our models are not without limitations.
+Like all theoretical models, our models are not without limitations.
 First, we assume that nucleotides are uniformly distributed according to GC-content.
 This is not true as nucleotide composition can vary significantly across the genome.
-Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}.
-However, our models can be adapted to investigate specific genomic loci, where information on the local GC-content and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
-Second, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}.
+\cmnt{Since our model primarily focuses on gene emergence at one locus, it can be used to predict evolutionary probabilities for different genomic loci.
+Furthermore, several predictions of our models are not qualitatively affected by GC-content.
+More realistic nucleotide composition distribution estimated from \textit{D.
+melanogaster} intergenic sequences, also qualitatively agree with GC-content based predictions.
+Thus, most predictions are qualitatively robust to small differences in parameters.
+Second, in our models, we assumed uniform mutation rate throughout the genome.
+However, recent studies have shown Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}}.
+Our models can be adapted to investigate specific genomic loci, where estimates of the local nucleotide composition and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
+Third, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}.
 The accuracy of the models' prediction will depend on the accuracy of parameters.
@@
 Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate.
-More importantly, different species may have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other.
+Mutation bias may have more significant effect on the predictions than mutation rate.
+Existing data suggest that different species have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other.
 Thus our inferences based on our chosen parameters will not universally apply to all species.
-Third, we assume that all mutations are independent.
+Fourth, we assume that occurrence of a point mutation is independent of another mutation.
 This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
-Fourth, we did not model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes.
+Fourth, we did not specifically model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes.
 For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}.
-We instead focused on minimal requirements for transcription and translation.
-However, our modeling framework allows inclusion of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}).
+\cmnt{To model this phenomenon in our analyses, we assumed that DNA loci feasible for transcript emergence are close to known \textit{cis-} regulatory elements \citep{RedFly2005,RedFly2007,RedFly2010,RedFly2018,RedFly2022}.
+Specifically, we assumed that transcription can be initiated at a probability that is equal to the proportion of such regulatory elements in the intergenic genome.
+This is a rough estimate but it can be updated when more specific data becomes available.
+For example, ATAC-seq data can provide reliable estimates of nucleosome depleted regions, that can act as transcription initiation sites \citep{ATAC-seq}.
+This analysis can be supplemented with data from techniques like native elongating transcript sequencing \citep[NET-seq;][]{NET-seq} and global run-on sequencing \citep[GRO-seq;][]{GRO-seq}, that can detect nascent transcripts.} Our modeling framework also allows incorporation of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}).
 Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation.
@@
 Our models are not an exception.
-They are not 100\% accurate but they are useful for making predictions that can lead to a more focused experimental validation.
-These models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution.
-Therefore our work opens up an opportunity for theoretical and computational biologists to develop models using our modeling framework, that more accurately describe their system of interest.
+They may not be 100\% accurate but they make useful predictions that can lead to a more focused experimental validation.
+\cmnt{Many of our predictions are qualitatively robust to small parameter changes.
+Some predictions are also in agreement with experimental data.
+For example, our prediction of transcript gain rate is only slightly inflated than the estimated rates using an empirical analysis of \textit{D.
+melanogaster} populations \citep{AnnaPeteTranscripts}.
+This inflated estimate could have possibly resulted from our relaxed requirements for a promoter, or non-inclusion of other unknown factors that may have affected population evolution.
+When supplemented with experimental data analysis our models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution, and can provide more accurate predictions.} Therefore our work opens up an opportunity for theoretical and computational biologists to design analyses using our modeling framework, that more accurately describe their system of interest.
 
@@
 We calculated the probability of finding a DNA sequence based on global nucleotide frequency distributions, given by the GC-content.
-We used a GC-content of 0.42 for our calculations.
-Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is: $S = 0.5\times$GC-content.
+Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is:
+
+$$S = \frac{0.5\times\text{GC\%}}{100}$$
+
 The probability of finding an \texttt{A} or a \texttt{T} is: $W = 0.5 - S$.
@@
 For example, the probability of finding the sequence \texttt{ATG} would be: $W\times W \times S$.
+
+\cmnt{We also estimated the probability of finding specific DNA sequences in a reference genome.
+Specifically, we calculated the frequencies of all 64 trimers and all 4096 hexamers in the genomic regions of \textit{D.
+melanogaster} that exist in open chromatin, and do not contain any known genes or regulatory elements (see \autoref{methRNA})}.
 
@@
 
+\cmnt{We also calculated all the above-defined probabilities ($P_\textit{feature-gain}$, $P_\textit{feature-loss}$ and $ P_\textit{feature-stay}$), using DNA trimer and hexamer frequencies from \textit{D.
+melanogaster}.
+In this case the probability of finding a nucleotide sequence depends on the trimer/hexamer distributions instead of GC-content, but the probability of mutational changes are only dependent on mutation bias.
+Trimers and hexamers would contain codons and polyA signals, respectively.}
+
 \subsection{Probabilities of finding, gaining, and losing an ORF}
@@
 \label{methRNA}
-To calculate the probabilities of finding, gaining, and losing transcription, we first defined the minimum requirements for transcription.
-These requirements are a promoter, either a TATA-box or an Initiator (Inr) element, and a polyadenylation signal.
-Each of these three DNA sequence motifs consists of more than one DNA sequence.
-
-\subsubsection{Definition of TATA-box, Inr and poly-A signal}
-
-\paragraph{TATA box:}
-
-The consensus sequence of TATA-box is \texttt{TATA [T/A] A [T/A] [G/A]} \citep{tata2} such that, any of the nucleotides enclosed within the square brackets, can exist in that specific position in the sequence.
-Another study used X-ray crystallography to identify mutations that are tolerated in TATA-box consensus sequence \citep{tata1}.
-Thus there are eight consensus sequences, and at seven positions, a non-consensus nucleotide is tolerated (\autoref{tatatab}).
-This makes a TATA-box feature set that contains 104 sequences.
-We defined the corresponding non-feature set by generating all 8-mer DNA sequences that do not belong to the TATA-box feature set.
-
-\begin{table}[!t]
-\centering
-\begin{tabular}{c c}
-\toprule
-Position  & Accepted nucleotides \\ \midrule
-1         & T$\gg$C$>$A,G        \\ \midrule
-2         & A$\gg$T               \\ \midrule
-3         & T$\gg$A,C             \\ \midrule
-4         & A$\gg$T               \\ \midrule
-5         & T/A                   \\ \midrule
-6         & A$\gg$G$>$C/T         \\ \midrule
-7         & A/T$>$G$>$C           \\ \midrule
-8         & G/A$>$C/T             \\ \bottomrule
-\end{tabular}
-\caption{Nucleotide residues that are tolerated at different positions in a TATA-box \citep[based on][]{tata1}}
-\label{tatatab}
-\end{table}
-
-\paragraph{Inr:} The consensus sequence of Inr is \texttt{[T/G/C] [T/G/C] CA [T/G/C] [T/A]}, which gives 54 sequences in the Inr feature set \citep{Inr}.
-We generated the corresponding non-feature set with all DNA 6-mers that are distinct from the 54 Inr sequences.
-
-\paragraph{poly-A signal:} There are four known poly-A signal variants -- \texttt{AATAAA}, \texttt{ATTAAA}, \texttt{AGTAAA}, and \texttt{TATAAA} that compose the poly-A signal feature set \citep{polyA}.
-The non-feature set consists of all other DNA 6-mers.
-
-\vspace{1\baselineskip}
-
-We calculated probabilities of finding, gaining and losing these features as we described in the previous section.
 
@@
 
-For a a DNA region of a length, to be transcribed it should have a promoter the beginning of the DNA region, and a poly-A signal at the end of it.
-The transcribed DNA region should also not have any poly-A signal sequences within it except the one at the end of it.
-A transcribed DNA of length $l$ nucleotides, contains $l-5$ 6-mers.
-Each of these 6-mers can be a poly-A signal with a probability, $P_\textit{polyA}$.
-Probability that none of these 6-mers is a poly-A signal is $(1-P_\textit{polyA})^{l-5}$.
+\cmnt{In this study we defined a transcription model that is not dependent on specific promoter sequences.
+Instead we defined a probability that transcription is initiated at a genomic locus.
+To this end, we first extracted the genomic coordinates of intergenic sequences in \textit{D.
+melanogaster} genome \citep[release 6.49, FlyBase;][]{flybase}.
+Using the data on the genomic locations of open chromatin in \textit{D.
+melanogaster} \citep{dmel_Chromatin}, we obtained open chromatin regions within intergenic genome.
+Next, we calculated the fraction of this open-intergenic genome occupied by \textit{cis-} regulatory elements \citep{RedFly2005,RedFly2007,RedFly2010,RedFly2018,RedFly2022}.
+A significant proportion of these elements include enhancers.
+Enhancers, that can promote transcription in their neighboring genomic regions \citep{enhancers}, are also known to promote \textit{de novo} gene emergence \citep{PacoEnhancers}.
+Thus we reasoned that proportion of \textit{cis-} regulatory elements in open intergenic chromatin would roughly approximate the likelihood of transcription initiation ($P_\textit{RNAinit}$).
+We do not model the exact sequences because data is not yet available on exact sequence requirements and mutational robustness of enhancers and other \textit{cis-} regulatory elements.
+Thus poly-A signal is the primary sequence determinant of transcription probabilities.
+Relatedly, open intergenic chromatin regions, that are not interrupted by \textit{cis-} regulatory elements are the potential sites for \textit{de novo} gene emergence.
+
+A poly-A signal marks the end of a transcript.
+The transcribed DNA region should thus have only one poly-A signal at its end.} A transcribed DNA of length $l$ nucleotides, contains $l-5$ hexamers (6nt long sequences).
+Each of these hexamers can be a poly-A signal with a probability, $P_\textit{polyA}$.
+Probability that none of these hexamers is a poly-A signal is $(1-P_\textit{polyA})^{l-5}$.
 Since we focus on protein coding genes in this work, we describe the length of transcript in terms of the length of protein it can encode.
 Assuming that untranslated regions have no effect on protein synthesis, the length of a transcript in number of nucleotides should be at least three times the length of the encoded protein in number of amino acids.
-Thus the probability that a DNA region produces a transcript that can harbor an ORF with $k$ codons ($P_\textit{RNA}$) is:
+Thus the probability that a DNA region produces a transcript that can harbor an ORF with $k$ codons ($P_\textit{RNA}(k)$) is:
 
 \begin{equation}
-P_\textit{RNA}(k) = P_\textit{polyA} \times (P_\textit{TATA} + P_\textit{Inr}) (1-P_\textit{polyA})^{3k-5}
+P_\textit{RNA}(k) = P_\textit{RNAinit} \times P_\textit{polyA} \times (1-P_\textit{polyA})^{3k-5}
 \label{eqrnaprob}
@@
 
-As we mentioned in the previous section, three requirements need to be met to produce a transcript of any given length.
-First, a promoter needs to be present.
-Second, a poly-A signal needs to be present at the end of the DNA region to be transcribed, and third, no poly-A signal should exist within this DNA region.
-Transcription can be gained if one of two of these requirements are already met and the third required feature emerges due to mutations.
-At the same time mutations do not destroy the existing features.
-It is possible that two or all the required features are missing and they emerge due to mutations.
+\cmnt{As we mentioned in the previous section, two requirements need to be met to produce a transcript of any given length.
+First, a poly-A signal needs to be present at the end of the DNA region to be transcribed, and second, no poly-A signal should exist within this DNA region.
+Transcription can be gained if one of these requirements are already met and the third required feature emerges due to mutations.
+At the same time mutations should not destroy the existing features.
+It is possible that both the required features are missing and they emerge simultaneously due to mutations.
 However such an event is highly improbable.
-Thus transcription gain can occur via three mechanisms.
+Thus transcription gain can occur via two mechanisms.
+In the first mechanism a poly-A signal is gained at the end of a sequence ($P_\textit{polyA-gain}$), while no poly-A sites are present at any $l-5$ (or $3k-5$) sites within the sequence, and none emerge ($P_\textit{nopolyA-stay} = 1 - P_\textit{polyA} - P_\textit{polyA-gain}$).
+In the second mechanism, a poly-A signal is present at the end of the sequence, and is not lost due to mutations.
+Meanwhile, one poly-A signal at any of the $3k-5$ sites within the sequence is lost ($P_\textit{polyA-loss}$), whereas the other $3k-6$ sites do not encode a poly-A signal and continue to remain so.
 
-\vspace{\baselineskip}
-In the first mechanism:
-\begin{itemize}
-\item A promoter (TATA-box or Inr) exists at the beginning of a DNA region and remains intact ($P_\textit{TATA-stay}$, $P_\textit{Inr-stay}$)
-\item No poly-A signal sequences are present within the DNA region and none emerges ($P_\textit{nopolyA-stay} = 1 - P_\textit{polyA} - P_\textit{polyA-gain}$)
-\item A poly-A signal emerges at the end of the DNA region ($P_\textit{polyA-gain}$)
-\end{itemize}
+The combined probability of these two transcription gain mechanism is defined as:
 
-\vspace{\baselineskip}
-In the second mechanism:
-\begin{itemize}
-\item A poly-A signal is present at the end of the DNA region remains intact 
-\item No poly-A signal sequences are present within the DNA region and none emerges
-\item A promoter emerges at the beginning of the DNA region ($P_\textit{TATA-gain}$, $P_\textit{Inr-gain}$)
-\end{itemize}
-
-\vspace{\baselineskip}
-Finally, in the third mechanism:
-\begin{itemize}
-\item A promoter exists at the beginning of a DNA region and remains intact
-\item A poly-A site is present at the end of the DNA region remains intact
-\item One poly-A site that was present in the DNA region, is lost. ($P_\textit{polyA-loss}$)
-\item At every other site in the DNA region, no poly-A signal exists and none emerges
-\end{itemize}
-
-The total gain probability ($P_\textit{RNA-gain}$) of a transcript that can harbor an ORF with $k$ codons  would be the sum of the probabilities of the three gain mechanisms:
-
-\begin{align}
-P_\textit{RNA-gain}(k)  = & \ (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber\\[1ex]
-%
-& + (P_\textit{TATA-gain} + P_\textit{Inr-gain})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \nonumber \\[1ex]
-& + (P_\textit{TATA-stay} + P_\textit{Inr-stay})\times P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
+\begin{equation}
+P_\textit{RNA-gain}(k) = P_\textit{RNAinit} \times \begin{pmatrix}
+P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \\[1em]
+ +\  P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
+\end{pmatrix}  
 \label{eqrnagain}
-\end{align}
-
-It is also possible that both the promoter and the poly-A signal were initially absent, and emerge via mutations. However, this probability is very small and is negligible.
+\end{equation}
 
@@
 
-Transcription is lost when either the promoter is lost or when the poly-A signal is lost.
-Since two kinds of promoters can initiate transcription, there can be three kinds of transcription units.
-First that contains only a TATA-box, second that contains only an Inr, and the third that contains both these promoters.
-All these three transcription units must also have a poly-A signal.
-To calculate the conditional probability of promoter loss, we calculated the fraction of transcription units ($Q$) that belong to each of these three classes.
-For example, the fraction of transcription units that are TATA-only ($Q_\textit{TATA-only}$) is:
+Transcription is lost when either the poly-A signal is lost or if one is gained within the transcribed sequence.
+The probability of transcription loss ($P_\textit{RNA-loss}$), given that already transcription exists, is thus defined as:
 
-\begin{equation*}
-Q_\textit{TATA-only} = \frac{P_\textit{TATA}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})}
-\end{equation*}
-
-\vspace{1ex}
-
-Here, the numerator denotes probability of finding a TATA-box but not an Inr, and the denominator denotes the probability of finding a promoter that can be a TATA-box, Inr, or both.
-Likewise, the fraction of Inr-only ($Q_\textit{Inr-only}$) and TATA-Inr promoters ($Q_\textit{both}$) are given by:
-
-\begin{align*}
-Q_\textit{Inr-only} & = \frac{P_\textit{Inr}\times(1 - P_\textit{Inr})}{(P_\textit{TATA} + P_\textit{Inr} - P_\textit{TATA}\times P_\textit{Inr})} \\[1ex]
-Q_\textit{both} & = 1 - Q_\textit{TATA-only} -Q_\textit{Inr-only}
-\end{align*}
-
-For a TATA-only transcription unit to lose a promoter, it should lose the TATA-box but simultaneously not gain an Inr.
-Likewise, an Inr-only transcription unit will lose a promoter when it loses the Inr but not gain a TATA-box.
-A transcription unit that has both the promoters must lose both of them.
-For all the three kinds of transcription units, transcription would be lost if poly-A signal is lost.
-
-The conditional probability of transcription loss ($P_\textit{RNA-loss}$), given that transcription exists, is thus defined as:
-
-\begin{align}
-P_\textit{RNA-loss}(k) & = Q_\textit{TATA-only} \times P_\textit{TATA-loss}\times(1-P_\textit{Inr-gain}) \nonumber\\[1pt]
- & \quad + Q_\textit{Inr-only}\times P_\textit{Inr-loss}\times(1-P_\textit{TATA-gain}) \nonumber \\[1pt]
- & \quad + Q_\textit{both}\times P_\textit{TATA-loss} \times P_\textit{Inr-loss} + P_\textit{polyA-loss}
+\begin{equation}
+P_\textit{RNA-loss}(k) = P_\textit{polyA-loss} + (3k-5)\times P_\textit{polyA-gain}
 \label{eqrnaloss}
-\end{align}
+\end{equation}
 
@@
 
-Transcription remains intact if neither the promoter, nor the poly-A signal is lost. More specifically, we define this probability ($P_\textit{RNA-stay}$) as:
+Transcription remains intact if neither the poly-A signal is not lost and none is gained within the sequence. Probability of this event ($P_\textit{RNA-stay}$) is defined as:
 
 \begin{equation}
-P_\textit{RNA-stay}(k) = (P_\textit{TATA-stay} + P_\textit{Inr-stay}) \times P_\textit{polyA-stay}
+P_\textit{RNA-stay}(k) = P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
 \label{eqrnastay}
 \end{equation}
-
+}
 \subsection{Interdependence of transcription and ORF probabilities}
@@
 
+\subsection{Calculation of fixation probabilities}
+
+We estimated gene fixation probabilities using Kimura's model \citep{KimuraFix}. Specifically we defined the probabilities of fixation and extinction of a proto-gene as shown in \autoref{pFix}:
+
+\begin{table}[H]
+\centering
+\begin{tabular}{c c c c}
+\toprule
+\textbf{Case} & \textbf{Gain of gene}  & \textbf{Loss of gene} & \textbf{Selection coefficient (s)}\\
+\midrule
+No selection on protogene & $\dfrac{1}{2N_e}$ & $\dfrac{1}{2N_e}$ & -- \\
+\midrule
+Selection on protogene & $\dfrac{1 - e^{-2s}}{1- e^{-4N_es}}$ & $\dfrac{1 - e^{2s}}{1- e^{4N_es}}$ & $\pm0.001$\\
+\bottomrule
+\end{tabular}
+\caption{Gene fixation/extinction probabilities with or without selection.
+Here $N_e$ denotes effective population size and $s$ denotes the selection co-efficient denoting the fitness advantage provided by the protein encoded in the proto-gene (positive valued for a beneficial protein, and negative valued for a deleterious protein).}
+\label{pFix}
+\end{table}
+
 \subsection{Probability of amino acid substitutions}

### 2023-02-24 Modeling/DeNovoProb.jl

@@ -266,7 +266,7 @@ crnagain = crnagain*promprob;
 
 dGC = findall(gcrange .∈ Ref([0.34:0.08:0.5;]));
 cxlen = [30:30:150;]
-klen = ncodons .∈ Ref(cxlen)
+klen = ncodons .∈ Ref(cxlen);
 
 # Gene gain/loss probabilities
 
@@ -321,11 +321,6 @@ onlyorfgain = (1 .- rnaprob .- rnagain).*orfgain;
 
 rnafirst2 = onlyrnagain.*(1 .- rnaloss).*(corfgain./(1 .-orfprob));
 orffirst2 = onlyorfgain.*(1 .- orfloss).*(crnagain./(1 .-rnaprob));
-
-# Exclusive RNA/ORF loss
-
-onlyrnaloss = crnaloss.*corfstay./corfprob;
-onlyorfloss = corfloss.*crnastay./crnaprob;
 
 
 # Plot single step trajectory probability versus ORF length
@@ -397,6 +392,12 @@ plot_whichfirstgcr2 = plot(100*gcrange,log2.(rnafirst2./orffirst2)[:,klen],
 );
 
 savefig(plot_whichfirstgcr2, figdir*"whoisfirstgcr2_promoterless.pdf")
+
+
+# Exclusive RNA/ORF loss
+
+onlyrnaloss = crnaloss.*corfstay./corfprob;
+onlyorfloss = corfloss.*crnastay./crnaprob;
 
 # Plot ORF loss versus RNA loss
 
@@ -590,8 +591,8 @@ savefig(plot_nsites,figdir*"pTotalSites_dmel.pdf")
 # Fixation probabilities #
 
 popsizes = 100 .* 2 .^ [1:10;];
-gainfix = hcat([genegain2X/(2*x) for x in popsizes]...);
-lossfix = hcat([genelossX/(2*x) for x in popsizes]...);
+# gainfix = hcat([genegain2X/(2*x) for x in popsizes]...);
+# lossfix = hcat([genelossX/(2*x) for x in popsizes]...);
 
 selcoef = 0.001
 
@@ -605,14 +606,14 @@ fixprobsPS = log.(gainfixPS)./log.(lossfixPS);
 fixprobsNS = log.(gainfixNS)./log.(lossfixNS);
 
 
-plot_fix = plot(ncodons,fixprobs[:,2:2:10],
-    xlabel = "ORF length (codons)",
[…]

### 2023-02-26 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/SM_deNovoRates.tex 	modified:   Manuscripts/deNovoRates.tex 	modified:   Manuscripts/refs.bib

@@
 \onehalfspacing
+
+\section*{Editor}
+
+Dear Editor,
+
+Enclosed please find a major revision of manuscript number MBE-23-0015, titled ``Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory''.
+We are very grateful to you for co-ordinating the review process.
+We are also thankful to both the reviewers for their detailed analysis of our manuscript, which allowed us to improve our own understanding of our results, and helped us to improve our work.
+We were able to adddress the reviewers' concerns by performing additional analyses of our data, as well as by revising our manuscript.
+
+Before embarking on a point-by-point reply to the reviewers' comments, we summarize the major structural changes we have made in our manuscript:
+
+\begin{itemize}
+\item We have revised our model based on several suggestions by the reviewers. We include the results from our older model in the supplementary material, for the reader's interest. 
+\item We have changed figure 1 to a three panel figure such that it shows probabilities of fixation under different selection conditions.
+\item In response to the reviewers' comments, we have also made several changes to the main text. We have highlighted these changes using a different font color.
+\end{itemize}
+
+We hope that this revision addresses all the concerns and makes our manuscript acceptable for publication. Thank you very much again for considering this manuscript.
+
+Yours sincerely,
+Bharat Ravi and Erich Bornberg-Bauer
 
@@
 
-anna's data (table 4). Can be estimated by total rnagain * total sites * generations per year * \% open chromatin (1-6)* \% open chromatin in intergenic region (35)
+We thank the reviewer for both suggestions, and have tried to address both.
+We have now expanded our discussion to speculate why empirical data may be different from the model's predictions (lines 528 -- 535, 556 -- 566).
+Specifically, we note that real parameters may be different from the ones estimated from currently available data.
+However, we also note that many predictions are not qualitatively affected by small changes in parameter values.
+Furthermore, we now suggest experiments that can provide data for better theoretical predictions (lines 5616 -- 566)
+
+As reviewer suggested we also compared our predictions with existing data.
+A recent study (yet unpublished in a peer-reviewed journal), uses coalescence theory to estimate the rates of transcript gain and loss in geographically isolated populations of \textit{Drosophila melanogaster} (\href{https://doi.org/10.1101/2023.02.13.528330}{Grandchamp et al.
+2023}).
+We compared the estimates of this study with ours and show that the two estimates do not differ by a large margin (lines 237 -- 246).
+Furthermore, we reason why the differences could occur (576 -- 585).
+We now cite the study mentioned by the reviewer (\href{https://doi.org/10.1093/molbev/msab368}{Lange et al., 2021}) to reason how the real effective population size could be different from the previous estimates, and that environmental conditions play a role in gene evolution.
+However, we haven't used this study for calculation of gene gain and loss rates because it does not explicitly report the evolution of new genes.
 
@@
 These observations suggests that polyA signals within the gene body (including ORF) can be disruptive.
-Second, many studies on \textit{de novo} gene discovery to our knowledge, employ a RNAseq library preparation kit that amplifies polyadenylated RNAs (\href{https://doi.org/10.7554/eLife.09977}{Neme and Tautz, 2016}; \href{https://doi.org/10.7554/eLife.47138}{Witt \textit{et al.}, 2019}; \href{https://doi.org/10.1093/molbev/msz300}{Majic and Payne, 2019}; \href{https://doi.org/10.1038/s41467-021-20911-3}{Blevins \textit{et al.}, 2021}).
+More importantly, many studies on \textit{de novo} gene discovery to our knowledge, employ a RNAseq library preparation kit that amplifies polyadenylated RNAs (\href{https://doi.org/10.7554/eLife.09977}{Neme and Tautz, 2016}; \href{https://doi.org/10.7554/eLife.47138}{Witt \textit{et al.}, 2019}; \href{https://doi.org/10.1093/molbev/msz300}{Majic and Payne, 2019}; \href{https://doi.org/10.1038/s41467-021-20911-3}{Blevins \textit{et al.}, 2021}).
 Thus most of the discovered \textit{de novo} genes indeed have a polyA tail (and must also have a polyadenylation signal).
-We now mention both these observations in the main text (lines XXX).
+We now mention this observations in the main text (lines 99 -- 104).
 
 We next focus on sequences that initiate transcription.
-Although several DNA sequence motifs corresponding to enhancers and core promoters have been identified, there is no set of sequence motifs that is deemed to be absolutely necessary.
-The best understood, and the most common core promoter motifs are TATA-box and Inr, which we included in our original analysis.
+Although several DNA sequence motifs corresponding to enhancers and core promoters have been identified, there is no set of sequence motifs that is deemed to be absolutely necessary for this process.
+The best understood, and the most common core promoter motifs are TATA-box and Inr, which we had included in our original analysis.
 We further note that although some sequences are indeed associated with enhancers, the actual functional enhancers are largely determined by epigenetic marks that vary significantly in different cell types, developmental stages, and during different environmental conditions.
@@
 Overall, we surmise that transcription can be initiated at any nucleosome depleted region but the stability of such transcripts is determined by polyA signals.
+We now explain these arguments in lines 95 -- 105.
 
-We assume that such nucleosome free regions are uniformly distributed in open chromatin.
-With the condition that a \textit{de novo} gene emerges in open chromatin regions, we now calculated the probability of transcript emergence based on poly-A signals alone, while ignoring the requirement of a promoter.
+We assume that such nucleosome free regions are uniformly distributed in open chromatin and these regions can initiate transcription at a basal rate.
+To this end, we first estimated the intergenic genome of \textit{D.
+melanogaster} that are in an open chromatin state.
+Next, we calculated the fraction of known cis-regulatory elements (including enhancer) that are present in the open intergenic chromatin regions.
+This fraction is our estimate of basal rate of transcription initiation.
+Using this estimate, we now calculated the probability of transcript emergence based on poly-A signals alone, while ignoring the requirement of a core promoter.
 We first asked if such an assumption would significantly change the results presented in our first draft (Figures 1 -- 3).
 We found that these results do not change significantly.
-The fundamental inferences remain the same.
+The fundamental inferences remain the same (for example, compare figures 1 and S1).
+We now mention the similarities and differences between the promoterless (new) and with-promoter (old) models in the main text (lines 234 -- 236, 276 -- 278, 308 -- 310).
 If we had data on every promoter/enhancer sequence, and if we incorporate them in the model, then the exact RNA probability values should lie in between those estimated by the promoterless model and TATA/Inr-only model.
-In the revised manuscript we now present the results from the promoterless model, and provide a suitable justification in the text (lines XXX).
-We include the old results based on TATA/Inr-only model in the supplementary data (Fig SXX).
+Therefore, in the revised manuscript we now present the results from the promoterless model, and provide a suitable justification in the text (lines 123 -- 126) including a new materials and methods section on how we inferred the general likelihood of transcription initiation (714 -- 733).
+We include the old results based on TATA/Inr-only model in the supplementary data (Fig S1 -- S3).
 
@@
 The reviewer raised an important concern.
-To address this concern, we performed an analysis of total number of available sites and the probability of gene emergence, focusing on the model organism \textit{D.melanogaster}.
-To this end, we first downloaded the genomic coordinates of all \textit{cis}-regulatory motifs from \href{http://redfly.ccr.buffalo.edu/}{REDfly database}.
-Next, we calculated the lengths of intergenic sequences that are not interrupted by any \textit{cis}-regulatory motif, and estimated the total number of sites available for genes of different lengths to emerge.
+To address this concern, we performed an analysis of total number of available sites and the probability of gene emergence, focusing on the model organism \textit{D.
+melanogaster}.
+To this end, we first downloaded the genomic coordinates of all \textit{cis}-regulatory motifs from \href{http://redfly.ccr.buffalo.edu/}{REDfly database} (please also see our reply to the previous comment).
+Next, we calculated the lengths of intergenic sequences that are not interrupted by any \textit{cis}-regulatory motif.
+This would constitute the genomic regions where new genes are free to emerge.
+We also estimated the total number of sites available for genes of different lengths to emerge.
 Finally, we calculated the total gene gain probability by multiplying the total number of feasible sites and per-site gene gain probability.
-We now briefly describe these findings in the main text, and in greater detail in the supplementary data (Figure SXX).
+Using the same principle, we calculated the rate of transcript gain that we compare with the population data from \href{https://doi.org/10.1101/2023.02.13.528330}{Grandchamp et al.
+2023}.
+We now briefly describe these findings in the main text (lines 237 -- 246, 577 -- 585).
 
 We would like to note that in the original analysis, we calculate the probabilities of gene loss and gain, at one genomic locus.
+We have now explicitly mentioned this in the revised manuscript (line 123).
 We also assume that this locus exists in an intergenic region devoid of any functional sequences.
-Therefore the original results will not be dependent on the number of feasible loci (or specific genomes).
-We have now clarified this assumption in the revised manuscript (lines XXX).
+Therefore the original results would not be dependent on the number of feasible loci.
 
@@
 From the frequency distribution of all the 4096 hexamers, we estimated the probability of finding (also gaining and losing) the polyadenylation signal sequences (that are 6 nucleotides long).
-Similarly, we calculated the probabilities of start and stop codons, using a distribution of trimers estimated from the \textit{D.melanogaster} intergenic regions.
+Similarly, we calculated the probabilities of start and stop codons, using a distribution of trimers estimated from the \textit{D.melanogaster} intergenic regions (lines 604 -- 607).
 With these probabilities, we repeated the analyses shown in the original figures 1A, 2A, 2C and 3A.
 
-We found that the the results from actual nucleotide distributions indeed differed from the ones obtained using total GC content of \textit{D.melanogaster} intergenic genome (0.403).
-We now include these results in the main manuscript along with the old figures.
-We also explain the details of the re-analysis and its rationale (as described in this response), in the main text.
-(lines XXX)
+We found that the the results from actual nucleotide distributions indeed differed from the ones obtained using total GC content of \textit{D.melanogaster} intergenic genome (0.41).
+We now include these results in the main manuscript with the predictions using GC-content.
+We also explain the details of the re-analysis and its rationale (as described in this response), at several places in the main text (lines 197 -- 206, 600 -- 603).
+We also use the oligomer based nucleotide frequency distributions to calculate the fixation probabilities and rate of gene/transcript gain.
 
@@
 
-I think that if the authors would follow at least some of these recommendations then they would be walking on much more solid ground and they would not need to use the ``all models are wrong'' quote, in favor of a more grounded implications of their findings.
-This would also mean that we would count with a much more solid point of argument for the discussion of the likelihood of de novo gene emergence, a field which tends to be too empirical and with little theory, which is why this paper would be a good addition to the literature if improvements are made.
+This is indeed a very useful suggestion.
+We now calculate the probability of fixation as well as the probability of extinction under the assumptions of (1) total neutrality, (2) weak positive selection, and (3) weak negative selection.
+We include these analyses in the new figure 1, and describe the rationale (lines 207 -- 232) and the method (lines 800 -- 802, new table 2) in the main text.
+
+{\itshape I think that if the authors would follow at least some of these recommendations then they would be walking on much more solid ground and they would not need to use the ``all models are wrong'' quote, in favor of a more grounded implications of their findings.
+This would also mean that we would count with a much more solid point of argument for the discussion of the likelihood of de novo gene emergence, a field which tends to be too empirical and with little theory, which is why this paper would be a good addition to the literature if improvements are made.}
+
+We thank the reviewer again for all their comments, and hope that we have addressed all their concerns satisfactorily.
 
@@
 
-Comments to the Author
+{\itshape Comments to the Author
 This paper describes a model to calculate the mutational probabilities of de novo gene origin and loss.
 Its conclusions are that certain modes of gene origin are more probable than others.
-I am no expert in these kinds of models or in de novo gene emergence, but I found these arguments interesting and valuable.
+I am no expert in these kinds of models or in de novo gene emergence, but I found these arguments interesting and valuable.}
 
-I have two major comments about the paper.
+We thank the reviewer for their comments and suggestions.
+
+{\itshape I have two major comments about the paper.
 The first is that I was surprised to see no arguments or formalism rooted in population genetics in this paper.
@@
 And while fixation probabilities only depend on the mutation rate, I think the time to fixation under neutrality does depend on the effective population size.
-So the authors' model as it stands does not make predictions about how long it takes for de novo genes to fix or be lost (though this extension may be easy to carry out).
+So the authors' model as it stands does not make predictions about how long it takes for de novo genes to fix or be lost (though this extension may be easy to carry out).}
 
-Another difficulty for me was the argument about whether the existence of a lineage specific gene really implies it evolved only in that lineage.
+We thank the reviewer for this important suggestion.
+We now calculate the probability of fixation as well as the probability of extinction under the assumptions of (1) total neutrality, (2) weak positive selection, and (3) weak negative selection.
+We include these analyses in the new figure 1, and describe the rationale (lines 207 -- 232) and the method (lines 796 -- 798, new table 2) in the main text.
+
+We now explicitly state that fixation under total neutrality depends solely on the mutation rate (lines 221 -- 222).
+We have also mentioned that time to fixation would be much smaller than time to gain a gene (lines 230 -- 232)
+
+{\itshape Another difficulty for me was the argument about whether the existence of a lineage specific gene really implies it evolved only in that lineage.
 The alternative being an earlier gain with losses in all lineages but one.
@@
 It would seem to me that this is vastly less likely under strict neutrality than a lineage specific gain.
-I may of course be wrong, but I think a very simplistic argument based on only two probabilities with no treatment of the actual phylogeny is not sufficient to convince me of the authors' argument.
+I may of course be wrong, but I think a very simplistic argument based on only two probabilities with no treatment of the actual phylogeny is not sufficient to convince me of the authors' argument.}
 
-Another minor comment from me is that in the discussion it might be valuable to address that at least some de novo genes become essential quickly (I would for example point to work by the Long lab, which is cited already).
-This would allow purifying selection to quickly preserve such genes against losses.
+We agree the reviewer's point that phylogenetic inferences should take into consideration branch lengths.
+To clarify our analyses we have now made explicit that we talk about populations (lines 148 -- 154) and not species that could have diverged over many millions of years.
+Thus our model is suitable for relatively shorter timescales.
+Furthermore, multiple gene gain and loss events within a lineage or branch would be highly rare, given the low probability of these events.
 
-Lastly, I couldn't find the actual parameter values used for the various probabilities in the paper.
-I know the authors will make the code available, but for the casual reader a table would be valuable.
+Thus given an observation that a population has a gene but two other populations do not have it, there can be two assumptions:
+
+1. Gene is gained in one population
+2. Gene was present in the ancestor and lost in two populations
+
+The parsimonious assumption would be that gene is only gained in one population.
+What we show is that the gene \textbf{can} be lost in two populations independently.
+It is indeed possible that these two populations in turn share a common ancestor that had already lost the gene.
+This event is more probable than two gene losses, which in turn strengthens our argument that gene losses are much more frequent than gene gain.
+We just show how frequent they are.
+We hope we have clarified the reviewer's concern.
+
+{\itshape Another minor comment from me is that in the discussion it might be valuable to address that at least some de novo genes become essential quickly (I would for example point to work by the Long lab, which is cited already).
+This would allow purifying selection to quickly preserve such genes against losses.}
+
+We thank the reviewer for their suggestion and cite the relevant work in lines 470 -- 471.
+
+{\itshape Lastly, I couldn't find the actual parameter values used for the various probabilities in the paper.
+I know the authors will make the code available, but for the casual reader a table would be valuable.}
+
+Our primary parameters are mutation rate and mutation rate bias (shown in Table 1).
+All other parameters are derived and we therefore do not mention their values explicitly.
+Furthermore, the derived parameters are complex functions of gene length and nucleotide composition, and therefore, vary significantly.
+We now include the frequencies of trimers and hexamers from \textit{D.
+melanogaster} intergenic regions as supplementary files.
 

### 2023-02-26 Manuscripts/SM_deNovoRates.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/SM_deNovoRates.tex 	modified:   Manuscripts/deNovoRates.tex 	modified:   Manuscripts/refs.bib

@@
 
-\section{Effect on GC content on the probability of gene gain and loss}
+\clearpage
+
+\section{Model without core promoters}
+
+\subsection{Effect on GC content on the probability of gene gain and loss}
 
@@
 
-\section{Exclusive ORF or transcription loss}
+\subsection{Exclusive ORF or transcription loss}
 
@@
 \caption{Exclusive ORF loss in proto-genes is more probable than exclusive transcription loss.
-The horizontal axis denotes the number of codons in the ORF \textbf{(A)} and GC-content \textbf{(B)}.
-The vertical axis shows the $\log_2$ transformed ratio of only ORF loss and only RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.}
+The vertical axis shows the $\log_2$ transformed ratio of only ORF loss and only RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.
+The horizontal axis denotes the number of codons in the ORF.
+Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
+melanogaster} intergenic genome.}
 \end{figure}
+
+\clearpage
 

### 2023-02-26 Manuscripts/deNovoRates.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/SM_deNovoRates.tex 	modified:   Manuscripts/deNovoRates.tex 	modified:   Manuscripts/refs.bib

@@
 \usepackage[margin=2.5cm]{geometry}
-\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec}
+\usepackage{amsmath,amssymb,graphicx,setspace,palatino,float,subcaption, booktabs, titlesec,lineno}
 \usepackage[colorlinks, allcolors = blue]{hyperref}
@@
 
-\begin{abstract}
+\linenumbers
+
+\section*{Abstract}
 New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence.
@@
 Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.
-\end{abstract}
 
@@
 
-$$ \text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$$
+\begin{equation}
+\text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})
+\label{loglogratio}
+\end{equation}
 
@@
 \caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$.
-The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.
-lack lines denote the probability values estimated from overall GC content and the orange lined denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
-melanogaster} intergenic genome.}
+The vertical axis in all panels shows the number of independent gene loss events with extinction (\autoref{genelosseq}) that can occur relative to one gene gain event with fixation (\autoref{loglogratio}, \autoref{pFix}), under \textbf{(A)} no selection, \textbf{(B)} weak positive selection and \textbf{(C)} weak negative selection on the expression of the protein from a proto-gene.
+Horizontal axes show the number of codons in the ORF.
+Black lines in panel \textbf{(A)} denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
+melanogaster} intergenic genome.
+In panels \textbf{(B)} and \textbf{(C)} we use the probabilitiy values from \textit{D.
+melanogaster} intergenic genome.
+$N_e$ denotes effective population size and $s$ denotes the selection co-efficient.}
 \label{gainlossprob}
@@
 To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals.
-First is the probability that any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$).
-The second probability describes the event that all individuals in a population have two copies of the proto-gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$).
+First is the probability that any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$, \autoref{pFix}).
+The second probability describes the event that all individuals in a population have two copies of the proto-gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$, \autoref{pFix}).
 We calculated the log-log ratio of the two probabilities under three scenarios (\autoref{pFix}), wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$).
 We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously.
-Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele.
-Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF.
+Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
+Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
 For example, a gene of any length is less likely to go extinct than it is fixed, in populations with at least 6400 individuals.
 Furthermore, a beneficial gene can be independently lost in more than two small populations with 400 individuals, only when it has more than 133 codons.
-When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population.}
+When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population (\hyperref[gainlossprob]{\autoref{gainlossprob}C}).
+We further note that time to fixation of a rare allele (for example, a newly emerged protogene) would be much smaller \citep[$4N_e$ generations;][]{KimuraFix} than the time to gain ($10^{14}$ -- $10^{20}$ generations) or lose it ($10^{7}$ -- $10^{8}$ generations).}
 
 Overall, our analysis suggests that a proto-gene expressed in only one population may not necessarily mean that it emerged for the first time in this population.
+\cmnt{This is also true for a model of gene emergence that requires a core promoter (TATA-box or Inr) for initiating transcription ({\color{blue} Figure S1}).
 
@@
 This estimate is slightly higher than the rate of transcript gain predicted from \textit{D.
-melanogaster} population transcriptomics data \citep[0.13 -- 0.34;][]{AnnaPeteTranscripts}.
+melanogaster} population transcriptomics data \citep[0.13 -- 0.34;][]{AnnaPeteTranscripts}}.
 
@@
 A positive value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
-In both panels, the horizontal axis denotes the number of codons in the ORF.}
+In both panels, the horizontal axis denotes the number of codons in the ORF.
+Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
+melanogaster} intergenic genome.}
 \label{whoisfirst}
@@
 We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
-In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Initiator element) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first ({\color{blue}Figure S2}).
 More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length.
@@
 melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length.
-Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.}
+Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.
+In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Inr) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first ({\color{blue}Figure S2})}
 
@@
 The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.
-Black lines denote the probability values estimated from overall GC content and the orange lined denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
+Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
 melanogaster} intergenic genome.}
@@
 
-We show that genes can be lost much more rapidly than they are gained, such that a gene can be independently lost two to three times in the time required for it to emerge.
+We show that when a protein product of a proto-gene doesn't affect organismal fitness, the gene can be lost much more rapidly than they are gained, such that it can be independently lost two to three times in the time required for it to emerge (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
 In a hypothetical case, a proto-gene may be present in a population (A) but is absent in two other geographically isolated populations (B and C).
@@
 Future phylogenetic studies on proto-genes should consider this finding for inferring the dynamics of gene gain and loss.
+\cmnt{Proto-genes that encode proteins that are beneficial to an organism's fitness are likely to be protected in the genome against gene loss \citep{Lee2019}.
+Our models indeed show that loss of even moderately beneficial genes is unlikely in large populations (\hyperref[gainlossprob]{\autoref{gainlossprob}B})}.
 
@@
 That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
-We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first).
+We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first, \autoref{whoisfirst}).
 Long non-coding RNAs may be a potential source of proto-genes but this evolutionary process may be constrained by a different set of factors.
@@
 
-We show that transcription loss is less likely than the loss of long ORFs.
+We show that transcription loss is less likely than the loss of long ORFs (\autoref{lossprob}).
 However, it may be difficult to find direct evidence for such a phenomenon.
@@
 Second, in our models, we assumed uniform mutation rate throughout the genome.
-However, recent studies have shown Mutation rates are also not uniform across the genome and can vary significantly \citep{mutbiasArabidopsis}}.
+However, recent studies have shown that mutation rates are also not uniform across the genome, and can vary significantly \citep{mutbiasArabidopsis}}.
 Our models can be adapted to investigate specific genomic loci, where estimates of the local nucleotide composition and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
@@
 melanogaster} populations \citep{AnnaPeteTranscripts}.
-This inflated estimate could have possibly resulted from our relaxed requirements for a promoter, or non-inclusion of other unknown factors that may have affected population evolution.
+This difference could have possibly resulted from our relaxed requirements for a promoter, or other unknown factors that may have affected population evolution \citep[for example, the actual effective population size and environmental conditions;][]{dmel3decades}.
 When supplemented with experimental data analysis our models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution, and can provide more accurate predictions.} Therefore our work opens up an opportunity for theoretical and computational biologists to design analyses using our modeling framework, that more accurately describe their system of interest.
@@
 
-We estimated gene fixation probabilities using Kimura's model \citep{KimuraFix}. Specifically we defined the probabilities of fixation and extinction of a proto-gene as shown in \autoref{pFix}:
+\newcommand{\specialcell}[2][c]{%
+  \bfseries\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+\cmnt{
+We estimated gene fixation probabilities using Kimura's model \citep{KimuraFix}. Specifically we defined the probabilities of fixation and extinction of a proto-gene as shown in \autoref{pFix}:}
 
@@
 \toprule
-\textbf{Case} & \textbf{Gain of gene}  & \textbf{Loss of gene} & \textbf{Selection coefficient (s)}\\
+\textbf{Case} & \specialcell{Gain of gene\\(Fixation, $P_\textit{fix}$)}  & \specialcell{Loss of gene\\ (Extinction, $P_\textit{ext}$)} & \specialcell{Selection\\coefficient ($s$)}\\
 \midrule

### 2023-02-27 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

@@
 
-We agree the reviewer's point that phylogenetic inferences should take into consideration branch lengths.
-To clarify our analyses we have now made explicit that we talk about populations (lines 148 -- 154) and not species that could have diverged over many millions of years.
+We are not sure if we understood the reviewer's point completely.
+We do agree that phylogenetic inferences should take into consideration branch lengths.
+We assume that branch lengths should have minimal effect when comparing gene gains and losses in populations that haven't diverged from each other over a long time.
+We now explicitly say that we analyse populations (lines 148 -- 154) and not species that could have diverged over many millions of years.
 Thus our model is suitable for relatively shorter timescales.
@@
 It is indeed possible that these two populations in turn share a common ancestor that had already lost the gene.
-This event is more probable than two gene losses, which in turn strengthens our argument that gene losses are much more frequent than gene gain.
-We just show how frequent they are.
+This event is more probable than two gene losses.
+This in turn strengthens our argument that gene losses are much more frequent than gene gain.
+Multiple independent losses is an extreme case that may be less feasible than a single loss followed by population divergence.
+Nevertheless it is possible, and can be more likely than a single gene gain event.
 We hope we have clarified the reviewer's concern.

### 2023-02-27 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

@@
 P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \\[1em]
- +\  P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
+ +\  P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-6} \times (3k-6) \times P_\textit{polyA-loss}
 \end{pmatrix}  
@@
 \begin{equation}
-P_\textit{RNA-stay}(k) = P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
+P_\textit{RNA-stay}(k) = P_\textit{RNAinit} \times P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
 \label{eqrnastay}

### 2023-03-01 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/deNovoRates.tex

@@
 \end{itemize}
+
+Our manuscript is longer than the optimal 20 pages.
+This is primarily because we include a detailed methods section which we believe is essential for a reader to understand the results.
+Methods are especially important for theoretical papers that are not based on standard protocols and analytical pipeline.
+Thus we request you to allow us keep the methods in the main text and not in the supplementary material.
 
@@
 To this end, we first estimated the intergenic genome of \textit{D.
-melanogaster} that are in an open chromatin state.
+melanogaster} that are in an open chromatin state (\href{https://doi.org/10.1186/1471-2164-15-988}{Milon et al., 2014}).
 Next, we calculated the fraction of known cis-regulatory elements (including enhancer) that are present in the open intergenic chromatin regions.
@@
 This event is more probable than two gene losses.
-This in turn strengthens our argument that gene losses are much more frequent than gene gain.
 Multiple independent losses is an extreme case that may be less feasible than a single loss followed by population divergence.
@@
 
+We will indeed make the code available when the work is published. Currently it is not accessible to the public because we use the same github repository to also keep track of the manuscript.
+
 \end{document}

### 2023-03-01 Manuscripts/deNovoRates.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/deNovoRates.tex

@@
 
+\renewcommand{\subsectionautorefname}{Methods section}
+\renewcommand{\subsubsectionautorefname}{Methods section}
+\renewcommand{\thesubsection}{\arabic{subsection}}
+
 \newcommand{\cmnt}[1]{{\color{purple} #1}}
@@
 \title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
-\author{Bharat Ravi Iyengar$^1$, Erich Bornberg-Bauer$^{1,2}$}
-\date{\small $^1$Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany\\ $^2$Max Planck-Institute for Biology T\"{u}bingen, T\"{u}bingen, Germany}
+\author{Bharat Ravi Iyengar$^{1,\dagger}$, Erich Bornberg-Bauer$^{1,2}$}
+\date{\small $^1$Institute for Evolution and Biodiversity, University of M\"{u}nster,\\ H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany\\[1ex] $^2$Department of Protein Evolution, Max Planck Institute for Biology T\"{u}bingen, Max-Planck-Ring 5, 72076 T\"{u}bingen, Germany \\[1ex] $^\dagger$ Corresponding author: b.ravi@uni-muenster.de}
 
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\section{Results}
+\section*{Results}
 
@@
 
-\subsection{How likely is gene loss?}
+\subsection*{How likely is gene loss?}
 
@@
 
-\subsection{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}
+\subsection*{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}
 
@@
 
-\subsection{Would extensive transcription loss suggest negative selection of toxic proteins?}
+\subsection*{Would extensive transcription loss suggest negative selection of toxic proteins?}
 
@@
 
-\subsection{Does mutation bias shape protein composition?}
+\subsection*{Does mutation bias shape protein composition?}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\section{Materials and Methods}
+\section*{Materials and Methods}
 
@@
 
-\section{Data availability}
+\section*{Data availability}
 We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
 Specifically, our model is implemented in the script \texttt{DeNovoProb.jl} which in turn uses the script \texttt{nucleotidefuncts.jl} for some basic functions.
+
+\section*{Acknowledgments}
+
+We thank Pete Czuppon (IEB, University of M\"{u}nster), for discussions on mathematical models.
 

### 2023-03-02 Modeling/antisenseGenes.jl
modified:   Modeling/antisenseGenes.jl

@@ -15,8 +15,38 @@ p_A = 0.256362;
 p_G = 0.267809;
 p_C = 0.271612;
 
+nfmat = zeros(4,1);
+
+nfmat[nucnames['A']] = 0.256362;
+nfmat[nucnames['T']] = 0.204217;
+nfmat[nucnames['G']] = 0.267809;
+nfmat[nucnames['A']] = 0.271612;
+
+function nucprobss(s,nfmat)
+    return prod([nfmat[nucnames[x]] for x in s])
+end
+
 allcodons = kmers(3)
 nostopcodons = nostop(allcodons);
+
+noATG = setdiff(allcodons, ["ATG"]);
+function ATGprobs(gccontent)
+    ATGprob = nucprob("ATG",gccontent)
+    ATGgain = featuregain(noATG,["ATG"],gccontent)
+    ATGloss = featuregain(["ATG"],noATG,gccontent)/ATGprob
+    ATGstay = featurestay(["ATG"],gccontent)
+    return [ATGprob, ATGgain, ATGloss, ATGstay]
+end
+
+## Stop codon
+
+function stopprobs(gccontent)
+    stopprob = sum([nucprob(x,gccontent) for x in stopvars])
+    stopgain = featuregain(nostopcodons,stopvars,gccontent)
+    stoploss = featuregain(stopvars,nostop,gccontent)/stopprob
+    stopstay = featurestay(stopvars,gccontent)
+    return [stopprob, stopgain, stoploss, stopstay]
+end
 
 syncodons = Dict(x => gencode[(gencode[:,3].== transnucs(x)) .& (gencode[:,1] .!= x),1] for x in nostopcodons)
 
@@ -38,8 +68,10 @@ end
 vjoin = (z) -> [join(x) for x in eachrow(z)]
 phcp = (x,y) -> permutedims(hcat(collect.(product(x,y))...))
 
-p_synsubs = [featuregain([x],syncodons[x],gccontent) for x in keys(syncodons)]
-p_simsubs = [featuregain([x],simcodons[x],gccontent) for x in keys(simcodons)]
+
+
+# p_synsubs = [featuregain([x],syncodons[x],gccontent) for x in keys(syncodons)]
+# p_simsubs = [featuregain([x],simcodons[x],gccontent) for x in keys(simcodons)]
 
 # Codon pairs that have a stop codon in frame -2
 # Example: 
[…]

### 2023-03-02 Modeling/nucleotidefuncts.jl

@@ -105,22 +105,40 @@ end
 
 
 
-nsub = zeros(4,4)
+nsub = zeros(4,4);
 
-nsub[1,2] = 0.056043956043956
-nsub[1,3] = 0.242857142857143
-nsub[1,4] = 0.074725274725275
-nsub[3,1] = 0.438461538461539
-nsub[3,2] = 0.075164835164835
-nsub[3,4] = 1 - sum(nsub[1,:]) - sum(nsub[3,1:2])
-nsub[2,1] = nsub[1,2]
-nsub[2,3] = nsub[1,4]
-nsub[2,4] = nsub[1,3]
-nsub[4,3] = nsub[3,4]
-nsub[4,2] = nsub[3,1]
-nsub[4,1] = nsub[3,2]
+nsub[1,2] = 0.056043956043956;
+nsub[1,3] = 0.242857142857143;
+nsub[1,4] = 0.074725274725275;
+nsub[3,1] = 0.438461538461539;
+nsub[3,2] = 0.075164835164835;
+nsub[3,4] = 1 - sum(nsub[1,:]) - sum(nsub[3,1:2]);
+nsub[2,1] = nsub[1,2];
+nsub[2,3] = nsub[1,4];
+nsub[2,4] = nsub[1,3];
+nsub[4,3] = nsub[3,4];
+nsub[4,2] = nsub[3,1];
+nsub[4,1] = nsub[3,2];
 
 mutrate = 7.8e-9
+
+
+nsubbac = zeros(4,4)
+
+nsubbac[1,2] = 0.0031813361611877;
+nsubbac[1,3] = 0.0095440084835631;
+nsubbac[1,4] = 0.0031813361611877;
+nsubbac[3,1] = 0.0689289501590668;
+nsubbac[3,2] = 0.907741251325557;
+nsubbac[3,4] = 1 - sum(nsubbac[1,:]) - sum(nsubbac[3,1:2]);
+nsubbac[2,1] = nsubbac[1,2];
+nsubbac[2,3] = nsubbac[1,4];
+nsubbac[2,4] = nsubbac[1,3];
+nsubbac[4,3] = nsubbac[3,4];
+nsubbac[4,2] = nsubbac[3,1];
+nsubbac[4,1] = nsubbac[3,2];
+
+mutratebac = 2e-8;
 
 function µA(mutrate,nsub)
     return 2*mutrate*sum(nsub[1,:])

### 2023-03-03 Modeling/antisenseGenes.jl
new file:   Manuscripts/Figures/pORFgain_antisense.svg 	new file:   Manuscripts/Figures/pORFloss_antisense.svg 	modified:   Modeling/antisenseGenes.jl

@@ -1,14 +1,24 @@
 using Plots
+using Measures
 cd(Base.source_dir())
 include("nucleotidefuncts.jl")
-stop = ["TAA","TAG","TGA"]
 
+cm2pt = (cm) -> 28.3465*cm
+figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
+colors = ["#FFCC00","#5599FF","#D40000","#000000"];
+
+default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
+guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
+
+stopcodons = ["TAA","TAG","TGA"]
 function nostop(codonset)
-	codonset[codonset .∉ Ref(stop)]
+	codonset[codonset .∉ Ref(stopcodons)]
 end
 
 gccontent = 0.493611; # all mRNA #
 # gccontent = 0.53942 # all CDS #
+
+(xProb, xGain, xLoss, xStay) = [i for i in 1:4];
 
 p_T = 0.204217;
 p_A = 0.256362;
@@ -41,10 +51,10 @@ end
 ## Stop codon
 
 function stopprobs(gccontent)
-    stopprob = sum([nucprob(x,gccontent) for x in stopvars])
-    stopgain = featuregain(nostopcodons,stopvars,gccontent)
-    stoploss = featuregain(stopvars,nostop,gccontent)/stopprob
-    stopstay = featurestay(stopvars,gccontent)
+    stopprob = sum([nucprob(x,gccontent) for x in stopcodons])
+    stopgain = featuregain(nostopcodons,stopcodons,gccontent)
+    stoploss = featuregain(stopcodons,nostopcodons,gccontent)/stopprob
+    stopstay = featurestay(stopcodons,gccontent)
     return [stopprob, stopgain, stoploss, stopstay]
 end
 
@@ -84,13 +94,13 @@ phcp = (x,y) -> permutedims(hcat(collect.(product(x,y))...))
 #   AGT
 
 
-stopneighborsR = [unique(vcat(frameXcpairs.(stop,-x)...), dims = 1) for x in 1:2];
+stopneighborsR = [unique(vcat(frameXcpairs.(stopcodons,-x)...), dims = 1) for x in 1:2];
 
 # Codon pairs inside an ORF, that have a stop codon in frame -x
[…]

### 2023-03-06 Modeling/antisenseGenes_supplement.jl
new file:   Modeling/antisenseGenes_supplement.jl

@@ -1,0 +1,416 @@
+using DelimitedFiles
+data, header = readdlm(joinpath(Base.source_dir(),"codons.txt"), '\t', header=true);
+header = replace.(header, r"[^A-Z,a-z]" => "");
+mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
+
+nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
+
+stopcodons = ["TAA","TAG","TGA"]
+function nostop(codonset)
+	codonset[codonset .∉ Ref(stopcodons)]
+end
+
+gccontent = 0.493611; # all mRNA #
+# gccontent = 0.53942 # all CDS #
+
+(xProb, xGain, xLoss, xStay) = [i for i in 1:4];
+
+p_T = 0.204217;
+p_A = 0.256362;
+p_G = 0.267809;
+p_C = 0.271612;
+
+nfmat = zeros(4,1);
+
+nfmat[nucnames['A']] = 0.256362;
+nfmat[nucnames['T']] = 0.204217;
+nfmat[nucnames['G']] = 0.267809;
+nfmat[nucnames['A']] = 0.271612;
+
+function nucprobss(s,nfmat)
+    return prod([nfmat[nucnames[x]] for x in s])
+end
+
+allcodons = kmers(3)
+nostopcodons = nostop(allcodons);
+
+noATG = setdiff(allcodons, ["ATG"]);
+function ATGprobs(gccontent)
+    ATGprob = nucprob("ATG",gccontent)
+    ATGgain = featuregain(noATG,["ATG"],gccontent)
+    ATGloss = featuregain(["ATG"],noATG,gccontent)/ATGprob
+    ATGstay = featurestay(["ATG"],gccontent)
+    return [ATGprob, ATGgain, ATGloss, ATGstay]
+end
+
+## Stop codon
+
+function stopprobs(gccontent)
+    stopprob = sum([nucprob(x,gccontent) for x in stopcodons])
+    stopgain = featuregain(nostopcodons,stopcodons,gccontent)
+    stoploss = featuregain(stopcodons,nostopcodons,gccontent)/stopprob
+    stopstay = featurestay(stopcodons,gccontent)
+    return [stopprob, stopgain, stoploss, stopstay]
+end
+
+syncodons = Dict(x => gencode[(gencode[:,3].== transnucs(x)) .& (gencode[:,1] .!= x),1] for x in nostopcodons)
+
[…]

### 2023-03-06 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl 	modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl 	modified:   Modeling/nucleotidefuncts.jl

@@ -2,6 +2,7 @@ using Combinatorics
 using Plots
 using Measures
 # using HypothesisTests
+using DataFrames
 
 include("nucleotidefuncts.jl")
 
@@ -18,6 +19,61 @@ gccontent = 0.41
 (xProb, xGain, xLoss, xStay) = [i for i in 1:4];
 
 mers6 = kmers(6);
+
+# Data based calculations #
+
+data, header = readdlm(joinpath(Base.source_dir(),"hexamers2.txt"), '\t', header=true);
+header = replace.(header, r"[^A-Z,a-z]" => "");
+mers6freq = DataFrame(data[:,[1,3]], header[[1,3]]);
+
+data, header = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t', header=true);
+header = replace.(header, r"[^A-Z,a-z]" => "");
+mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
+
+nprob6 = x -> mers6freq.Frequency[mers6freq.kmer .== x][1]
+nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
+
+function featuregain3(set1,set2)
+    if(isempty(set1) || isempty(set2))
+        @warn "empty feature sets, returning 0"
+        return 0
+    end
+    v1 = sum([nprob3(s1)*sum([mprob(s1,s2) for s2 in set2])
+                for s1 in set1])
+    return v1
+end
+
+function featurestay3(set1)
+    v =  sum(
+        [nprob3(s1)*(
+            sum([mprob(s1,s2) for s2 in set1]) +
+            (1-µA(mutrate,nsub))^numAT(s1)*
+            (1-µG(mutrate,nsub))^(3-numAT(s1))
+            ) for s1 in set1]
+        )
+    return v
+end
+
+function featuregain6(set1,set2)
+    if(isempty(set1) || isempty(set2))
+        @warn "empty feature sets, returning 0"
+        return 0
+    end
+    v1 = sum([nprob6(s1)*sum([mprob(s1,s2) for s2 in set2])
+                for s1 in set1])
+    return v1
+end
+
+function featurestay6(set1)
+    v =  sum(
+        [nprob6(s1)*(
+            sum([mprob(s1,s2) for s2 in set1]) +
+            (1-µA(mutrate,nsub))^numAT(s1)*
+            (1-µG(mutrate,nsub))^(6-numAT(s1))
[…]

### 2023-03-06 Modeling/antisenseGenes.jl

@@ -10,12 +10,6 @@ colors = ["#FFCC00","#5599FF","#D40000","#000000"];
 default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
 guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
 
-data, header = readdlm(joinpath(Base.source_dir(),"trimers1CDS.txt"), '\t', header=true);
-header = replace.(header, r"[^A-Z,a-z]" => "");
-mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
-
-nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
-
 stopcodons = ["TAA","TAG","TGA"]
 function nostop(codonset)
 	codonset[codonset .∉ Ref(stopcodons)]
@@ -253,6 +247,8 @@ function orfprobs(ATG,stop,k)
     orfstay = ATG[xStay]*stop[xStay]*(nostopstay)^(k-2)
     return [orfprob; orfgain; orfloss; orfstay]
 end
+
+include("antisenseGenes_supplement.jl")
 
 ncodons = [30:300;];
 gcrange = [0.3:0.05:0.6;];

### 2023-03-06 Modeling/antisenseGenes_supplement.jl

@@ -1,134 +1,111 @@
 using DelimitedFiles
-data, header = readdlm(joinpath(Base.source_dir(),"codons.txt"), '\t', header=true);
-header = replace.(header, r"[^A-Z,a-z]" => "");
-mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
 
-nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
+normalize = x -> x/sum(x)
 
-stopcodons = ["TAA","TAG","TGA"]
-function nostop(codonset)
-	codonset[codonset .∉ Ref(stopcodons)]
+codonfreq = readdlm(joinpath(Base.source_dir(),"codons.txt"), '\t');
+codonfreq[codonfreq[:,1] .∈ Ref(stopcodons),2] .= 0;
+codonfreq[:,2] = codonfreq[:,2]/sum(codonfreq[:,2]);
+
+
+dcfreqdata = readdlm(joinpath(Base.source_dir(),"dicodons.txt"), '\t');
+cc = z -> [x[z:z+2] for x in dcfreqdata[:,1]]
+
+dcfreqdata[(cc(1) .∈ Ref(stopcodons)) .| (cc(3) .∈ Ref(stopcodons)) ,2] .= 0;
+
+dicodonfreq = hcat(kmers(6),zeros(4^6,1));
+for x in eachrow(dcfreqdata)
+    dicodonfreq[dicodonfreq[:,1] .== x[1],2] .= x[2];
 end
 
-gccontent = 0.493611; # all mRNA #
-# gccontent = 0.53942 # all CDS #
+dicodonfreq[:,2] = normalize(dicodonfreq[:,2]);
 
-(xProb, xGain, xLoss, xStay) = [i for i in 1:4];
+nprob3 = x -> codonfreq[codonfreq[:,1] .== x,2]
+nprob6 = x -> dicodonfreq[dicodonfreq[:,1] .== x,2]
 
-p_T = 0.204217;
-p_A = 0.256362;
-p_G = 0.267809;
-p_C = 0.271612;
-
-nfmat = zeros(4,1);
-
-nfmat[nucnames['A']] = 0.256362;
-nfmat[nucnames['T']] = 0.204217;
-nfmat[nucnames['G']] = 0.267809;
-nfmat[nucnames['A']] = 0.271612;
-
-function nucprobss(s,nfmat)
-    return prod([nfmat[nucnames[x]] for x in s])
+function featuregain3(set1,set2)
[…]

### 2023-03-06 Modeling/nucleotidefuncts.jl

@@ -227,59 +227,3 @@ function featurestay(set1,gccontent)
         )
     return v
 end
-
-using DataFrames
-# Data based calculations #
-
-data, header = readdlm(joinpath(Base.source_dir(),"hexamers2.txt"), '\t', header=true);
-header = replace.(header, r"[^A-Z,a-z]" => "");
-mers6freq = DataFrame(data[:,[1,3]], header[[1,3]]);
-
-data, header = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t', header=true);
-header = replace.(header, r"[^A-Z,a-z]" => "");
-mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
-
-nprob6 = x -> mers6freq.Frequency[mers6freq.kmer .== x][1]
-nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
-
-function featuregain3(set1,set2)
-    if(isempty(set1) || isempty(set2))
-        @warn "empty feature sets, returning 0"
-        return 0
-    end
-    v1 = sum([nprob3(s1)*sum([mprob(s1,s2) for s2 in set2])
-                for s1 in set1])
-    return v1
-end
-
-function featurestay3(set1)
-    v =  sum(
-        [nprob3(s1)*(
-            sum([mprob(s1,s2) for s2 in set1]) +
-            (1-µA(mutrate,nsub))^numAT(s1)*
-            (1-µG(mutrate,nsub))^(3-numAT(s1))
-            ) for s1 in set1]
-        )
-    return v
-end
-
-function featuregain6(set1,set2)
-    if(isempty(set1) || isempty(set2))
-        @warn "empty feature sets, returning 0"
-        return 0
-    end
-    v1 = sum([nprob6(s1)*sum([mprob(s1,s2) for s2 in set2])
-                for s1 in set1])
-    return v1
-end
-
-function featurestay6(set1)
-    v =  sum(
-        [nprob6(s1)*(
-            sum([mprob(s1,s2) for s2 in set1]) +
-            (1-µA(mutrate,nsub))^numAT(s1)*
-            (1-µG(mutrate,nsub))^(6-numAT(s1))
-            ) for s1 in set1]
-        )
-    return v
-end

### 2023-03-06 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl 	modified:   Modeling/antisenseGenes_supplement.jl 	modified:   Modeling/hexamers2.txt 	modified:   Modeling/nucleotidefuncts.jl 	modified:   Modeling/trimers2.txt

@@ -22,58 +22,8 @@ mers6 = kmers(6);
 
 # Data based calculations #
 
-data, header = readdlm(joinpath(Base.source_dir(),"hexamers2.txt"), '\t', header=true);
-header = replace.(header, r"[^A-Z,a-z]" => "");
-mers6freq = DataFrame(data[:,[1,3]], header[[1,3]]);
-
-data, header = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t', header=true);
-header = replace.(header, r"[^A-Z,a-z]" => "");
-mers3freq = DataFrame(data[:,[1,3]], header[[1,3]]);
-
-nprob6 = x -> mers6freq.Frequency[mers6freq.kmer .== x][1]
-nprob3 = x -> mers3freq.Frequency[mers3freq.kmer .== x][1]
-
-function featuregain3(set1,set2)
-    if(isempty(set1) || isempty(set2))
-        @warn "empty feature sets, returning 0"
-        return 0
-    end
-    v1 = sum([nprob3(s1)*sum([mprob(s1,s2) for s2 in set2])
-                for s1 in set1])
-    return v1
-end
-
-function featurestay3(set1)
-    v =  sum(
-        [nprob3(s1)*(
-            sum([mprob(s1,s2) for s2 in set1]) +
-            (1-µA(mutrate,nsub))^numAT(s1)*
-            (1-µG(mutrate,nsub))^(3-numAT(s1))
-            ) for s1 in set1]
-        )
-    return v
-end
-
-function featuregain6(set1,set2)
-    if(isempty(set1) || isempty(set2))
-        @warn "empty feature sets, returning 0"
-        return 0
-    end
-    v1 = sum([nprob6(s1)*sum([mprob(s1,s2) for s2 in set2])
-                for s1 in set1])
-    return v1
-end
-
-function featurestay6(set1)
-    v =  sum(
-        [nprob6(s1)*(
-            sum([mprob(s1,s2) for s2 in set1]) +
-            (1-µA(mutrate,nsub))^numAT(s1)*
-            (1-µG(mutrate,nsub))^(6-numAT(s1))
-            ) for s1 in set1]
-        )
-    return v
-end
+f6= readdlm(joinpath(Base.source_dir(),"hexamers2.txt"), '\t');
[…]

### 2023-03-06 Modeling/antisenseGenes_supplement.jl

@@ -6,7 +6,6 @@ codonfreq = readdlm(joinpath(Base.source_dir(),"codons.txt"), '\t');
 codonfreq[codonfreq[:,1] .∈ Ref(stopcodons),2] .= 0;
 codonfreq[:,2] = codonfreq[:,2]/sum(codonfreq[:,2]);
 
-
 dcfreqdata = readdlm(joinpath(Base.source_dir(),"dicodons.txt"), '\t');
 cc = z -> [x[z:z+2] for x in dcfreqdata[:,1]]
 
@@ -19,95 +18,51 @@ end
 
 dicodonfreq[:,2] = normalize(dicodonfreq[:,2]);
 
-nprob3 = x -> codonfreq[codonfreq[:,1] .== x,2]
-nprob6 = x -> dicodonfreq[dicodonfreq[:,1] .== x,2]
+trimerfreq = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t');
 
-function featuregain3(set1,set2)
-    if(isempty(set1) || isempty(set2))
-        @warn "empty feature sets, returning 0"
-        return 0
-    end
-    v1 = sum([nprob3(s1)*sum([mprob(s1,s2) for s2 in set2])
-                for s1 in set1])
-    return v1
-end
-
-function featurestay3(set1)
-    v =  sum(
-        [nprob3(s1)*(
-            sum([mprob(s1,s2) for s2 in set1]) +
-            (1-µA(mutrate,nsub))^numAT(s1)*
-            (1-µG(mutrate,nsub))^(3-numAT(s1))
-            ) for s1 in set1]
-        )
-    return v
-end
-
-function featuregain6(set1,set2)
-    if(isempty(set1) || isempty(set2))
-        @warn "empty feature sets, returning 0"
-        return 0
-    end
-    v1 = sum([nprob6(s1)*sum([mprob(s1,s2) for s2 in set2])
-                for s1 in set1])
-    return v1
-end
-
-function featurestay6(set1)
-    v =  sum(
-        [nprob6(s1)*(
-            sum([mprob(s1,s2) for s2 in set1]) +
-            (1-µA(mutrate,nsub))^numAT(s1)*
-            (1-µG(mutrate,nsub))^(6-numAT(s1))
-            ) for s1 in set1]
-        )
-    return v
-end
-
-noATG = setdiff(allcodons, ["ATG"]);
-function ATGprobsX(gccontent)
[…]

### 2023-03-06 Modeling/nucleotidefuncts.jl

@@ -227,3 +227,48 @@ function featurestay(set1,gccontent)
         )
     return v
 end
+
+nprob3 = (x,t3) -> t3[t3[:,1] .== x,2][1]
+nprob6 = (x,t6) -> t6[t6[:,1] .== x,2][1]
+
+function featuregain3(set1,set2,t3)
+    if(isempty(set1) || isempty(set2))
+        @warn "empty feature sets, returning 0"
+        return 0
+    end
+    v1 = sum([nprob3(s1,t3)*sum([mprob(s1,s2) for s2 in set2])
+                for s1 in set1])
+    return v1[1]
+end
+
+function featurestay3(set1,t3)
+    v =  sum(
+        [nprob3(s1,t3)*(
+            sum([mprob(s1,s2) for s2 in set1]) +
+            (1-µA(mutrate,nsub))^numAT(s1)*
+            (1-µG(mutrate,nsub))^(3-numAT(s1))
+            ) for s1 in set1]
+        )
+    return v[1]
+end
+
+function featuregain6(set1,set2,t6)
+    if(isempty(set1) || isempty(set2))
+        @warn "empty feature sets, returning 0"
+        return 0
+    end
+    v1 = sum([nprob6(s1,t6)*sum([mprob(s1,s2) for s2 in set2])
+                for s1 in set1])
+    return v1[1]
+end
+
+function featurestay6(set1,t6)
+    v =  sum(
+        [nprob6(s1,t6)*(
+            sum([mprob(s1,s2) for s2 in set1]) +
+            (1-µA(mutrate,nsub))^numAT(s1)*
+            (1-µG(mutrate,nsub))^(6-numAT(s1))
+            ) for s1 in set1]
+        )
+    return v[1]
+end

### 2023-03-06 Modeling/antisenseGenes.jl
modified:   Manuscripts/Figures/pORFgain_antisense.svg 	modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl

@@ -5,7 +5,8 @@ include("nucleotidefuncts.jl")
 
 cm2pt = (cm) -> 28.3465*cm
 figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
-colors = ["#FFCC00","#5599FF","#D40000","#000000"];
+colors = ["#FFCC00","#5599FF","#D40000","#000000","#754473"];
+lstyles = [:solid,:dash,:dot]
 
 default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
 guidefont = font(13,"Helvetica"),framestyle = :box, legend = false);
@@ -78,7 +79,12 @@ end
 vjoin = (z) -> [join(x) for x in eachrow(z)]
 phcp = (x,y) -> permutedims(hcat(collect.(product(x,y))...))
 
+allaacodonpairs = phcp(nostopcodons,nostopcodons);
 
+csetR = [allaacodonpairs[eachrow(allaacodonpairs) .∉ Ref(eachrow(stopNbrWithin[x])),:] for x in 1:2];
+
+Rcodon = (codonpair,x) -> reverse(comp(join(codonpair)[3-x+1:6-x]))
+cdn2aanum = (cdn) -> aanames[transnucs(cdn)]
 
 # p_synsubs = [featuregain([x],syncodons[x],gccontent) for x in keys(syncodons)]
 # p_simsubs = [featuregain([x],simcodons[x],gccontent) for x in keys(simcodons)]
@@ -248,16 +254,14 @@ function orfprobs(ATG,stop,k)
     return [orfprob; orfgain; orfloss; orfstay]
 end
 
-include("antisenseGenes_supplement.jl")
-
 ncodons = [30:300;];
-gcrange = [0.3:0.05:0.6;];
+gcrange = [0.3:0.1:0.6;];
 
 orfvalsITG = zeros(length(gcrange),length(ncodons),4);
 orfvalsONS, orfvalsORS, orfvalsOSS = [zeros(length(gcrange),length(ncodons),3,4) for i = 1:3];
 
-PStpR = zeros(size(gcrange));
-GstpR, LstpR = [zeros(length(gcrange),3) for i = 1:3];
+PstpR = zeros(3,length(gcrange));
+GstpR, LstpR = [zeros(3,3,length(gcrange)) for i = 1:2];
 
 for g in eachindex(gcrange)
[…]

### 2023-03-06 Modeling/antisenseGenes_supplement.jl

@@ -19,6 +19,7 @@ end
 dicodonfreq[:,2] = normalize(dicodonfreq[:,2]);
 
 trimerfreq = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t');
+trimerfreq[:,2] = normalize(trimerfreq[:,2])
 
 function ATGprobsX(t3)
     ATGprob = nprob3("ATG",t3)
@@ -41,9 +42,9 @@ end
 function stopprobsoverlapX(t3,t6)
     # Probability of finding a stop codon
 
-    p_stopWithin = [sum(nprob6.(vjoin(stopNbrWithin[x]),t6)) for x in 1:2];
+    p_stopWithin = [sum([nprob6(y,t6) for y in vjoin(stopNbrWithin[x])]) for x in 1:2];
 
-    p_stopR0 = sum(nprob3.(stopneighborsR0,t3));
+    p_stopR0 = sum([nprob3(y,t3) for y in stopneighborsR0]);
     push!(p_stopWithin,p_stopR0);
 
     # Probability of gaining a stop codon
@@ -62,8 +63,7 @@ function stopprobsoverlapX(t3,t6)
 
     s_stopWithin = [featurestay6(vjoin(stopNbrWithin[x]),t6) for x in 1:2];
 
-    push!(s_stopWithin,featurestay3(stopneighborsR0));
-
+    push!(s_stopWithin,featurestay3(stopneighborsR0,t3));
 
     return(hcat(p_stopWithin,g_stopWithin,l_stopWithin, s_stopWithin))
 end
@@ -152,4 +152,49 @@ function pstopselX(sdict,stopvals,t3,t6)
     stay[3] = staysel0X(stopneighborsR0,sdict,t3)
 
     return hcat(pstops,gain,loss,stay)
+end
+
+function aadivframeR_X(frm::Int,sdict::Dict,t6)
+    d = 0
+    p = 0
+    for Z in eachrow(csetR[frm])
+        stZ = phcp(vcat(Z[1],sdict[Z[1]]),vcat(Z[2],sdict[Z[2]]))
+        rcdn = Rcodon(Z,frm)
+        stZ2 = stZ[(eachrow(stZ) .!= Ref(Z)) .& ([Rcodon(y,frm) for y in eachrow(stZ)] .∉ Ref(stopcodons)) ,:]
+        if(isempty(stZ2))
+            dZ=0
+        else
[…]

### 2023-03-07 Modeling/antisenseGenes.jl
modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl 	new file:   Modeling/bacteria_mutbias.txt 	deleted:    Modeling/codons.txt 	deleted:    Modeling/dicodons.txt 	renamed:    Modeling/hexamers2.txt -> Modeling/dmel_intergenic_hexamers.txt 	renamed:    Modeling/trimers2.txt -> Modeling/dmel_intergenic_trimers.txt 	new file:   Modeling/dmel_mutbias.txt 	new file:   Modeling/dmel_orf_codonfreq.txt 	new file:   Modeling/dmel_orf_dicodonfreq.txt 	modified:   Modeling/nucleotidefuncts.jl 	new file:   Modeling/scer_intergenic_trimers.txt 	new file:   Modeling/sc

@@ -81,11 +81,6 @@ phcp = (x,y) -> permutedims(hcat(collect.(product(x,y))...))
 
 allaacodonpairs = phcp(nostopcodons,nostopcodons);
 
-csetR = [allaacodonpairs[eachrow(allaacodonpairs) .∉ Ref(eachrow(stopNbrWithin[x])),:] for x in 1:2];
-
-Rcodon = (codonpair,x) -> reverse(comp(join(codonpair)[3-x+1:6-x]))
-cdn2aanum = (cdn) -> aanames[transnucs(cdn)]
-
 # p_synsubs = [featuregain([x],syncodons[x],gccontent) for x in keys(syncodons)]
 # p_simsubs = [featuregain([x],simcodons[x],gccontent) for x in keys(simcodons)]
 
@@ -115,6 +110,11 @@ nostopneighbors = [unique(vcat(frameXcpairs.(nostopcodons,-x)...), dims = 1) for
 nostopNbr0 = nostopcodons[nostopcodons .∉ Ref(stopneighborsR0)];
 
 nostopNbrWithin = [nostopneighbors[x][(nostopneighbors[x][:,1] .∉ Ref(stopcodons)) .& (nostopneighbors[x][:,2] .∉ Ref(stopcodons)),:] for x in 1:2];
+
+csetR = [allaacodonpairs[eachrow(allaacodonpairs) .∉ Ref(eachrow(stopNbrWithin[x])),:] for x in 1:2];
+
+Rcodon = (codonpair,x) -> reverse(comp(join(codonpair)[3-x+1:6-x]))
+cdn2aanum = (cdn) -> aanames[transnucs(cdn)]
 
 
 function stopprobsoverlap(gccontent)
@@ -467,31 +467,3 @@ push!(divergence_PurSel_Rel_X,aadivframe0_X(simcodons,codonfreq))
 
 divergence_PurSel_Max_X = [aadivframeR_X(x,syncodons,dicodonfreq) for x in 1:2]
 push!(divergence_PurSel_Max_X,aadivframe0_X(syncodons,codonfreq))
-
-# # INTRONS #
-
-# # Intron frame frequency (dmel-6.46) #
-
-# intfrm2 = 0.423237
-# intfrm1 = 0.281180
-# intfrm0 = 0.295583
-
-# # Intron nucleotide frequency (dmel-6.46) #
-# first6 = [
-#     0.000763	0.998862	0.000264	0.000083;
[…]

### 2023-03-07 Modeling/antisenseGenes_supplement.jl

@@ -2,23 +2,24 @@ using DelimitedFiles
 
 normalize = x -> x/sum(x)
 
-codonfreq = readdlm(joinpath(Base.source_dir(),"codons.txt"), '\t');
-codonfreq[codonfreq[:,1] .∈ Ref(stopcodons),2] .= 0;
-codonfreq[:,2] = codonfreq[:,2]/sum(codonfreq[:,2]);
+codonfreq = readdlm(joinpath(Base.source_dir(),"dmel_orf_codonfreq.txt"), '\t');
+dicodonfreq = readdlm(joinpath(Base.source_dir(),"dmel_orf_dicodonfreq.txt"), '\t');
+# codonfreq[codonfreq[:,1] .∈ Ref(stopcodons),2] .= 0;
+# codonfreq[:,2] = normalize(codonfreq[:,2]);
 
-dcfreqdata = readdlm(joinpath(Base.source_dir(),"dicodons.txt"), '\t');
-cc = z -> [x[z:z+2] for x in dcfreqdata[:,1]]
+# dcfreqdata = readdlm(joinpath(Base.source_dir(),"dmel_orf_dicodonfreq.txt"), '\t');
+# cc = z -> [x[z:z+2] for x in dcfreqdata[:,1]]
 
-dcfreqdata[(cc(1) .∈ Ref(stopcodons)) .| (cc(3) .∈ Ref(stopcodons)) ,2] .= 0;
+# dcfreqdata[(cc(1) .∈ Ref(stopcodons)) .| (cc(4) .∈ Ref(stopcodons)) ,2] .= 0;
 
-dicodonfreq = hcat(kmers(6),zeros(4^6,1));
-for x in eachrow(dcfreqdata)
-    dicodonfreq[dicodonfreq[:,1] .== x[1],2] .= x[2];
-end
+# dicodonfreq = hcat(kmers(6),zeros(4^6,1));
+# for x in eachrow(dcfreqdata)
+#     dicodonfreq[dicodonfreq[:,1] .== x[1],2] .= x[2];
+# end
 
-dicodonfreq[:,2] = normalize(dicodonfreq[:,2]);
+# dicodonfreq[:,2] = normalize(dicodonfreq[:,2]);
 
-trimerfreq = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t');
+trimerfreq = readdlm(joinpath(Base.source_dir(),"dmel_intergenic_trimers.txt"), '\t');
[…]

### 2023-03-07 Modeling/nucleotidefuncts.jl

@@ -105,38 +105,9 @@ end
 
 
 
-nsub = zeros(4,4);
-
-nsub[1,2] = 0.056043956043956;
-nsub[1,3] = 0.242857142857143;
-nsub[1,4] = 0.074725274725275;
-nsub[3,1] = 0.438461538461539;
-nsub[3,2] = 0.075164835164835;
-nsub[3,4] = 1 - sum(nsub[1,:]) - sum(nsub[3,1:2]);
-nsub[2,1] = nsub[1,2];
-nsub[2,3] = nsub[1,4];
-nsub[2,4] = nsub[1,3];
-nsub[4,3] = nsub[3,4];
-nsub[4,2] = nsub[3,1];
-nsub[4,1] = nsub[3,2];
+nsub, nucsmbt = readdlm("dmel_mutbias.txt",'\t',header = true);
 
 mutrate = 7.8e-9
-
-
-nsubbac = zeros(4,4)
-
-nsubbac[1,2] = 0.0031813361611877;
-nsubbac[1,3] = 0.0095440084835631;
-nsubbac[1,4] = 0.0031813361611877;
-nsubbac[3,1] = 0.0689289501590668;
-nsubbac[3,2] = 0.907741251325557;
-nsubbac[3,4] = 1 - sum(nsubbac[1,:]) - sum(nsubbac[3,1:2]);
-nsubbac[2,1] = nsubbac[1,2];
-nsubbac[2,3] = nsubbac[1,4];
-nsubbac[2,4] = nsubbac[1,3];
-nsubbac[4,3] = nsubbac[3,4];
-nsubbac[4,2] = nsubbac[3,1];
-nsubbac[4,1] = nsubbac[3,2];
 
 mutratebac = 2e-8;
 

### 2023-03-07 Modeling/DeNovoProb.jl
modified:   Modeling/DeNovoProb.jl

@@ -22,8 +22,8 @@ mers6 = kmers(6);
 
 # Data based calculations #
 
-f6= readdlm(joinpath(Base.source_dir(),"hexamers2.txt"), '\t');
-f3 = readdlm(joinpath(Base.source_dir(),"trimers2.txt"), '\t');
+f6= readdlm(joinpath(Base.source_dir(),"dmel_intergenic_hexamers.txt"), '\t');
+f3 = readdlm(joinpath(Base.source_dir(),"dmel_intergenic_trimers.txt"), '\t');
 
 ## TATA box
 # Consensus sequence = TATAWAWR

### 2023-03-07 Modeling/antisenseGenes.jl
renamed:    Manuscripts/Figures/gainProbs_dmel.svg -> Manuscripts/Figures/M1_main/gainProbs_dmel.svg 	renamed:    Manuscripts/Figures/pGainLossFix.svg -> Manuscripts/Figures/M1_main/pGainLossFix.svg 	renamed:    Manuscripts/Figures/pGainLossFix2.svg -> Manuscripts/Figures/M1_main/pGainLossFix2.svg 	renamed:    Manuscripts/Figures/pLoss_promoterlessX.svg -> Manuscripts/Figures/M1_main/pLoss_promoterlessX.svg 	renamed:    Manuscripts/Figures/pLossgcr_promoterless.svg -> Manuscripts/Figures/M1_main/pLossgcr_promoterless.svg 	renamed:    Manuscripts/Figures/proteinComposition.svg -> Manuscripts/Fi

@@ -3,6 +3,8 @@ using Measures
 cd(Base.source_dir())
 include("nucleotidefuncts.jl")
 
+organism = "dmel"
+
 cm2pt = (cm) -> 28.3465*cm
 figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
 colors = ["#FFCC00","#5599FF","#D40000","#000000","#754473"];
@@ -16,29 +18,12 @@ function nostop(codonset)
 	codonset[codonset .∉ Ref(stopcodons)]
 end
 
-gccontent = 0.493611; # all mRNA #
-# gccontent = 0.53942 # all CDS #
-
 (xProb, xGain, xLoss, xStay) = [i for i in 1:4];
-
-p_T = 0.204217;
-p_A = 0.256362;
-p_G = 0.267809;
-p_C = 0.271612;
-
-nfmat = zeros(4,1);
-
-nfmat[nucnames['A']] = 0.256362;
-nfmat[nucnames['T']] = 0.204217;
-nfmat[nucnames['G']] = 0.267809;
-nfmat[nucnames['A']] = 0.271612;
-
-function nucprobss(s,nfmat)
-    return prod([nfmat[nucnames[x]] for x in s])
-end
 
 allcodons = kmers(3)
 nostopcodons = nostop(allcodons);
+
+nsub, nucsmbt = readdlm(organism*"_mutbias.txt",'\t',header = true);
 
 noATG = setdiff(allcodons, ["ATG"]);
 function ATGprobs(gccontent)
@@ -335,7 +320,7 @@ for f = 1:3
         mng>10 ? d = 2 : d = 3
         plots_gain[f,s] = plot(ncodons, ydatag[:,1],
             linecolor = colors[1],
-            yticks = round.(range(lbg,stop=ubg,length = 4))
+            yticks = round.(range(lbg,stop=ubg,length = 4),digits=2)
         );
 
         ydatal = rat[:,:,s,2]
@@ -355,10 +340,10 @@ for f = 1:3
 end
 
 pG = plot(plots_gain..., size = (width = cm2pt(27.5), height = cm2pt(25)));
-savefig(pG, figdir*"pORFgain_antisense.pdf")
+savefig(pG, figdir*"pORFgain_antisense_"*organism*"GC.pdf")
 
 pL = plot(plots_loss..., size = (width = cm2pt(27.5), height = cm2pt(25)));
[…]

### 2023-03-07 Modeling/antisenseGenes_supplement.jl

@@ -2,8 +2,8 @@ using DelimitedFiles
 
 normalize = x -> x/sum(x)
 
-codonfreq = readdlm(joinpath(Base.source_dir(),"dmel_orf_codonfreq.txt"), '\t');
-dicodonfreq = readdlm(joinpath(Base.source_dir(),"dmel_orf_dicodonfreq.txt"), '\t');
+codonfreq = readdlm(joinpath(Base.source_dir(),organism*"_orf_codonfreq.txt"), '\t');
+dicodonfreq = readdlm(joinpath(Base.source_dir(),organism*"_orf_dicodonfreq.txt"), '\t');
 # codonfreq[codonfreq[:,1] .∈ Ref(stopcodons),2] .= 0;
 # codonfreq[:,2] = normalize(codonfreq[:,2]);
 
@@ -19,7 +19,7 @@ dicodonfreq = readdlm(joinpath(Base.source_dir(),"dmel_orf_dicodonfreq.txt"), '\
 
 # dicodonfreq[:,2] = normalize(dicodonfreq[:,2]);
 
-trimerfreq = readdlm(joinpath(Base.source_dir(),"dmel_intergenic_trimers.txt"), '\t');
+trimerfreq = readdlm(joinpath(Base.source_dir(),organism*"_intergenic_trimers.txt"), '\t');
 trimerfreq[:,2] = normalize(trimerfreq[:,2])
 
 function ATGprobsX(t3)

### 2023-03-08 Manuscripts/M1/deNovoRates_old.tex
renamed:    Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex -> Manuscripts/M1/Response_to_Reviewers_Ravi2023MBE.tex 	renamed:    Manuscripts/SM_deNovoRates.tex -> Manuscripts/M1/SM_deNovoRates.tex 	renamed:    Manuscripts/deNovoRates.tex -> Manuscripts/M1/deNovoRates.tex 	renamed:    Manuscripts/deNovoRates_old.tex -> Manuscripts/M1/deNovoRates_old.tex 	renamed:    Manuscripts/deNovoStudiesRNAseqInfo.csv -> Manuscripts/M1/deNovoStudiesRNAseqInfo.csv 	renamed:    Manuscripts/mybst.bst -> Manuscripts/M1/mybst.bst 	renamed:    Manuscripts/refs.bib -> Manuscripts/M1/refs.bib 	new file:   Manuscr

--- 2022-12-22 Manuscripts/deNovoRates2.tex
@@
 In the first scenario, the ORF remains intact whereas the transcription is lost.
-The probability of this scenario is described by:
-$P_\textit{ORF-stay \vline\ RNA} \times P_\textit{RNA-loss \vline\ ORF}/P_\textit{ORF}$.
-The probability of the second scenario, where transcription stays intact while ORF is lost, is: $P_\textit{ORF-loss} \times P_\textit{RNA-stay}/P_\textit{RNA}$.
-We find that the first scenario is more plausible for ORFs that have fewer than 34 codons, which is in agreement with our previous analysis (\autoref{lossprob}).
-Also, the probability values of scenario one and two, obtained from this stringent analysis, were not significantly different from the transcription ($P_\textit{RNA-loss}$) and ORF loss ($P_\textit{ORF-loss}$) probabilities, respectively.
+The probability of this scenario ($P_\textit{onlyORF-loss}$) is described by:
+
+\vspace{-1ex}
+
+\begin{equation}
+P_\textit{onlyRNA-loss} = \frac{P_\textit{ORF-stay \vline\ RNA}}{P_\textit{ORF \vline \ RNA}} \times P_\textit{RNA-loss \vline\ ORF}
+\end{equation}
+
+We note that the probability that an ORF is not lost due to mutations ($P_\textit{ORF-stay \vline \ RNA}$), includes the probability that the ORF already exists ($P_\textit{ORF \vline \ RNA}$).
+To describe the conditional probability of ORF remaining undisrupted, given that it already exists, we divide $P_\textit{ORF-stay \vline \ RNA}$ by $P_\textit{ORF \vline \ RNA}$.
+
+In the second scenario, transcription stays intact while ORF is lost. The probability of this scenario is:
+
+\vspace{-1em}
+\begin{equation}
+P_\textit{onlyORF-loss} = \frac{P_\textit{RNA-stay \vline\ ORF}}{P_\textit{RNA \vline \ ORF}} \times P_\textit{ORF-loss \vline\ RNA}
+\end{equation}
+
+We find that the second scenario (only ORF loss) is more plausible for all the analysed ORFs lengths.
+Furthermore, the ratio of $P_\textit{onlyORF-loss}$ and $P_\textit{onlyRNA-loss}$ has nearly identical values and trend across different ORF lengths (data not shown), as the ratio of $P_\textit{ORF-loss}$ and $P_\textit{RNA-loss}$ shown in \autoref{lossprob}.
 
@@
 
-$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} - P_\textit{poly-A})}{1- P_\textit{stop}}$$
+$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} + P_\textit{poly-A})}{1- P_\textit{stop}}$$
 

### 2023-03-08 Manuscripts/M2/antisense_genes.tex
renamed:    Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex -> Manuscripts/M1/Response_to_Reviewers_Ravi2023MBE.tex 	renamed:    Manuscripts/SM_deNovoRates.tex -> Manuscripts/M1/SM_deNovoRates.tex 	renamed:    Manuscripts/deNovoRates.tex -> Manuscripts/M1/deNovoRates.tex 	renamed:    Manuscripts/deNovoRates_old.tex -> Manuscripts/M1/deNovoRates_old.tex 	renamed:    Manuscripts/deNovoStudiesRNAseqInfo.csv -> Manuscripts/M1/deNovoStudiesRNAseqInfo.csv 	renamed:    Manuscripts/mybst.bst -> Manuscripts/M1/mybst.bst 	renamed:    Manuscripts/refs.bib -> Manuscripts/M1/refs.bib 	new file:   Manuscr

--- 2023-03-01 Manuscripts/deNovoRates.tex
@@
 \section*{Abstract}
-New protein coding genes can emerge from genomic regions that previously did not contain any genes, via a process called \textit{de novo} gene emergence.
-To synthesize a protein, DNA must be transcribed as well as translated.
-Both processes need certain DNA sequence features.
-Stable transcription requires promoters and a polydenylation signal, while translation requires at least an open reading frame (ORF).
-We develop mathematical models based on mutation probabilities, and the assumption of neutral evolution, to find out how quickly genes emerge and are lost.
-We also investigate the effect of the order by which DNA features evolve, and if sequence composition is biased by mutation rate.
-We rationalize how genes are lost much more rapidly than they emerge, and how genes with long ORFs preferentially arise in regions that are already transcribed.
-Our study not only answers some fundamental questions on the topic of \textit{de novo} emergence but also provides a modeling framework for future studies.
 
@@
 
-Organisms evolve new traits by expanding their functional genome.
-Evolution of new genes is one of the ways by which new traits can emerge.
-The definition of a gene is complicated, and has been changing constantly \citep{genepostencode}.
-We use the following working definition of a gene: a gene is a region of the genome that gives rise to a functional product, that is an RNA or a protein.
-The next big challenge is to define what a function is, and how to classify gene functions.
-One of the easiest definition of a gene product's function is its ability to improve the survival of the organism under one or more environmental conditions \citep{deNovoFunction}.
-Proteins perform the most diverse kinds of molecular functions ranging from catalysis of biochemical reactions to formation of cellular structures \citep{berg}, whereas RNAs that do not encode proteins like mRNAs, or participate in protein synthesis like rRNAs and tRNAs, are mostly involved in regulation of gene expression \citep{fncel,Statello2021}.
-
-In this study we focus on the evolution of genes that encode new proteins.
-New protein coding genes frequently arise via duplication of existing protein coding genes.
-These duplicated genes then genetically and functionally diverge from their parent genes by accumulating mutations \citep{Long2003,IAD}.
-Recently it has been shown that new protein coding genes can arise independent of gene duplication, from genomic sequences that did not previously encode any protein.
-This phenomenon is called \textit{de novo} gene emergence \citep{Tautz2011,Zhao2014,EBB-F1000,Vakirlis2017,vanOss}.
-\textit{De novo} protein coding genes can emerge from intergenic sequences, non-coding RNA genes, introns, and even regions that partially overlap with existing protein coding genes \citep{Tautz2011,Zhao2014,Vakirlis2017,vanOss,Prabh2019}.
-To express a protein, a genomic sequence should be transcribed as well as translated.
-Therefore, the gene needs sequence features that enable both these processes.
-
-\cmnt{The primary requirement for transcription is recruitment of RNA polymerase.
-This process is facilitated by DNA sequences called \citep{Promoters}.
-A part of the promoter, called the core promoter is the region that determines the start of transcription \citep{corepromoters}.
-Transcription can also be initiated without the requirement of a defined promoter sequence, and this process can occur throughout the genome \cite{PervasiveTranscription}.
-Even though pervasive transcription is widespread in eukaryotic genomes, most RNAs are quickly eliminated by RNA degrading enzymes \citep{RNAturnover}.
-Most eukaryotic mRNAs and long non-coding RNAs, are polyadenylated at their 3' termini, which makes them resistant to degradation.
-Polyadenylation, which also marks the end of transcription \citep{polyAterm}, is facilitated by sequences known as polyadenylation signals \citep{polyA}.
-In prokaryotes, the end of transcription is determined by sequences known as terminators \citep{ProkTerm}.
-Termination at an appropriate site is another fundamental requirement for a productive transcription.}
-
-The second major requirement for protein expression is translation of the transcribed mRNA.
-The most fundamental requirement for translation is the presence of an open reading frame (ORF).
-Usually, an mRNA needs additional features to initiate protein synthesis.
-These features include ribosome binding sites in prokaryotes \citep{RBS}, and Kozak consensus sequences in eukaryotes \citep{kozak,kozakDroso,kozakHuman}.
-
-Although gain of transcription and translation features guarantees \textit{de novo} birth of a protein coding gene, it does not ensure that the gene would persist in the genome for many generations.
-The newly born gene can lose the features as easily as it gained them unless it has been fixed in the genome, for example via evolutionary selection.
-Specifically, if the protein synthesized by the \textit{de novo} gene provides a fitness advantage to the host organism, it will undergo positive selection \citep{deNovoFunction}.
-Simultaneously, the protein should have low toxicity and cost of synthesis, to survive purifying selection.
-A common mechanism by which protein mediated toxicity occurs is by misfolding and aggregation of proteins \citep{misfold1,misfold2}.
-
-In this study we develop probabilistic models of \textit{de novo} gene emergence \cmnt{in eukaryotes}.
-We focus on \textit{de novo} genes emerging from non-genic sequences, also known as ``proto-genes'' \citep{protogenes,vanOss}.
-We model \textit{de novo} gene emergence as a two step process.
-In the first step, a non-genic DNA sequence gains transcriptional features that allows it to express an untranslated (non-coding) RNA.
-In the second step this non-coding RNA gene acquires translational features that allows it to express a protein.
-Alternatively, translational features can emerge in the non-genic DNA before transcriptional features.
-Specifically, we calculate the probabilities by which polyadenylation signals and ORFs emerge, and are lost, based on the rates at which different kinds of DNA mutations naturally occur (mutation bias).
-With these probabilities we estimate whether transcriptional and translational features preferentially evolve in a specific order.
-Using a similar approach, we calculate how random mutations affect protein composition.
-In our models, we assume that the proteins expressed from these proto-genes provide no fitness advantage and are not toxic to the host organism.
-Thus our models are based on neutral evolution.
-Our models predict that there is indeed a preferred sequence of DNA feature evolution during \textit{de novo} emergence, even under the assumption of neutrality.
-
 \section*{Results}
 
-In this work, we developed mathematical models to estimate the rates and probabilities of \textit{de novo} gene emergence, as well as gene loss.
-A proto-gene emerges from non-genic DNA, when the latter mutates to gain sequence features necessary for transcription and translation \citep{protogenes,vanOss}.
-Both transcription and translation are complex processes involving many biomolecular complexes that work in concert.
-Here, we focus on the minimal requirements for these processes to occur.
-\cmnt{Experimental data show that transcription in eukaryotes can be initiated genome wide, especially in regions proximal to enhancers, and more specifically, in regions depleted of nucleosomes \citep{enhancers}.
-Thus transcription of many proto-genes can be initiated at a basal transcription rate, and may not require specific core promoters.
-Although transcription may be initiated without a core promoter, a stable RNA product often requires a poly-A tail that also facilitates RNA export to cytoplasm \citep{polyAterm,polyAexp}.
-This assumption is supported by experimental data, where the majority of detectably expressed proto-genes have polyadenylated RNAs \citep{Neme2016,Witt2019,PacoEnhancers,albaYeastdenovo}.
-Thus, we define a poly-A signal as the primary requirement for transcription such that it should exist but only at the end of a transcribed region.} For translation, we only require the gene to have an open reading frame (ORF).
-We only focus on intronless genes in this study.
-
-Using simple probability models, we calculate the likelihood of finding the sequence features that facilitate \textit{de novo} gene emergence by random chance, and the rate at which these features are gained and lost due to random mutations.
-These probability models are essentially described by two kinds of probability.
-First is the probability of finding a DNA sequence feature (such as the poly-A signal or an ORF; \autoref{methprob}).
-This probability depends on the nucleotide distribution \cmnt{that can be roughly approximated by the GC-content}.
-The second kind of probability is the transition probability, that is the probability of a DNA sequence mutating to another (\autoref{methgain}).
-Transition probabilities that describe the gain and loss of DNA sequence features, depend on mutation rate, mutation bias, and \cmnt{nucleotide composition}.
-Our model of DNA sequence evolution is in principle similar to a theoretical model described in a previous study \citep{promoterWaiting}.
-In this study, the authors have calculated the time required for different transcription factor binding sites to emerge in gene promoters.
-We extend this principle to calculate not only the probability of sequence emergence but also the likelihood that an emerged sequence will remain intact or be lost (\autoref{methfeatures}).
-
-\cmnt{In our model we primarily focus on \textit{de novo} gene emergence at one genomic locus.
-We assume that transcription is initiated at this locus with a probability of 0.12.
-We based this estimate on the fraction of intergenic open chromatin in \textit{Drosophila melanogaster} that is occupied by \textit{cis}-regulatory elements (\autoref{methRNA})}.
-We used a default GC-content of 42\% in our calculations which is reasonably close to the total genomic GC-content of human \citep[41\%,][]{Merchant2007}, \textit{Drosophila melanogaster} \citep[41.6\%,][]{flybase}, and \textit{Saccharomyces cerevisiae} \citep[38\%,][]{yeastGC}.
-Because GC content can vary between different genomic loci, we also performed of our calculations with other values of GC-content.
-\cmnt{We note that nucleotide composition in any DNA region may not be uniformly distributed.
-Therefore, we also included calculations using a more realistic approximation of nucleotide distribution, using frequencies of short DNA sequences in the intergenic genome of \textit{D.melanogaster}.
-Specifically, we calculated the frequency of all DNA trimers for codon probabilites, and of all DNA hexamers for poly-A signal probabilities (\autoref{methRNA}).}
-
-We used a mutation rate of $7.8\times10^{-9}$ mutations per nucleotide position per generation, which corresponds to spontaneous mutation rate estimated in a \textit{D.
-melanogaster} population that was subjected to periodic population bottlenecks \cite[mutation accumulation line,][]{drosophilamutrate}.
-We used mutation bias data (\autoref{mutbias}) from the same study \citep{drosophilamutrate} which was in agreement with another previous study on human pseudogenes \citep{humanmutrate}.
-
-Using a similar approach, we predict how the nucleotide distribution and random mutations shape protein composition and evolution.
-
-Overall, our models define a null hypothesis under which protein coding DNA sequences evolve neutrally through mutation pressure alone.
-
-\subsection*{How likely is gene loss?}
-
-If a gene is found in one population but not populations, then the most parsimonious assumption is that this gene emerged only in one specific population.
-We asked what is the chance that the gene was present in an ancestral population but was lost in all lineages except one.
-To this end we calculated the probability of gene emergence and gene loss (\autoref{methfeatures}).
-Specifically, the probability of gene gain can be defined as the sum of three probabilities.
-First is the probability that an ORF already exists and is not lost due to mutations ($P_\textit{ORF-stay}$, \autoref{methORF}), while a mutation causes transcription to emerge ($P_\textit{RNA-gain}$, \autoref{methRNA}).
-Second is the probability that an ORF emerges ($P_\textit{ORF-gain}$, \autoref{methORF}) in a region of DNA that is already transcribed and continues to be transcribed ($P_\textit{RNA-stay}$, \autoref{methRNA}).
-Third is the probability that neither of the two features already exist and both emerge at the same time due to mutations (this probability is very small and is negligible).
-We found that the probability of transcription gain is dependent on the existence of an ORF, such that it is higher when the ORF is already present.
-This dependence exists because an ORF does not have stop codons (\texttt{TAA}, \texttt{TAG} and \texttt{TGA}) in its sequence.
-This in turn, restricts the number of positions where a poly-A signal, that has a \texttt{TAA} in its sequence, can exist (\autoref{rnaorfindependence}).
-Because of the same reason, the probability of ORF gain is higher when the DNA is already transcribed.
-Therefore, we additionally calculated the conditional probabilities of transcription gain ($P_\textit{RNA-gain \vline\ ORF}$) and ORF gain ($P_\textit{ORF-gain \vline\ RNA}$) given the condition that an ORF and a transcript already exists, respectively.
-We thus define the total probability of gene gain ($P_\textit{gene-gain}^*$) as:
-
-\begin{equation}
-P_\textit{gene-gain}^* = P_\textit{RNA-gain \vline\ ORF} \times P_\textit{ORF-stay} + P_\textit{ORF-gain \vline\ RNA} \times P_\textit{RNA-stay} + P_\textit{RNA-gain}\times P_\textit{ORF-gain}
-\label{genegaineq}
-\end{equation}
-
-\autoref{genegaineq} describes the total probability of gene gain that includes the probability that the gene was not previously present.
-To more precisely address the question of whether a gene that was absent in an ancestor but emerges in a descendant, we calculated the corresponding conditional probability of gene gain:
-
-\begin{equation}
-P_\textit{gene-gain} = P_\textit{gene-gain}^*/(1-P_\textit{ORF} - P_\textit{RNA})
-\label{genegaineq2}
-\end{equation}
-
-We note that the difference between the total probability ($P_\textit{gene-gain}^*$) and the conditional probability ($P_\textit{gene-gain}$) of gene gain, is very small.
-
-Next, we calculated the probability of gene loss, given the gene (transcription and ORF) is already present($P_\textit{gene-loss}$).
-It is the sum of the probabilities that transcription ($P_\textit{RNA-loss}$, \autoref{methRNA}) or the ORF is lost due to mutations ($P_\textit{ORF-loss}$, \autoref{methORF}).
-Specifically:
-
-\begin{equation}
-P_\textit{gene-loss} = P_\textit{RNA-loss} + P_\textit{ORF-loss}
-\label{genelosseq}
-\end{equation}
-
-The probability that a gene is lost $n$ times independently, is $(P_\textit{gene-loss})^{n}$.
-To find out how many independent gene loss events are as likely as a single gene gain event, we calculate the ratio of logarithms (log-log ratio) of gene gain and gene loss for proto-genes with an ORF ranging from 30 -- 300 codons in length.
-
-\begin{equation}
-\text{Gene losses per gene gain} = \log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})
-\label{loglogratio}
-\end{equation}
-
-For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event.
-We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF.
-We found that proto-genes with a GC-content of 42\% and ORFs longer than 40 codons, can be lost two times independently in the time frame of one gene gain event (log-log ratio = 2, \hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-For proto-genes with GC-contents of 34\% and 50\%, the minimum number of codons required for a log-log ratio of 2, are 42 and 33 respectively.
-To better understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content ({\color{blue} Figure S4}), and that contain ORFs of different lengths (codons).
-We found that the effect of GC-content on the log-log ratio depends on the length of the ORF.
-Specifically, the log-log ratio steadily increases with GC-content ranging from 30\% to 60\%, when the proto-gene contains an ORF with 30 codons.
-When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases.
-For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range).
-\cmnt{Because GC-content may not accurately describe the nucleotide composition of a locus, we calculated gene gain and loss probabilities using frequencies of DNA trimers and hexamers from the intergenic genome of \textit{D.
-melanogaster}.
-Specifically, we used the frequencies of hexamers to estimate the stationary, gain and loss probabilities of poly-A signals (which is also a hexamer) and the RNA.
-Likewise, we used the frequencies of trimers to estimate the stationary, gain and loss probabilities of different codons and the ORF.
-Using these values, we found that protogenes with more than 49 codons can be independently lost twice, in the time frame of one gene gain event (orange line, \hyperref[gainlossprob]{\autoref{gainlossprob}A}).}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/pGainLossFix2.pdf}
-\caption{Genes are more likely to be independently lost twice, than being born once.
-The vertical axis in all panels shows the number of independent gene loss events with extinction (\autoref{genelosseq}) that can occur relative to one gene gain event with fixation (\autoref{loglogratio}, \autoref{pFix}), under \textbf{(A)} no selection, \textbf{(B)} weak positive selection and \textbf{(C)} weak negative selection on the expression of the protein from a proto-gene.
-Horizontal axes show the number of codons in the ORF.
-Black lines in panel \textbf{(A)} denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
-melanogaster} intergenic genome.
-In panels \textbf{(B)} and \textbf{(C)} we use the probabilitiy values from \textit{D.
-melanogaster} intergenic genome.
-$N_e$ denotes effective population size and $s$ denotes the selection co-efficient.}
-\label{gainlossprob}
-\vspace{1ex}
-\hrule
-\end{figure}
-
-\cmnt{Natural selection and effective population size ($N_e$) determine the rate at which new genes spread in a population \citep{KimuraFix}.
-To understand the combined effect of mutations, and selection we calculated two fixation probabilities for populations consisting different number of diploid individuals.
-First is the probability that any one individual in a population gains a gene ($2N_e \times P_\textit{gene-gain}$), and this gene eventually spreads in entire the population via selection and population dynamics ($P_\textit{fix}$, \autoref{pFix}).
-The second probability describes the event that all individuals in a population have two copies of the proto-gene, one copy of the gene is lost in an individual ($2N_e \times P_\textit{gene-loss}$), and the gene eventually goes extinct from the entire population due to population dynamics ($P_\textit{ext}$, \autoref{pFix}).
-We calculated the log-log ratio of the two probabilities under three scenarios (\autoref{pFix}), wherein the gene provides no fitness advantage (neutral, $s=0$), the gene is marginally beneficial (positive, $s = 0.001$), and the gene is marginally deleterious (negative, $s = -0.001$).
-We note that gene gain and loss probabilities due to mutation are so small that it is unlikely that two or more individuals gain or lose a gene simultaneously.
-Under the neutral scenario, the probability of fixation of an allele is same as the probability of the mutation that gives rise to that allele (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-Extinction of a marginally beneficial gene is more likely than its fixation only when populations are small or when the gene has a long ORF (\hyperref[gainlossprob]{\autoref{gainlossprob}B}).
-For example, a gene of any length is less likely to go extinct than it is fixed, in populations with at least 6400 individuals.
-Furthermore, a beneficial gene can be independently lost in more than two small populations with 400 individuals, only when it has more than 133 codons.
-When a gene is mildly deleterious, its is very likely to go extinct such that independent loss in more than two populations is more likely than its fixation in one population (\hyperref[gainlossprob]{\autoref{gainlossprob}C}).
-We further note that time to fixation of a rare allele (for example, a newly emerged protogene) would be much smaller \citep[$4N_e$ generations;][]{KimuraFix} than the time to gain ($10^{14}$ -- $10^{20}$ generations) or lose it ($10^{7}$ -- $10^{8}$ generations).}
-
-Overall, our analysis suggests that a proto-gene expressed in only one population may not necessarily mean that it emerged for the first time in this population.
-\cmnt{This is also true for a model of gene emergence that requires a core promoter (TATA-box or Inr) for initiating transcription ({\color{blue} Figure S1}).
-
-Despite low probability of gene gain at one particular locus ($\sim 10^{-15}$ per generation), the overall rate of gene gain throughout the genome could be much higher.
-To this end we calculated the total gain rate of genes of any length, at any locus in \textit{D.
-melanogaster} intergenic genome.
-We found this rate to be $2.8\times 10^{-5}$ new genes per generation, that corresponds to approximately $7.5\times 10^{-5}$ new genes per year \citep[assuming \textit{D.
-melanogaster} generation time of two weeks;][]{drosophilaModelOrg}.
-We performed an analogous analysis on the rate of gain of new transcripts and found this rate to be 1.53 transcripts per year.
-This estimate is slightly higher than the rate of transcript gain predicted from \textit{D.
-melanogaster} population transcriptomics data \citep[0.13 -- 0.34;][]{AnnaPeteTranscripts}}.
-
-\subsection*{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}
-
-For a proto-gene to emerge from a non-genic DNA sequence, both transcription and ORF need to emerge.
-That is, probability of gene emergence is equal to the product of probabilities of transcription gain and ORF gain.
-Thus it may appear that the order of the occurrence of these two events does not matter.
-However, gene emergence is much more likely when one of the two features already exists (\autoref{genegaineq}) and therefore it has two possible trajectories -- ORF emerges first (ORF-first) or transcription emerges first (RNA-first).
-Furthermore, emergence of an ORF is more likely when the proto-gene is already transcribed, and \textit{vice versa} (\autoref{rnaorfindependence}).
-To understand whether \textit{de novo} gene emergence has a preferred trajectory, we calculated two probabilities.
-First, the probability that an RNA exists and mutations cause a gain of an ORF but no disruption of transcription.
-This probability denotes the trajectory where RNA emerges first ($P_\textit{RNA-first}$, \autoref{rnafirsteq1}).
-
-\begin{equation}
-P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
-\label{rnafirsteq1}
-\end{equation}
-
-The second probability ($P_\textit{ORF-first}$, \autoref{orffirsteq1}) that denotes the trajectory where ORF emerges first, is the probability that an ORF already exists and mutations cause a gain of transcription but do not disrupt the ORF.
-
-\begin{equation}
-P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}
-\label{orffirsteq1}
-\end{equation}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/whoisfirstX.pdf}
-\caption{Proto-genes preferentially emerge RNA-first.
-The vertical axis shows the $\log_2$ transformed ratio of the probabilities of the RNA-first and the ORF-first trajectories, described as \textbf{(A)} a single-step process (Equations \ref{rnafirsteq1} \& \ref{orffirsteq1}) and \textbf{(B)} a two-step process (Equations \ref{rnafirsteq2} \& \ref{orffirsteq2}).
-A positive value suggests that RNA-first trajectory is more likely than ORF-first trajectory, and \textit{vice versa}.
-In both panels, the horizontal axis denotes the number of codons in the ORF.
-Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
-melanogaster} intergenic genome.}
-\label{whoisfirst}
-\vspace{1ex}
-\hrule
-\end{figure}
-
-\cmnt{We calculated the log-transformed ratio of $P_\textit{RNA-first}$ and $P_\textit{ORF-first}$, such that a positive value of the ratio would mean that ORF gain in an already transcribed DNA (RNA-first) is more likely than transcription gain for an untranscribed ORF (ORF-first).
-In other words, the RNA-first trajectory is more feasible.
-Likewise, a negative value of the ratio would suggest that the ORF-first trajectory is more feasible.
-We found that proto-genes with ORFs of all the investigated lengths, preferentially emerge RNA-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}).
-More specifically, the likelihood of the RNA-first trajectory relative to that of the ORF-first trajectory  increases with increasing ORF length.
-This is true for loci with three different values of GC content (34\%, 42\%, 50\%).
-However, our complementary analysis using DNA oligomer frequencies from \textit{D.
-melanogaster} intergenic genome shows that the likelihood of RNA-first trajectory relative to that of ORF-first trajectory, reduces with increasing ORF length.
-Nonetheless, RNA-first trajectory is still the more feasible trajectory of \textit{de novo} gene emergence.
-In a more stringent scenario of \textit{de novo} emergence, where a core promoter (TATA box or Inr) is necessary for transcription, proto-genes containing short ORFs can emerge ORF-first ({\color{blue}Figure S2})}
-
-A more stringent definition of the RNA-first trajectory would describe a two-step process.
-In the first step, an RNA emerges in an untranscribed region of DNA but an ORF does not emerge.
-In the second step an ORF emerges, while transcription stays intact (\autoref{rnafirsteq2}).
-Likewise, a ORF-first trajectory can be defined by a two-step probability.
-In the first step ORF emerges in an untranscribed DNA region, and in the second step transcription emerges, while the ORF remains intact (\autoref{orffirsteq2}).
-
-\begin{align}
-P_\textit{RNA-first}' & = \underbrace{P_\textit{RNA-gain} \times (1-P_\textit{ORF} -P_\textit{ORF-gain})}_\textit{first step} \times \underbrace{P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}/(1-P_\textit{ORF})}_\textit{second step} \label{rnafirsteq2} \\[1em]
-P_\textit{ORF-first}' & = \underbrace{P_\textit{ORF-gain} \times (1-P_\textit{RNA} -P_\textit{RNA-gain})}_\textit{first step} \times \underbrace{P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}/(1-P_\textit{RNA})}_\textit{second step} \label{orffirsteq2}
-\end{align}
-
-\cmnt{Even with this stringent definition, \textit{de novo} emergence via the RNA-first trajectory is more probable than via the ORF-first trajectory (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
-For all the three GC-content values, as well the specific analysis with \textit{D.
-melanogaster} genome, the two-step RNA-first trajectory becomes increasingly more probable than the two-step ORF-first trajectory, with increasing ORF length.}
-
-\subsection*{Would extensive transcription loss suggest negative selection of toxic proteins?}
-
-\begin{figure}[!b]
-\centering
-\hrule
-\vspace{1ex}
-\includegraphics[scale=0.5]{Figures/pLoss_promoterlessX.pdf}
-\caption{ORF loss in proto-genes is more probable than transcription loss.
-The horizontal axis denotes the number of codons in the ORF.
-The vertical axis shows the $\log_2$ transformed ratio of ORF loss and RNA loss probabilities, such that a positive value means ORF loss is more probable than RNA loss, and \textit{vice versa}.
-Black lines denote the probability values estimated from overall GC content (34\%, 42\% and 50\%) and the orange line denotes the probability values estimated from trimer and hexamer frequencies in \textit{D.
-melanogaster} intergenic genome.}
-\label{lossprob}
-\end{figure}
-
-Proto-genes that do not provide any fitness benefit to an organism may not be fixed in populations via natural selection.
-These genes may be lost due to mutation pressure.
-Some newborn proto-genes can also encode toxic proteins, that may aggregate or interfere with physiology in some other way.
-These genes would thus be eliminated from the population genomes via negative selection.
-We note that an ORF is lost if the start codon is mutated, the stop codon is mutated to an amino acid encoding codon (non-stop mutation), or if an amino acid encoding codon is mutated to a stop codon (premature stop/non-sense mutation).
-However, it is likely that a non-stop mutation or a non-sense mutation, can still result in translation of a protein (extended or truncated, respectively).
-Furthermore, non-stop mutations can also lead to cellular toxicity \citep{nonstop}.
-Thus ORF loss does not ensure elimination of toxic proteins, which in turn suggests that transcription loss may more effectively inactivate the associated genes.
-
-\cmnt{To understand which is the most probable mechanism of gene loss, we compared the probabilities of ORF loss ($P_\textit{ORF-loss}$) and transcription loss ($P_\textit{RNA-loss}$).
-We found that ORF loss is more probable than transcription loss, especially so when the ORFs are long (\hyperref[lossprob]{\autoref{lossprob}A}).
-If transcription is strictly dependent on a core promoter (TATA-box or Inr), small genes (30 codons) with low GC content ($>$34\%) can be preferentially lost by loss of transcription ({\color{blue} Figure S3}).
-When the organismal fitness is affected only by the protein product of a proto-gene, then both RNA loss and ORF loss would affect the fitness equally.
-Thus the likelihood of gene extinction through ORF loss or RNA loss depends solely on the probability of these events.
-
-Our analysis of gene loss mechanism considers the loss of one feature (for example transcription) without requiring the other feature to remain intact (for example the ORF).
-With more stringent analysis where we consider the loss of only one feature but not the other, we find a the that exclusive ORF loss is more likely than exclusive RNA loss ({\color{blue}Figure S5}).
-The difference between the likelihood of the two gene loss mechanisms is more remarkable when we consider exclusive loss of one feature.}
-
-Overall, our analysis suggests widespread transcription loss may be indicative of a negative selection on protein expression.
-
-\subsection*{Does mutation bias shape protein composition?}
-
-\begin{figure}[!t]
-\centering
-\includegraphics[scale=0.5]{Figures/proteinComposition.pdf}
-
-\caption{Mutation bias and GC-content shape protein composition.
-Panels \textbf{(A)}, \textbf{(B)} and \textbf{(C)} show expected frequency, gain probability (\autoref{aagaineq}), and loss probability (\autoref{aalosseq}) for different amino acids (horizontal axes).
-The vertical axis in all these three subplots show the normalized values of these measures such that their sum over all amino acids is equal to one.
-Panel \textbf{(D)} shows the directionality (colored tiles, $\textit{D}_{x\to y}$, \autoref{directionality}) of the substitution from an amino acid (vertical axis) to another (horizontal axis).
-Positive values of $\textit{D}_{x\to y}$ are shown in shades of red, whereas negative values are shown in shades of blue.
-We only show substitutions that can be realized by a single nucleotide mutation but we include all possible nucleotide mutations in the directionality calculations for these substitutions.
-In all panels, amino acids are ordered based on their chemical similarity \citep{PMBEC}.}
-\label{aafreq}
-
-\vspace{1ex}
-\hrule
-\end{figure}
-
-In previous sections, we showed that mutation bias affects the rate of \textit{de novo} gene emergence and loss.
-We next turned our attention to whether this bias affects the composition (and thereby the chemistry) of proteins encoded by proto-genes.
-To this end, we first asked if the expected frequency of different amino acids is uniform.
-For a GC-content of 42\% and with DNA trimer frequencies from \textit{D.
-melanogaster} intergenic genome, we found that it is not uniform.
-Amino acids like leucine (L) and serine (S) have a higher expected frequency than other amino acids.
-On the other hand, amino acids like methionine (M) and tryptophan (W) are less probable (\autoref{aafreq}{\color{blue}A}).
-This non-uniformity is to a great extent determined by number of degenerate codons for an amino acid.
-However, nucleotide composition also determines the frequency of an amino acid.
-For example, leucine, arginine and serine, all have six codons each.
-However, leucine is more likely to be encoded than serine in a random stretch of DNA, given a uniform GC-content of 42\% (\autoref{aafreq}{\color{blue}A}).
-The same is true for the frequencies of these amino acids estimated from the DNA trimer distribution.
-Overall, our analysis is roughly in agreement with many previous studies \citep{aasubOhta,aafreq3,aafreq4}.
-However, notable differences exist between our analysis and that of the previous studies because the latter primarily focused on characterized proteins and not random proto-gene derived protein sequences.
-
-Next, we aimed to find out if some amino acid substitutions are more probable than the others.
-To this end, we calculated the gain probability (\autoref{aagaineq}) for every amino acid.
-Specifically for every amino acid $x$, we calculated the average probability that it substitutes any of the other nineteen amino acids, due to mutations.
-More precisely, the total probability that an amino acid $y$ will mutate to an amino acid $x$ is the product of the expected frequency of amino acid $y$ ($P_y$), and the substitution rate from $x$ to $y$ ($\mu_{y\to x}$).
-The gain probability for an amino acid $x$ is then defined as:
-
-\begin{equation}
-G_x = \sum_{b \neq a} P_y \times \mu_{y\to x}
-\label{aagaineq}
-\end{equation}
-
-We found that amino acid gain probabilities were also non-uniform across different amino acids (\autoref{aafreq}{\color{blue}B}) but they had a strong positive correlation with the expected frequency (GC = 42\%, Pearson $\rho = 0.955, P<10^{-10}$; \textit{D.
-melanogaster} trimer frequencies, Pearson $\rho = 0.841, P<10^{-10}$).
-
-Next, we calculated how easily is an amino acid mutated to any other amino acid.
-Specifically, the conditional loss probability for an amino acid ($L_x$), given the amino acid already exists, is defined as:
-
-\begin{equation}
-L_x = \sum_{x \neq y} \mu_{x\to y}
-\label{aalosseq}
-\end{equation}
-
-The loss probability was also not uniformly distributed (\autoref{aafreq}{\color{blue}C}) but it did not significantly correlate with the expected frequency of the amino acids (GC = 42\%, Pearson $\rho = -0.255, P = 0.2818$; \textit{D.
-melanogaster} trimer frequencies, Pearson $\rho = -0.356, P=0.1232$).
-
-Next, we investigated if some amino acid substitutions are more common than the others.
-As expected, amino acid substitutions that require more than one nucleotide change were much less probable than those needing just a single nucleotide change.
-Therefore, we focused our analysis on amino acid substitutions whose total probabilities were more than that of a double nucleotide mutation (note that this procedure identifies amino acid substitutions that are possible via single nucleotide mutations but it does not exclude multinucleotide mutations from the calculation of amino acid substitution probabilities).
-Although single nucleotide mutations are more probable than double mutations, the different single mutations do not occur at the same rate due to mutation rate bias.
-Thus an amino acid substitution (for example K$\to$E) may not be as likely to occur as the reverse substitution (E$\to$K).
-That is substitutions between a pair of amino acids may have a directionality.
-Although many previous studies have explored the likelihood of different amino acid substitutions \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}, none to our knowledge have focused primarily on the directionality of these substitutions.
-To verify if such a directionality exists, we calculated the log transformed ratio of forward and reverse substitution probability, $D_{x\to y}$, for every pair of amino acids, $x$ and $y$.
-A positive value of $D_{x\to y}$ means that $x$ to $y$ substitution is more likely than $y$ to $x$ substitution, and \textit{vice versa}.
-
-\begin{equation}
-\textit{D}_{x\to y} = \log_2\left(\frac{\mu_{x\to y}}{\mu_{y\to x}}\right)
-\label{directionality}
-\end{equation}
-
-As we mentioned in the previous paragraph, we restricted this analysis to substitutions that can be achieved via single nucleotide mutations.
-We found that most substitutions indeed had a preferred direction ($\textit{D}_{xy} \neq 0$,  \autoref{aafreq}{\color{blue}D}).
-The median absolute value of $\textit{D}_{xy}$ was 0.905 and the maximum value was 2.836.
-That is, in 38 out of 75 amino acid pairs we analysed, one direction of substitution was at least 1.87 times more probable than the other.
-For example, E$\to$K substitution is 1.8 times more probable than K$\to$E substitution.
-This may appear as a small number but many such unidirectional substitutions occurring in an evolving protein could indicate a directional selection.
-\cmnt{We note that that actual frequencies of specific substitutions depend on the initial distribution of different codons in the ORF sequence.}
-
-A previous study found that mutations that cause hydrophobic amino acids to appear on protein surfaces, can lead to evolution of protein dimers \citep{hydrophobicRatchet}.
-The authors also suggested that such an evolutionary process may be widespread because mutation bias tends to facilitate emergence of hydrophobic amino acids.
-They based this argument on another study that estimated mutation rate bias in bacteria \citep{bacteriaMutbias}.
-The authors \citep{hydrophobicRatchet} further suggested that hydrophobic amino acids are more frequently found in random protein sequences.
-
-We asked if our model also makes similar predictions.
-To this end, we first calculated the percentage of amino acids in a protein sequence that are expected to be hydrophobic.
-A popular hydrophobicity scale of amino acids is based on solubility of an amino acid in water or ethanol \citep{hydropathy}.
-However, this scale does not classify tryptophan as a hydrophobic amino acid.
-Another hydrophobicity scale estimated by a different study \citep{WWhydropathy} is more biologically realistic and is based on rates of transfer of different amino acids between water and a hydrophobic medium.
-This study analysed two hydrophobic media -- a lipid bilayer and octanol, and also considered the effect of peptide bonds in the calculations.
-Therefore we used the hydrophobicity scale by \cite{WWhydropathy} for classifying amino acids such that hydrophobic amino acids have a negative free energy change of transfer from water to hydrophobic media.
-Conversely, a hydrophilic amino acid would have a positive value of the free energy change.
-Based the octanol hydrophobicity scale, we classified the following amino acids as hydrophobic -- cysteine (C), phenylalanine (F), isoleucine (I), leucine (L), methionine (M), valine (V), tryptophan (W) and tyrosine (Y).
-With the lipid bilayer hydrophobicity scale, valine is not classified as hydrophobic, possibly because it does not integrate well in the bilayer.
-Therefore we used the octanol scale for our calculations.
-
-We found that $\sim$40\% of amino acids in a protein sequence are expected to be hydrophobic (with a GC-content of 42\% as well as with \textit{D.
-melanogaster} trimer distribution).
-This finding is broadly in agreement with the previous study \citep{hydrophobicRatchet}.
-We note that cytosolic proteins may require more than 42\% of their constituent amino acids to be hydrophobic, in order to fold efficiently \citep{Dill1985}.
-Next, we asked if a protein sequence in general tends to be hydrophobic.
-Specifically, we calculated expected hydrophobicity ($\bar{\alpha}$, \autoref{expHydropathy}) of a protein sequence, based on expected frequency of different amino acids ($P_x$, \autoref{aafreq}) and their hydrophobicity values ($\alpha_x$).
-
-\begin{equation}
-\bar{\alpha} = \sum_{x} P_x \alpha_x
-\label{expHydropathy}
-\end{equation}
-
-We found that the expected hydropathy is higher than zero (0.38 with GC\% = 42; 0.36 with \textit{D.
-melanogaster} trimer frequencies), which suggests that random protein sequences are not on an average, hydrophobic.
-
-To find out if random mutations indeed cause hydrobphobic amino acids to occur in protein sequences, we calculated the probability that any mutation substitutes a non-hydrophobic amino acid with a hydrophobic amino acid ($P_{\alpha\textit{-gain}}$, \autoref{alphagain}).
-Likewise, we calculated the probability that any mutation causes a hydrophobic to non-hydrophobic amino acid substitution ($P_{\alpha\textit{-loss}}$, \autoref{alphaloss}).
-
-\begin{align}
-P_{\alpha\textit{-gain}} & = \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_y \times \mu_{y\to x}\label{alphagain}\\[1em]
-%
-P_{\alpha\textit{-loss}} &= \sum_{\genfrac{}{}{0pt}{1}{x}{\,|\,\alpha_x>0}} \sum_{\genfrac{}{}{0pt}{1}{y}{\,|\,\alpha_y \geq 0}} P_x \times \mu_{x\to y}\label{alphaloss}
-\end{align}
-
-Next, we calculated the ratio of $P_{\alpha\textit{-gain}}$ and $P_{\alpha\textit{-loss}}$, and found that it is slightly more than 1 (1.114 with GC\% = 42; 1.015 with \textit{D.
-melanogaster} trimer frequencies).
-This suggests that gain of a hydrophobic amino acid is slightly more probable than its loss.
-We performed a complementary analysis where we calculated the average change in hydrophobicity due to any random mutation ($\bar{\Delta\alpha}$), defined by:
-
-\begin{equation}
-\bar{\Delta\alpha} = \frac{\displaystyle\sum_{x} \sum_{y} \mu_{x\to y} \times (\alpha_y - \alpha_x)}{\displaystyle\sum_{x} \sum_{y} \mu_{x\to y}}
-\end{equation}
-
-We found the average change in hydrophobicity to be slightly less than zero ($-0.0798$), which suggests that on an average mutations cause a very small increase in hydrophobicity.
-We reiterate that hydrophobic amino acids have a negative value hydrophobicity and thus a negative change in hydrophobicity also denotes a shift towards a more hydrophobic protein.
-To focus exclusively on non-hydrophobic to hydrophobic amino acid substitutions, we excluded amino acid pairs where both amino acids are either hydrophobic or are non-hydrophobic.
-That is, we only analysed pairs of amino acids that have opposite signs of hydrophobicity.
-With this focused analysis, we find that the average change in hydrophobicity is comparable ($-0.0793$).
-For both the analyses, we note that the actual change in hydrophobicity will depend on the composition of the ancestral protein sequence.
-
-Overall our analyses show that random mutations may preferentially cause hydrophobic amino acids to accumulate in protein sequence but this effect may be very small.
-
 \section*{Discussion}
-
-In this work, we addressed some fundamental questions about \textit{de novo} emergence of protogenes.
-Broadly, we asked how quickly these genes can emerge and be lost, whether their birth and death has a preferred trajectory of mutational events, and how the composition of their protein sequence is determined.
-To this end, we developed mathematical models and used them to address specific questions.
-These models are based on mutation probability and mutation rate bias, and represent neutral evolution where DNA sequences are not evolving under natural selection.
-Thus our models provide an opportunity to test the alternative hypothesis that some proto-genes are evolving under selection.
-
-We show that when a protein product of a proto-gene doesn't affect organismal fitness, the gene can be lost much more rapidly than they are gained, such that it can be independently lost two to three times in the time required for it to emerge (\hyperref[gainlossprob]{\autoref{gainlossprob}A}).
-In a hypothetical case, a proto-gene may be present in a population (A) but is absent in two other geographically isolated populations (B and C).
-Two evolutionary scenarios could explain this observation.
-In the first scenario, the proto-gene simply emerged for the first time in population A.
-The second scenario posits that the proto-gene was present in the ancestral population but was lost in both population B and population C.
-Our models suggest that the second scenario is more likely than is assumed.
-\cmnt{This inference can be further extended to the level of closely related species where the divergence times are relatively shorter than the timescale of multiple gain and loss events within a lineage}.
-Future phylogenetic studies on proto-genes should consider this finding for inferring the dynamics of gene gain and loss.
-\cmnt{Proto-genes that encode proteins that are beneficial to an organism's fitness are likely to be protected in the genome against gene loss \citep{Lee2019}.
-Our models indeed show that loss of even moderately beneficial genes is unlikely in large populations (\hyperref[gainlossprob]{\autoref{gainlossprob}B})}.
-
-Within the limitations of our model assumptions, we answer the long standing question of which trajectory of mutational events leads to \textit{de novo} emergence.
-That is, whether ORF emerges first or transcription emerges first \citep{EBB-F1000}.
-We show that, in the absence of selection, genes are likely to emerge in existing transcripts (transcription emerges first, \autoref{whoisfirst}).
-Long non-coding RNAs may be a potential source of proto-genes but this evolutionary process may be constrained by a different set of factors.
-A previous study has shown that new transcripts frequently arise in regions overlapping existing genes.
-These new transcripts can harbor ORFs which in turn leads to \textit{de novo} gene emergence \citep{albaYeastdenovo}.
-This is an example of RNA-first trajectory.
-However, the ORFs present in these genes may be under indirect selection pressure that acts on the overlapping gene sequences.
-Therefore, they may be less frequently mutated or lost.
-
-Although our model is based on the assumption of neutral evolution, one can speculate the effects of positive selection on the proto-gene.
-For example, if an ORF encodes a protein that is beneficial to the organism, then selection would cause the fixation of the gene as soon as the it emerges via RNA-first trajectory.
-On the other hand, the selection would have no effect on an ORF that is not transcribed, as long as it is does not harbor a functional DNA sequence (such as regulatory elements), and is not under indirect selection (for example, if it overlaps with an existing gene).
-The ORF may mutate before transcription emerges and the corresponding mutated protein may not be beneficial anymore.
-Thus positive selection may further increase the likelihood of the RNA-first trajectory.
-
-We show that transcription loss is less likely than the loss of long ORFs (\autoref{lossprob}).
-However, it may be difficult to find direct evidence for such a phenomenon.
-Specifically, it is difficult to infer from data if transcription was lost or never emerged.
-If the protein encoded by an ORF is indeed toxic, it would be silenced or purged rapidly in microevolutionary time scales.
-One way of inferring negative selection from transcription loss may involve correlation of ORF sequence variation and transcription status.
-Mutations within an ORF sequence are much more probable than mutations that disrupt transcription or an ORF.
-This is so because the entire ORF sequence can have a higher number of mutable sites than a promoter.
-If an ORF variant is untranscribed in many populations (species or taxa), it is possibly toxic.
-
-A large volume of work exists on the likelihood of amino acid substitutions and many of these models are widely used in phylogenetics studies \citep{aasubOhta,PAM,blosum,submat92,submat92j,submat01,submat05,submat07,submat08}.
-However, commonly used amino acid substitution models such as PAM and BLOSUM \citep{PAM,blosum} do not focus on directionality of mutations (the substitution matrices are symmetric).
-Our aim was to highlight how proteins can evolve under mutation pressure and that mutation bias causes some substitutions to occur more frequently than the others.
-Furthermore, we do not base our substitution probability calculations on known protein sequences but rather on random protein sequences that are more likely to to be encoded in proto-genes.
-Using our models, we show that for a pair of amino acids, one direction of substitution can be significantly more likely than the other.
-If a substitution occurs more frequently than it is expected to occur in our neutral model, then directional evolution can be a possible explanation for the observation.
-We also show that random mutations can lead to a small increase in a protein's hydrophobicity.
-Hydrophobic amino acids can facilitate folding \citep{Dill1985}, but can also cause proteins to aggregate \citep{hydrophobicRatchet}.
-Thus the effect of hydrophobicity increasing mutations on a protein's stability and toxicity, is dependent on where they occur in the protein structure.
-
-Like all theoretical models, our models are not without limitations.
-First, we assume that nucleotides are uniformly distributed according to GC-content.
-This is not true as nucleotide composition can vary significantly across the genome.
-\cmnt{Since our model primarily focuses on gene emergence at one locus, it can be used to predict evolutionary probabilities for different genomic loci.
-Furthermore, several predictions of our models are not qualitatively affected by GC-content.
-More realistic nucleotide composition distribution estimated from \textit{D.
-melanogaster} intergenic sequences, also qualitatively agree with GC-content based predictions.
-Thus, most predictions are qualitatively robust to small differences in parameters.
-Second, in our models, we assumed uniform mutation rate throughout the genome.
-However, recent studies have shown that mutation rates are also not uniform across the genome, and can vary significantly \citep{mutbiasArabidopsis}}.
-Our models can be adapted to investigate specific genomic loci, where estimates of the local nucleotide composition and mutation rate can more accurately determine the \textit{de novo} gene evolution in these loci.
-Third, our models are based on parameters such as mutation rate and mutation bias, which we obtained from published data on human and \textit{D.melanogaster} genomes \citep{humanmutrate,drosophilamutrate}.
-The accuracy of the models' prediction will depend on the accuracy of parameters.
-However, mutation rate itself only determines the values of different probabilities.
-Our results, which are based on probability ratios, will not significantly change with a different value of mutation rate.
-Mutation bias may have more significant effect on the predictions than mutation rate.
-Existing data suggest that different species have a different spectrum of mutations \citep{joshmutbias}, especially for species that are highly distant from each other.
-Thus our inferences based on our chosen parameters will not universally apply to all species.
-Fourth, we assume that occurrence of a point mutation is independent of another mutation.
-This is not the case for tandem mutations, where multinucleotide mutations occur more frequently than their expected frequency under the assumption of mutational independence \citep{MNM}.
-Fourth, we did not specifically model the effect of many regulatory sequence features such as enhancers, transcription factor binding sites, 5' and 3' untranslated regions, and Kozak consensus sequence, that can influence the evolution of proto-genes.
-For example \textit{de novo} emergence is very likely to occur near enhancers \citep{PacoEnhancers}.
-\cmnt{To model this phenomenon in our analyses, we assumed that DNA loci feasible for transcript emergence are close to known \textit{cis-} regulatory elements \citep{RedFly2005,RedFly2007,RedFly2010,RedFly2018,RedFly2022}.
-Specifically, we assumed that transcription can be initiated at a probability that is equal to the proportion of such regulatory elements in the intergenic genome.
-This is a rough estimate but it can be updated when more specific data becomes available.
-For example, ATAC-seq data can provide reliable estimates of nucleosome depleted regions, that can act as transcription initiation sites \citep{ATAC-seq}.
-This analysis can be supplemented with data from techniques like native elongating transcript sequencing \citep[NET-seq;][]{NET-seq} and global run-on sequencing \citep[GRO-seq;][]{GRO-seq}, that can detect nascent transcripts.} Our modeling framework also allows incorporation of any number of sequence features in the calculations as long as their sequences can be defined (\autoref{methfeatures}).
-Finally, we ignore insertions, deletions and transpositions as mechanisms of mutation.
-Insertions and deletions (indels) are reported to be less frequent than substitutions \citep{drosophilamutrate}, and both indels and transpositions cause larger sequence alterations than point mutations.
-Our modeling methodology also allows incorporation of these mechanisms as a source of mutations.
-
-A famous aphorism about theoretical models says ``all models are wrong but some are useful'' \citep{GEBox}.
-Our models are not an exception.
-They may not be 100\% accurate but they make useful predictions that can lead to a more focused experimental validation.
-\cmnt{Many of our predictions are qualitatively robust to small parameter changes.
-Some predictions are also in agreement with experimental data.
-For example, our prediction of transcript gain rate is only slightly inflated than the estimated rates using an empirical analysis of \textit{D.
-melanogaster} populations \citep{AnnaPeteTranscripts}.
-This difference could have possibly resulted from our relaxed requirements for a promoter, or other unknown factors that may have affected population evolution \citep[for example, the actual effective population size and environmental conditions;][]{dmel3decades}.
-When supplemented with experimental data analysis our models can be made more complex to accommodate diverse molecular mechanisms driving gene evolution, and can provide more accurate predictions.} Therefore our work opens up an opportunity for theoretical and computational biologists to design analyses using our modeling framework, that more accurately describe their system of interest.
 
@@
 
-\subsection{Probabilities of finding, gaining, and losing a DNA sequence}
-\label{methbasic}
-
-\subsubsection{Probability of finding a DNA sequence}
-\label{methprob}
-
-We calculated the probability of finding a DNA sequence based on global nucleotide frequency distributions, given by the GC-content.
+%\subsection{Probabilities of finding, gaining, and losing a DNA sequence}
+%\label{methbasic}
+%
+%\subsubsection{Probability of finding a DNA sequence}
+%\label{methprob}
+%
+%We calculated the probability of finding a DNA sequence based on global nucleotide frequency distributions, given by the GC-content.
 Specifically, the probability of finding either a \texttt{G} or a \texttt{C} is:
-
-$$S = \frac{0.5\times\text{GC\%}}{100}$$
-
-The probability of finding an \texttt{A} or a \texttt{T} is: $W = 0.5 - S$.
+%
+%$$S = \frac{0.5\times\text{GC\%}}{100}$$
+%
+%The probability of finding an \texttt{A} or a \texttt{T} is: $W = 0.5 - S$.
 Using these values, we calculated the probability of finding a DNA sequence motif by chance.
 For example, the probability of finding the sequence \texttt{ATG} would be: $W\times W \times S$.
-
-\cmnt{We also estimated the probability of finding specific DNA sequences in a reference genome.
+%
+%\cmnt{We also estimated the probability of finding specific DNA sequences in a reference genome.
 Specifically, we calculated the frequencies of all 64 trimers and all 4096 hexamers in the genomic regions of \textit{D.
 melanogaster} that exist in open chromatin, and do not contain any known genes or regulatory elements (see \autoref{methRNA})}.
-
-\subsubsection{Probability of gaining a DNA sequence}
-\label{methgain}
-
-We calculated the probability of gaining a DNA sequence due to mutations using GC-content, mutation rate and mutation bias.
+%
+%\subsubsection{Probability of gaining a DNA sequence}
+%\label{methgain}
+%
+%We calculated the probability of gaining a DNA sequence due to mutations using GC-content, mutation rate and mutation bias.
 Specifically, we calculated the probability that a DNA sequence does not exist, and it emerges due to specific nucleotide mutations.
@@
 Overall, the chance of \texttt{CTA} mutating to \texttt{ATG} would be:
-
-$$SW^2 \times \lambda^2 \times \mu_{\texttt{G:C}\to\texttt{T:A}} \times \mu_{\texttt{A:T}\to\texttt{G:C}}$$
-
-Next, we calculated the probability that every nucleotide triplet that is not \texttt{ATG}, mutates to \texttt{ATG}.
+%
+%$$SW^2 \times \lambda^2 \times \mu_{\texttt{G:C}\to\texttt{T:A}} \times \mu_{\texttt{A:T}\to\texttt{G:C}}$$
+%
+%Next, we calculated the probability that every nucleotide triplet that is not \texttt{ATG}, mutates to \texttt{ATG}.
 This can happen via one, two, or three nucleotide mutations.
 The sum of all these mutation probabilities is the probability of \texttt{ATG} gain.
-
-Using the same principle we calculated the gain probability of any DNA sequence motif (of any length or composition).
+%
+%Using the same principle we calculated the gain probability of any DNA sequence motif (of any length or composition).
 We excluded insertions and deletions as a mechanism of gain of small DNA sequences that we analysed in this study.
-
-\subsubsection{Probability of losing a DNA sequence}
-
-We calculated the loss of a DNA sequence motif using the same principle we used for calculating gain probabilities.
+%
+%
+%\subsubsection{Probability of losing a DNA sequence}
+%
+%We calculated the loss of a DNA sequence motif using the same principle we used for calculating gain probabilities.
 However, we defined loss probability as a conditional probability, that is we assume that the DNA sequence of interest already exists in the genome.
@@
 We use conditional loss probabilities by default, because usually one is interested in finding out how quickly an existing DNA sequence can erode.
-
-We used this method to calculate the loss probability of any DNA sequence motif, and we excluded insertions and deletions from this calculation.
-
-\subsection{Probabilities of finding, gaining, and losing DNA features}
-\label{methfeatures}
-
-Usually, a specific function is encoded in DNA by several DNA sequences.
+%
+%We used this method to calculate the loss probability of any DNA sequence motif, and we excluded insertions and deletions from this calculation.
+%
+%\subsection{Probabilities of finding, gaining, and losing DNA features}
+%\label{methfeatures}
+%
+%Usually, a specific function is encoded in DNA by several DNA sequences.
 For example, translation stop is encoded by three codons (\texttt{TGA}, \texttt{TAG}, \texttt{TAA}).
@@
 For example, the non-feature set of stop codons would be all the other 61 codons.
-
-The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set (\autoref{methprob}).
-
-The probability that a DNA feature is gained via mutations, is the sum of probabilities of every non-feature sequence mutating to any feature sequence.
+%
+%The probability that a DNA feature exists, is the sum of probabilities of every DNA sequence in that set (\autoref{methprob}).
+%
+%The probability that a DNA feature is gained via mutations, is the sum of probabilities of every non-feature sequence mutating to any feature sequence.
 If $F$ denotes the feature set, and $\mu_{y\to x}$ denotes the probability of a DNA sequence $y$ mutating to a DNA sequence $x$ (see \autoref{methgain}), then:
-
-\begin{equation}
-P_\textit{feature-gain} = \sum_{x \in F} \sum_{y \notin F} P_y \times \mu_{y\to x}
-\end{equation}
-
-The probability that a DNA feature is lost via mutations is a conditional probability that given a feature exists, it mutates to any of the non-feature sequences.
-
-\begin{equation}
-P_\textit{feature-loss} = \frac{\displaystyle\sum_{x \in F} \sum_{y \notin F} P_x \times \mu_{x\to y}}{\displaystyle\sum_{x \in F} P_x}
-\end{equation}
-
-Because a feature set usually has many DNA sequences, a mutation can change a feature sequence such that the resulting sequence is also a part of the feature set.
+%
+%\begin{equation}
+%P_\textit{feature-gain} = \sum_{x \in F} \sum_{y \notin F} P_y \times \mu_{y\to x}
+%\end{equation}
+%
+%The probability that a DNA feature is lost via mutations is a conditional probability that given a feature exists, it mutates to any of the non-feature sequences. 
+%
+%\begin{equation}
+%P_\textit{feature-loss} = \frac{\displaystyle\sum_{x \in F} \sum_{y \notin F} P_x \times \mu_{x\to y}}{\displaystyle\sum_{x \in F} P_x}
+%\end{equation}
+%
+%Because a feature set usually has many DNA sequences, a mutation can change a feature sequence such that the resulting sequence is also a part of the feature set.
 Thus we defined the probability ($P_\textit{feature-stay}$) that a feature does not erode because of mutations.
@@
 First is the probability that no mutation occurs in the DNA sequence ($P_0$), and the second probability describes the event where the mutated sequence remains a part of the feature set.
-
-\begin{equation}
-P_\textit{feature-stay} = P_0 + \sum_{x \in F} \sum_{\genfrac{}{}{0pt}{1}{y \in F}{y\neq x}} \mu_{x\to y}
-\end{equation}
-
-The probability that no mutation occurs ($P_0$) in a DNA sequence of length $k$ is described by Poisson distribution.
-
-$$P_0 = 1-e^{-k\lambda}$$
-
-Because the mutation rate is biased (\autoref{mutbias}), the probability that no mutation occurs in a DNA sequence depends on its composition.
-
-The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is thus described as:
-
-$$\lambda_\texttt{AT} = 2\times\lambda\times(\mu_{\texttt{A:T}\to\texttt{T:A}} + \mu_{\texttt{A:T}\to\texttt{G:C}} + \mu_{\texttt{A:T}\to\texttt{C:G}})$$
-
-Likewise, the probability that a \texttt{G} or a \texttt{C} mutates ($\lambda_\texttt{GC}$) is:
-
-\vspace{-1ex}
-
-$$\lambda_\texttt{GC} = 2\times\lambda\times(\mu_{\texttt{G:C}\to\texttt{A:T}} + \mu_{\texttt{G:C}\to\texttt{T:A}} + \mu_{\texttt{G:C}\to\texttt{C:G}})$$
-
-(Note that the general mutation rate, $\lambda$, is an average of $\lambda_\texttt{AT}$ and $\lambda_\texttt{GC}$.)
-
-\vspace{1\baselineskip}
-
-Thus the probability that a sequence of length $k$, containing $m$ number of \texttt{A} and \texttt{T}, does not mutate is:
-
-\vspace{-1ex}
-
-$$P_0 = (1-e^{-m\lambda_\texttt{AT}})\times(1-e^{-(k-m)\lambda_\texttt{GC}})$$
-
-\cmnt{We also calculated all the above-defined probabilities ($P_\textit{feature-gain}$, $P_\textit{feature-loss}$ and $ P_\textit{feature-stay}$), using DNA trimer and hexamer frequencies from \textit{D.
+%
+%\begin{equation}
+%P_\textit{feature-stay} = P_0 + \sum_{x \in F} \sum_{\genfrac{}{}{0pt}{1}{y \in F}{y\neq x}} \mu_{x\to y}
+%\end{equation}
+%
+%
+%The probability that no mutation occurs ($P_0$) in a DNA sequence of length $k$ is described by Poisson distribution. 
+%
+%$$P_0 = 1-e^{-k\lambda}$$
+%
+%Because the mutation rate is biased (\autoref{mutbias}), the probability that no mutation occurs in a DNA sequence depends on its composition. 
+%
+%The probability that an \texttt{A} or a \texttt{T} mutates ($\lambda_\texttt{AT}$), is thus described as:
+%
+%$$\lambda_\texttt{AT} = 2\times\lambda\times(\mu_{\texttt{A:T}\to\texttt{T:A}} + \mu_{\texttt{A:T}\to\texttt{G:C}} + \mu_{\texttt{A:T}\to\texttt{C:G}})$$
+%
+%Likewise, the probability that a \texttt{G} or a \texttt{C} mutates ($\lambda_\texttt{GC}$) is: 
+%
+%\vspace{-1ex}
+%
+%$$\lambda_\texttt{GC} = 2\times\lambda\times(\mu_{\texttt{G:C}\to\texttt{A:T}} + \mu_{\texttt{G:C}\to\texttt{T:A}} + \mu_{\texttt{G:C}\to\texttt{C:G}})$$
+%
+%(Note that the general mutation rate, $\lambda$, is an average of $\lambda_\texttt{AT}$ and $\lambda_\texttt{GC}$.)
+%
+%\vspace{1\baselineskip}
+%
+%Thus the probability that a sequence of length $k$, containing $m$ number of \texttt{A} and \texttt{T}, does not mutate is:
+%
+%\vspace{-1ex}
+%
+%$$P_0 = (1-e^{-m\lambda_\texttt{AT}})\times(1-e^{-(k-m)\lambda_\texttt{GC}})$$
+%
+%
+%\cmnt{We also calculated all the above-defined probabilities ($P_\textit{feature-gain}$, $P_\textit{feature-loss}$ and $ P_\textit{feature-stay}$), using DNA trimer and hexamer frequencies from \textit{D.
 melanogaster}.
@@
 Trimers and hexamers would contain codons and polyA signals, respectively.}
+%
 
@@
 
-\subsection{Probabilities of finding, gaining, and losing transcription}
-
-\label{methRNA}
-
-\subsubsection{Probability of transcription}
-
-\cmnt{In this study we defined a transcription model that is not dependent on specific promoter sequences.
-Instead we defined a probability that transcription is initiated at a genomic locus.
-To this end, we first extracted the genomic coordinates of intergenic sequences in \textit{D.
-melanogaster} genome \citep[release 6.49, FlyBase;][]{flybase}.
-Using the data on the genomic locations of open chromatin in \textit{D.
-melanogaster} \citep{dmel_Chromatin}, we obtained open chromatin regions within intergenic genome.
-Next, we calculated the fraction of this open-intergenic genome occupied by \textit{cis-} regulatory elements \citep{RedFly2005,RedFly2007,RedFly2010,RedFly2018,RedFly2022}.
-A significant proportion of these elements include enhancers.
-Enhancers, that can promote transcription in their neighboring genomic regions \citep{enhancers}, are also known to promote \textit{de novo} gene emergence \citep{PacoEnhancers}.
-Thus we reasoned that proportion of \textit{cis-} regulatory elements in open intergenic chromatin would roughly approximate the likelihood of transcription initiation ($P_\textit{RNAinit}$).
-We do not model the exact sequences because data is not yet available on exact sequence requirements and mutational robustness of enhancers and other \textit{cis-} regulatory elements.
-Thus poly-A signal is the primary sequence determinant of transcription probabilities.
-Relatedly, open intergenic chromatin regions, that are not interrupted by \textit{cis-} regulatory elements are the potential sites for \textit{de novo} gene emergence.
-
-A poly-A signal marks the end of a transcript.
-The transcribed DNA region should thus have only one poly-A signal at its end.} A transcribed DNA of length $l$ nucleotides, contains $l-5$ hexamers (6nt long sequences).
-Each of these hexamers can be a poly-A signal with a probability, $P_\textit{polyA}$.
-Probability that none of these hexamers is a poly-A signal is $(1-P_\textit{polyA})^{l-5}$.
-Since we focus on protein coding genes in this work, we describe the length of transcript in terms of the length of protein it can encode.
-Assuming that untranslated regions have no effect on protein synthesis, the length of a transcript in number of nucleotides should be at least three times the length of the encoded protein in number of amino acids.
-Thus the probability that a DNA region produces a transcript that can harbor an ORF with $k$ codons ($P_\textit{RNA}(k)$) is:
-
-\begin{equation}
-P_\textit{RNA}(k) = P_\textit{RNAinit} \times P_\textit{polyA} \times (1-P_\textit{polyA})^{3k-5}
-\label{eqrnaprob}
-\end{equation}
-
-\subsubsection{Probability of gaining transcription}
-
-\cmnt{As we mentioned in the previous section, two requirements need to be met to produce a transcript of any given length.
-First, a poly-A signal needs to be present at the end of the DNA region to be transcribed, and second, no poly-A signal should exist within this DNA region.
-Transcription can be gained if one of these requirements are already met and the third required feature emerges due to mutations.
-At the same time mutations should not destroy the existing features.
-It is possible that both the required features are missing and they emerge simultaneously due to mutations.
-However such an event is highly improbable.
-Thus transcription gain can occur via two mechanisms.
-In the first mechanism a poly-A signal is gained at the end of a sequence ($P_\textit{polyA-gain}$), while no poly-A sites are present at any $l-5$ (or $3k-5$) sites within the sequence, and none emerge ($P_\textit{nopolyA-stay} = 1 - P_\textit{polyA} - P_\textit{polyA-gain}$).
-In the second mechanism, a poly-A signal is present at the end of the sequence, and is not lost due to mutations.
-Meanwhile, one poly-A signal at any of the $3k-5$ sites within the sequence is lost ($P_\textit{polyA-loss}$), whereas the other $3k-6$ sites do not encode a poly-A signal and continue to remain so.
-
-The combined probability of these two transcription gain mechanism is defined as:
-
-\begin{equation}
-P_\textit{RNA-gain}(k) = P_\textit{RNAinit} \times \begin{pmatrix}
-P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \\[1em]
- +\  P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-6} \times (3k-6) \times P_\textit{polyA-loss}
-\end{pmatrix}  
-\label{eqrnagain}
-\end{equation}
-
-\subsubsection{Probability of losing transcription}
-
-Transcription is lost when either the poly-A signal is lost or if one is gained within the transcribed sequence.
-The probability of transcription loss ($P_\textit{RNA-loss}$), given that already transcription exists, is thus defined as:
-
-\begin{equation}
-P_\textit{RNA-loss}(k) = P_\textit{polyA-loss} + (3k-5)\times P_\textit{polyA-gain}
-\label{eqrnaloss}
-\end{equation}
-
-\subsubsection{Probability that transcription remains intact}
-
-Transcription remains intact if neither the poly-A signal is not lost and none is gained within the sequence. Probability of this event ($P_\textit{RNA-stay}$) is defined as:
-
-\begin{equation}
-P_\textit{RNA-stay}(k) = P_\textit{RNAinit} \times P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
-\label{eqrnastay}
-\end{equation}
-}
-\subsection{Interdependence of transcription and ORF probabilities}
-
-\label{rnaorfindependence}
-
-All the four poly-A signal variants contain a \texttt{TAA} in their sequence (position 3 -- 5), which is also a stop codon.
-Since stop codons cannot exist inside an ORF, presence of an ORF reduces the possible number of poly-A signal sites in a DNA region.
-Specifically, a poly-A signal can exist in any of the three reading frames but it can overlap with the ORF only in one frame.
-Thus, out of $3k-5$ positions for a 6-mer (length of a poly-A signal) in an ORF with $k$ codons, $k$ positions cannot harbor a poly-A signal (\autoref{eqrnaprob}).
-Furthermore a poly-A signal cannot overlap with the start codon, thus reducing two more positions from all sites where a poly-A signal can exist (one position is already counted -- if a poly-A signal overlaps with the first codon in the second frame, then the second codon becomes a stop codon).
-Overall, the possible number poly-A sites on an ORF with $k$ codons is $2k-5$.
-Thus for a gene to emerge via gain of transcription, the DNA should not contain poly-A sites at these $2k-5$ positions.
-As a consequence, the probability of transcription gain for a specific transcript length, is higher when the DNA region contains an ORF than when it does not (\autoref{eqrnagain}).
-We have ignored the probability of poly-A signal overlapping with a stop codon because it makes a very small difference in the probability of RNA gain.
-This is so because at least one stop codon (\texttt{TAA}) allows a poly-A signal to overlap with it in three frames.
-
-Conversely to the effect of an existing ORF on transcription gain, stop codons are less probable in an existing transcript that does not contain any poly-A signal sequences within its sequence, than in an untranscribed DNA region.
-Specifically, the probability that a 6-mer contains a stop codon in the third position is same as the probability of finding a stop codon ($P_\textit{stop}$).
-If poly-A signals are excluded from these 6-mers, then the probability of finding a stop codon is $P_\textit{stop} - P_\textit{poly-A}$.
-Conversely, lack of poly-A signals will increase the probability of amino acid coding codons and thereby their probability of mutating into a stop codon.
-Furthermore, the likelihood that an ORF is undisrupted my premature nonsense mutations, is lower when the DNA is already transcribed.
-Specifically:
-
-$$P_\textit{stop-gain \vline\ RNA} = \frac{P_\textit{stop-gain}\times(1- P_\textit{stop} + P_\textit{poly-A})}{1- P_\textit{stop}}$$
-
-Overall the evolutionary dynamics of transcription and ORF gain, are not independent of each other.
-That is, presence of one feature makes the gain of the other more likely.
-The dependence becomes more prominent with increasing number of codons in the ORF of the proto-gene.
-
-\subsection{Calculation of fixation probabilities}
-
-\newcommand{\specialcell}[2][c]{%
-  \bfseries\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-\cmnt{
-We estimated gene fixation probabilities using Kimura's model \citep{KimuraFix}. Specifically we defined the probabilities of fixation and extinction of a proto-gene as shown in \autoref{pFix}:}
-
-\begin{table}[H]
-\centering
-\begin{tabular}{c c c c}
-\toprule
-\textbf{Case} & \specialcell{Gain of gene\\(Fixation, $P_\textit{fix}$)}  & \specialcell{Loss of gene\\ (Extinction, $P_\textit{ext}$)} & \specialcell{Selection\\coefficient ($s$)}\\
-\midrule
-No selection on protogene & $\dfrac{1}{2N_e}$ & $\dfrac{1}{2N_e}$ & -- \\
-\midrule
-Selection on protogene & $\dfrac{1 - e^{-2s}}{1- e^{-4N_es}}$ & $\dfrac{1 - e^{2s}}{1- e^{4N_es}}$ & $\pm0.001$\\
-\bottomrule
-\end{tabular}
-\caption{Gene fixation/extinction probabilities with or without selection.
-Here $N_e$ denotes effective population size and $s$ denotes the selection co-efficient denoting the fitness advantage provided by the protein encoded in the proto-gene (positive valued for a beneficial protein, and negative valued for a deleterious protein).}
-\label{pFix}
-\end{table}
-
-\subsection{Probability of amino acid substitutions}
-
-We calculated the probability of amino acid substitutions based on nucleotide substitution in the corresponding codons.
-In this case, multiple codons that code for the same amino acid constitute a feature set (\autoref{methfeatures}).
-Similarly when analysing hydrobhobic to non-hydrophobic substitutions, all codons that represent hydrophobic amino acids form a feature set (and \textit{vice versa}).
-
 \section*{Data availability}
@@
 \section*{Acknowledgments}
-
-We thank Pete Czuppon (IEB, University of M\"{u}nster), for discussions on mathematical models.
 

### 2023-03-08 Manuscripts/M2/antisense_genes.tex
new file:   Manuscripts/Figures/M2_main/frames.svg 	modified:   Manuscripts/M2/antisense_genes.tex

@@
 
-\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
+\title{Antisense transcripts can quickly evolve to encode proteins}
 \author{Bharat Ravi Iyengar$^{1,\dagger}$, Erich Bornberg-Bauer$^{1,2}$}
@@
 \onehalfspacing
+\def\figdir{../Figures/M1_main}
 
@@
 \section*{Results}
+
+We developed mathematical models to estimate the probabilities of ORF emergence and loss, in DNA regions antisense to existing protein coding ORFs.
+For any stretch of DNA to be an ORF, its sequence should contain 3$n$ nucleotides (with $n$ being at least 3), with a start codon that marks its beginning, and exactly one stop codon that marks its end.
+Specifically, the absence of any stop codons within the DNA sequence is the most important factor in determining the likelihood of an ORF to exist.
+That is because the likelihood of a premature stop codon increases exponentially with the ORF's length, whereas likelihood of a start codon and a terminal stop codon are unaffected by the ORF's length.
+
+We estimate the gain and loss probabilities of antisense ORFs using the existing (sense) ORF as a reference.
+Specifically, we analyse the effect of the composition of the sense ORF, and of mutations in its sequence, to determine how likely it is for an antisense ORFs to exist, and to be gained or lost.
+Antisense ORFs can overlap with the sense ORFs in three different frames.
+In frame 0, the codons in the antisense ORF exactly overlap the codons in the sense ORF.
+In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of sense ORF by one and two nucleotide positions, respectively.
+In frame 0, the sequence of an antisense codon is determined by the sequence of the exactly overlapping sense codon.
+However, for frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons.
 
@@
 We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
-Specifically, our model is implemented in the script \texttt{DeNovoProb.jl} which in turn uses the script \texttt{nucleotidefuncts.jl} for some basic functions.
+Specifically, our model is implemented in the scripts \texttt{antisenseGenes.jl}, and  \texttt{antisenseGenes\_supplement.jl}, which in turn depend on the script \texttt{nucleotidefuncts.jl} for some basic functions.
 

### 2023-03-09 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/Figures/M2_main/Divergence_dmel.svg 	modified:   Manuscripts/Figures/M2_main/Divergence_scer.svg 	new file:   Manuscripts/Figures/M2_main/pORF_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_scerGC.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Manuscripts/M2/refs.bib 	modified:   Modeling/antisenseGenes.jl

@@
 We developed mathematical models to estimate the probabilities of ORF emergence and loss, in DNA regions antisense to existing protein coding ORFs.
-For any stretch of DNA to be an ORF, its sequence should contain 3$n$ nucleotides (with $n$ being at least 3), with a start codon that marks its beginning, and exactly one stop codon that marks its end.
-Specifically, the absence of any stop codons within the DNA sequence is the most important factor in determining the likelihood of an ORF to exist.
-That is because the likelihood of a premature stop codon increases exponentially with the ORF's length, whereas likelihood of a start codon and a terminal stop codon are unaffected by the ORF's length.
+These models are defined by two kinds of probability.
+The first is the probability of finding a certain kind of DNA sequence, for example an ORF.
+This stationary probability depends on the nucleotide composition of the DNA region that can be roughly approximated by GC content or by the frequencies of short DNA sequences (k-mers).
+The second kind of probability describes the mutational change of a sequence to a different kind of sequence.
+For example, gain or loss of an ORF.
+This transition probability depends on the mutation rate and mutation bias, in addition to nucleotide composition.
+We estimate these parameters primarily from the data on the yeast, \textit{Saccharomyces cerevisiae} \citep{scermutrate}.
+Our choice is motivated by the fact that yeast is a convenient model organism for laboratory experimental studies that can be used to validate theoretical predictions.
+We also perform analogous analyses using data obtained from \textit{Drosophila melanogaster} \citep[][Supplementary section XX]{drosophilamutrate}.
 
-We estimate the gain and loss probabilities of antisense ORFs using the existing (sense) ORF as a reference.
+We estimate the stationary and transition probabilities of antisense ORFs using the existing (sense) ORF as a reference.
 Specifically, we analyse the effect of the composition of the sense ORF, and of mutations in its sequence, to determine how likely it is for an antisense ORFs to exist, and to be gained or lost.
@@
 In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of sense ORF by one and two nucleotide positions, respectively.
-In frame 0, the sequence of an antisense codon is determined by the sequence of the exactly overlapping sense codon.
-However, for frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons.
+Thus in frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons (dicodons).
+Due to this sequence overlap, the evolution of antisense ORFs would be constrained by the evolutionary selection pressures on the sense ORF.
+Furthermore, these constraints would be different for antisense ORFs located in the three different frames.
+We analysed the evolution of antisense ORFs when the sense ORF is under three different levels of purifying selection.
+The first level describes an absence of purifying selection, where any kind of mutation except a non-sense mutation (gain of stop codon) in the sense ORF is tolerated.
+The second level describes a weak purifying selection that allows synonymous mutations (nucleotide mutations that do not change the encoded amino acid), as well as mutations where an amino acid is substituted by a chemically similar amino acid (for example, aspartic acid to glutamic acid).
+Finally, in the third level that describes a strong purifying selection, only synonymous mutations are tolerated in the sense ORF.
+
+\subsection*{Antisense ORFs are more likely exist in frame 1}
+
+For any stretch of DNA to be an ORF, its sequence should contain 3$n$ nucleotides (with $n$ being at least 3), with a start codon that marks its beginning, and exactly one stop codon that marks its end.
+The absence of any stop codon within the DNA sequence is the most important factor in determining the existence of an ORF.
+That is because the likelihood of a premature stop codon increases exponentially with the ORF's length, whereas the likelihoods of a start codon and a terminal stop codon are independent of the ORF's length.
+Furthermore, overlapping DNA region between the sense and antisense ORF would mostly consist of non-terminal (internal) codons.
+
+Based on these considerations, we determined the probability of finding an antisense ORF.
+To this end, we calculated the probability of finding a stop codon the three antisense frames, given the nucleotide composition of sense ORF including the condition that no stop codon exists in the sense ORF.
+Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF.
+It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons, respectively.
+We thus calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons.
+We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF.
+Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames.
+We did so for four different values of GC content.
+We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases.
+The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%.
+Even for these exceptional cases, the probability of an antisense ORF in frame 1 was no less than 91\% of the corresponding ORF probabilities in other frames.
+
+To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC content and lengths.
+We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition.
+We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 were more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs had less than 39 codons and a GC content of 60\%.
+
+GC content may not accurately describe the nucleotide composition of a locus.
+Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs.
+Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome.
 
@@
 \end{table}
-
-We calculated nucleotide substitution probabilities based on mutation rate and mutation rate bias data.
-Specifically, we  used a mutation rate ($\lambda$) of $7.8\times10^{-9}$ mutations per nucleotide position per generation \citep{drosophilamutrate}.
-We derived our mutation bias parameters from two published studies, the first on \textit{Drosophila melanogaster} \citep{drosophilamutrate}, and the second on humans \citep{humanmutrate}.
-\autoref{mutbias} shows the exact values of mutation bias probabilities that we used in this study.
 

### 2023-03-09 Modeling/antisenseGenes.jl

@@ -3,11 +3,11 @@ using Measures
 cd(Base.source_dir())
 include("nucleotidefuncts.jl")
 
-organism = "dmel"
+organism = "scer"
 
 cm2pt = (cm) -> 28.3465*cm
-figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/");
-colors = ["#FFCC00","#5599FF","#D40000","#000000","#754473"];
+figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/M2_main/");
+colors = ["#FFCC00","#5599FF","#D40000","#754473","#000000"];
 lstyles = [:solid,:dash,:dot]
 
 default(linecolor = :black, linewidth = 2, tickfont = font(10,"Helvetica"), 
@@ -24,6 +24,9 @@ allcodons = kmers(3)
 nostopcodons = nostop(allcodons);
 
 nsub, nucsmbt = readdlm(organism*"_mutbias.txt",'\t',header = true);
+if organism == "scer"
+    mutrate = 1.7e-10
+end
 
 noATG = setdiff(allcodons, ["ATG"]);
 function ATGprobs(gccontent)
@@ -381,6 +384,20 @@ for f = 1:3
     end
 
 end
+
+pORF = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n (Frame 1)", size = (width = cm2pt(12), height = cm2pt(10)));
+
+for q in 1:4
+    plot!(pORF,ncodons,log2.(orfvalsONS[q,:,1,1]./orfvalsITG[q,:,1]),
+        linecolor = colors[q]
+    );
+end
+
+plot!(pORF,ncodons,log2.(orfvalsONS_X[:,1,1]./orfvalsITG_X[:,1]),
+        linecolor = colors[5]
+);
+
+savefig(pORF, figdir*"pORF_antisense_"*organism*".pdf")
 
 pGD = plot(plots_gain_dmel..., layout = (1,3), size = (width = cm2pt(27.5), height = cm2pt(9)));
 savefig(pGD, figdir*"pORFgain_antisense_"*organism*"WC.pdf")

### 2023-03-10 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/Figures/M2_main/frames.svg 	modified:   Manuscripts/Figures/M2_main/pORF_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_dmelWC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerWC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_dmelGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_dmelWC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisens

@@
 In frame 0, the codons in the antisense ORF exactly overlap the codons in the sense ORF.
-In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of sense ORF by one and two nucleotide positions, respectively.
+In frames 1 and 2, the codons in the antisense ORF are shifted towards the 5' end of the sense ORF by one and two nucleotide positions, respectively.
 Thus in frames 1 and 2, the sequence of an antisense codon is determined by two partially overlapping sense codons (dicodons).
@@
 Based on these considerations, we determined the probability of finding an antisense ORF.
-To this end, we calculated the probability of finding a stop codon the three antisense frames, given the nucleotide composition of sense ORF including the condition that no stop codon exists in the sense ORF.
+To this end, we calculated the probability of finding a stop codon in the three antisense frames, given the nucleotide composition of the sense ORF, including the condition that no stop codon exists in the sense ORF.
 Specifically, a stop codon can exist in frame 0 wherever the reverse complementary codons (\texttt{TTA},\texttt{CTA},\texttt{TCA}) exist in the sense ORF.
-It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons, respectively.
-We thus calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons.
+It can exist in frames 1 and 2, overlapping with one of the 128 and 192 dicodons in the sense ORF, respectively.
+We next calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons.
 We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF.
@@
 The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%.
-Even for these exceptional cases, the probability of an antisense ORF in frame 1 was no less than 91\% of the corresponding ORF probabilities in other frames.
+Even for these exceptional cases, the probability of an antisense ORF in frame 1 is no less than 91\% of the corresponding ORF probabilities in the other frames.
 
@@
 We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition.
-We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 were more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs had less than 39 codons and a GC content of 60\%.
+We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 are more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs have a GC content of 60\% and less than 39 codons.
 
@@
 Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome.
+With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to occur than intergenic ORF of the same lengths.
+For all ORF lengths, antisense ORFs in frame 3 are less likely to occur than intergenic ORFs.
 

### 2023-03-10 Modeling/antisenseGenes.jl

@@ -385,7 +385,7 @@ for f = 1:3
 
 end
 
-pORF = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n (Frame 1)", size = (width = cm2pt(12), height = cm2pt(10)));
+pORF = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n (Frame 1)", size = (width = cm2pt(11), height = cm2pt(10)));
 
 for q in 1:4
     plot!(pORF,ncodons,log2.(orfvalsONS[q,:,1,1]./orfvalsITG[q,:,1]),
@@ -393,11 +393,17 @@ for q in 1:4
     );
 end
 
-plot!(pORF,ncodons,log2.(orfvalsONS_X[:,1,1]./orfvalsITG_X[:,1]),
-        linecolor = colors[5]
-);
+savefig(pORF, figdir*"pORF_antisense_"*organism*"GC.pdf")
 
-savefig(pORF, figdir*"pORF_antisense_"*organism*".pdf")
+pORFX = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n log2(antisense/intergenic)", size = (width = cm2pt(11), height = cm2pt(10)));
+
+for q = 1:3
+    plot!(pORFX,ncodons,log2.(orfvalsONS_X[:,q,1]./orfvalsITG_X[:,1]),
+            linestyle = lstyles[q]
+    );
+end
+
+savefig(pORFX, figdir*"pORF_antisense_"*organism*"WC.pdf")
 
 pGD = plot(plots_gain_dmel..., layout = (1,3), size = (width = cm2pt(27.5), height = cm2pt(9)));
 savefig(pGD, figdir*"pORFgain_antisense_"*organism*"WC.pdf")

### 2023-03-10 Manuscripts/M2/antisense_genes.tex
new file:   Manuscripts/Figures/M2_main/pORF_antisense_dmel.svg 	modified:   Manuscripts/Figures/M2_main/pORF_antisense_scer.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl

@@
 
-\renewenvironment{abstract}{%
-  \normalsize
-  \textbf{\large\abstractname\\[1em]} % with a normal space
-}
-
 \title{Antisense transcripts can quickly evolve to encode proteins}
@@
 \onehalfspacing
-\def\figdir{../Figures/M1_main}
+\def\figdir{../Figures/M2_main/pdf}
 
@@
 
-\subsection*{Antisense ORFs are more likely exist in frame 1}
+\subsection*{Antisense ORFs are more likely to exist in frame 1}
 
@@
 Furthermore, overlapping DNA region between the sense and antisense ORF would mostly consist of non-terminal (internal) codons.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.5]{\figdir/pORF_antisense_scer.pdf}
+\caption{Antisense ORFs in frame 1 are more likely to exist than intergenic ORFs of identical lengths.
+Panel \textbf{(A)} shows the probability of antisense ORFs in frame 1 relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis).
+Line colors indicate the GC-content of the ORFs (yellow = 30\%, blue = 40\%, red = 50\%, purple = 60\%).
+We do not show antisense ORFs in frames 0 and 2 because their probabilities are identical to that of intergenic ORFs.
+Panel \textbf{(B)} shows the probability of antisense ORFs relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from yeast genome.
+Frames 0, 1 and 2 are denoted by dotted, solid and dashed lines, respectively.
+Horizontal axis in both panels shows the length of the ORFs.
+For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
+\label{pORF}
+\end{figure}
 
@@
 Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome.
-With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to occur than intergenic ORF of the same lengths.
-For all ORF lengths, antisense ORFs in frame 3 are less likely to occur than intergenic ORFs.
+With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to exist than intergenic ORF of the same lengths.
+For all ORF lengths, antisense ORFs in frame 3 are less likely to exist than intergenic ORFs.
+
+Our complementary analyses using \textit{D.
+melanogaster} as a model for mutation bias, produced similar results.
+However, when we computed the ORF probabilities using the frequencies of codons, dicodons and intergenic trimers from  \textit{D.
+melanogaster}, we found that frame 0 was most likely to harbor an antisense ORF.
+
+\subsection*{Antisense ORFs are more likely to emerge in, and less likely to be lost from frame 1}
+
+We next analysed how likely it is for antisense ORFs to emerge, when they are not already present.
+To this end, we calculated gain probability of antisense ORFs in each of the three frames, and under three different intensities of purifying selection.
+We also calculated the probability of ORF gain in the intergenic regions.
+We found that antisense ORFs are less likely to emerge in frames 0 and 2, than ORFs in intergenic regions, for all ORF lengths and GC-content.
+With increasing intensity of purifying selection, the emergence of these antisense ORFs became increasingly less likely than that of intergenic ORFs.
 

### 2023-03-10 Modeling/antisenseGenes.jl

@@ -3,7 +3,7 @@ using Measures
 cd(Base.source_dir())
 include("nucleotidefuncts.jl")
 
-organism = "scer"
+organism = "dmel"
 
 cm2pt = (cm) -> 28.3465*cm
 figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/M2_main/");
@@ -302,6 +302,26 @@ for k in eachindex(ncodons)
     end
 end
 
+pORF = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n (Frame 1)", size = (width = cm2pt(11), height = cm2pt(10)));
+
+for q in 1:4
+    plot!(pORF,ncodons,log2.(orfvalsONS[q,:,1,1]./orfvalsITG[q,:,1]),
+        linecolor = colors[q]
+    );
+end
+
+savefig(pORF, figdir*"pORF_antisense_"*organism*"GC.pdf")
+
+pORFX = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n log2(antisense/intergenic)", size = (width = cm2pt(11), height = cm2pt(10)));
+
+for q = 1:3
+    plot!(pORFX,ncodons,log2.(orfvalsONS_X[:,q,1]./orfvalsITG_X[:,1]),
+            linestyle = lstyles[q]
+    );
+end
+
+savefig(pORFX, figdir*"pORF_antisense_"*organism*"WC.pdf")
+
 plots_gain, plots_loss = [Array{Plots.Plot{Plots.GRBackend}}(undef,3,3) for i = 1:2]; 
 
 rat = zeros(length(ncodons),4,3,2);
@@ -384,26 +404,6 @@ for f = 1:3
     end
 
 end
-
-pORF = plot(xlabel = "ORF length (codons)", ylabel = "Relative ORF Probability\n (Frame 1)", size = (width = cm2pt(11), height = cm2pt(10)));
-
-for q in 1:4
-    plot!(pORF,ncodons,log2.(orfvalsONS[q,:,1,1]./orfvalsITG[q,:,1]),
-        linecolor = colors[q]
-    );
-end
-
-savefig(pORF, figdir*"pORF_antisense_"*organism*"GC.pdf")
-
[…]

### 2023-03-21 Manuscripts/M2/antisense_genes.tex
new file:   Manuscripts/Figures/M2_main/pORFgain_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFgain_antisense_scerWC.svg 	new file:   Manuscripts/Figures/M2_main/pORFloss_antisense_scer.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_scerGC.svg 	modified:   Manuscripts/Figures/M2_main/pORFloss_antisense_scerWC.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl

@@
 
+\usepackage[section]{placeins}
 \usepackage[usenames,dvipsnames]{xcolor}
@@
 The first is the probability of finding a certain kind of DNA sequence, for example an ORF.
-This stationary probability depends on the nucleotide composition of the DNA region that can be roughly approximated by GC content or by the frequencies of short DNA sequences (k-mers).
+This stationary probability depends on the nucleotide composition of the DNA region that can be roughly approximated by GC-content or by the frequencies of short DNA sequences (k-mers).
 The second kind of probability describes the mutational change of a sequence to a different kind of sequence.
@@
 We do not show antisense ORFs in frames 0 and 2 because their probabilities are identical to that of intergenic ORFs.
-Panel \textbf{(B)} shows the probability of antisense ORFs relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from yeast genome.
+Panel \textbf{(B)} shows the probability of antisense ORFs relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from the yeast genome.
 Frames 0, 1 and 2 are denoted by dotted, solid and dashed lines, respectively.
@@
 We next calculated the probability of an antisense stop codon based on the probability of the corresponding sense codons and dicodons.
-We calculated the probability of finding a start codon using the general GC content of the locus, ignoring any effect of the sense ORF.
+We calculated the probability of finding a start codon using the general GC-content of the locus, ignoring any effect of the sense ORF.
 Using the start and stop codon probabilities, we estimated the probability of finding an antisense ORF of different lengths in each of the three frames.
-We did so for four different values of GC content.
+We did so for four different values of GC-content.
 We found that frame 1 has the highest likelihood of harboring an antisense ORF in 99.2\% of cases.
-The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC content of 60\%.
+The only exceptions are ORFs shorter than 39 codons present in a DNA region with a GC-content of 60\%.
 Even for these exceptional cases, the probability of an antisense ORF in frame 1 is no less than 91\% of the corresponding ORF probabilities in the other frames.
 
-To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC content and lengths.
+To understand whether antisense ORFs are more likely to be found than ORFs in intergenic regions, we compared the probabilities of antisense ORFs with that of intergenic ORFs of identical GC-content and lengths.
 We note that that the probability of intergenic ORFs is not constrained by existing ORFs and is thus purely dependent on nucleotide composition.
-We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 are more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs have a GC content of 60\% and less than 39 codons.
+We found that the probability of antisense ORFs in frames 0 and 2 were identical to that of the corresponding intergenic ORFs, which means that antisense ORFs in frame 1 are more likely to exist than intergenic ORFs except in the 0.8\% cases where the ORFs have a GC-content of 60\% and less than 39 codons.
 
-GC content may not accurately describe the nucleotide composition of a locus.
+GC-content may not accurately describe the nucleotide composition of a locus.
 Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs.
@@
 
-\subsection*{Antisense ORFs are more likely to emerge in, and less likely to be lost from frame 1}
+\subsection*{Antisense overlap can facilitate ORF emergence and reduce ORF loss}
+
+\begin{figure}[!b]
+\centering
+\includegraphics[scale=0.5]{\figdir/pORFgain_antisense_scer.pdf}
+\caption{Antisense overlap can facilitate ORF emergence.
+Panel \textbf{(A)} shows the probability of ORF emergence in the three antisense frames (left to right) relative to that in intergenic regions ($\log_2$ ratio, vertical axis), at different intensities of purifying selection (top to bottom).
+Line colors indicate the GC-content of the ORFs.
+Panel \textbf{(B)} shows the ORFs gain probability in the three antisense frames relative to that in intergenic regions ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from the yeast genome.
+Dotted, solid and dashed lines, denote the zero, weak and strong purifying selection, respectively.
+Horizontal axis in all panels shows the length of the ORFs.
+For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
+\label{pORFgain}
+\end{figure}
 
@@
 We found that antisense ORFs are less likely to emerge in frames 0 and 2, than ORFs in intergenic regions, for all ORF lengths and GC-content.
-With increasing intensity of purifying selection, the emergence of these antisense ORFs became increasingly less likely than that of intergenic ORFs.
+In contrast, the emergence of antisense ORF emergence in frame 1 is often more likely than that of intergenic ORFs, especially when the ORFs are long.
+
+Increasing the intensity of purifying selection, reduces the emergence likelihood of antisense ORFs in all three frames.
+However, emergence of antisense ORFs in frame 1 is still more likely than emergence of intergenic ORFs, even under strong purifying selection.
+Specifically, the minimum ORF length where frame 1 antisense ORFs are more likely to emerge than intergenic ORFs, increases with GC-content and the intensity of selection.
+For example, in the absence of purifying selection and a GC-content of 30\%, this length is smaller than the smallest investigated ORF length of 30 codons.
+At the same intensity of selection, this length is 46 codons when the GC-content is 60\%.
+ORFs can emerge in antisense frame 1 more frequently than in intergenic regions under strong purifying selection, with GC-contents of 30\% and 60\%, if they are longer 49 and 109 codons, respectively.
+
+\begin{figure}[!b]
+\centering
+\includegraphics[scale=0.5]{\figdir/pORFloss_antisense_scer.pdf}
+\caption{Antisense overlap can reduce ORF loss.
+Panel \textbf{(A)} shows the probability of ORF loss in the three antisense frames (left to right) relative to that in intergenic regions ($\log_2$ ratio, vertical axis), at different intensities of purifying selection (top to bottom).
+Line colors indicate the GC-content of the ORFs.
+Panel \textbf{(B)} shows the ORFs loss probability in the three antisense frames relative to that in intergenic regions ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from the yeast genome.
+Dotted, solid and dashed lines, denote the zero, weak and strong purifying selection, respectively.
+Horizontal axis in all panels shows the length of the ORFs.
+For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
+\label{pORFloss}
+\end{figure}
+
+Our analysis of ORF gain probabilities using the frequencies of DNA oligomers (codons, dicodons and intergenic trimers), also shows that antisense ORFs are very likely to emerge in frame 1.
+ORFs longer than 30 (shortest investigated length), 60 and 69 codons are more likely to emerge in antisense frame 1 than in intergenic regions, when the purifying selection is absent, weak and strong, respectively.
+Interestingly, this analysis revealed that although antisense ORFs are less likely to emerge in frame 2 than in frame 1, they can emerge more frequently than intergenic ORFs.
+Specifically when the purifying selection is absent, weak and strong, ORFs that are more likely to emerge in antisense frame 2 than in intergenic regions, contain at least 30, 43 and 82 codons, respectively.
+
+Purifying selection reduces the number of tolerated mutations in a DNA locus.
+We note again that our lowest level of purifying selection also disallows nonsense mutations from occurring in the sense ORFs.
+We thus hypothesized that overlap with a sense ORF may protect the antisense ORFs from being lost.
+To this end, we calculated ORF loss probabilities for different ORF lengths, GC-content, and intensities of purifying selection.
+In an analogous analysis, we used codon, dicodon, and intergenic trimer frequencies, instead of GC-content, to calculate ORF loss probabilities.
+Our analyses show that antisense ORFs are indeed protected from loss due to overlap with existing ORFs.
+That is, the loss probability of antisense ORFs is less than that of intergenic ORFs.
+This protection against loss increases with increasing intensity of purifying selection.
+However, selection does not protect the ORFs in all the three antisense frames equally.
+Antisense ORFs are more likely to be preserved when they exist in frame 1.
+
+Overall, our analyses suggest that antisense overlap with an existing ORF facilitates emergence of new ORFs, and protects the existing antisense ORFs from being lost.
+A complementary analysis with parameters estimated from \textit{D.
+melanogaster} further supports this finding.
+
+%frame 1 vs intergenic 
+%
+%no sel:
+% 30
+% 30
+% 33
+% 46
+%
+%ws: 
+% 43
+% 54
+% 70
+% 93
+% 
+%ss:
+%  49
+%  63
+%  81
+% 109
+
+%genomic:
+%ns, ws, ss : 30, 60, 69
 
@@
 \section*{Data availability}
-We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution.
-Specifically, our model is implemented in the scripts \texttt{antisenseGenes.jl}, and  \texttt{antisenseGenes\_supplement.jl}, which in turn depend on the script \texttt{nucleotidefuncts.jl} for some basic functions.
+We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution. Specifically, our model uses the following scripts:
+
+\vspace{-2em}
+\begin{itemize}\setlength{\itemsep}{-1pt}
+\item \texttt{antisenseGenes.jl} (main script)
+\item \texttt{antisenseGenes\_supplement.jl} \\(calculations using codon, dicodon, and intergenic trimer frequencies)
+\item \texttt{nucleotidefuncts.jl} (dependency for basic functions)
+\end{itemize}
 

### 2023-03-21 Modeling/antisenseGenes.jl

@@ -3,10 +3,10 @@ using Measures
 cd(Base.source_dir())
 include("nucleotidefuncts.jl")
 
-organism = "dmel"
+organism = "scer"
 
 cm2pt = (cm) -> 28.3465*cm
-figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/M2_main/");
+figdir = joinpath(Base.source_dir(),"../Manuscripts/Figures/M2_main/pdf/");
 colors = ["#FFCC00","#5599FF","#D40000","#754473","#000000"];
 lstyles = [:solid,:dash,:dot]
 
@@ -107,34 +107,28 @@ cdn2aanum = (cdn) -> aanames[transnucs(cdn)]
 
 function stopprobsoverlap(gccontent)
     # Probability of finding a stop codon
-
-    p_stopWithin = [sum(nucprob.(vjoin(stopNbrWithin[x]),gccontent)) for x in 1:2];
-
+    p_stopR12 = [sum(nucprob.(vjoin(stopNbrWithin[x]),gccontent)) for x in 1:2];
     p_stopR0 = sum(nucprob.(stopneighborsR0,gccontent));
-    push!(p_stopWithin,p_stopR0);
+    p_stopWithin = vcat(p_stopR0,p_stopR12);
 
     # Probability of gaining a stop codon
 
-    g_stopWithin = [featuregain(vjoin(nostopNbrWithin[x]),vjoin(stopNbrWithin[x]), gccontent) for x in 1:2]
-
-    push!(g_stopWithin,featuregain(nostopNbr0,stopneighborsR0,gccontent));
+    g_stopR12 = [featuregain(vjoin(nostopNbrWithin[x]),vjoin(stopNbrWithin[x]), gccontent) for x in 1:2]
+    g_stopR0 = featuregain(nostopNbr0,stopneighborsR0,gccontent);
+    g_stopWithin = vcat(g_stopR0,g_stopR12);
 
     # Probability of losing a stop codon (conditional)
 
-    l_stopWithin = [featuregain(vjoin(stopNbrWithin[x]),vjoin(nostopNbrWithin[x]), gccontent) for x in 1:2]
+    l_stopR12 = [featuregain(vjoin(stopNbrWithin[x]),vjoin(nostopNbrWithin[x]), gccontent) for x in 1:2]
+    l_stopR0 = featuregain(stopneighborsR0,nostopNbr0,gccontent);
+	l_stopWithin = vcat(l_stopR0,l_stopR12)./p_stopWithin; 
 
[…]

### 2023-03-21 Modeling/antisenseGenes_supplement.jl

@@ -43,28 +43,25 @@ end
 function stopprobsoverlapX(t3,t6)
     # Probability of finding a stop codon
 
-    p_stopWithin = [sum([nprob6(y,t6) for y in vjoin(stopNbrWithin[x])]) for x in 1:2];
-
+    p_stopR12 = [sum([nprob6(y,t6) for y in vjoin(stopNbrWithin[x])]) for x in 1:2];
     p_stopR0 = sum([nprob3(y,t3) for y in stopneighborsR0]);
-    push!(p_stopWithin,p_stopR0);
+    p_stopWithin = vcat(p_stopR0,p_stopR12);
 
     # Probability of gaining a stop codon
 
-    g_stopWithin = [featuregain6(vjoin(nostopNbrWithin[x]),vjoin(stopNbrWithin[x]),t6) for x in 1:2]
-
-    push!(g_stopWithin,featuregain3(nostopNbr0,stopneighborsR0,t3));
+    g_stopR12 = [featuregain6(vjoin(nostopNbrWithin[x]),vjoin(stopNbrWithin[x]),t6) for x in 1:2]
+    g_stopR0 = featuregain3(nostopNbr0,stopneighborsR0,t3);
+    g_stopWithin = vcat(g_stopR0,g_stopR12);
 
     # Probability of losing a stop codon (conditional)
 
-    l_stopWithin = [featuregain6(vjoin(stopNbrWithin[x]),vjoin(nostopNbrWithin[x]),t6) for x in 1:2]
+    l_stopR12 = [featuregain6(vjoin(stopNbrWithin[x]),vjoin(nostopNbrWithin[x]),t6) for x in 1:2]
+    l_stopR0 = featuregain3(stopneighborsR0,nostopNbr0,t3);
+	l_stopWithin = vcat(l_stopR0,l_stopR12)./p_stopWithin; 
 
-    push!(l_stopWithin,featuregain3(stopneighborsR0,nostopNbr0,t3));
-
-    l_stopWithin = l_stopWithin./p_stopWithin;
-
-    s_stopWithin = [featurestay6(vjoin(stopNbrWithin[x]),t6) for x in 1:2];
-
-    push!(s_stopWithin,featurestay3(stopneighborsR0,t3));
+    s_stopR12 = [featurestay6(vjoin(stopNbrWithin[x]),t6) for x in 1:2];
[…]

### 2023-03-22 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl

@@
 
-%frame 1 vs intergenic 
-%
-%no sel:
-% 30
-% 30
-% 33
-% 46
-%
-%ws: 
-% 43
-% 54
-% 70
-% 93
-% 
-%ss:
-%  49
-%  63
-%  81
-% 109
+\subsection*{Antisense ORF sequence can diverge even with weak purifying selection on sense ORFs}
 
-%genomic:
-%ns, ws, ss : 30, 60, 69
+In the previous sections, purifying selection on the sense ORF can affect the emergence and loss of antisense ORFs.
+We next asked if this purifying selection can also constrain the diversification of the proteins encoded by antisense ORF sequences.
+To this end, we first calculated the chemical distance ($\delta$) between an amino acid and the substituted amino acid that results due to a mutation.
+For this calculation we used a distance matrix that we derived from an experimentally estimated amino acid similarity matrix reported in a previous study \citep{PMBEC}.
+Next, we calculated the average chemical difference ($\bar{\delta}$) introduced by a random mutation, weighted by the probability of the corresponding mutations.
+Specifically, if $i$ denotes the original codon, $j$ denotes the substituted codon, $P_i$ denotes the probability of finding codon-$i$, $\mu_{ij}$ denotes the probability of codon-$i$ mutating to codon-$j$, and $\delta_{ij}$ denotes the chemical difference between the amino acids encoded by these codons, then the average chemical difference is defined by the following equation:
+
+\begin{equation}
+\bar{\delta} = \frac{\displaystyle\sum_i P_i \sum_j \mu_{ij}\delta_{ij}}{\displaystyle\sum_i P_i \sum_j \mu_{ij}}
+\end{equation}
 

### 2023-03-22 Modeling/antisenseGenes.jl

@@ -421,13 +421,19 @@ function aadivframeR(frm::Int,sdict::Dict,gccontent)
         rcdn = Rcodon(Z,frm)
         stZ2 = stZ[(eachrow(stZ) .!= Ref(Z)) .& ([Rcodon(y,frm) for y in eachrow(stZ)] .∉ Ref(stopcodons)) ,:]
         if(isempty(stZ2))
-            dZ=0
+            dZ = 0
+            mZ = 0
         else
-            dZ = sum([PBMEC2[cdn2aanum(rcdn),cdn2aanum(Rcodon(y,frm))] for y in eachrow(stZ2)])
+            dZZ = [PBMEC2[cdn2aanum(rcdn),cdn2aanum(Rcodon(y,frm))] for y in eachrow(stZ2)]
+
+            mZZ = [mprob(rcdn,Rcodon(y,frm)) for y in eachrow(stZ2)]
+
+            dZ = sum(dZZ.*mZZ)
+            mZ = sum(mZZ)
         end
         pZ = nucprob(join(Z),gccontent)
         d += abs(dZ)*pZ
-        p += pZ
+        p += pZ*mZ
     end
     if p==0
         return 0
@@ -443,13 +449,39 @@ function aadivframe0(sdict::Dict,gccontent)
         stZ = sdict[Z]
         stZ2 = stZ[stZ .!= Z]
         if(isempty(stZ2))
-            dZ=0
+            dZ = 0
+            mZ = 0
         else
-            dZ = sum([PBMEC2[cdn2aanum(Z),cdn2aanum(y)] for y in stZ2])
+            dZZ = [PBMEC2[cdn2aanum(Z),cdn2aanum(y)] for y in stZ2]
+            mZZ = [mprob(Z,y) for y in stZ2]
+
+            dZ = sum(dZZ.*mZZ)
+            mZ = sum(mZZ)
         end
         pZ = nucprob(Z,gccontent)
         d += abs(dZ)*pZ
-        p += pZ
+        p += pZ*mZ
+    end
+    if p==0
+        return 0
+    else
+        return d/p
+    end
+end
+
+function divall(gcc)
+    d = 0
+    p = 0
+    for Z in nostopcodons
+        dZZ = [PBMEC2[cdn2aanum(Z),cdn2aanum(y)] for y in nostopcodons]
+        mZZ = [mprob(Z,y) for y in nostopcodons]
+
[…]

### 2023-03-22 Modeling/antisenseGenes_supplement.jl

@@ -160,13 +160,20 @@ function aadivframeR_X(frm::Int,sdict::Dict,t6)
         rcdn = Rcodon(Z,frm)
         stZ2 = stZ[(eachrow(stZ) .!= Ref(Z)) .& ([Rcodon(y,frm) for y in eachrow(stZ)] .∉ Ref(stopcodons)) ,:]
         if(isempty(stZ2))
-            dZ=0
+            dZ = 0
+            mZ = 0
         else
-            dZ = sum([PBMEC2[cdn2aanum(rcdn),cdn2aanum(Rcodon(y,frm))] for y in eachrow(stZ2)])
+            dZZ = [PBMEC2[cdn2aanum(rcdn),cdn2aanum(Rcodon(y,frm))] for y in eachrow(stZ2)]
+
+            mZZ = [mprob(rcdn,Rcodon(y,frm)) for y in eachrow(stZ2)]
+
+            dZ = sum(dZZ.*mZZ)
+            mZ = sum(mZZ)
+
         end
         pZ = nprob6(join(Z),t6)
         d += abs(dZ)*pZ
-        p += pZ
+        p += pZ*mZ
     end
     if p==0
         return 0
@@ -182,13 +189,38 @@ function aadivframe0_X(sdict::Dict,t3)
         stZ = sdict[Z]
         stZ2 = stZ[stZ .!= Z]
         if(isempty(stZ2))
-            dZ=0
+            dZ = 0
+            mZ = 0
         else
-            dZ = sum([PBMEC2[cdn2aanum(Z),cdn2aanum(y)] for y in stZ2])
+            dZZ = [PBMEC2[cdn2aanum(Z),cdn2aanum(y)] for y in stZ2]
+            mZZ = [mprob(Z,y) for y in stZ2]
+
+            dZ = sum(dZZ.*mZZ)
+            mZ = sum(mZZ)
         end
         pZ = nprob3(join(Z),t3)
         d += abs(dZ)*pZ
-        p += pZ
+        p += pZ*mZ
+    end
+    if p==0
+        return 0
+    else
+        return d/p
+    end
+end
+
+function divallX(t3)
+    d = 0
+    p = 0
+    for Z in nostopcodons
+        dZZ = [PBMEC2[cdn2aanum(Z),cdn2aanum(y)] for y in nostopcodons]
+        mZZ = [mprob(Z,y) for y in nostopcodons]
+
+        dZ = sum(dZZ.*mZZ)
[…]

### 2023-03-23 Modeling/antisenseGenes.jl
modified:   Modeling/antisenseGenes.jl 	modified:   Modeling/antisenseGenes_supplement.jl

@@ -414,10 +414,11 @@ savefig(pLD, figdir*"pORFloss_antisense_"*organism*"WC.pdf")
 
 
 function aadivframeR(frm::Int,sdict::Dict,gccontent)
-    d = 0
-    p = 0
+    pvec,dvec = [zeros(size(csetR[frm],1),1) for i = 1:2]
+    ind = 0
     for Z in eachrow(csetR[frm])
-        stZ = phcp(vcat(Z[1],sdict[Z[1]]),vcat(Z[2],sdict[Z[2]]))
+        ind += 1
+            stZ = phcp(vcat(Z[1],sdict[Z[1]]),vcat(Z[2],sdict[Z[2]]))
         rcdn = Rcodon(Z,frm)
         stZ2 = stZ[(eachrow(stZ) .!= Ref(Z)) .& ([Rcodon(y,frm) for y in eachrow(stZ)] .∉ Ref(stopcodons)) ,:]
         if(isempty(stZ2))
@@ -432,20 +433,24 @@ function aadivframeR(frm::Int,sdict::Dict,gccontent)
             mZ = sum(mZZ)
         end
         pZ = nucprob(join(Z),gccontent)
-        d += abs(dZ)*pZ
-        p += pZ*mZ
+        dvec[ind] = abs(dZ)*pZ
+        pvec[ind] = pZ*mZ
     end
+    p = sum(pvec)
+    dm = sum(dvec)/p
+    dv = sum((dvec - dm.*pvec).^2)/p
     if p==0
-        return 0
+        return [0,0]
     else
-        return d/p
+        return [dm,sqrt(dv)]
     end
 end
 
 function aadivframe0(sdict::Dict,gccontent)
-    d = 0
-    p = 0
+    pvec,dvec = [zeros(size(nostopNbr0)) for i = 1:2]
+    ind = 0
     for Z in nostopNbr0
+        ind += 1
         stZ = sdict[Z]
         stZ2 = stZ[stZ .!= Z]
         if(isempty(stZ2))
@@ -459,20 +464,24 @@ function aadivframe0(sdict::Dict,gccontent)
             mZ = sum(mZZ)
         end
         pZ = nucprob(Z,gccontent)
-        d += abs(dZ)*pZ
-        p += pZ*mZ
+        dvec[ind] = abs(dZ)*pZ
+        pvec[ind] = pZ*mZ
     end
+    p = sum(pvec)
+    dm = sum(dvec)/p
[…]

### 2023-03-23 Modeling/antisenseGenes_supplement.jl

@@ -153,9 +153,10 @@ function pstopselX(sdict,stopvals,t3,t6)
 end
 
 function aadivframeR_X(frm::Int,sdict::Dict,t6)
-    d = 0
-    p = 0
+    pvec,dvec = [zeros(size(csetR[frm],1),1) for i = 1:2]
+    ind = 0
     for Z in eachrow(csetR[frm])
+        ind +=1
         stZ = phcp(vcat(Z[1],sdict[Z[1]]),vcat(Z[2],sdict[Z[2]]))
         rcdn = Rcodon(Z,frm)
         stZ2 = stZ[(eachrow(stZ) .!= Ref(Z)) .& ([Rcodon(y,frm) for y in eachrow(stZ)] .∉ Ref(stopcodons)) ,:]
@@ -172,20 +173,24 @@ function aadivframeR_X(frm::Int,sdict::Dict,t6)
 
         end
         pZ = nprob6(join(Z),t6)
-        d += abs(dZ)*pZ
-        p += pZ*mZ
+        dvec[ind] = abs(dZ)*pZ
+        pvec[ind] = pZ*mZ
     end
+    p = sum(pvec)
+    dm = sum(dvec)/p
+    dv = sum((dvec - dm.*pvec).^2)/p
     if p==0
-        return 0
+        return [0,0]
     else
-        return d/p
+        return [dm,sqrt(dv)]
     end
 end
 
 function aadivframe0_X(sdict::Dict,t3)
-    d = 0
-    p = 0
+    pvec,dvec = [zeros(size(nostopNbr0)) for i = 1:2]
+    ind = 0
     for Z in nostopNbr0
+        ind +=1
         stZ = sdict[Z]
         stZ2 = stZ[stZ .!= Z]
         if(isempty(stZ2))
@@ -199,32 +204,39 @@ function aadivframe0_X(sdict::Dict,t3)
             mZ = sum(mZZ)
         end
         pZ = nprob3(join(Z),t3)
-        d += abs(dZ)*pZ
-        p += pZ*mZ
+        dvec[ind] = abs(dZ)*pZ
+        pvec[ind] = pZ*mZ
     end
+    p = sum(pvec)
+    dm = sum(dvec)/p
+    dv = sum((dvec - dm.*pvec).^2)/p
     if p==0
-        return 0
+        return [0,0]
     else
-        return d/p
+        return [dm,sqrt(dv)]
     end
 end
 
 function divallX(t3)
-    d = 0
[…]

### 2023-03-24 Manuscripts/M2/antisense_genes.tex
modified:   Manuscripts/Figures/M2_main/Divergence_scer.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl

--- 2023-03-21 Manuscripts/M2/antisense_genes.tex
@@
 
-%frame 1 vs intergenic 
-%
-%no sel:
-% 30
-% 30
-% 33
-% 46
-%
-%ws: 
-% 43
-% 54
-% 70
-% 93
-% 
-%ss:
-%  49
-%  63
-%  81
-% 109
+\subsection*{Overlap with sense ORFs promotes divergence of antisense ORFs in frame 2}
 
-%genomic:
-%ns, ws, ss : 30, 60, 69
+In the previous sections, we showed that purifying selection on the sense ORF can affect the emergence and loss of antisense ORFs.
+We next asked if this purifying selection can also constrain the diversification of the proteins encoded by antisense ORF sequences.
+To this end, we first calculated the chemical distance ($\delta$) between any two amino acids.
+For this calculation we used a distance matrix that we derived from an experimentally estimated amino acid similarity matrix reported in a previous study \citep{PMBEC}.
+Next, we calculated the average chemical difference ($\bar{\delta}$) introduced by a random mutation, weighted by the probability of different mutations.
+Using $\bar{\delta}$ as a measure of divergence, we estimated the extent to which antisense ORFs in the three frames, can diverge as a result of mutations, and due to purifying selection on the sense ORFs.
+Likewise, we also calculated the divergence of intergenic ORFs as a consequence of random mutations.
 
+We found that frame 2 allows maximum divergence of antisense ORFs, under both weak and strong purifying selection on the sense ORF.
+Antisense ORFs in frame 0 diverge the least.
+Interestingly, strong selection on sense ORF increases the divergence of antisense ORFs in frame 2.
+The reason could be that the few mutations that do occur under strong purifying selection, cause a relatively higher increase in divergence than the more numerous mutations that are allowed to occur under weak purifying selection.
+We also found that the divergence of frame 2 antisense ORFs was higher than than that of intergenic ORFs under both selection regimes.
+We note this result does not mean that intergenic ORFs can diverge less than antisense ORFs.
+Evolution of intergenic ORFs is not constrained by another DNA sequence.
+However, as long as the mutants do not affect the organismal fitness, evolution would not be biased towards divergence increasing mutations.
+Thus random mutations in intergenic ORFs could also consist of many synonymous and chemistry preserving mutations, that are probably disallowed in frame 2 antisense ORFs due to purifying selection on sense ORFs.
+
+In contrast to frame 2, the divergence of antisense ORFs in the other two frames decreased with increasing strength of purifying selection on the sense ORF.
+For example, frame 0 antisense ORFs did not diverge at all when the sense ORF was under strong purifying selection.
+Antisense ORFs in frames 0 and 1 also diverged less than intergenic ORFs under both selection regimes.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[scale=0.5]{\figdir/Divergence_scer.pdf}
+\caption{Antisense ORFs can diversify when sense ORFs are under purifying selection.
+Vertical axis denotes the divergence ($\bar{\delta}$) of antisense ORFs due to a random mutation when the sense ORF is under weak (left) or strong (right) purifying selection.
+Horizontal axes denote the three antisense frames.
+Colored bars denote divergence values of antisense ORFs with different GC-content, and black bars denote the diversity values calculated using frequencies of short DNA sequences from the yeast genome.
+Filled circles that are similarly color coded, denote the divergence of intergenic ORFs due to mutations.}
+\label{divergence}
+\end{figure}
 \section*{Discussion}
@@
 
+\subsection{Divergence of protein sequence}
+Specifically, if $i$ denotes the original codon, $j$ denotes the substituted codon, $P_i$ denotes the probability of finding codon-$i$, $\mu_{ij}$ denotes the probability of codon-$i$ mutating to codon-$j$, and $\delta_{ij}$ denotes the chemical difference between the amino acids encoded by these codons, then the average chemical difference is defined by the following equation:
+
+\begin{equation}
+\bar{\delta} = \frac{\displaystyle\sum_i P_i \sum_j \mu_{ij}\delta_{ij}}{\displaystyle\sum_i P_i \sum_j \mu_{ij}}
+\end{equation}
+
 \section*{Data availability}

### 2023-03-24 Modeling/antisenseGenes.jl

@@ -536,50 +536,46 @@ bwf = 1/7;
 
 pDivRel = plot(
     xlabel = "Frame",
-    ylabel = "Divergence", 
-    size = (width = cm2pt(11.5), height = cm2pt(10)),   
 );
-
-
-for j = 1:3
-    bar!(pDivRel,[j+x*bwf for x in 1:5],alldivRel[:,j],
-        fill = colors,
-        bar_width = bwf,
-        linecolor = nothing
-    )
-    plot!(pDivRel,[j+x*bwf for x in 1:5],alldivNoS,
-        markersize = 2,
-        markershape = :circle,
-        markercolor = :white,
-        linecolor = nothing
-    )
-end
-
-xticks!(pDivRel,[1:3;] .+ 3*bwf, string.([0,1,2]));
-
-savefig(pDivRel,figdir*"DivergenceRel_"*organism*".pdf")
-
-
 pDivMax = plot(
     xlabel = "Frame",
-    ylabel = "Divergence", 
-    size = (width = cm2pt(11.5), height = cm2pt(10)),   
 );
 
 for j = 1:3
-    bar!(pDivMax,[j+x*bwf for x in 1:5],alldivMax[:,j],
+    bar!(pDivRel,[j+x*bwf for x in 1:5],alldivRel_avg[:,j],
         fill = colors,
         bar_width = bwf,
         linecolor = nothing
     )
-    plot!(pDivMax,[j+x*bwf for x in 1:5],alldivNoS,
-        markersize = 2,
+    plot!(pDivRel,[j+x*bwf for x in 1:5],alldivNoS[1,:],
+        markersize = 3,
         markershape = :circle,
-        markercolor = :white,
+        markercolor = colors,
+        markerstrokecolor = :white,
+        markerstrokewidth = 2.5,
+        linecolor = nothing
+    )
+    bar!(pDivMax,[j+x*bwf for x in 1:5],alldivMax_avg[:,j],
+        fill = colors,
+        bar_width = bwf,
+        linecolor = nothing
+    )
+    plot!(pDivMax,[j+x*bwf for x in 1:5],alldivNoS[1,:],
+        markersize = 3,
+        markershape = :circle,
+        markercolor = colors,
+        markerstrokecolor = :white,
+        markerstrokewidth = 2.5,
         linecolor = nothing
     )
 end
 
[…]
